---
title: "The structured view: programmatic access to regression tables"
description: >
  `as_structured()` exposes the typed representation behind every
  `table_regression()` output: a numeric body with split CI columns,
  per-column metadata, row-role indices, and the format
  specification -- the same contract spicy's own Excel, gt,
  tinytable, flextable, and clipboard engines consume. Filter
  coefficients programmatically, aggregate raw values, or build a
  custom renderer.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{The structured view: programmatic access to regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

`table_regression()` returns a *display* representation: a character
table with `(ref.)` annotations, en-dashes, bracketed `[LL, UL]`
intervals, and APA-formatted p-values. That is the right thing to
print and the right thing to publish -- and the wrong thing to
compute on. The companion accessor `as_structured()` returns the
*typed* view underneath: a fully numeric body, confidence intervals
pre-split into their own columns, `NA` where a cell is structurally
empty, plus the metadata a renderer needs to reconstruct the
display. It is the same contract spicy's own output engines (Excel,
gt, tinytable, flextable, clipboard) consume internally -- nothing
about it is second-class.

Two sibling tools serve different needs, and choosing well saves
work. `broom::tidy()` returns the *statistical* long form -- one row
per term and estimate type per model, with `estimate`, `std.error`,
`conf.low`, `p.value`. A default table carries a single estimate
type, so that is one row per coefficient; requesting standardized
betas or AMEs adds a row per term for each. It is the right input
for meta-analysis, further computation on estimates, or ggplot. `as_structured()` returns the *display
geometry* -- rows and columns exactly as the rendered table lays
them out, factor headers and fit statistics included -- and is the
right input for filtering what a reader will see, post-processing
numbers in table shape, or rendering the table with an engine spicy
does not ship. This vignette covers the second tool. The shared
mechanics of `table_regression()` are in
[*Publication-ready regression
tables*](table-regression.html); the class-by-class map is
[*Supported models*](table-regression-supported-models.html).

## The schema on one model

Start from the display view, so the typed view has something to be
compared against:

```{r schema}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
tbl <- table_regression(fit)
tbl
```

Every display trait named above is visible here: `(ref.)` markers
and en-dashes on the reference rows, blank factor header rows, one
bracketed `[LL, UL]` column, APA p-values (`<.001`, no leading
zero). The typed view underneath:

```{r components}
s <- as_structured(tbl)
names(s)
```

The centrepiece is `body`: a data frame with the `Variable` column
and one numeric column per displayed statistic. The confidence
interval that prints as one bracketed column arrives split into
`LL` / `UL`:

```{r body}
s$body
```

Read it against the printed table and the structure becomes
obvious: factor header rows (`sex:`, `smoking:`) and reference
rows carry `NA` across the statistics -- structurally empty, not
missing data -- and the fit-statistics rows (`n`, `R²`, ...) put
their value in the first numeric column. The row-role indices name
each of these regions so you never have to guess from the text:

```{r indices}
s$reference_rows
s$factor_header_rows
s$fit_stat_rows
s$level_rows
```

`col_meta` describes each column: which token produced it, which
model it belongs to, its display precision, and -- for p-value
columns -- the APA style and below-threshold rule, plus per-row
`fit_stat_overrides` where a fit-statistic row departs from the
column's precision (put to work in the renderer below).
`format_spec` carries the global defaults (decimal mark, digits,
CI level):

```{r meta}
s$col_meta$B
s$format_spec$ci_level
```

That is seven of the twelve components; `spanners` and
`reference_models_by_row` follow in the multi-model section below.
Of the remaining three, `ci_pairs` records which `LL`/`UL`
columns form a confidence-interval pair (here, columns 4 and 5
under the label `95% CI`), while `outcome_row` and
`outcome_labels_by_col` locate and label the optional outcome
header row of multi-outcome tables and are empty here. The
complete schema, component by component, is in `?as_structured`.

## Filtering and aggregating

Because the body is numeric, ordinary data-frame idioms work. Keep
the coefficient rows a reader would call significant:

```{r filter}
s$body[which(s$body$p < 0.05), c("Variable", "B", "p")]
```

Note the `which()`: the structurally empty rows are `NA`, and
`which()` drops them where a bare logical index would keep them as
`NA` rows. To restrict attention to coefficient rows before any
filter, exclude the structural regions:

```{r coefrows}
structural <- c(s$reference_rows, s$factor_header_rows, s$fit_stat_rows)
coef_rows <- setdiff(seq_len(nrow(s$body)), structural)
s$body[coef_rows, c("Variable", "B", "95% CI: LL", "95% CI: UL")]
```

Aggregation is just as direct -- the mean absolute coefficient, the
range of CI widths:

```{r aggregate}
mean(abs(s$body$B[coef_rows]))
range(s$body[coef_rows, "95% CI: UL"] - s$body[coef_rows, "95% CI: LL"])
```

## Multi-model tables: spanners and per-model columns

With several models, columns are prefixed by the model label and
two more components become useful. `spanners` maps each model label
to its column indices in `body`; `reference_models_by_row` records,
for each reference row, which models actually contain the factor --
the rule renderers use to decide where the reference en-dash
belongs. To see it discriminate, the `Minimal` model below omits
`smoking`:

```{r multi}
fit2 <- lm(wellbeing_score ~ age + sex + smoking + bmi, data = sochealth)
fit0 <- lm(wellbeing_score ~ age + sex, data = sochealth)
s2 <- as_structured(
  table_regression(list(Minimal = fit0, Base = fit, Extended = fit2))
)
s2$spanners
names(s2$body)
s2$reference_models_by_row
```

Rows 4 and 7 are the two reference rows (`Female (ref.)`,
`No (ref.)`). All three models contain `sex`, so row 4 lists all
three; only `Base` and `Extended` contain `smoking`, so row 7
lists those two -- a renderer draws the reference en-dash in
exactly those models' columns and leaves the `Minimal` cells
blank.

Model-wise extraction follows from the spanner map:

```{r modelwise}
s2$body[, c(1, s2$spanners$Extended)]
```

## Building your own renderer

The structured view carries everything a renderer needs. A compact
`knitr::kable()` rendering takes a few lines: each column formats at
its `col_meta` precision, and the `fit_stat_overrides` seen earlier
supply the per-row exceptions -- that is what renders the `n` row as
`1175` rather than `1175.00`:

```{r kable}
render_kable <- function(tbl) {
  s <- as_structured(tbl)
  out <- s$body
  for (nm in names(out)[-1]) {
    meta <- s$col_meta[[nm]]
    prec <- meta$precision
    if (is.null(prec)) prec <- s$format_spec$digits
    row_prec <- rep(prec, nrow(out))
    for (ov in meta$fit_stat_overrides) {
      if (!is.null(ov$precision)) row_prec[ov$row] <- ov$precision
    }
    out[[nm]] <- vapply(
      seq_len(nrow(out)),
      function(i) {
        if (is.na(out[[nm]][i])) {
          ""
        } else {
          formatC(out[[nm]][i], format = "f", digits = row_prec[i])
        }
      },
      character(1)
    )
  }
  knitr::kable(out, align = c("l", rep("r", ncol(out) - 1)))
}
render_kable(tbl)
```

Anything more ambitious -- a `DT::datatable()` with significant
rows highlighted, a LaTeX booktabs layout, an HTML widget --
follows the same pattern: `body` for the cells, `col_meta` for
per-column formatting, the row indices for styling regions, and
`format_spec` for the global conventions. The note and title stay
available on the original object (`attr(tbl, "note")`,
`attr(tbl, "title")`), so a custom renderer can reproduce spicy's
self-documenting footer verbatim.

## The contract

Three properties make the structured view safe to build on:

* **Engine parity.** The built-in engines consume this exact
  structure: what you read in `body` is, cell for cell, what the
  Excel workbook contains and what the gt table displays at its
  precision. This parity is pinned by tests.
* **Stability tier.** `as_structured()` sits in the *stabilising*
  tier of the API (see `?spicy`, section *API stability*):
  components will not be silently renamed or change semantics
  within `0.y.z`, and additions are announced in `NEWS.md`.
* **Version guard.** Objects built by a spicy version before the
  structured contract existed are refused with an actionable
  message rather than mis-read.

For the statistical long form -- estimates with standard errors and
unformatted p-values, one row per term and estimate type across
models -- reach for `broom::tidy()` on the same object; for one-line model summaries,
`broom::glance()`. The three views are built from the same frames,
so their numbers always agree.
