---
title: "Cramer's V, Phi, and association measures"
description: >
  Calculate Cramer's V, Phi, Goodman-Kruskal Gamma, Kendall's Tau-b,
  Somers' D, and other effect sizes for contingency tables in R, with
  confidence intervals and p-values.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cramer's V, Phi, and association measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spicy)
```

spicy provides a coherent set of effect size and association measures
for contingency tables, covering nominal and ordinal variables. This
vignette explains which measure to use depending on the measurement
level of your variables, and how to obtain confidence intervals and
p-values for chi-squared-based and rank-based statistics.

## Choosing the right measure

The table below summarizes the recommended measures by variable type.

| Variable types | Recommended measure | Function |
|---|---|---|
| Nominal x Nominal | Cramer's V | `cramer_v()` |
| Nominal x Nominal | Contingency Coefficient | `contingency_coef()` |
| Nominal x Nominal (2x2) | Phi | `phi()` |
| Ordinal x Ordinal | Kendall's Tau-b | `kendall_tau_b()` |
| Ordinal x Ordinal (rectangular) | Stuart's Tau-c | `kendall_tau_c()` |
| Ordinal x Ordinal | Goodman-Kruskal Gamma | `gamma_gk()` |
| Ordinal x Ordinal (asymmetric) | Somers' D | `somers_d()` |
| Nominal (asymmetric, PRE) | Lambda | `lambda_gk()` |
| Nominal (asymmetric, PRE) | Goodman-Kruskal Tau | `goodman_kruskal_tau()` |
| Nominal (asymmetric, PRE) | Uncertainty Coefficient | `uncertainty_coef()` |
| 2x2 table | Yule's Q | `yule_q()` |

PRE = Proportional Reduction in Error. These measures quantify how much
knowing one variable reduces prediction error for the other.

Cramer's V and the contingency coefficient are defined for tables of
any dimension. For a nominal-by-ordinal pair they are the standard
choice: the ordering is simply ignored, because the rank-based measures
require both variables to be ordinal.

All functions accept a contingency table (class `table`, typically from
`xtabs()` or `table()`).

## Quick overview with assoc_measures()

`assoc_measures()` computes all available measures at once:

```{r assoc-all}
tbl <- xtabs(~ smoking + education, data = sochealth)
assoc_measures(tbl)
```

Directional variants are labelled `R|C` (the row variable is treated
as dependent, i.e. predicted from the column variable) and `C|R` (the
reverse); `symmetric` variants single out no dependent variable.

`assoc_measures()` computes every measure the table dimensions allow,
regardless of measurement level: the ordinal rows above treat the
factor level order of `smoking` (No < Yes) as a substantive ordering.
This is useful for exploratory analysis, but read only the rows that
match your variable types. For reporting, pick the measure that
matches them.

## Nominal variables

Chi-squared-based measures ignore any ordering of the categories, which
makes them the reference choice when at least one of the two variables
is nominal.

### Cramer's V

Cramer's V measures the strength of association between two categorical
variables. It ranges from 0 (no association) to 1 (perfect association).

```{r cramer}
tbl <- xtabs(~ smoking + education, data = sochealth)
cramer_v(tbl)
```

In this pair, `smoking` is nominal and `education` is ordinal: Cramer's
V treats `education` as an unordered set of categories, which is the
standard treatment for a nominal-by-ordinal pair. When *both* variables
are ordinal, the rank-based measures presented below use more of the
information in the data.

Pass `detail = TRUE` for the confidence interval and p-value. The
p-value tests the null hypothesis of no association using the Pearson
chi-squared test. The chi-squared-based measures (Cramer's V, Phi, the
contingency coefficient) have no standard asymptotic standard error, so
the `SE` column prints `–` for them; the rank-based measures later in
this vignette report their asymptotic SE in that column.

```{r cramer-detail}
cramer_v(tbl, detail = TRUE)
```

### Phi coefficient

For 2x2 tables, Phi is equivalent to Cramer's V. spicy implements Phi
as \(\sqrt{\chi^2 / n}\), matching the DescTools and PSPP conventions
-- the value is always non-negative and reports only the *strength*
of association, not its direction. SPSS itself signs Phi on 2x2
tables (there it equals the Pearson correlation between the two
variables coded 0/1), so SPSS output can show a negative value of the
same magnitude. To recover the signed direction, compute the Pearson
correlation between the two binary variables explicitly (e.g.,
`cor(x, y)` after coding both 0/1). The p-value tests H0: no
association (Pearson chi-squared test).

```{r phi}
tbl_22 <- xtabs(~ smoking + physical_activity, data = sochealth)
phi(tbl_22, detail = TRUE)
```

### Contingency coefficient

The contingency coefficient is an alternative to Cramer's V. Its upper
bound depends on the table dimensions, which makes it harder to compare
across tables of different sizes. The p-value tests H0: no association
(Pearson chi-squared test).

```{r contingency}
contingency_coef(tbl, detail = TRUE)
```

## Ordinal variables

When both variables are ordinal (ordered factors), measures that account
for the ordering are more appropriate than Cramer's V. Note that these
measures are not on a common scale: on the health-by-education table
used throughout this section, Cramer's V is 0.176, Tau-b is 0.205 and
Gamma is 0.310 for the very same association. Compare magnitudes within
a measure, never across measures.

### Goodman-Kruskal Gamma

Gamma ranges from -1 to +1. It ignores tied pairs, which tends to
overestimate the strength of association when there are many ties: on
any table, |Gamma| is at least as large as |Tau-b|. Like the other
measures built on the difference between concordant and discordant
pairs (Tau-b, Tau-c, Somers' D), its sign carries the direction of
association.

```{r gamma}
tbl_ord <- xtabs(~ self_rated_health + education, data = sochealth)
gamma_gk(tbl_ord, detail = TRUE)
```

A positive value means that higher values on one variable tend to occur
with higher values on the other. The p-value tests H0: Gamma = 0 using
a Wald z-test.

### Kendall's Tau-b

Tau-b adjusts for ties and ranges from -1 to +1. It is generally
preferred over Gamma for square or near-square tables. The p-value
tests H0: Tau-b = 0 (Wald z-test).

```{r tau-b}
kendall_tau_b(tbl_ord, detail = TRUE)
```

### Stuart's Tau-c

Stuart's Tau-c -- the label spicy shares with SAS, after Stuart
(1953) -- is also known as Kendall's Tau-c, the name SPSS and PSPP
print for the same statistic. It is similar to Tau-b but adjusts for
rectangular tables where the number of rows and columns differ. The
p-value tests H0: Tau-c = 0 (Wald z-test).

```{r tau-c}
kendall_tau_c(tbl_ord, detail = TRUE)
```

### Somers' D

Somers' D is an asymmetric measure: it distinguishes between a dependent
and an independent variable. By default, the row variable is treated as
dependent (D(R|C)). The p-value tests H0: D = 0 (Wald z-test). One
disclosure worth knowing: the association line printed below the table
by `cross_tab()` reports the *symmetric* variant of Somers' D, while
the standalone `somers_d()` defaults to the row-dependent D(R|C) — the
two values differ by construction.

```{r somers}
somers_d(tbl_ord, detail = TRUE)
```

### How the p-values compare with SPSS and PSPP

spicy's Wald z-tests divide each estimate by the displayed asymptotic
standard error, which is estimated *without* assuming independence.
SPSS and PSPP test the ordinal measures differently: they use the
standard error computed *under* the independence hypothesis (which is
why their "Approx. T" column shows the same value for Gamma, Tau-b,
Tau-c and Somers' D), and SPSS refers Goodman-Kruskal's Tau to a
chi-squared distribution rather than a z-test. Estimates and standard
errors match SPSS and PSPP; the p-values can differ, especially near
the null. For the non-negative PRE measures below (Lambda,
Goodman-Kruskal's Tau, the uncertainty coefficient) the Wald z-test is
approximate in a further way: H0 places the parameter on the boundary
of its range, so read those p-values as indicative rather than exact.

## Asymmetric (PRE) measures

These measures answer a specific question: how much does knowing the
column variable reduce our error in predicting the row variable (or vice
versa)? Each function takes a `direction` argument. `lambda_gk()` and
`uncertainty_coef()` default to `direction = "symmetric"`, which
combines the two prediction directions rather than singling out a
dependent variable; `goodman_kruskal_tau()` defaults to
`direction = "row"` (the column variable predicts the row variable).
Pass `direction = "row"` or `direction = "column"` explicitly when your
question is directional.

### Lambda

Lambda measures the proportional reduction in classification error. The
default is the symmetric variant; here we also show the row-dependent
variant, where education (the column variable) predicts self-rated
health (the row variable):

```{r lambda}
lambda_gk(tbl_ord, detail = TRUE)
lambda_gk(tbl_ord, direction = "row", detail = TRUE)
```

The row-dependent result illustrates a well-known caveat: Lambda can
equal zero even when the variables are associated, if the modal
category of the dependent variable does not change across the
categories of the predictor. That is exactly what happens here --
"Good" is the modal health category at every education level, so the
row-dependent Lambda is exactly 0 even though the same table shows a
clear association (Cramer's V = 0.176, chi-squared p < .001). When the
estimate is zero its asymptotic SE is zero as well, and no test is
printed (`–`). Otherwise the p-value tests H0: Lambda = 0 (Wald
z-test).

### Goodman-Kruskal Tau

Tau measures the proportional reduction in error when predicting the row
variable from the column variable, using the full distribution (not just
the mode). This matches its default `direction = "row"`; pass
`direction = "column"` for the reverse prediction. The p-value tests
H0: Tau = 0 (Wald z-test).

```{r gk-tau}
goodman_kruskal_tau(tbl_ord, detail = TRUE)
```

### Uncertainty coefficient

The uncertainty coefficient (Theil's U) is based on entropy. It measures
how much knowing one variable reduces uncertainty about the other. By
default it reports the symmetric coefficient; pass
`direction = "row"` or `direction = "column"` for the asymmetric
variants. The p-value tests H0: U = 0 (Wald z-test).

```{r uncertainty}
uncertainty_coef(tbl_ord, detail = TRUE)
```

## Yule's Q

Yule's Q is defined for 2x2 tables only. It ranges from -1 to +1 and
is equivalent to Gamma for 2x2 tables. The p-value tests H0: Q = 0
(Wald z-test).

```{r yule}
tbl_22 <- xtabs(~ smoking + physical_activity, data = sochealth)
yule_q(tbl_22, detail = TRUE)
```

## Automatic selection in cross_tab()

`cross_tab()` can automatically select an appropriate measure via
`assoc_measure = "auto"` (the default). When both variables are ordered
factors, it picks Kendall's Tau-b; otherwise it uses Cramer's V.

```{r cross-tab-auto}
# Nominal: Cramer's V
cross_tab(sochealth, smoking, education)

# Ordinal: Kendall's Tau-b (automatic)
cross_tab(sochealth, self_rated_health, education)
```

You can override the automatic choice:

```{r cross-tab-override}
cross_tab(sochealth, self_rated_health, education, assoc_measure = "gamma")
```

## Confidence intervals

Most functions report a confidence interval via `detail = TRUE`, with
two exceptions: `contingency_coef()` never reports one (no standard
asymptotic SE exists for the contingency coefficient), and
`somers_d(direction = "symmetric")` reports the estimate only. Cramer's
V and Phi also omit the interval silently in degenerate cases (estimate
exactly 0, or n <= 3). The confidence level defaults to 95% and can be
changed with `conf_level` -- here both levels on the
smoking-by-education table from the beginning of the vignette:

```{r ci-level}
cramer_v(tbl, detail = TRUE)
cramer_v(tbl, detail = TRUE, conf_level = 0.99)
```

To drop the confidence interval, pass `conf_level = NULL`; the result
keeps the estimate, the SE column, and the p-value:

```{r ci-null}
cramer_v(tbl, detail = TRUE, conf_level = NULL)
```

## Controlling decimal places

When `detail = FALSE` (the default), functions return a plain numeric
scalar, so R's own formatting rules apply. When `detail = TRUE`, the
result uses a custom print method that defaults to 3 decimal places.
Pass `digits` to change the precision of the estimate, SE, and CI
columns. The p-value follows APA-style formatting independent of
`digits`: 3 decimal places with the leading zero stripped (`.045`)
or `<.001` below the threshold:

```{r digits}
cramer_v(tbl, detail = TRUE, digits = 4)
```

The same `digits` argument works for `assoc_measures()`:

```{r digits-table}
assoc_measures(tbl, digits = 2)
```

You can also store a result and re-display it with a different
precision without recalculating:

```{r digits-print}
res <- cramer_v(tbl, detail = TRUE)
print(res, digits = 5)
```

## See also

- See `vignette("frequency-tables")` for `cross_tab()`, which reports
  these measures inline through its `assoc_measure` argument.
- See `vignette("table-categorical")` for association measures inside
  publication-ready categorical summary tables.

## References

* Agresti, A. (2002). *Categorical Data Analysis* (2nd ed.). Wiley.
* Brown, M. B., &amp; Benedetti, J. K. (1977). Sampling behavior of tests
  for correlation in two-way contingency tables. *Journal of the
  American Statistical Association*, 72(358), 309–315.
* Goodman, L. A., &amp; Kruskal, W. H. (1954). Measures of association
  for cross classifications. *Journal of the American Statistical
  Association*, 49(268), 732–764.
* Kendall, M. G. (1938). A new measure of rank correlation.
  *Biometrika*, 30(1–2), 81–93.
* Liebetrau, A. M. (1983). *Measures of Association*. Sage.
* Somers, R. H. (1962). A new asymmetric measure of association for
  ordinal variables. *American Sociological Review*, 27(6), 799–811.
* Stuart, A. (1953). The estimation and comparison of strengths of
  association in contingency tables. *Biometrika*, 40(1–2), 105–110.
* Theil, H. (1970). On the estimation of relationships involving
  qualitative variables. *American Journal of Sociology*, 76(1),
  103–154.
* Yule, G. U. (1900). On the association of attributes in statistics.
  *Philosophical Transactions of the Royal Society of London, Series
  A*, 194, 257–319.
