---
title: "Categorical predictors"
description: >
  How categorical predictors enter a regression table and how to work
  with them: dummy coding and the reference level, joint
  (multiple-degree-of-freedom) tests of a factor, R's polynomial
  contrasts for ordered factors, the scores-versus-dummies decision for
  ordinal predictors, successive-difference coding, and why continuous
  predictors should not be categorized.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Categorical predictors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

Every regression vignette in this collection is organised by the type
of *outcome* — continuous, binary, count, ordinal, multinomial,
survival. This one is organised by the type of *predictor*. How a
categorical predictor is coded, tested, simplified, and presented is
the same decision whether the model is an `lm()`, a logistic
regression, or a Cox model, so it has one home rather than a copy in
every family vignette. The examples use the bundled `sochealth` survey
data (`?sochealth`) with `lm()` and `glm()` fits; everything shown
carries over to the other supported classes.

The running order follows the analysis itself: how a factor enters
the model, how to choose what it is compared against, how to test
whether it matters at all, what changes when its levels are ordered,
and — the reverse question — why a continuous predictor should not be
turned into a factor.

## How a factor enters the model

R expands a factor with `k` levels into `k − 1` **indicator (dummy)
variables**, one per non-reference level; the remaining level becomes
the **reference** absorbed by the intercept (Gelman, Hill &amp;
Vehtari 2020, §10.4). `table_regression()` groups the indicator rows
under the parent variable and renders the reference level explicitly,
with `(ref.)` and an en dash in the statistic columns:

```{r basic}
fit <- lm(wellbeing_score ~ age + employment_status, data = sochealth)
table_regression(fit)
```

Each coefficient is a comparison **with the reference level, holding
the other predictors constant** — not an overall effect of the
variable. Reading the rows against that reference: unemployed respondents score
on average 3.6 points lower on the 0–100 wellbeing index than
*employed* respondents of the same age (B = -3.60, 95% CI
[-6.17, -1.03], p = .006); for students and inactive respondents the
data show no clear difference *from the employed* (p = .441 and
.854). Nothing in this table compares students with the unemployed —
that contrast is the difference between the two coefficients, and its
standard error is not printed here. If one specific non-reference
contrast is the research question, the cleanest route is to make one
of its ends the reference.

## Choosing and changing the reference level

The reference level is a **presentation choice, not a modeling one**:
refitting with another reference is the same model reparameterized —
identical fit, likelihood, AIC, and fitted values; only the
comparisons displayed change (Gelman, Hill &amp; Vehtari 2020, §10.4).

```{r relevel}
d2 <- sochealth
d2$employment_status <- relevel(d2$employment_status, ref = "Unemployed")
fit_u <- lm(wellbeing_score ~ age + employment_status, data = d2)
c(AIC(fit), AIC(fit_u))
table_regression(fit_u, keep = "employment_status")
```

Every group now sits 3.6 to 4.7 points *above* the unemployed — the
same fit read from the other end. Two practical rules:

* **Do not let alphabetical order decide.** `factor()` sorts levels
  alphabetically by default, so the reference is whichever category
  happens to sort first — rarely the one you would have chosen, and a
  silent source of confusing signs when data come from text files.
* **Choose a reference that makes every displayed contrast
  readable**: the largest category (stable comparisons), an
  unexposed / control category (epidemiology), or the substantively
  natural baseline. A rare category makes a poor reference — every
  contrast inherits its noise.

## Does the factor matter at all? The joint test

The per-level p-values above cannot answer "does employment status
matter?": each one tests a single level against the reference, and
the set of them changes with the reference choice. The question about
the factor *as a whole* is a **joint test of all `k − 1` coefficients
at once** — a multiple-degree-of-freedom test, invariant to the
reference (Harrell 2015, §2.6). Harrell's rule is to run it *before*
interpreting any individual contrast: when levels straddle
significance, as here, the joint test is the honest headline.

The change statistics of a nested comparison give exactly that test —
adding the factor to a model that lacks it. One prerequisite: change
statistics are only valid when both models are fit to the same rows,
and R's listwise deletion silently drops *different* rows when the
added variable carries its own missing values — so the code below
first restricts the data to cases complete on every variable
involved. `table_regression()` enforces this: models with different
`nobs` are refused, with instructions to refit on the common subset.

```{r joint}
d_cc <- sochealth[complete.cases(
  sochealth[, c("wellbeing_score", "age", "employment_status")]
), ]
m0 <- lm(wellbeing_score ~ age, data = d_cc)
m1 <- lm(wellbeing_score ~ age + employment_status, data = d_cc)
table_regression(list(m0, m1), nested = TRUE, show_columns = c("b", "p"))
```

The `F-change` row is the joint 3-df test: F = 3.13, p = .025 —
employment status improves the model, driven by the unemployed
contrast. For `glm` fits, the `partial_chi2` column reports the same
joint question per term as a likelihood-ratio chi-square with its df,
without needing a second model (see the main vignette's *Term-level
partial chi-square*).

The rule cuts the other way too:

```{r joint-region}
d_r <- sochealth[complete.cases(
  sochealth[, c("wellbeing_score", "age", "region")]
), ]
anova(lm(wellbeing_score ~ age,          data = d_r),
      lm(wellbeing_score ~ age + region, data = d_r))
```

For `region`, the joint test gives F(5, 1193) = 0.89, p = .48 — no
evidence that region matters at all (`anova()` on two nested fits is
the same F-change test the table reports). Fitting the factor anyway
and scanning its five per-level p-values for a "significant" contrast
to report alone would be a multiplicity artifact — "if the overall
test is not significant, it can be dangerous to rely on individual
pairwise comparisons" (Harrell 2015, §2.6).

## Ordered factors: R's polynomial contrasts

An **ordinal** predictor — levels with a natural order, like
`self_rated_health` (Poor &lt; Fair &lt; Good &lt; Very good) —
raises a coding decision that R quietly makes for you. For an
`ordered` factor, R does *not* use dummy coding: it uses **orthogonal
polynomial contrasts**, and the table shows trend components instead
of level comparisons. spicy flags this the first time it happens and
the footer names the coding:

```{r ordered}
fit_srh <- lm(wellbeing_score ~ age + self_rated_health,
              data = sochealth)
table_regression(fit_srh, show_columns = c("b", "se", "p"))
```

These rows answer trend questions across the ordered levels: `.L`
(linear) is the dominant component here — wellbeing rises steeply
across health ratings — while `.Q` (quadratic, B = -3.93, p &lt; .001)
says the rise is **concave**: the steps get smaller toward the top of
the scale. `.C` (cubic, p = .235) adds nothing. This decomposition is
a legitimate reading — trend tests are exactly what polynomial
contrasts are for — but it is *not* the level-versus-reference
reading most regression tables intend. For that, convert to a plain
factor (`factor(x, ordered = FALSE)`), as the message suggests and as
the other vignettes in this collection do.

## Scores or dummies?

An ordinal predictor poses a substantive decision: should it enter
as **numeric scores** (1, 2, 3, … — one slope, read as "per step up
the scale") or as **dummies** (k − 1 coefficients, no order
imposed)?
The scores model is nested in the dummy model, so a likelihood-ratio
test compares them: fit both codings and let the data say whether the
level coefficients depart from a linear trend in the scores (Agresti
2007, §4.4.3). `nested = TRUE` runs the comparison. The two examples
below land on opposite sides of the decision.

**Case 1 — the scores suffice.** Education (3 levels) predicting
smoking:

```{r scores-education}
sh <- sochealth
sh$education  <- factor(sh$education, ordered = FALSE)
sh$educ_score <- as.numeric(sh$education)
sh <- sh[complete.cases(sh[, c("smoking", "sex", "age", "education")]), ]

g_lin <- glm(smoking ~ sex + age + educ_score, data = sh,
             family = binomial)
g_fac <- glm(smoking ~ sex + age + education,  data = sh,
             family = binomial)
table_regression(list(g_lin, g_fac), nested = TRUE,
                 show_columns = c("b", "p"))
```

Freeing education from the linear trend buys a chi-square of 0.03 on
1 df, p = .852 — no evidence the dummies improve on the scores — and
AIC agrees (1198.9 vs 1200.9). The scores model is simpler to report
(one coefficient: B = -0.45 per education step, on the log-odds
scale) and more powerful against a genuine trend, because it spends
one parameter where the dummies spend two. With 3 levels the test has
a familiar counterpart: it is the same 1-df question as the Wald test
on the `.Q` contrast of the ordered coding (z-test p = .852 here; in
a GLM the Wald and likelihood-ratio answers agree asymptotically, and
exactly in a linear model) — the quadratic component *is* the
departure from linearity.

**Case 2 — linearity is rejected.** Self-rated health (4 levels)
predicting wellbeing:

```{r scores-srh}
d3 <- sochealth[complete.cases(
  sochealth[, c("wellbeing_score", "age", "self_rated_health")]
), ]
d3$srh_score  <- as.numeric(d3$self_rated_health)
d3$srh_factor <- factor(d3$self_rated_health, ordered = FALSE)

s_lin <- lm(wellbeing_score ~ age + srh_score,  data = d3)
s_fac <- lm(wellbeing_score ~ age + srh_factor, data = d3)
table_regression(list(s_lin, s_fac), nested = TRUE,
                 show_columns = c("b", "p"))
```

Here the test rejects the single slope: F-change = 12.96 on 2 df,
p &lt; .001. The dummy coefficients show why: the steps from Poor are
+14.97 (Fair), +28.01 (Good), +35.13 (Very good) — increments of
about 15, 13, then 7 points. The effect flattens at the top of the
scale, exactly the concavity the `.Q` contrast detected above (the
2-df test here is identical to the joint test of `.Q` and `.C`).
Forcing one slope of 10.96 per step would overstate the last step and
understate the first. Keep the dummies.

Three caveats keep the decision honest (Harrell 2015, §§2.6 and
2.7.3; Grambsch &amp; O'Brien 1991):

* **A non-significant test does not prove linearity** — in modest
  samples several codings fit adequately, and integer scores assume
  the steps are *equally spaced*, a substantive claim about the
  scale, not a statistical default.
* **Test-then-simplify inflates error.** Reporting the scores model
  as if it had been chosen in advance makes the subsequent test's
  distribution wrong. Grambsch and O'Brien showed, for the
  quadratic-versus-linear case, that the pre-specified 2-df joint
  test remains close to optimal *even when the truth is linear*; the
  same logic carries over to the factor's full-df test — so when the
  factor is a control variable rather than the focus, the dummies
  cost little and assume nothing.
* **The stakes rise with the number of levels.** With a handful of
  levels — three to five, in Harrell's account — the variable is
  usually modeled as a polytomous factor: the dummies spend a
  parameter or two and assume nothing. The advantage of scores grows
  when `k` is larger or the model is dense relative to the sample.

## A third coding: successive differences

Dummy coding answers "how far is each level from the reference";
scores answer "how much per step, forced equal". A third question —
"how big is **each** step?" — has its own coding: MASS's
successive-difference contrasts, whose coefficients are the
differences between *adjacent* levels (Venables &amp; Ripley 2002).
Harrell (2015, §2.7.3) reaches the same reading with threshold
indicators `[X ≥ level]`; the two parameterizations give identical
fits.

```{r sdif}
d3$srh_sdif <- d3$self_rated_health
contrasts(d3$srh_sdif) <- MASS::contr.sdif(4)
table_regression(
  lm(wellbeing_score ~ age + srh_sdif, data = d3),
  keep = "srh_sdif"
)
```

Each row is one step of the scale: Fair sits 14.97 points above
Poor, Good 13.04 above Fair, Very good 7.12 above Good — the
flattening gradient read directly, with a standard error per step.
This is the natural display when the steps themselves are the
finding. (Two notes: the rows group under the factor but carry no
`(ref.)` row — no reference level exists under this coding; and the
joint test of the factor is unchanged — same model, different
questions asked of it.)

## The reverse temptation: categorizing a continuous predictor

Turning a *categorical* variable into a number is sometimes
justified (the scores above); turning a **continuous** predictor into
categories almost never is (Harrell 2015, §2.4.1). `sochealth` ships both `bmi` and a conventional
`bmi_category` (normal weight / overweight / obesity), so the cost
can be shown directly:

```{r categorize}
d4 <- sochealth[complete.cases(
  sochealth[, c("wellbeing_score", "bmi", "bmi_category")]
), ]
d4$bmi_category <- factor(d4$bmi_category, ordered = FALSE)  # per-level dummies, as above
b_cont <- lm(wellbeing_score ~ bmi,          data = d4)
b_cat  <- lm(wellbeing_score ~ bmi_category, data = d4)
table_regression(list(b_cont, b_cat), show_columns = c("b", "p"),
                 show_fit_stats = c("nobs", "r2", "AIC"))
```

The continuous fit explains twice the variance with half the
parameters (R² = .020 with 1 df against .010 with 2 df; AIC 9884.2
vs 9898.1). That is Harrell's argument in one table: categorization
discards the variation *within* intervals — every BMI from 25.0 to
29.9 is treated as the same exposure — and assumes wellbeing jumps
at the cut-points and is flat between them, a far stronger assumption
than linearity. The estimates also stop being portable: the
"Overweight vs Normal weight" coefficient (B = -1.69, p = .082)
depends on where the sample sits inside the 25.0–29.9 interval, where
the per-unit slope (B = -0.60 points per BMI unit, p &lt; .001) does
not. And if a *nonlinear* effect is the worry, the remedy is a
flexible continuous fit — a polynomial or a spline — not bins:

```{r quadratic}
AIC(b_cont, update(b_cont, . ~ . + I(bmi^2)))
```

The quadratic term adds nothing (AIC 9885.9 against 9884.2 for the
linear fit), so the linear fit stands.

Cut-points chosen *after looking at the outcome* do more damage
still: p-values and CIs for "optimal" cut-points are invalid
without special corrections, and such cut-points fail to replicate
across studies (Harrell 2015, §2.4.1). Categories remain fine for
*descriptive* tables and for presentation-layer grouping — the
mistake is putting them in the model when the continuous measurement
exists.

## Presentation

The presentation layer is independent of the coding decisions above.
`labels` renames variables and levels; `reference_style =
"annotation"` lifts the reference into the factor header for compact
tables; `keep` / `drop` filter rows by regex after all statistics are
computed (see the main vignette's *Display options* and *Filtering
displayed coefficients*):

```{r presentation}
table_regression(
  fit,
  labels = c(age = "Age (years)",
             employment_status = "Employment status"),
  reference_style = "annotation"
)
```

## See also

- `vignette("table-regression", package = "spicy")` for the shared
  table mechanics (variance estimators, nested layouts, filtering,
  output formats).
- `vignette("table-regression-ordinal", package = "spicy")` for
  ordinal *outcomes* — the proportional-odds model, its tests and
  relaxations.
- `vignette("table-regression-multinomial", package = "spicy")` for
  the scores-vs-dummies decision inside a multinomial model, where
  each freed coding costs one parameter per equation.

## References

* Agresti, A. (2007). *An Introduction to Categorical Data Analysis*
  (2nd ed.). Wiley.
* Gelman, A., Hill, J., &amp; Vehtari, A. (2020). *Regression and
  Other Stories*. Cambridge University Press.
* Grambsch, P. M., &amp; O'Brien, P. C. (1991). The effects of
  transformations and preliminary tests for non-linearity in
  regression. *Statistics in Medicine*, 10(5), 697–709.
* Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.).
  Springer.
* Venables, W. N., &amp; Ripley, B. D. (2002). *Modern Applied
  Statistics with S* (4th ed.). Springer.
