---
title: "Getting started with spicy"
description: >
  Get started with spicy for descriptive statistics, variable
  inspection, frequency tables, cross-tabulations, association
  measures, categorical and continuous summary tables, regression
  coefficient tables, and codebooks in R. A tidyverse-friendly
  alternative to SPSS and Stata for survey and labelled data
  workflows.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with spicy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spicy)
```

spicy is an R package for descriptive statistics and data analysis,
designed for data science and survey research workflows. It covers
variable inspection, frequency tables, cross-tabulations with
chi-squared tests and effect sizes, and publication-ready summary
tables, offering functionality similar to Stata or SPSS but within a
tidyverse-friendly R environment. This vignette walks through the core
workflow using the bundled `sochealth` dataset, a simulated
social-health survey with 1200 respondents and 24 variables.

## Inspect your data

`varlist()` (or its shortcut `vl()`) gives a compact overview of every
variable in a data frame: name, label, representative values, class,
number of distinct values, valid observations, and missing values.
In RStudio or Positron, calling `varlist(mydata)` with the default
`tbl = FALSE` opens an interactive viewer - this is the most common
usage in practice. Here we use `tbl = TRUE` to produce static output
for the vignette:

```{r varlist}
varlist(sochealth, tbl = TRUE)
```

You can also select specific columns with tidyselect syntax:

```{r varlist-select}
varlist(sochealth, starts_with("bmi"), income, weight, tbl = TRUE)
```

## Frequency tables

`freq()` produces frequency tables with counts, percentages, valid
percentages (shown by default whenever missing values are present),
and optionally cumulative percentages.

```{r freq}
freq(sochealth, education)
```

Weighted frequencies use the `weights` argument. With `rescale = TRUE`,
the total weighted N matches the unweighted N:

```{r freq-weighted}
freq(sochealth, education, weights = weight, rescale = TRUE)
```

## Cross-tabulations

`cross_tab()` crosses two categorical variables. By default it shows
counts, a chi-squared test, and Cramer's V:

```{r crosstab}
cross_tab(sochealth, smoking, education)
```

Add percentages with `percent`:

```{r crosstab-pct}
cross_tab(sochealth, smoking, education, percent = "column")
```

Group by a third variable with `by`:

```{r crosstab-by}
cross_tab(sochealth, smoking, education, by = sex)
```

When both variables are ordered factors, `cross_tab()` automatically
selects an ordinal measure (Kendall's Tau-b) instead of Cramer's V:

```{r crosstab-ordinal}
cross_tab(sochealth, self_rated_health, education)
```

## Association measures

For a quick overview of all available association statistics, pass a
contingency table to `assoc_measures()`:

```{r assoc-measures}
tbl <- xtabs(~ smoking + education, data = sochealth)
assoc_measures(tbl)
```

Individual functions such as `cramer_v()`, `gamma_gk()`, or
`kendall_tau_b()` return a scalar by default. Pass `detail = TRUE` for
the confidence interval and p-value:

```{r cramer-detail}
cramer_v(tbl, detail = TRUE)
```

## Summary tables

`table_categorical()` covers grouped or one-way summary tables for
categorical variables:

```{r table-categorical}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education
)
```

`table_continuous()` summarizes continuous variables, either overall
or by a categorical `by` variable; when `by` is supplied it also
reports a group-comparison test (a Welch test by default):

```{r table-continuous}
table_continuous(
  sochealth,
  select = c(bmi, life_sat_health),
  by = education
)
```

`table_continuous_lm()` covers the same reporting territory when you
want to stay in a linear-model framework, for example with robust or
cluster-robust standard errors, case weights, or additive covariate
adjustment:

```{r table-continuous-lm}
table_continuous_lm(
  sochealth,
  select = c(wellbeing_score, bmi),
  by = sex,
  vcov = "HC3"
)
```

By default every `table_*` helper formats to APA conventions. To match
a journal instead, pass `style =` — `"jama"`, `"nejm"`, `"lancet"`,
`"annals"`, `"apa"`, `"aer"`; `spicy_style_names()` lists them all. Set
it once for a whole document with `options(spicy.style = "nejm")`, or
compose a custom variant with `spicy_style()`, for example
`spicy_style("lancet", ci_sep = " to ")`. The same styles apply to the
regression tables below. French typographic conventions are not a
journal style: they come with the language, through
`options(spicy.language = "fr")`.

For detailed guidance, see the dedicated articles on
`table_categorical()`, `table_continuous()`, and
`table_continuous_lm()`, and the final reporting overview tying the
summary tables together.

## Regression tables

`table_regression()` reports the full coefficient table for one or
several fitted models — more than thirty classes, from `lm()` /
`glm()` to mixed-effects, ordinal, survival and Bayesian engines (the
full map is the [Supported models](https://amaltawfik.github.io/spicy/articles/table-regression-supported-models.html)
article) — with
APA formatting by default and the same journal styles on demand
(`style = "jama"`, `"nejm"`, `"lancet"` and more), factor grouping with
reference rows, robust variance, standardised coefficients, average
marginal effects, hierarchical comparisons, and side-by-side
multi-model layouts:

```{r table-regression}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit)
```

For detailed guidance, start with the core `table_regression()`
article; the Learn more section below lists the companion articles by
model family.

## Row-wise summaries

`mean_n()`, `sum_n()`, and `count_n()` compute row-wise statistics
across selected columns, with explicit control over missing values.
By default, a row mean or sum requires every selected value to be
valid; `min_valid` relaxes this to a minimum count (or proportion) of
valid values. The rows selected below include respondents with missing
satisfaction items so the difference is visible: `mean_sat` (no
`min_valid`) turns `NA` as soon as one item is missing, `sum_sat` is
computed from the valid items as long as at least three are present
(and turns `NA` in the last row, which has only two), and
`count_n(special = "NA")` counts the missing items themselves:

```{r mean-n}
sochealth |>
  dplyr::mutate(
    mean_sat  = mean_n(select = starts_with("life_sat")),
    sum_sat   = sum_n(select = starts_with("life_sat"), min_valid = 3),
    n_missing = count_n(select = starts_with("life_sat"), special = "NA")
  ) |>
  dplyr::select(starts_with("life_sat"), mean_sat, sum_sat, n_missing) |>
  dplyr::slice(c(1, 2, 43, 82, 455)) |>
  as.data.frame()
```

## Learn more

This is the vignette the package ships. Every walk-through below is an
article on the [spicy website](https://amaltawfik.github.io/spicy/),
kept alongside the reference pages and rebuilt with each release.

**Explore and tabulate**

- [Explore variables and build codebooks](https://amaltawfik.github.io/spicy/articles/variable-exploration.html)
  — inspect variables, labels and missingness (`varlist()`,
  `code_book()`, `label_from_names()`).
- [Frequency tables and cross-tabulations](https://amaltawfik.github.io/spicy/articles/frequency-tables.html)
  — `freq()` and `cross_tab()` in depth (weights, simulation,
  labelled data).
- [Cramer's V, Phi, and association measures](https://amaltawfik.github.io/spicy/articles/association-measures.html)
  — choosing the right effect size for a contingency table.

**Summary tables**

- [Categorical summary tables](https://amaltawfik.github.io/spicy/articles/table-categorical.html)
  and [Continuous summary tables](https://amaltawfik.github.io/spicy/articles/table-continuous.html)
  — the Table 1 / 2 builders (APA by default, other journal styles via
  `style =`).
- [One outcome across several groupings](https://amaltawfik.github.io/spicy/articles/table-outcome.html)
  — one continuous outcome across several groupings, one block of rows
  per grouping.
- [Model-based continuous summary tables](https://amaltawfik.github.io/spicy/articles/table-continuous-lm.html)
  — model-based group comparisons (robust SE, weights, covariate
  adjustment).
- [Summary tables from a survey design](https://amaltawfik.github.io/spicy/articles/survey-tables.html)
  — the same two tables from a `survey` design: design-based standard
  errors, degrees of freedom and tests (`table_continuous_svy()`,
  `table_categorical_svy()`).

**Regression tables**

- [Publication-ready regression tables](https://amaltawfik.github.io/spicy/articles/table-regression.html)
  — the core guide, including the univariable screen
  (`table_regression_uv()`).
- [Supported models](https://amaltawfik.github.io/spicy/articles/table-regression-supported-models.html)
  — the class-by-class capability map (more than thirty model
  classes).
- [The structured view](https://amaltawfik.github.io/spicy/articles/as-structured.html)
  — the typed view behind every regression table: filter, aggregate,
  or build a custom renderer.
- [Mixed-effects regression tables](https://amaltawfik.github.io/spicy/articles/table-regression-mixed.html)
  — mixed-effects (multilevel) models, random effects as table rows,
  ICC.
- [GEE regression tables](https://amaltawfik.github.io/spicy/articles/table-regression-gee.html)
  — GEE (population-averaged) models, working-correlation choice via
  QIC, sandwich inference.
- [Count and two-part regression tables](https://amaltawfik.github.io/spicy/articles/table-regression-counts.html)
  — Poisson, negative-binomial, zero-inflated and hurdle models.
- [Survival regression tables](https://amaltawfik.github.io/spicy/articles/table-regression-survival.html)
  — Cox hazard ratios and accelerated failure time models.
- [Ordinal regression tables](https://amaltawfik.github.io/spicy/articles/table-regression-ordinal.html)
  — proportional-odds models, thresholds, per-category marginal
  effects.
- [Multinomial regression tables](https://amaltawfik.github.io/spicy/articles/table-regression-multinomial.html)
  — multinomial logit with outcome categories as columns.
- [Bayesian regression tables](https://amaltawfik.github.io/spicy/articles/table-regression-bayesian.html)
  — posterior medians, credible intervals, convergence checks.
- [Categorical predictors](https://amaltawfik.github.io/spicy/articles/categorical-predictors.html)
  — reference levels, joint tests, contrasts — across all of the
  above.

**Putting it together**

- [Summary tables for reporting](https://amaltawfik.github.io/spicy/articles/summary-tables-reporting.html)
  — the end-to-end reporting sequence tying the four `table_*` helpers
  together.
- `?mean_n`, `?sum_n`, `?count_n` — row-wise summaries with
  minimum-valid-values rules.
