---
title: "Summary tables for APA-style reporting"
description: >
  Learn when to use table_categorical() and table_continuous() for
  APA-style reporting in R, how their shared arguments fit together,
  and which output format to choose for console, Quarto, Word, or
  Excel workflows.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summary tables for APA-style reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spicy)
```

`table_categorical()` and `table_continuous()` are the two main summary
table helpers in spicy. They share the same high-level grammar:
choose variables with `select`, optionally split the table with `by`,
apply readable labels, and pick an output format that matches your
reporting workflow. This vignette focuses on that shared logic rather
than repeating every function-specific option.

## Choose the right function

Use the function that matches the type of variables you want to report:

| Function | Use for | Optional `by` | Typical additions |
|:--|:--|:--|:--|
| `table_categorical()` | Factors, labelled categorical variables, grouped frequency-style summaries | Yes | Chi-squared test, association measure, confidence interval |
| `table_continuous()` | Numeric or continuous variables | Yes | Group-comparison test, statistic, effect size |

In practice:

- use `table_categorical()` for smoking status, education, or activity;
- use `table_continuous()` for BMI, income, or scale scores;
- keep `by` for the grouping variable you want to compare across.

## A shared reporting grammar

Both functions follow the same core pattern:

```{r grammar-categorical}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  labels = c("Smoking status", "Regular physical activity"),
  output = "tinytable"
)
```

```{r grammar-continuous}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  labels = c(
    bmi = "Body mass index",
    wellbeing_score = "Well-being score",
    life_sat_health = "Satisfaction with health"
  ),
  output = "tinytable"
)
```

The same mental model works in both cases:

- `select` chooses the reported variables;
- `by` defines the grouping structure;
- `labels` cleans up the row labels;
- `output` decides how the result is rendered or exported.

## A practical reporting sequence

A common report contains both table types, often with the same grouping
variable. For example, you might first summarize categorical health
behaviors, then summarize continuous well-being indicators.

### Categorical variables

```{r report-categorical}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  labels = c(
    "Smoking status",
    "Regular physical activity",
    "Visited a dentist in the last 12 months"
  ),
  output = "gt"
)
```

### Continuous variables

```{r report-continuous}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  labels = c(
    bmi = "Body mass index",
    wellbeing_score = "Well-being score",
    life_sat_health = "Satisfaction with health"
  ),
  p_value = TRUE,
  effect_size = TRUE,
  output = "gt"
)
```

This keeps the reporting structure consistent while still using the
function that fits each variable type.

## Choose the output format

Both functions support the same reporting formats:

| Output | Best use |
|:--|:--|
| `"default"` | Quick console review in plain ASCII |
| `"tinytable"` | Quarto or R Markdown documents |
| `"gt"` | HTML output with styled reporting tables |
| `"flextable"` | Word-oriented workflows |
| `"excel"` | Spreadsheet handoff or downstream editing |
| `"word"` | Direct `.docx` export |
| `"clipboard"` | Fast pasting into another application |

Pick the output based on where the table is going, not on the analysis
itself. The underlying selection and grouping pattern stays the same.

## Keep the detailed options in the function-specific articles

The dedicated articles go deeper into each function:

- `table_categorical()` covers missing values, level filtering,
  association measures, and one-way frequency-style tables.
- `table_continuous()` covers grouped descriptive statistics,
  parametric and nonparametric tests, and effect sizes.

Use this vignette as the final reporting overview, then consult the
function-specific articles when you need the detailed controls.
