---
title: "Summary tables from a survey design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summary tables from a survey design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
survey_ok <- requireNamespace("survey", quietly = TRUE)
knitr::opts_chunk$set(eval = survey_ok)
```

```{r, echo = FALSE, eval = !survey_ok, results = "asis"}
cat("The **survey** package is not installed, so the examples below are not run.\n")
```

When a sample is not a simple random sample -- when it is stratified,
clustered, weighted to a population, or carried by replicate weights --
the mean is still a weighted mean, but everything around it changes.
The standard error depends on the strata and the clusters, the
degrees of freedom are a count of sampling units rather than of rows,
and the tests are corrected for the design. None of that can be
recovered from a column of weights.

spicy has two functions for that case, and they compute nothing
themselves: `table_continuous_svy()` and `table_categorical_svy()`
delegate every number to the **survey** package (Lumley, 2004, 2010)
and put spicy's
restitution layer around it -- the same columns, the same nine output
routes, the same typed view, plus a footer that says what design
produced the numbers.

```{r setup, message = FALSE}
library(spicy)
library(survey)
data(api)
```

Three designs from survey's own `api` data, used throughout:

```{r designs}
dclus1 <- svydesign(id = ~dnum, weights = ~pw, data = apiclus1, fpc = ~fpc)
dstrat <- svydesign(
  id = ~1, strata = ~stype, weights = ~pw, data = apistrat, fpc = ~fpc
)
rclus1 <- as.svrepdesign(dclus1)
```

## Continuous variables

```{r cont-basic}
table_continuous_svy(dclus1, select = c(api00, api99))
```

Three sentences under the table, and each is there because the number
above it cannot be read without them. The design is a one-stage
cluster sample of 15 school districts with a finite population
correction; the standard errors come from Taylor linearisation; the
interval is a *t* interval on 14 degrees of freedom -- the number of
clusters minus one, not the 183 schools.

That last point is worth dwelling on. `confint()` on a survey object
uses the *normal* distribution by default:

```{r cont-df}
m <- svymean(~api00, dclus1)
confint(m)                    # survey's default: normal
confint(m, df = degf(dclus1)) # what the table reports
```

`svyciprop()`, `svyquantile()`, `svyttest()` and `regTermTest()` all
take their degrees of freedom from the design, not from the row
count, and so do the tables -- everywhere. (Each spends them as its
statistic requires: `svyttest()` uses one on the estimated
difference, so its *t* sits on `degf(design) - 1` -- 13 here.) The
footer says so rather than leaving the reader to find out.

### Choosing the statistics

`show_columns` takes the tokens of `table_continuous()`, plus `"se"`
(the design-based standard error) and `"deff"` (the design effect),
and minus `"med_ci"`:

```{r cont-columns}
table_continuous_svy(
  dclus1,
  select = api00,
  show_columns = c("m", "se", "ci", "med_iqr", "n", "weighted_n", "deff"),
  deff = TRUE
)
```

The design effect of 9.35 is the point of the whole exercise: the
variance of this cluster sample is nine times what a simple random
sample of 183 schools would have given. A table that ignored the
design would report a standard error three times too small.

`"med_ci"` is refused rather than approximated. The exact interval
`table_continuous()` reports inverts a binomial sign test, and its
coverage guarantee rests on independent observations -- which a
clustered or stratified sample is not. The estimand (the population
median) is not the problem; the interval construction is. For a
design-based interval on that same estimand, call
`survey::svyquantile(interval.type = )` on the design itself.

### Groups

`by = ` cuts one domain per group. survey recomputes the degrees of
freedom on the sampling units each domain retains, so a grouped table
generally carries a different *df* per row, and the footer gives the
span:

```{r cont-by}
table_continuous_svy(dclus1, select = api00, by = stype, statistic = TRUE)
```

The comparison is one design-based test on the whole design, not a set
of pairwise ones: `svyttest()` with two observed groups, the Wald *F*
of `regTermTest()` with three or more, and `svyranktest()` under
`test = "nonparametric"`. Under a design there is no Welch / Student
distinction -- the variance is the design's -- so `test = "student"`
warns and behaves like the default.

## Categorical variables

```{r cat-basic}
table_categorical_svy(dclus1, select = c(stype, awards))
```

`n` is the *observed* count: 144 schools of type E were sampled. The
`78.7%` beside it is an *estimate* of the population share, and the two
answer different questions -- which is why the table shows both and
the footer gives the estimated population (6194) beside the sample
size (183).

```{r cat-ci}
table_categorical_svy(
  dclus1,
  select = stype,
  proportion_ci = TRUE,
  deff = TRUE
)
```

The percentage always comes from `svymean()` and the interval from
`svyciprop()`. That is deliberate: `svyciprop()` estimates on the
transformed scale its method names, so a percentage taken from there
would move -- in the thirteenth decimal -- with `ci_method`, which is
a property of the interval and not of the proportion.

```{r cat-by}
table_categorical_svy(dclus1, select = stype, by = sch.wide)
```

The `p` is `svychisq()`: Pearson's chi-square with the Rao-Scott
second-order correction, referred to *F*(ndf, `degf(design)`).
`chisq_statistic` offers `"Chisq"`, `"Wald"`, `"adjWald"` and
`"saddlepoint"` as well. Two of survey's seven are refused:
`"lincom"`, whose numerical integration `?pchisqsum` documents as
failing when the upper tail approaches machine epsilon, and
`"wls-score"`, which has no reporting convention here.

Association measures -- Cramer's *V*, phi, tau-b/c, gamma, Somers'
*D*, lambda -- are absent, and that absence is a decision. None has an
established design-based variance, and the intervals
`table_categorical()` gives them assume simple random sampling. The
design-based measure of association is the test in the `p` column; for
an effect size, model it:
`table_regression(survey::svyglm(...))`.

## Replicate weights

A replicate-weights design goes through the same functions. The point
estimates depend on the weights alone and are therefore identical; the
standard errors are not, and should not be:

```{r rep}
table_continuous_svy(rclus1, select = api00)
```

## Weights, or a design?

This is the question the two families answer differently, and the
difference is an estimand rather than an approximation.

`table_continuous(weights = )` implements the
**frequency-expansion** convention of decision 17: a weight is a
number of copies, and `SD` has denominator `sum(w) - 1`. With integer
weights the statistics are exactly those of the data repeated that
many times.

`table_continuous_svy()` implements the **sampling-weight**
convention: a weight is a number of units represented, and `SD` is
`sqrt(survey::svyvar())`, whose denominator is `n - 1` on weights
normalised to sum to `n`.

`rescale = TRUE` is the bridge between them, and it is an identity.
Writing `w' = w * n / sum(w)`, so that `sum(w') = n`:

```
sum(w' (x - xbar)^2) / (sum(w') - 1)
  = (n / sum(w)) * sum(w (x - xbar)^2) / (n - 1)
  = n / (n - 1) * sum(w (x - xbar)^2) / sum(w)
```

and the right-hand side is what `survey::svyvar()` computes. So on a
design that declares nothing but weights, the two tables agree:

```{r bridge}
d_iid <- svydesign(id = ~1, weights = ~pw, data = apiclus1)

svy <- table_continuous_svy(d_iid, select = api00, output = "long")
wtd <- table_continuous(
  apiclus1, select = api00, weights = pw, rescale = TRUE, output = "long"
)
c(svy = svy$sd, weighted_rescaled = wtd$sd)
```

Under the default `rescale = FALSE` they do not, and that is the
estimand boundary, not a bug:

```{r bridge-default}
expanded <- table_continuous(
  apiclus1, select = api00, weights = pw, output = "long"
)
c(svy = svy$sd, weighted_default = expanded$sd)
```

The *mean* is continuous across both regimes: `sum(w x) / sum(w)` does
not move when the weights are rescaled.

And the whole point of the design functions is that a design carries
more than weights. On the real cluster design, the same mean has a
standard error the weights alone cannot produce:

```{r bridge-se}
c(
  iid_design = table_continuous_svy(d_iid, select = api00, output = "long")$se,
  cluster_design = table_continuous_svy(dclus1, select = api00, output = "long")$se
)
```

Passing a design to `table_continuous()`, or a data frame to
`table_continuous_svy()`, is an error naming the other function. There
is no silent coercion in either direction.

## Quantiles

`qrule` chooses the rule, and the footer always names the one in
force. The default `"math"` estimates the quantile of the *population*,
`inf{x : F(x) >= p}`. `"spicy"` switches to the type-7 interpolation
`table_continuous()` uses, for a reader who needs the two tables to
agree cell for cell; anything else -- including a function -- is
handed to `survey::svyquantile()` untouched.

```{r qrule}
q <- function(rule) {
  table_continuous_svy(
    dclus1, select = api00, qrule = rule,
    show_columns = c("med", "q1", "q3"), output = "long"
  )[, c("q1", "median", "q3")]
}
rbind(math = q("math"), spicy = q("spicy"), hf7 = q("hf7"))
```

## Downstream

Both tables carry the typed view the rest of the package uses, so
`as_structured()` and `inline()` work as they do elsewhere:

```{r structured}
tbl <- table_categorical_svy(dclus1, select = stype, by = sch.wide)
s <- as_structured(tbl)
s$body[, c("Variable", "Yes n", "Yes %", "p", ".row_role")]
names(s$spanners)
```

```{r inline}
inline(tbl, "stype", level = "E", column = "pct", model = "Yes")
inline(tbl, "stype", column = "p")
```

## Rendering

Every engine of the family is available: `"tinytable"`, `"gt"`,
`"flextable"`, `"word"`, `"excel"`, `"clipboard"`, plus
`"data.frame"` / `"long"` for the raw frame.

```{r gt, eval = survey_ok && requireNamespace("gt", quietly = TRUE)}
table_continuous_svy(
  dstrat,
  select = c(api00, api99),
  by = stype,
  output = "gt"
)
```

## What is not here yet

Two-phase, pps, database-backed and multiframe designs are refused
with a classed error rather than approximated: the delegation map was
measured on `svydesign()` and `as.svrepdesign()` designs, and a
plausible wrong number is worse than a refusal. Call survey directly
for those, and say which design you need.

## References

* Lumley, T. (2004). Analysis of complex survey samples. *Journal of
  Statistical Software*, 9(1), 1--19.
* Lumley, T. (2010). *Complex surveys: A guide to analysis using R*.
  John Wiley & Sons.
