---
title: "Summary tables from a survey design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summary tables from a survey design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
survey_ok <- requireNamespace("survey", quietly = TRUE)
knitr::opts_chunk$set(eval = survey_ok)
```

```{r, echo = FALSE, eval = !survey_ok, results = "asis"}
cat("The **survey** package is not installed, so the examples below are not run.\n")
```

When a sample is not a simple random sample -- when it is stratified,
clustered, weighted to a population, or carried by replicate weights --
the mean is still a weighted mean, but everything around it changes.
The standard error depends on the strata and the clusters, the
degrees of freedom are a count of sampling units rather than of rows,
and the tests are corrected for the design. None of that can be
recovered from a column of weights.

spicy has two functions for that case, and they compute nothing
themselves: `table_continuous_svy()` and `table_categorical_svy()`
delegate every number to the **survey** package (Lumley, 2004, 2010)
and put spicy's
restitution layer around it -- the same columns, the same nine output
routes, the same typed view, plus a footer that says what design
produced the numbers.

```{r setup, message = FALSE}
library(spicy)
library(survey)
data(api)
```

Three designs from survey's own `api` data, used throughout:

```{r designs}
dclus1 <- svydesign(id = ~dnum, weights = ~pw, data = apiclus1, fpc = ~fpc)
dstrat <- svydesign(
  id = ~1, strata = ~stype, weights = ~pw, data = apistrat, fpc = ~fpc
)
rclus1 <- as.svrepdesign(dclus1)
```

## Continuous variables

```{r cont-basic}
table_continuous_svy(dclus1, select = c(api00, api99))
```

Three sentences under the table, and each is there because the number
above it cannot be read without them. The design is a one-stage
cluster sample of 15 school districts with a finite population
correction; the standard errors come from Taylor linearisation; the
interval is a *t* interval on 14 degrees of freedom -- the number of
clusters minus one, not the 183 schools.

That last point is worth dwelling on. `confint()` on a survey object
uses the *normal* distribution by default:

```{r cont-df}
m <- svymean(~api00, dclus1)
confint(m)                    # survey's default: normal
confint(m, df = degf(dclus1)) # what the table reports
```

`svyciprop()`, `svyquantile()`, `svyttest()` and `regTermTest()` all
take their degrees of freedom from the design, not from the row
count, and so do the tables -- everywhere. (Each spends them as its
statistic requires: `svyttest()` uses one on the estimated
difference, so its *t* sits on `degf(design) - 1` -- 13 here.) The
footer says so rather than leaving the reader to find out.

### Choosing the statistics

`show_columns` takes the tokens of `table_continuous()`, plus `"se"`
(the design-based standard error) and `"deff"` (the design effect),
and minus `"med_ci"`:

```{r cont-columns}
table_continuous_svy(
  dclus1,
  select = api00,
  show_columns = c("m", "se", "ci", "med_iqr", "n", "weighted_n", "deff"),
  deff = TRUE
)
```

The design effect of 9.35 is the point of the whole exercise: the
variance of this cluster sample is nine times what a simple random
sample of 183 schools would have given. A table that ignored the
design would report a standard error three times too small.

`"med_ci"` is refused rather than approximated. The exact interval
`table_continuous()` reports inverts a binomial sign test, and its
coverage guarantee rests on independent observations -- which a
clustered or stratified sample is not. The estimand (the population
median) is not the problem; the interval construction is. For a
design-based interval on that same estimand, call
`survey::svyquantile(interval.type = )` on the design itself.

### Groups

`by = ` cuts one domain per group. survey recomputes the degrees of
freedom on the sampling units each domain retains, so a grouped table
generally carries a different *df* per row, and the footer gives the
span:

```{r cont-by}
table_continuous_svy(dclus1, select = api00, by = stype, statistic = TRUE)
```

The comparison is one design-based test on the whole design, not a set
of pairwise ones: `svyttest()` with two observed groups, the Wald *F*
of `regTermTest()` with three or more, and `svyranktest()` under
`test = "nonparametric"`. Under a design there is no Welch / Student
distinction -- the variance is the design's -- so `test = "student"`
warns and behaves like the default.

The intervals are referred to those degrees of freedom, which is worth
saying because survey's own shortcut is not: `svyby(vartype = "ci")`
builds its interval with `confint.default`, i.e. a normal quantile, and
so returns a slightly narrower interval than the tables here. The two
differ by the ratio of `qnorm(0.975)` to `qt(0.975, degf(design))` --
about 3% on the 14 degrees of freedom of `dclus1`. Neither is a
mistake; they answer to different reference distributions, and this
family answers to the design's.

## Categorical variables

```{r cat-basic}
table_categorical_svy(dclus1, select = c(stype, awards))
```

`n` is the *observed* count: 144 schools of type E were sampled. The
`78.7%` beside it is an *estimate* of the population share, and the two
answer different questions -- which is why the table shows both and
the footer gives the estimated population (6194) beside the sample
size (183).

```{r cat-ci}
table_categorical_svy(
  dclus1,
  select = stype,
  proportion_ci = TRUE,
  deff = TRUE
)
```

The percentage always comes from `svymean()` and the interval from
`svyciprop()`. That is deliberate: `svyciprop()` estimates on the
transformed scale its method names, so a percentage taken from there
would move -- in the thirteenth decimal -- with `ci_method`, which is
a property of the interval and not of the proportion.

```{r cat-by}
table_categorical_svy(dclus1, select = stype, by = sch.wide)
```

The `p` is `svychisq()`: Pearson's chi-square with the Rao-Scott
second-order correction, referred to *F*(ndf, `degf(design)`).
`chisq_statistic` offers `"Chisq"`, `"Wald"`, `"adjWald"` and
`"saddlepoint"` as well. Two of survey's seven are refused:
`"lincom"`, whose numerical integration `?pchisqsum` documents as
failing when the upper tail approaches machine epsilon, and
`"wls-score"`, which has no reporting convention here.

Association measures -- Cramer's *V*, phi, tau-b/c, gamma, Somers'
*D*, lambda -- are absent, and that absence is a decision. None has an
established design-based variance, and the intervals
`table_categorical()` gives them assume simple random sampling. The
design-based measure of association is the test in the `p` column; for
an effect size, model it:
`table_regression(survey::svyglm(...))`.

## Replicate weights

A replicate-weights design goes through the same functions. The point
estimates depend on the weights alone and are therefore identical; the
standard errors are not, and should not be:

```{r rep}
table_continuous_svy(rclus1, select = api00)
```

## Weights, or a design?

This is the question the two families answer differently, and the
difference is an estimand rather than an approximation.

`table_continuous(weights = )` implements the
**frequency-expansion** convention of decision 17: a weight is a
number of copies, and `SD` has denominator `sum(w) - 1`. With integer
weights the statistics are exactly those of the data repeated that
many times.

`table_continuous_svy()` implements the **sampling-weight**
convention: a weight is a number of units represented, and `SD` is
`sqrt(survey::svyvar())`, whose denominator is `n - 1` on weights
normalised to sum to `n`.

`rescale = TRUE` is the bridge between them, and it is an identity.
Writing `w' = w * n / sum(w)`, so that `sum(w') = n`:

```
sum(w' (x - xbar)^2) / (sum(w') - 1)
  = (n / sum(w)) * sum(w (x - xbar)^2) / (n - 1)
  = n / (n - 1) * sum(w (x - xbar)^2) / sum(w)
```

and the right-hand side is what `survey::svyvar()` computes. So on a
design that declares nothing but weights, the two tables agree:

```{r bridge}
d_iid <- svydesign(id = ~1, weights = ~pw, data = apiclus1)

svy <- table_continuous_svy(d_iid, select = api00, output = "long")
wtd <- table_continuous(
  apiclus1, select = api00, weights = pw, rescale = TRUE, output = "long"
)
c(svy = svy$sd, weighted_rescaled = wtd$sd)
```

Under the default `rescale = FALSE` they do not, and that is the
estimand boundary, not a bug:

```{r bridge-default}
expanded <- table_continuous(
  apiclus1, select = api00, weights = pw, output = "long"
)
c(svy = svy$sd, weighted_default = expanded$sd)
```

The *mean* is continuous across both regimes: `sum(w x) / sum(w)` does
not move when the weights are rescaled.

And the whole point of the design functions is that a design carries
more than weights. On the real cluster design, the same mean has a
standard error the weights alone cannot produce:

```{r bridge-se}
c(
  iid_design = table_continuous_svy(d_iid, select = api00, output = "long")$se,
  cluster_design = table_continuous_svy(dclus1, select = api00, output = "long")$se
)
```

Passing a design to `table_continuous()`, or a data frame to
`table_continuous_svy()`, is an error naming the other function. There
is no silent coercion in either direction.

## Regression under a design

`table_regression()` reads a design-based fit the same way, and
delegates the same way: the coefficients, the variance and the degrees
of freedom all come from survey.

Three model classes are supported, on a linearised and on a
replicate-weights design alike: `survey::svyglm()`, `survey::svyolr()`
for a cumulative-link model, and `survey::svycoxph()` for a
design-weighted Cox model.

```{r reg-glm}
gl <- svyglm(api00 ~ ell + meals + stype, design = dstrat)
table_regression(gl)
```

Three things in that footer are worth naming.

The design line says what the design is and how many degrees of freedom
the table's own tests use. That number is the *model's* residual df --
what survey writes on the fit, and what `survey::regTermTest()` takes as
its denominator -- not the design's own. It is read, never re-derived:
the engines of survey do not share one expression, and a Cox fit ends
one degree of freedom above what the same arithmetic would give for a
`svyglm`. For this linear fit:

```{r reg-df}
c(design = degf(dstrat), model = df.residual(gl))
```

The variance label names the estimator the design actually uses. On a
replicate design there is no linearisation anywhere, and the footer says
so:

```{r reg-rep}
table_regression(
  svyglm(api00 ~ ell, design = rclus1),
  show_columns = c("b", "ci", "p")
)
```

And the counts are both there: the observed `n` and the `Weighted n`
the estimates describe, the same pair the descriptive tables print in
their header.

### Ordinal and Cox fits

A `svyolr()` fit gets its cut-points as a Thresholds block, and its
average marginal effects one column per response category -- averaged
over the population, not over the sample:

```{r reg-olr}
apistrat$grade <- ordered(cut(apistrat$api00, c(0, 600, 700, 1000)))
dg <- svydesign(
  id = ~1, strata = ~stype, weights = ~pw, data = apistrat, fpc = ~fpc
)
table_regression(svyolr(grade ~ ell + stype, design = dg))
```

A `svycoxph()` fit reports hazard ratios, the events, and the
concordance:

```{r reg-cox}
data(pbc, package = "survival")
pbc$randomized <- with(pbc, !is.na(trt) & trt > 0)
bias <- glm(randomized ~ age * edema, data = pbc, family = binomial)
pbc$sw <- 1 / predict(bias, type = "response")
dpbc <- svydesign(
  id = ~1, prob = ~sw, strata = ~edema, data = subset(pbc, randomized)
)
cx <- svycoxph(
  survival::Surv(time, status > 0) ~ log(bili) + protime + albumin,
  design = dpbc
)
table_regression(cx, exponentiate = TRUE, show_columns = c("b", "ci", "p"))
```

### What a design fit refuses

Statistics that need a likelihood are absent rather than approximated:
`AIC()`, `BIC()`, `logLik()` and the pseudo-R-squared family are not
defined for a design-weighted fit, and asking for them by token prints
nothing rather than a number without an interpretation. The one
exception is `svyglm`, whose `AIC` row is the design-based criterion of
Lumley & Scott -- and `show_fit_stats = "eff_p"` reports the effective
number of parameters of the design beside it.

A robust variance is refused for every design class: the design is the
variance authority, and the way to change the estimator is to change the
design.

```{r reg-refuse, error = TRUE}
table_regression(gl, vcov = "HC3")
```

The RMST and risk-difference columns are refused for a design-based Cox
fit. The estimands are not what is in question -- they are validated
against exact oracles for an unweighted Cox model -- but their
uncertainty comes from resampling subjects, and resampling rows ignores
the strata and the clusters the design declares. Use `survey::svykm()`
for a marginal survival curve, and `survey::regTermTest()` to test a
term.

Average marginal effects are refused for a Cox fit, design-based or not:
a proportional-hazards model has no natural response scale to average an
effect on.

## Quantiles

`qrule` chooses the rule, and the footer always names the one in
force. The default `"math"` estimates the quantile of the *population*,
`inf{x : F(x) >= p}`. `"spicy"` switches to the type-7 interpolation
`table_continuous()` uses, for a reader who needs the two tables to
agree cell for cell; anything else -- including a function -- is
handed to `survey::svyquantile()` untouched.

```{r qrule}
q <- function(rule) {
  table_continuous_svy(
    dclus1, select = api00, qrule = rule,
    show_columns = c("med", "q1", "q3"), output = "long"
  )[, c("q1", "median", "q3")]
}
rbind(math = q("math"), spicy = q("spicy"), hf7 = q("hf7"))
```

## Downstream

Both tables carry the typed view the rest of the package uses, so
`as_structured()` and `inline()` work as they do elsewhere:

```{r structured}
tbl <- table_categorical_svy(dclus1, select = stype, by = sch.wide)
s <- as_structured(tbl)
s$body[, c("Variable", "Yes n", "Yes %", "p", ".row_role")]
names(s$spanners)
```

```{r inline}
inline(tbl, "stype", level = "E", column = "pct", model = "Yes")
inline(tbl, "stype", column = "p")
```

## Rendering

Every engine of the family is available: `"tinytable"`, `"gt"`,
`"flextable"`, `"word"`, `"excel"`, `"clipboard"`, plus
`"data.frame"` / `"long"` for the raw frame.

```{r gt, eval = survey_ok && requireNamespace("gt", quietly = TRUE)}
table_continuous_svy(
  dstrat,
  select = c(api00, api99),
  by = stype,
  output = "gt"
)
```

## What is not here yet

Two-phase, pps, database-backed and multiframe designs are refused
with a classed error rather than approximated: the delegation map was
measured on `svydesign()` and `as.svrepdesign()` designs, and a
plausible wrong number is worse than a refusal. Call survey directly
for those -- and if one of them matters for your work, an issue on the
package's tracker saying which design you need is the fastest way to
move it up the roadmap.

## References

* Lumley, T. (2004). Analysis of complex survey samples. *Journal of
  Statistical Software*, 9(1), 1--19.
* Lumley, T. (2010). *Complex surveys: A guide to analysis using R*.
  John Wiley & Sons.
* Lumley, T., & Scott, A. (2015). AIC and BIC for modeling with complex
  survey data. *Journal of Survey Statistics and Methodology*, 3(1),
  1--18.
