---
title: "Multinomial regression tables"
description: >
  Publication-ready tables for multinomial logistic regression in R with
  `table_regression()`: outcome categories as columns, odds ratios (what
  Stata labels relative-risk ratios), per-category average marginal
  effects, changing the reference category, nested comparison by
  likelihood-ratio test, and the `nnet::multinom` engine plus the
  `mlogit` discrete-choice engine.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multinomial regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_nnet <- requireNamespace("nnet", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_nnet
)
# One column group per outcome category makes wide tables -- that is
# the point of the layout; let the console render them unwrapped.
options(width = 130)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_mlogit <- have_nnet && requireNamespace("mlogit", quietly = TRUE)
have_broom  <- have_nnet && requireNamespace("broom", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(nnet)
```

This vignette covers **multinomial logistic regression** — models for a
*nominal* outcome, a categorical response whose categories have no
natural order: employment status, party choice, transport mode. The
companion vignette [*Publication-ready regression
tables*](table-regression.html) covers the shared mechanics (`vcov`,
`ci_level`, output formats, multi-model layouts, broom integration),
and [*Ordinal regression tables*](table-regression-ordinal.html) covers
the case where the categories *are* ordered. Order alone does not
decide the model: the ordinal model buys parsimony — one slope per
predictor instead of $J-1$ — at the price of the parallel-regressions
constraint, and when that constraint is implausible, the multinomial
model is the standard unconstrained alternative (Long & Freese 2014),
at the cost of extra coefficients and category-by-category
interpretation. An approximate check compares the two fits directly by
likelihood ratio or AIC. Here we focus on what is
specific to nominal outcomes: one column of effects per outcome
category, the choice of reference category, and why a coefficient's
sign need not be the direction of its effect on a probability.

`table_regression()` supports two engines:

* **`nnet::multinom()`** — the baseline-category logit model on
  characteristics of the *person* (case-specific covariates); the
  social-science standard.
* **`mlogit::mlogit()`** — McFadden's conditional logit on
  characteristics of the *alternatives* (discrete-choice data).

The running example is employment status in the bundled `sochealth`
survey (`?sochealth`) — four categories with `Employed` by far the most
frequent:

```{r data}
d <- sochealth
table(d$employment_status)
```

One recoding first. `education` is stored as an *ordered* factor, and R
gives ordered factors polynomial contrasts by default — the model would
estimate `.L` (linear) and `.Q` (quadratic) components that fit
identically but are awkward to report. Applied tables conventionally
show treatment dummies against a reference level (Fox & Weisberg 2019),
so we convert to an unordered factor, preserving the level
sequence:

```{r recode}
d$education <- factor(as.character(d$education),
                      levels = c("Lower secondary", "Upper secondary",
                                 "Tertiary"))
```

## The baseline-category logit model in one paragraph

For an outcome with $J$ categories, the model fits $J-1$ logit
equations simultaneously by maximum likelihood, each comparing one
category against a common reference (the *baseline*):
$\log[P(Y = j)/P(Y = J)] = \alpha_j + x^\top\beta_j$ (Agresti 2013, ch. 8).
Two consequences shape the table. Each predictor gets **one coefficient
per non-reference outcome** — there is no single "effect of education",
only its effect on each comparison — and the $J-1$ fitted equations
determine the logit for *every* pair of categories, so nothing is lost
by picking one baseline. Fitting separate binary logits instead of the
simultaneous model is consistent but less efficient; the loss is minor
when the baseline is the most frequent category (Begg & Gray 1984). In
the simultaneous fit the baseline changes nothing about the model, but
it does set which contrasts the table displays — and contrasts against
a well-populated category are precisely estimated, which is why Stata's
`mlogit` defaults to the most frequent outcome. `multinom()` uses the
factor's first level — here `Employed`, which is also the most
frequent, so R's positional default coincides with the frequency-based
choice: large, so the displayed contrasts are precise, and
substantively meaningful.

## Basic table

`trace = FALSE` only silences the optimizer's iteration log; unlike
`polr()`, `multinom()` needs no `Hess = TRUE`.

```{r basic}
fit <- multinom(employment_status ~ age + sex + education,
                data = d, trace = FALSE)
table_regression(fit)
```

This is the layout multinomial results are published in: predictors as
rows, **one column group per non-reference outcome category**, so a
predictor's effects on the $J-1$ comparisons sit side by side on its
row (Long & Freese 2014) — the arrangement most journals print. Raw
`mlogit` and NOMREG output instead stack the same $J-1$ equations as
row blocks; side by side is the publication convention. Reading it:

* Each column group is one **equation against the reference**: the
  `Tertiary` cell under `Unemployed` is the tertiary-education
  coefficient in the *Unemployed-versus-Employed* logit. The footer
  names the baseline once — `Reference outcome: Employed.` — Stata's
  "base outcome" line; every cell in the table answers "compared to
  `Employed`".
* The table is wide by construction, so the default trims each group
  to **B, SE, p** — the same compaction applied to multi-model tables.
  Atomic tokens restore anything, e.g.
  `show_columns = c("b", "ci", "p")` for CIs.
* To label the comparisons explicitly — the style some journals ask
  for — relabel the spanners:
  `outcome_labels = c("Student vs Employed", "Unemployed vs Employed",
  "Inactive vs Employed")`.
* The `(Intercept)` row is one intercept per equation — the log-odds
  of each category against `Employed` at the reference levels and
  age 0. It anchors the equations rather than carrying substantive
  interest.
* Inference is **Wald-z** (`df = Inf`), matching `summary()`-based
  practice for ML fits and Stata `mlogit`; `nnet` itself prints no
  p-values.
* The model-fit block reports **n** and **AIC**, once, under the first
  group — they belong to the model, not to an equation.

Before reading any cell, the whole model earns its keep: against the
intercept-only fit, the likelihood-ratio chi-squared is 41.9 on 12
degrees of freedom (p < .001) — the test Stata prints in its `mlogit`
header and the first number Long & Freese read. McFadden's pseudo-R²
is 1 − (−1242.9)/(−1263.8) = 0.017: a comparative index, not a share
of explained variance (Long & Freese 2014) — the contrast between the
tiny value and the decisive test is exactly why it should not be read
as one.

One scope note: `multinom` fits report **classical Wald-z inference
only** — a robust or cluster-robust `vcov` request is refused with an
explanatory error rather than silently ignored.

## Odds ratios — and what Stata calls them

`exponentiate = TRUE` puts the estimates and CI bounds on the ratio
scale:

```{r or}
table_regression(fit, exponentiate = TRUE, show_columns = c("b", "ci", "p"))
```

Each predictor cell is an **odds ratio against the reference outcome**
(the exponentiated intercepts are baseline *odds* — each category's
odds versus `Employed` at the reference levels and age 0 — not
ratios): compared with lower-secondary graduates of the same age and
sex, tertiary graduates' odds of being unemployed rather than employed
are 0.29 times as large — a 71% reduction in those odds, the education
gradient the table is really about. Formally it is the odds ratio of the *conditional*
distribution restricted to the two categories in the comparison
(Agresti 2013).

One terminology note, because readers will meet it: Stata's
`mlogit, rrr` labels these same numbers **relative-risk ratios** —
$e^\beta$ is also the multiplicative change in the probability ratio
of outcome $j$ versus the baseline — while SAS prints "Odds Ratio
Estimates" for its generalized logits. Same quantity, two readings;
`spicy` labels it `OR`, the reading that generalizes from binary
logistic regression.

## Changing the reference category

The reference is a *presentation* choice, not a modeling one: refit
with another baseline and it is the same model, reparameterized —
same likelihood, same AIC, same fitted probabilities; only the
comparisons displayed change.

```{r relevel}
d2 <- d
d2$employment_status <- relevel(d2$employment_status, ref = "Unemployed")
fit_unemp <- multinom(employment_status ~ age + sex + education,
                      data = d2, trace = FALSE)
c(AIC(fit), AIC(fit_unemp))
table_regression(fit_unemp, exponentiate = TRUE, show_columns = c("b", "p"))
```

Against `Employed`, education looked irrelevant for the `Student`
contrast; against `Unemployed`, the whole `Tertiary` row is strong
(OR 3.94, p < .001 in the `Student` group; even the `Inactive` group
reaches p = .005; only `Upper secondary` in the `Inactive` group stays
inconclusive). The p-values moved because the *questions* changed, not
the model — a multinomial table only ever shows $J-1$ of the
$J(J-1)/2$ pairwise comparisons, and any of the others can be
recovered by releveling (Agresti 2013; Long 1997). Report the baseline
that makes your substantive comparisons direct.

## Does a predictor matter at all?

The per-cell p-values just moved with the baseline, so none of them
answers the question a reviewer asks first. The baseline-invariant
answer is the **joint likelihood-ratio test** that all $J-1$
coefficients of a predictor are zero — the first testing step of Long
& Freese (2014):

```{r joint-lrt}
fit_noeduc <- multinom(employment_status ~ age + sex, data = d,
                       trace = FALSE)
anova(fit_noeduc, fit)
```

Education matters decisively — chi-squared 36.29 on 6 degrees of
freedom (2 parameters in each of 3 equations), p < .001 — even though
its `Student`-equation cells looked unconvincing one table ago. The
trap runs in both directions: sex shows one tempting cell
(`Unemployed`, p = .172) but is jointly null (chi-squared 2.26 on
3 df, p = .520). Individually weak cells can be jointly overwhelming,
and one suggestive cell among $J-1$ can be noise. `car::Anova(fit)`
runs this test for every predictor in one call — the equivalent of
Stata's `mlogtest, lr`.

## Can two categories be combined?

Before deciding how to report $J$ categories, the textbook workflow
asks whether all $J$ are distinguishable at all (Long & Freese 2014):
if no covariate separates outcomes $m$ and $n$, the data cannot tell
them apart, and the model may be estimating noise between them. The
likelihood-ratio version restricts the sample to the two categories
and tests a binary logit against the intercept-only model (Cramer &
Ridder 1991):

```{r combine}
sub <- droplevels(subset(d, employment_status %in% c("Student", "Inactive")))
y <- as.integer(sub$employment_status == "Student")
anova(glm(y ~ 1, binomial(), sub),
      glm(y ~ age + sex + education, binomial(), sub), test = "LRT")
```

`Student` and `Inactive` are not separated — chi-squared 6.95 on 4
degrees of freedom, p = .139. Running all six pairs tells the honest
story of these data: only `Unemployed` is sharply distinguished from
the rest (versus `Employed`, chi-squared 33.6; versus `Student`, 22.6;
both p < .001), while `Employed`, `Student`, and `Inactive` are not
mutually separated by age, sex, and education. Two cautions close the
step: failing to reject does not prove interchangeability — the
smaller cells give the test little power — and combining is ultimately
a substantive decision: employment states remain distinct constructs
even when these covariates do not separate them.

## The sign trap, and marginal effects

The most common misreading of a multinomial table is to treat a
coefficient's sign as the direction of the effect on that category's
*probability*. The two need not agree: for a continuous $x_k$, the
marginal effect on $P(Y=m)$ is
$P_m(\beta_{k,m} - \sum_{j=1}^{J} \beta_{k,j} P_j)$, with
$\beta_{k,J} = 0$ for the baseline so the sum runs over all $J$
categories — the coefficient *minus a probability-weighted average of
every category's coefficient* — so its sign can differ from the
coefficient's, and can even change across the range of $x_k$ (Long &
Freese 2014; Wulff 2015). For a factor, the AME below is instead the
average difference in predicted probabilities between levels; the same
competition logic applies. The categories compete for probability
mass.

The probability-scale summary is the per-category **average marginal
effect** (AME): one AME column per outcome category, next to that
category's coefficients:

```{r ame}
table_regression(fit, show_columns = c("b", "ame"))
```

Three things to read off, using the `Tertiary` row:

* The **reference category is back** — as a last, AME-only column
  group. `Employed` has no coefficients (it is the baseline of every
  equation, so its `B` cells are empty), yet it carries the largest
  *positive* probability effect in the table (+0.14 for `Tertiary`,
  second in magnitude only to `Unemployed`'s −0.16). The log-odds side
  never shows this; the probability side must, because the effects on
  all $J$ categories **sum to exactly zero** across each row (the $J$
  probabilities sum to 1, so whatever probability a predictor moves
  out of one category must land in the others; displayed values can
  miss 0 only by rounding) — the mass that tertiary education moves
  out of `Unemployed` lands mainly in `Employed`.
* Cells are **on the probability scale**: −0.16 is a drop of 16
  **percentage points** in the probability of unemployment for
  tertiary relative to lower-secondary education, averaged over the
  observed covariates (`marginaleffects::avg_slopes()`; Williams
  2012). Add `"ame_se"`, `"ame_ci"`, or `"ame_p"` to `show_columns`
  for delta-method inference on these effects.
* Here every AME happens to share its coefficient's sign; do not count
  on it. When an outcome's coefficient sits close to the
  probability-weighted average of all the coefficients, the
  probability effect can flip sign — the reason Wulff (2015)
  recommends never interpreting a multinomial model from its
  coefficient signs alone.

An AME is an average of *changes*; the concrete interpretation device
of Long & Freese (2014) is the **predicted probabilities** those
changes move between, evaluated at named profiles:

```{r profiles}
profiles <- data.frame(
  education = levels(d$education),
  sex       = "Female",
  age       = mean(d$age[!is.na(d$employment_status)])
)
round(predict(fit, newdata = profiles, type = "probs"), 3)
```

Read the `Unemployed` column down the education gradient: 0.235,
0.121, 0.084. For this profile — a woman of average age — the −0.16
AME is a fall from roughly one in four to under one in ten; the
baseline matters as much as the change. (The profile's own gap,
−0.151, sits close to the sample-averaged −0.160 — they answer
slightly different questions.) Base `predict()` returns no standard
errors for these predictions;
`marginaleffects::avg_predictions()` supplies delta-method inference
on the same quantities.

## An ordinal predictor: scores or dummies?

`education` is ordered — Lower secondary < Upper secondary < Tertiary —
and the dummy coding above ignores that. The alternative is to enter it
as **numeric scores** (1, 2, 3): one slope per equation instead of two
dummies, a model that is both simpler and nested in the factor model.
Agresti (2007, sec. 4.4.3) gives the workflow: fit both codings and let
a likelihood-ratio test judge whether the category coefficients depart
from a linear trend in the scores. `nested = TRUE` runs exactly that
comparison:

```{r scores}
d$educ_score <- as.numeric(d$education)   # 1 / 2 / 3
fit_lin <- multinom(employment_status ~ age + sex + educ_score,
                    data = d, trace = FALSE)
table_regression(list(fit_lin, fit), nested = TRUE,
                 show_columns = c("b", "p"))
```

The change rows read: freeing education from the linear trend buys a
chi-squared of 4.52 on 3 degrees of freedom (one per equation),
p = .211 — no evidence the dummies improve on the scores, and AIC
agrees (2514.3 vs 2515.8). On Agresti's grounds the scores model is
preferable when it fits: simpler to report, and more powerful against
a genuine trend, because it concentrates that trend in one parameter
per equation instead of two — an advantage that reverses if the true
pattern is non-monotone (Agresti 2007).

Three caveats keep the decision honest. A non-significant test does
not *prove* linearity — in modest samples several codings fit
adequately (Agresti 2007). Integer scores assume the steps are equally
spaced, which is a substantive claim about education levels, not a
statistical default (Agresti 2007, sec. 2.5.4). And testing first,
then reporting the scores model as if it had been chosen in advance,
inflates the Type I error of the subsequent test (Grambsch & O'Brien
1991). With only three levels the dummies cost one extra parameter per
equation, so Harrell (2015) recommends simply keeping them. Present the LRT as a transparent simplification decision, not a
proof.

## Independence of irrelevant alternatives, briefly

The multinomial logit assumes the odds between any two categories do
not depend on what other alternatives exist — the *independence of
irrelevant alternatives* (IIA), of red-bus/blue-bus fame (the critique
goes back to Debreu 1960). Formal tests exist (Hausman & McFadden
1984; Small & Hsiao 1985), but simulation evidence shows them to be
unsatisfactory for applied work — different tests reach conflicting
verdicts on the same data (Cheng & Long 2007) — and Long & Freese
(2014) do not recommend them. The working advice remains McFadden's
(1974): use the model when the alternatives are "distinct and weighed
independently" — not close substitutes. For a fixed, exhaustive set
like employment status, where no alternative is ever added or removed,
IIA is rarely the practical concern it is in choice modeling; where
alternatives genuinely substitute (two bus lines), the remedy is a
different model (nested or mixed logit), not a mechanical test.

## Discrete choice: `mlogit`

`multinom()` models the chooser; McFadden's **conditional logit**
models the *choices*. When covariates describe the alternatives —
the price of each travel mode, the catch rate of each fishing mode —
the coefficient is one per *attribute*, shared across alternatives
(McFadden 1974; Croissant 2020). The `mlogit` engine renders both
kinds of covariate in one table. The classic `Fishing` data: 1182
anglers choose among four modes; `price` and `catch` vary by
alternative, while `income_k` (income in thousands of dollars)
describes the angler:

```{r mlogit, eval = have_mlogit, message = FALSE}
library(mlogit)
data("Fishing", package = "mlogit")
Fishing$income_k <- Fishing$income / 1000
Fish <- dfidx(Fishing, varying = 2:9, choice = "mode", shape = "wide")

fit_choice <- mlogit(mode ~ price + catch | income_k, data = Fish)
table_regression(fit_choice, exponentiate = TRUE,
                 show_columns = c("b", "ci", "p"))
```

The table comes in **two segments**, the presentation Stata's
`asclogit` prints. The `Alternative-invariant` section holds the
model's centerpiece: the attribute coefficients (`price`, `catch`),
one row each, shared across alternatives — one more dollar on a mode's
own price multiplies the odds of choosing that mode over any other
alternative by 0.98, other modes' prices held fixed, *whichever mode
it is*. Then one section per non-reference alternative (`boat`,
`charter`, `pier`) collects that alternative's intercept and its
**case-specific** coefficients: `income_k` enters as it would under
`multinom()`, one coefficient per comparison against the reference —
the footer's `Reference alternative: beach.` names it, so each section
reads "this alternative versus beach". The footer's `n` counts **choice
situations** (1182 anglers), not the 4728 rows of the long-format
data: each angler contributes one choice, and that is the model's
sample size.

Three guardrails: average marginal effects are refused for `mlogit`
(`marginaleffects` has no slopes method for its one-row-per-choice
structure); the robust `vcov` family is `CR0`–`CR3` only, with one
cluster value per choice situation, not per long-format row; and
`HC*` is refused outright — `sandwich::vcovHC()` mis-scales the
sandwich for mlogit's per-choice-situation scores.

## Several models side by side

Multinomial fits can be compared like fits of any other model class —
here the education gradient before and after adjustment. With several
models the column groups are needed for the *models*, so the table
falls back to the one-row-per-comparison layout (`Student: Tertiary`
rows), as does `nested = TRUE` above:

```{r multi}
fit_min <- multinom(employment_status ~ education, data = d, trace = FALSE)
table_regression(list(Unadjusted = fit_min, Adjusted = fit),
                 exponentiate = TRUE, show_columns = c("b", "p"))
```

Adjustment changes nothing here: the education block is identical to
two decimals across the columns — `Tertiary` is OR 0.29 in the
`Unemployed`-versus-`Employed` equation both times — because age and
sex, at best weakly related to employment status in this sample (their
smallest p is .097), have nothing to confound the education gradient
with.

## Output formats

Everything above used the default console output. The same table
renders as a raw data frame, a long broom-style tibble, or — with the
corresponding Suggests package — a rich `gt`, `flextable`,
`tinytable`, Excel, or Word table:

```{r outputs-df}
head(table_regression(fit, exponentiate = TRUE, output = "data.frame"))
```

```{r outputs-gt, eval = build_rich_tables && have_nnet}
table_regression(fit, show_columns = c("b", "ame"), output = "gt")
```

[`broom::tidy()`](https://broom.tidymodels.org) always returns the
long frame, one row per `(term, estimate_type)`, whatever the display
layout — the per-outcome structure is carried in the term labels
(`"Student: age"`), and AME rows cover all $J$ categories including
the reference:

```{r broom, eval = have_broom}
broom::tidy(table_regression(fit, show_columns = c("b", "ame")))
```

## References

* Agresti, A. (2007). *An Introduction to Categorical Data Analysis*
  (2nd ed.). Wiley.
* Agresti, A. (2013). *Categorical Data Analysis* (3rd ed.). Wiley.
* Begg, C. B., & Gray, R. (1984). Calculation of polychotomous
  logistic regression parameters using individualized regressions.
  *Biometrika*, 71(1), 11–18.
* Cheng, S., & Long, J. S. (2007). Testing for IIA in the multinomial
  logit model. *Sociological Methods & Research*, 35(4), 583–600.
* Croissant, Y. (2020). Estimation of random utility models in R: The
  mlogit package. *Journal of Statistical Software*, 95(11), 1–41.
* Debreu, G. (1960). Review of R. D. Luce, *Individual Choice
  Behavior*. *American Economic Review*, 50(1), 186–188.
* Fox, J., & Weisberg, S. (2019). *An R Companion to Applied
  Regression* (3rd ed.). Sage.
* Grambsch, P. M., & O'Brien, P. C. (1991). The effects of
  transformations and preliminary tests for non-linearity in
  regression. *Statistics in Medicine*, 10(5), 697–709.
* Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.).
  Springer.
* Hausman, J. A., & McFadden, D. (1984). Specification tests for the
  multinomial logit model. *Econometrica*, 52(5), 1219–1240.
* Long, J. S. (1997). *Regression Models for Categorical and Limited
  Dependent Variables*. Sage.
* Long, J. S., & Freese, J. (2014). *Regression Models for Categorical
  Dependent Variables Using Stata* (3rd ed.). Stata Press.
* McFadden, D. (1974). Conditional logit analysis of qualitative
  choice behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics*
  (pp. 105–142). Academic Press.
* Small, K. A., & Hsiao, C. (1985). Multinomial logit specification
  tests. *International Economic Review*, 26(3), 619–627.
* Williams, R. (2012). Using the margins command to estimate and
  interpret adjusted predictions and marginal effects. *The Stata
  Journal*, 12(2), 308–331.
* Wulff, J. N. (2015). Interpreting results from the multinomial logit
  model: Demonstrated by foreign market entry. *Organizational
  Research Methods*, 18(2), 300–325.
