---
title: "Categorical summary tables"
description: >
  Build categorical summary tables in R with table_categorical(),
  including grouped cross-tabulations, effect sizes, confidence
  intervals, and export to gt, tinytable, flextable, Excel, or Word.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Categorical summary tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

`table_categorical()` builds publication-ready categorical tables for
reports and articles in social science and data science research. With
`by`, it produces grouped cross-tabulation tables with chi-squared
\(p\)-values, effect sizes, confidence intervals, and multi-level
headers. Without `by`, it produces one-way frequency-style tables for
the selected variables. Export to gt, tinytable, flextable, Excel, or
Word. This vignette walks through the main features.

## Basic usage

For grouped tables, provide a data frame, one or more selected
variables, and a grouping variable:

```{r basic}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education
)
```

The default output is `"default"`, which prints a styled ASCII table
to the console, folded into successive panels when it is wider than
the console. Reading the table: each level of `by` gets a pair of
columns, the count `n` and the **column percentage** computed within
that group -- percentages sum to 100 down each education group (for
smoking within "Lower secondary": 68.6 + 29.9 + 1.5 = 100.0) -- and
the `Total` pair is the margin over all groups combined. In the last
panel, `p` is the *p*-value of the chi-squared test of association
between the row variable and `by`, computed once per variable on its
full cross-table, and the final column is the association measure --
here Cramer's V (covered below). Use `output = "data.frame"` for a
plain data frame suitable for further processing.

## One-way tables

Omit `by` to build a frequency-style table for the selected variables:

```{r oneway}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity)
)
```

`select` itself is optional: when omitted, the table includes every
eligible categorical column in the data -- factor, character,
logical, and labelled (haven) columns -- excluding the `by` column,
mirroring the select-less defaults of `table_continuous()` and
`table_continuous_lm()`. An explicit `select` is taken verbatim, so
numeric-coded categorical variables can still be tabulated by naming
them.

## Output formats

`table_categorical()` supports several output formats. The table below
summarizes the options:

| Format | Description |
|---|---|
| `"default"` | Styled ASCII table in the console (default) |
| `"data.frame"` | Wide data frame, one row per level |
| `"long"` | Long data frame, one row per level x group |
| `"gt"` | Formatted gt table |
| `"tinytable"` | Formatted tinytable |
| `"flextable"` | Formatted flextable |
| `"excel"` | Excel file (requires `excel_path`) |
| `"clipboard"` | Copy to clipboard |
| `"word"` | Word document (requires `word_path`) |

The three rendered formats (`"gt"`, `"tinytable"`, `"flextable"`) are
demonstrated in the final section, together with the file exports.

### Data frame output

Use `output = "data.frame"` for a wide data frame with one row per
level: two character identifier columns (`Variable`, `Level`),
numeric columns carrying the counts and full-precision percentages,
and the chi-squared statistic, `df`, *p*-value, and association
measure repeated on every row of a variable's block. Use
`output = "long"` for a long format with one row per level x group,
the `Total` margin included as a group:

```{r data-frame}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  output = "data.frame"
)
```

## Custom labels

By default, `table_categorical()` labels each row block with the
variable's label attribute when one is present (e.g. data imported
with `haven`), and with the column name otherwise -- the same
auto-detection as `table_continuous()` and `table_continuous_lm()`.
Use the `labels` argument to override either: a **named character
vector** keyed by column name in `data`. Only listed columns are
relabelled; the others keep their attribute label or column name.
(Unnamed positional label vectors, accepted before 0.13.0, now raise
an error.)

```{r labels}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  labels = c(
    smoking           = "Smoking status",
    physical_activity = "Regular physical activity"
  )
)
```

## Association measures and confidence intervals

`table_categorical()` picks the association measure per row variable
based on the variable type (`assoc_measure = "auto"`, the default):

* **2x2** (binary row variable vs. binary `by`) -> `phi`,
* both ordered factors -> Kendall's `tau_b`,
* otherwise -> Cramer's `V`.

When the chosen measures differ across rows, the column header
collapses to `"Effect size"` and an APA-style `Note.` line documents
which measure was used for each variable.

Seven measures are available: `"cramer_v"`, `"phi"`, `"gamma"`,
`"tau_b"`, `"tau_c"`, `"somers_d"`, and `"lambda"`; `assoc_measure =
"none"` drops the column entirely. See `?table_categorical` for the
dispatch details and `vignette("association-measures", package =
"spicy")` for definitions and guidance on choosing among them.
Override the automatic choice with a single string for uniform
application, or with a named vector to mix measures per row:

```{r assoc-measure}
# Uniform: same measure for every row variable
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  assoc_measure = "lambda"
)
```

A significant chi-squared *p*-value next to a lambda of exactly .00
is not a contradiction. Goodman-Kruskal lambda measures the
proportional reduction in the error of predicting the row variable
once the group is known, and it is exactly 0 whenever the modal
category is the same in every group: here "No" is the most frequent
answer at all three education levels (68.6%, 77.0%, 83.0%), so
knowing education never changes the best single guess, even though
the distributions clearly differ (hence the significant chi-squared
test).

In a named vector, variables you do not name keep the `"auto"`
choice, so only the overrides need to be listed. Here `"auto"` would
pick Cramer's V for `smoking` (binary x ordered, not 2x2) and
Kendall's Tau-b for `self_rated_health` (ordered x ordered); the
named vector keeps the former and replaces the latter with
Goodman-Kruskal Gamma:

```{r assoc-measure-named}
# Named vector: override "auto" for one variable only
table_categorical(
  sochealth,
  select = c(smoking, self_rated_health),
  by = education,
  assoc_measure = c(self_rated_health = "gamma")
)
```

`smoking` keeps its automatic Cramer's V (.14) while
`self_rated_health` switches from the automatic Tau-b (.20) to Gamma
(.31) -- Gamma is larger on the same table because it ignores tied
pairs. Since the two measures differ, the header collapses to
`"Effect size"` and the `Note.` line documents the per-row choice.

Add confidence intervals with `assoc_ci = TRUE`. In the rendered
formats (`gt`, `tinytable`, `flextable`, `word`) the CI is shown
inline after the measure, e.g. `.14 [.08, .19]` (demonstrated in the
final section). In the default console table and in the wide data
formats (`"data.frame"`, `"excel"`, `"clipboard"`), separate
`CI lower` and `CI upper` columns are added; in the long format
(`"long"`) the bounds appear as `ci_lower` / `ci_upper`:

```{r ci-data}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  assoc_ci = TRUE,
  output = "data.frame"
)
```

## Weighted tables

Pass survey weights with the `weights` argument. By default
(`rescale = FALSE`) the weights are used as-is; `rescale = TRUE`
rescales them so the total weighted N equals the number of
observations (here 1200, against a raw weight sum of 1196.474).
Displayed counts are weighted counts rounded to integers at display
time -- the SPSS Crosstabs convention -- while the machine formats
(`"data.frame"`, `"long"`) carry the exact fractional weighted
counts:

```{r weighted}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  weights = "weight",
  rescale = TRUE
)
```

## Handling missing values

By default, missing values are displayed as a "(Missing)" category
(`drop_na = FALSE`), so the percentages sum over every observation.
Set `drop_na = TRUE` to remove them. The removal happens before each
cross-tabulation, variable by variable: each row variable keeps its
own complete cases (on itself and on `by`), so the total N can differ
across variables within the same table -- this is available-case
analysis per cross-table, not listwise deletion over the whole
selection. The removal is disclosed in a table note rather than
silent:

```{r missing}
table_categorical(
  sochealth,
  select = income_group,
  by = education,
  drop_na = TRUE
)
```

With several selected variables the note lists each variable's
removals separately: selecting `smoking` and `physical_activity`
together under `drop_na = TRUE` keeps 1175 and 1200 complete cases
respectively, in the same table.

## Filtering and reordering levels

Use `levels_keep` to display only specific levels. The order you
specify controls the display order, which is useful for placing
"(Missing)" first to highlight missingness:

```{r levels-keep}
table_categorical(
  sochealth,
  select = income_group,
  by = education,
  drop_na = FALSE,
  levels_keep = c("(Missing)", "Low", "High")
)
```

`levels_keep` filters the display only: counts, percentages, the
chi-squared test, and the association measure are all still computed
on the full cross-table. That is why the displayed percentages no
longer sum to 100 (within "Lower secondary": 1.1 + 33.3 + 8.0 =
42.4 -- the hidden "Lower middle" and "Upper middle" levels still
count in the denominator), and why *p* and Tau-b are identical to the
unfiltered table. To recompute the statistics on a subset of levels,
filter the data before calling `table_categorical()`.

## Formatting options

Control the number of digits for percentages, p-values, and the
association measure:

```{r formatting}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  percent_digits = 2,
  p_digits = 4,
  v_digits = 3
)
```

`p_digits` drives both the displayed precision of the `p` column and
the small-*p* threshold (`p_digits = 3` -> `<.001`, `p_digits = 4`
-> `<.0001`), matching `table_continuous()` and
`table_continuous_lm()`.

## Decimal alignment

By default (`align = "decimal"`) numeric columns are aligned on the
decimal mark, the standard scientific-publication convention
(SPSS, SAS, LaTeX `siunitx`). Numeric cells are pre-padded with
figure-spaces (U+2007, spaces exactly one digit wide) so that every
string in a column has the same width with the decimal mark at the
same internal position; centring those uniform-width strings then
stacks the decimal points vertically. The same pad-then-centre
strategy is applied on every rendering engine (`gt`, `tinytable`,
`flextable`, `word`, ASCII print) for a homogeneous rendering --
same single-font policy as `table_regression()`. The clipboard is
the exception: its payload is delimited text meant to be parsed, and
a padded number would paste as text next to an unpadded number, so
its cells travel unpadded. The native
`gt::cols_align_decimal()` and `tinytable::style_tt(align = "d")`
primitives are deliberately not used: the former renders visually
right-aligned and the latter centres each cell on its own value
rather than on the decimal mark, which would be inconsistent with
the other engines.

`"center"` and `"right"` apply literal alignment:

```{r align}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = sex,
  align = "right"
)
```

## Tidying for downstream pipelines

`table_categorical()` returns an object that can be coerced to a
plain `data.frame` / `tbl_df` (stripping the spicy formatting
attributes) or piped into `broom::tidy()` / `broom::glance()` for
any downstream tidyverse-stats workflow:

```{r tidy-glance}
out <- table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = sex
)

# One row per (variable x level x group) with broom-style columns
# (outcome, level, group, n, proportion). The synthetic Total
# margin is excluded so each observation is counted once.
broom::tidy(out)

# One row per outcome with the omnibus chi-squared test and the
# chosen association measure (test_type, statistic, df, p.value,
# assoc_type, assoc_value, assoc_ci_lower / assoc_ci_upper, n_total).
broom::glance(out)
```

## Rendered outputs and export

The rendered formats produce publication-ready tables for HTML and
Word workflows. The `"gt"` format produces a table with APA-style
borders, column spanners, and decimal alignment:

```{r gt, eval = build_rich_tables}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = c(smoking, physical_activity, dentist_12m),
    by = education,
    output = "gt"
  )
)
```

The `"tinytable"` format applies the same layout conventions through
the lightweight tinytable engine (here a two-group table by sex):

```{r tinytable, eval = build_rich_tables}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = sex,
  output = "tinytable"
)
```

With `assoc_ci = TRUE`, the rendered formats show the confidence
interval inline after the association measure:

```{r ci-rendered, eval = build_rich_tables}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = c(smoking, physical_activity),
    by = education,
    assoc_ci = TRUE,
    output = "gt"
  )
)
```

For Excel export, provide a file path:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  output = "excel",
  excel_path = "my_table.xlsx"
)
```

For Word, use `output = "word"`:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  output = "word",
  word_path = "my_table.docx"
)
```

You can also copy directly to the clipboard for pasting into a
spreadsheet or a text editor:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  output = "clipboard"
)
```

## See also

- See `vignette("table-continuous", package = "spicy")` for the
  continuous counterpart with classical group-comparison tests.
- See `vignette("table-continuous-lm", package = "spicy")` for
  model-based continuous summary tables.
- See `vignette("table-regression", package = "spicy")` for the
  full coefficient table from one or several fitted `lm` / `glm`
  models (APA Table 3).
- See `vignette("summary-tables-reporting", package = "spicy")` for
  a cross-function reporting workflow that ties the four
  summary-table helpers together along the APA Table 1 / 2 / 3
  sequence.
