---
title: "Publication-ready regression tables"
description: >
  Coefficient tables for `lm` and `glm` fits with APA-aligned
  formatting, classical / heteroskedasticity-consistent /
  cluster-robust / bootstrap / jackknife variance, Average
  Marginal Effects with Satterthwaite-corrected degrees of
  freedom under cluster-robust variance, four standardisation
  methods, per-coefficient effect sizes with noncentral CIs,
  multiple-comparison adjustment, focal-predictor filtering,
  side-by-side and hierarchical layouts, and output to console,
  gt, tinytable, flextable, Excel, Word, or clipboard.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publication-ready regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)

# Complete-case subset for the cluster-robust examples — the cluster
# vector must align with the rows actually used by `lm()`, so we
# drop the 25 rows with at least one NA in the predictor set up front.
sochealth_cc <- na.omit(
  sochealth[, c("wellbeing_score", "bmi", "age", "sex",
                "smoking", "education", "region")]
)
```

`table_regression()` produces a coefficient summary table from
one or several fitted `lm()` or `glm()` models. The output is
publication-ready by default and follows APA Manual 7 (American
Psychological Association 2020, Tables 7.13–7.15) formatting
conventions: paired estimate-and-CI columns, APA p-values without
leading zero, factor levels grouped under their parent variable,
fit statistics at the foot of the table, and a self-documenting
note line stating the variance estimator and any methodological
choice that affected the rendered values.

The function is **fit-first**: you pass already-fitted models, not
raw data and a formula. Internally each row of the table is
represented by an `(model_id, term, estimate_type)` triplet, so
the same object exports cleanly to long format for downstream
work (`broom::tidy()`) and to a one-row-per-model glance summary
(`broom::glance()`).

Two methodological choices deserve highlighting up front:

* **Average Marginal Effects under cluster-robust variance with
  Satterthwaite-corrected df.** When `vcov` is set to `"CR0"` /
  `"CR1"` / `"CR2"` / `"CR3"` and the user requests AME columns,
  the function constructs the closed-form linear contrast
  representing each AME and inverts it through
  [`clubSandwich::linear_contrast()`][clubSandwich::linear_contrast]
  with `test = "Satterthwaite"`. The B coefficient and the AME
  therefore share the same inferential regime — t-distribution
  with the same df — in the same table. The z-asymptotic
  alternative is anti-conservative for few clusters (Pustejovsky
  and Tipton 2018).
* **Transparency on standardised coefficients with non-additive
  terms.** When `standardized != "none"` and the model contains
  an interaction or a transformed term (`I()`, `poly()`, `log()`,
  `splines::ns()`), the function emits a classed `spicy_caveat`
  warning at runtime AND prints a method-specific caveat in the
  table footer (Aiken and West 1991; Cohen, Cohen, West, and
  Aiken 2003 §7.7). The table is rendered; the limitation is
  exposed at the point of use.

`table_regression()` supports `lm` and `glm` (binomial / Poisson /
Gamma / inverse.gaussian / quasi families with any link). The
*Generalised linear models* section below covers the glm-specific
argument semantics. Mixed-effects models are on the roadmap.

## Basic usage

Pass a fitted `lm()` object. The default rendering returns a
single-model table with `B`, `SE`, `95% CI`, and `p` columns plus a
fit-statistics footer (`n`, `R²`, `Adj.R²`):

```{r basic}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit)
```

Reading the table:

* Each predictor occupies one row; factor predictors are grouped
  under their parent variable name with one row per level. The
  reference level carries the `(ref.)` annotation and an em-dash
  across the statistic columns, so the substantive comparison is
  visible in a single glance (NEJM / BMJ clinical convention).
* The `95% CI` column reports a symmetric Wald interval at the
  level set by `ci_level` (default 0.95). The CI header tracks
  the chosen level — switching to `ci_level = 0.99` re-labels the
  column accordingly.
* The footer line names the variance estimator in plain English so
  the reader can find the inferential regime without leaving the
  table. The estimator switches with `vcov` (next section).

## Heteroskedasticity-consistent variance

Set `vcov = "HC*"` for sandwich-style standard errors via
[`sandwich::vcovHC()`][sandwich::vcovHC]. The valid types are `HC0`
through `HC5`; `HC3` is the small-sample-friendly default
(Long and Ervin 2000):

```{r hc}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, vcov = "HC3")
```

The footer reads "*Std. errors: heteroskedasticity-robust (HC3)*";
the column header for the confidence interval automatically tracks
`ci_level`.

## Cluster-robust variance

For clustered observations, `vcov = "CR*"` requests
cluster-robust variance via
[`clubSandwich::vcovCR()`][clubSandwich::vcovCR], with the cluster
identifier supplied through `cluster`. Three forms are accepted, in
order of preference:

* **Formula** -- `cluster = ~region`. The variables are looked up
  in `model.frame(fit)` first, then in the model's original `data`
  argument. Recommended: independent of the dataset's name,
  composable for multi-way clustering (`cluster = ~region:year`).
* **String** -- `cluster = "region"`. Single column name resolved
  the same way as the formula. Convenient but cannot express
  interactions.
* **Vector** -- `cluster = df$region`. An atomic vector of length
  `nobs(fit)`. Use this when the cluster key is **derived on the
  fly** (`cluster = interaction(df$region, df$year)`,
  `cluster = as.integer(format(df$date, "%Y"))`), or pulled from a
  different dataset with matching row order.

Bare unquoted names (`cluster = region`) are **not** accepted --
they would require non-standard evaluation that breaks under
programmatic use (function wrapping, loops, dynamic column choice).
Use `~region` or `"region"` instead.

```{r cr}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth_cc)
table_regression(
  fit,
  vcov = "CR2",
  cluster = ~region
)
```

`CR2` (the Bell-McCaffrey adjustment) is the recommended default
under few clusters (Pustejovsky and Tipton 2018; Imbens and Kolesár
2016). Coefficient inference uses
[`clubSandwich::coef_test()`][clubSandwich::coef_test] with
Satterthwaite-corrected degrees of freedom -- visible in the footer
when AME is also requested (next section).

## Average Marginal Effects with Satterthwaite df

Add `"ame"` to `show_columns` to display the average marginal effect of
each predictor. Use `"ame_ci"` and `"ame_p"` for the corresponding CI
and p-value columns; the shortcut `"all_ame"` expands to
`c("ame", "ame_se", "ame_ci", "ame_p")`. When the variance estimator is
cluster-robust, `table_regression()` constructs the linear contrast
representing each AME and inverts it through
`clubSandwich::linear_contrast()` with the Satterthwaite framework, so
B and AME share the same t-distribution with the same df:

```{r ame-cr}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth_cc)
table_regression(
  fit,
  vcov = "CR2",
  cluster = ~region,
  show_columns = c("b", "se", "ci", "p", "ame", "ame_ci", "ame_p")
)
```

Note that `"p"` always refers to the B (or β) coefficient, never to
the AME. To display the AME-specific p-value, include `"ame_p"` in
`show_columns`. Placing `"ame"` after `"p"` makes the "which p
belongs to what" reading unambiguous.

The footer now reads "*AME inference: t-distribution with
Satterthwaite-corrected df (Pustejovsky and Tipton 2018) via
`clubSandwich::linear_contrast()`*". For non-CR variance estimators
the AME column delegates to
[`marginaleffects::avg_slopes()`][marginaleffects::avg_slopes] with a
df argument coherent with B (residual df for classical and HC,
asymptotic z for bootstrap and jackknife).

## Standardised coefficients

`standardized` controls the standardisation method. Four are
available; the choice is consequential and the differences are
documented in the literature (Cohen et al. 2003 §3.4; Gelman 2008):

* `"refit"` — refit the model on z-scored outcome and predictors.
  This is the gold-standard convention used by SPSS `REGRESSION` and
  Stata `regress, beta`. Both numeric and dummy-coded predictors
  enter the refit on the same scale.
* `"posthoc"` — algebraic rescaling `β = B × SD(X) / SD(Y)`, applied
  to the original fit. Numerically identical to `"refit"` for purely
  linear-additive Gaussian models; preferred when refitting is
  expensive (large `n` × `p`) or when `lm()` was wrapped in a
  pipeline that resists re-execution.
* `"basic"` — algebraic, but factor dummies keep their 0/1 scale
  rather than being z-scored. Useful when factor levels carry
  meaningful base rates that scale-free standardisation would
  obscure.
* `"smart"` — Gelman's (2008) recommendation: numeric predictors
  divided by `2 × SD(X)`; binary predictors centred only. The
  resulting β is the predicted change in `Y` for one within-sample
  standard deviation of `X`, on the original `Y` scale, which keeps
  factor and numeric coefficients on roughly comparable footing.

When `standardized != "none"`, the `"beta"` token is auto-injected
into `show_columns` immediately after `"b"`:

```{r beta}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, standardized = "refit")
```

The standardised column is labelled `β` in the rendered table; the
unstandardised `B` stays alongside so both can be reported (the
convention recommended by APA Manual 7 §7.13 for transparency).

For models with interactions or transformed predictors, the function
emits a `spicy_caveat` warning AND prints a method-specific caveat
line in the footer (Aiken and West 1991; Cohen et al. 2003 §7.7).
The standardised coefficient of a product term has no closed-form
interpretation as a "one-SD change in X" effect, and the footer
makes that limitation explicit at the point of use.

## Per-coefficient effect sizes

Three partial effect-size tokens are available — Cohen's `f²`
(`partial_f2`), Pearson's partial `η²` (`partial_eta2`), and the
Olejnik–Algina bias-corrected partial `ω²` (`partial_omega2`).
Each estimate has a confidence interval derived from
noncentral-`F` inversion (Smithson 2003; Steiger 2004), exposed as
a separate `<token>_ci` column. The group shortcuts `"all_f2"`,
`"all_eta2"`, `"all_omega2"` expand to the point estimate and its
CI pair in one go:

```{r partial}
fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sochealth)
table_regression(
  fit,
  show_columns = c("b", "p", "all_eta2", "all_omega2")
)
```

Methodology of the per-coefficient effect sizes:

* The partial F-test is computed on a Type-II ANOVA reference
  (`car::Anova`), which respects the principle of marginality and
  is the SAS / SPSS default for unbalanced designs.
* The `ω²` point estimate is bias-corrected via the Olejnik and
  Algina (2003) formula `((F − 1) × df1) / (F × df1 + N − df1)`,
  yielding a less-biased small-sample estimator than partial `η²`.
* The CI bounds are obtained by inverting the noncentrality
  parameter of the F-distribution at the lower and upper
  confidence level (Steiger 2004 §4). The Steiger inversion bounds
  always bracket the corresponding bias-corrected point estimate
  even when the lower bound clips at zero (a common occurrence for
  near-null terms).
* For factor predictors with `k` levels, the partial F-test is the
  joint `(k − 1)` df Wald test, so the same effect-size value is
  broadcast across all non-reference dummy rows; the reference row
  shows an em-dash.

## Multiple-comparison adjustment

`p_adjust` applies a family-wise or false-discovery-rate adjustment
to the p-values via [`stats::p.adjust()`][stats::p.adjust]. The
family is the model's full coefficient set (intercept and reference
rows excluded); `B` and `AME` are adjusted as independent families
within the same call. Available methods are the same as in
`stats::p.adjust()`: `"none"` (default), `"holm"`, `"hochberg"`,
`"hommel"`, `"bonferroni"`, `"BH"` / `"fdr"`, `"BY"`.

```{r p-adjust}
fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sochealth)
table_regression(fit, p_adjust = "bonferroni")
```

The footer documents the chosen method and the family size; the
SE column is unchanged. `p_adjust` is orthogonal to `vcov`:
combining a robust SE with a family-wise correction is fully
supported, e.g. `vcov = "HC3", p_adjust = "holm"`. The adjustment
is applied **before** `keep` / `drop` filtering, so the family
stays the model's full coefficient set and the displayed adjusted
p-values reflect the right denominator regardless of which subset
is shown.

A methodological note. Adjusting the p-values of every
coefficient of a single regression model is *not* the standard
convention in social-science or clinical reporting (Rothman 1990;
Gelman, Hill, and Yajima 2012; Greenland 2017; Harrell 2015 §5.4;
APA Manual 7 §6.46). Each coefficient tests a scientifically
distinct hypothesis on a distinct predictor, which is not the
situation that family-wise procedures were designed for. The
default `"none"` reflects this consensus.

Adjustment is nonetheless legitimate in three contexts:

1. **Mass screening** with many candidate predictors and no prior
   hypothesis — typically `"BH"` (Benjamini–Hochberg, false
   discovery rate).
2. **Pre-registered multi-endpoint confirmatory designs** —
   typically `"holm"` (Holm's step-down, strong family-wise
   error rate control).
3. **When a reviewer, an editor, or a statistical analysis plan
   explicitly requests it** — apply the requested method and
   document it in the footer.

The argument is exposed under the same *transparency* rule used
for `standardized`: the tool is here, the methodological choice
is yours, and the footer makes the choice visible to the reader.

## Filtering displayed coefficients

`keep` and `drop` accept regular expressions matched against
coefficient names (as returned by [stats::coef()]). They are
mutually exclusive — pick `keep` to whitelist focal predictors, or
`drop` to hide a few control variables. Multiple patterns combine
with logical OR.

```{r keep}
fit <- lm(wellbeing_score ~ age + sex + smoking + bmi + education,
          data = sochealth)
table_regression(fit, keep = c("^smoking", "^bmi$"))
```

```{r drop}
table_regression(fit, drop = "^education")
```

Together with `p_adjust`, this is the article-ready workflow:
adjust on the full model, display only the rows the reader cares
about — `table_regression(fit, p_adjust = "BH", keep = "^treatment")`.

## Multiple models side by side

Pass a list of `lm()` fits. The default column layout places each
model in its own panel under a centered **spanner label** showing
the model name; sub-columns (`B / SE / p`) are shared across the
spanner. When dependent variables differ across models and the
user did not supply labels (`model_labels = ` or named list), the
bare DV name is lifted into the spanner automatically:

```{r multi}
m_wellbeing <- lm(wellbeing_score ~ age + sex + smoking,
                   data = sochealth)
m_bmi       <- lm(bmi             ~ age + sex + smoking,
                   data = sochealth)
table_regression(list(m_wellbeing, m_bmi))
```

When all models share the same DV, the DV appears in the title.
Use a named list -- e.g. `list(Crude = m1, Adjusted = m2)` -- to
set the spanner labels explicitly; pass `model_labels = c(...)`
to override the names from the list.

## Hierarchical / nested regression

Set `nested = TRUE` to add **in-table change-statistic rows**
(APA Table 7.13 / Stata `esttab` / SPSS Model Summary convention).
Each adjacent pair (M2 vs M1, M3 vs M2, ...) contributes one
column of change stats below `R² / Adj.R²`; the FIRST model
column gets em-dashes (no previous model to compare to):

```{r nested}
m1 <- lm(wellbeing_score ~ age + sex,                 data = sochealth_cc)
m2 <- lm(wellbeing_score ~ age + sex + smoking,       data = sochealth_cc)
m3 <- lm(wellbeing_score ~ age + sex + smoking + bmi, data = sochealth_cc)
table_regression(list(m1, m2, m3), nested = TRUE)
```

Default change tokens auto-injected: `c("r2_change", "f_change",
"p_change")` for `lm` (APA hierarchical-regression standard),
`c("lrt_change", "p_change")` for `glm` (Hosmer & Lemeshow §3.5).
Customise via `show_fit_stats`; the order of tokens controls the
order of the rows. Other change tokens are available:
`"adj_r2_change"`, `"f2_change"`, `"deviance_change"`,
`"aic_change"` / `"aicc_change"` / `"bic_change"`.

Validation is strict: identical `nobs` AND identical response
variable across all models, otherwise a `spicy_invalid_input`
error explains that R's listwise deletion may produce different
`n` per model and suggests refitting on the common subset (the
reason for `sochealth_cc` being prepared at the top of the
vignette).

## Generalised linear models (`glm`)

`table_regression()` accepts any `glm()` fit. Inference defaults
to the `z`-asymptotic Wald regime — the convention used by
`summary.glm()`, Stata's `logit, or`, and SPSS `LOGISTIC
REGRESSION`. The table title becomes family-aware ("Logistic
regression", "Poisson regression", "Probit regression", …) and
the default fit-statistics block swaps in `nobs`,
`pseudo_r2_mcfadden`, `pseudo_r2_nagelkerke`, and `AIC` instead
of `R²` and `Adj.R²`:

```{r glm-basic}
fit <- glm(am ~ mpg + wt, data = mtcars, family = binomial)
table_regression(fit)
```

### Response-scale display: `exponentiate = TRUE`

Set `exponentiate = TRUE` to switch the `B` column to the
response scale and rebrand its header per family and link: `OR`
for `binomial(logit)`, `IRR` for `poisson(log)`, `HR` for
`binomial(cloglog)`, `RR` for `binomial(log)`, `MR` for
`Gamma(log)`, and the generic `exp(B)` otherwise. The standard
error follows the delta-method approximation
`SE_OR = OR × SE_log-odds` (the Stata `logit, or` convention).
The test statistic and the p-value are invariant under any
monotone transformation, so they remain on the link scale and
match the unexponentiated table verbatim:

```{r glm-or}
table_regression(fit, exponentiate = TRUE)
```

### Term-level partial chi-square

`partial_chi2` is the glm analog of `partial_f2`: for each model
term, the partial likelihood-ratio chi-square via
`drop1(test = "LRT")` (SAS PROC LOGISTIC `TYPE3`; Long & Freese
2014 §3.5; Allison "TYPE3"). Rendered as `value (df)` so factor
terms (k−1 df) and numeric terms (1 df) read at a glance:

```{r glm-partial-chi2}
mt <- mtcars; mt$cyl <- factor(mt$cyl)
fit2 <- glm(am ~ mpg + cyl, data = mt, family = binomial)
table_regression(fit2, show_columns = c("b", "partial_chi2", "p"))
```

### Standardised coefficients: `refit` and the new `pseudo`

For `glm`, `standardized = "refit"` z-scores numeric *predictors*
only and refits the model — the response stays on its observed
scale because the link function is fixed. This is the
"x-standardization" convention (Long and Freese 2014 §4.3.4).
The other algebraic methods (`"posthoc"`, `"basic"`, `"smart"`)
apply X-only scaling using the same algebra as in the `lm` case.

`standardized = "pseudo"` (`glm` only) is the Menard (2004, 2011)
*fully* standardised coefficient, scaling by `SD(X) / SD(Y*)` where
`Y*` is the latent variable on the link scale and
`SD(Y*) = sqrt(var(linear-predictor) + var_link)` with `var_link` =
π²/3 for logit, 1 for probit, π²/6 for cloglog. Defined for binomial
families; non-binomial returns NA with a `spicy_caveat`:

```{r glm-pseudo}
table_regression(fit, standardized = "pseudo")
```

### Average Marginal Effects (AME)

The `ame` token returns response-scale marginal effects via
`marginaleffects::avg_slopes()`. For `glm`, AME is computed as
the average of `dE[Y|X]/dx` over the observed sample (so for
logistic regression the displayed AME is in probability units, not
log-odds). Under cluster-robust variance (`vcov = "CR2"` etc.), the
inference uses Satterthwaite df from
`clubSandwich::coef_test()` on the dominant underlying coefficient
— Pustejovsky & Tipton (2018) §4 approximation for nonlinear
contrasts:

```{r glm-ame}
table_regression(fit, show_columns = c("b", "p", "ame", "ame_ci", "ame_p"))
```

### Profile-likelihood CIs: `ci_method = "profile"`

The default is `ci_method = "wald"` (symmetric `estimate ± z × SE`,
matching `summary.glm`). For `glm` you can also request
`ci_method = "profile"`, which calls `MASS::confint.glm()` to
compute the profile-likelihood CI — asymmetric, exact for
likelihood-based inference, and the gold standard under sparse data
or near-boundary estimates (Venables & Ripley *MASS* §7.2). The
estimate, SE, statistic, and p-value all remain Wald;
`"profile"` only refines the CI bounds:

```{r glm-profile}
table_regression(fit, ci_method = "profile")
```

### Hierarchical glm (LRT)

`nested = TRUE` on a list of nested glm fits adds a "Model
comparison" block whose default tokens are `c("LRT", "p")` — the
APA-conventional layout for hierarchical logistic regression
(Hosmer & Lemeshow §3.5; Long & Freese 2014 §3.6). The chi-square
statistic comes from `anova(test = "LRT")`. `r2_change` and `F` are
not defined for glm and return em-dashes if explicitly requested:

```{r glm-nested}
m1 <- glm(am ~ mpg,                 data = mtcars, family = binomial)
m2 <- glm(am ~ mpg + wt,            data = mtcars, family = binomial)
m3 <- glm(am ~ mpg + wt + factor(cyl), data = mtcars, family = binomial)
table_regression(list(m1, m2, m3), nested = TRUE)
```

### Gaussian glm caveat

A `glm()` with `family = gaussian` and `link = "identity"` is
mathematically equivalent to `lm()` but lacks the variance-explained
effect-size family (`partial_f2 / η² / ω²`) and the Satterthwaite-
corrected AME path. Following the *transparency over rejection*
rule, spicy accepts the fit and emits a `spicy_caveat` suggesting a
refit with `lm()`.

## Significance stars

Stars are off by default. APA 7 §6.46 explicitly discourages
asterisks-only reporting in favour of exact p-values, and ASA's
post-2019 guidance (Wasserstein, Schirm, and Lazar 2019)
reinforces the same point. Set `stars = TRUE` for the APA preset
(`*** p < .001, ** p < .01, * p < .05`) or pass a named numeric
vector for custom thresholds:

```{r stars}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, stars = TRUE)
```

Stars suffix the B column (or β when standardisation is requested);
the threshold mapping is auto-documented in the footer. The `p`
column itself remains unstarred so the numeric value stays readable.

## Display knobs

`labels` accepts a named character vector keyed by either formula term
labels (`"sex"`) or coefficient names (`"sexMale"`), so individual
contrast rows can be relabelled. `intercept_position` switches the
intercept to the bottom (Stata convention). `reference_style =
"annotation"` lifts the reference level into the factor header
(`"sex: [ref: Female]"`) and drops the explicit reference row, for
compact tables. `decimal_mark = ","` switches to European decimal
notation and automatically changes the CI separator to `";"` to avoid
ambiguity:

```{r polish}
fit <- lm(wellbeing_score ~ age + sex + education, data = sochealth)
table_regression(
  fit,
  labels = c(
    age            = "Age (years)",
    sex            = "Sex",
    education      = "Education"
  ),
  reference_style = "annotation",
  intercept_position = "last",
  decimal_mark = ","
)
```

## Output formats

The default print method draws an ASCII table to the console. The
same content is available as a plain `data.frame` (`output =
"data.frame"`), a long broom-style data.frame (`output = "long"`), or
as a rich-format table for `gt`, `tinytable`, `flextable`, `excel`,
`word`, or `clipboard`:

```{r data-frame}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "data.frame"
)
str(out)
```

```{r long}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "long"
)
out[, c("model_id", "term", "estimate_type", "estimate",
        "std.error", "p.value")]
```

```{r gt-output, eval = build_rich_tables}
pkgdown_dark_gt(
  table_regression(
    lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
    output = "gt"
  )
)
```

## broom integration

`broom::tidy()` returns a long tibble with one row per
`(model_id, term, estimate_type)` and broom-canonical column
names (`estimate`, `std.error`, `conf.low`, `conf.high`,
`statistic`, `p.value`). `broom::glance()` returns one row per
`(model_id, outcome)` with model-level statistics; `df.residual`
is preserved as a numeric so cluster-robust Satterthwaite df
flows through verbatim:

```{r broom}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
out <- table_regression(fit, standardized = "refit")

broom::tidy(out)
broom::glance(out)
```

The long format is the right entry point when the table is one
step in a larger pipeline — saving to disk for the manuscript
appendix, faceting by subgroup, or feeding a downstream
post-estimation analysis. The `tidy()` output keeps the bilateral
`estimate_type` column so the same data frame can hold rows for
B, β, AME, and per-coefficient effect-size estimates without
ambiguity.

## See also

- `vignette("table-continuous-lm", package = "spicy")` for the
  one-predictor-by-many-outcomes counterpart (estimated marginal
  means, contrast or slope, four effect-size families).
- `vignette("summary-tables-reporting", package = "spicy")` for an
  end-to-end reporting workflow combining the spicy summary-table
  helpers.

## References

Aiken, L. S., and West, S. G. (1991).
*Multiple Regression: Testing and Interpreting Interactions*.
Sage.

American Psychological Association (2020).
*Publication Manual of the American Psychological Association*
(7th ed.). Section 6.46.

Cohen, J., Cohen, P., West, S. G., and Aiken, L. S. (2003).
*Applied Multiple Regression / Correlation Analysis for the
Behavioral Sciences* (3rd ed.). Lawrence Erlbaum.

Gelman, A. (2008). Scaling regression inputs by dividing by two
standard deviations. *Statistics in Medicine*, 27(15), 2865–2873.

Gelman, A., Hill, J., and Yajima, M. (2012). Why we (usually) don't
have to worry about multiple comparisons. *Journal of Research on
Educational Effectiveness*, 5(2), 189–211.

Greenland, S. (2017). For and against methodologies: Some
perspectives on recent causal and statistical inference debates.
*European Journal of Epidemiology*, 32(1), 3–20.

Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.).
Springer. Section 5.4 on multiplicity.

Hosmer, D. W., Lemeshow, S., and Sturdivant, R. X. (2013).
*Applied Logistic Regression* (3rd ed.). Wiley.

Imbens, G. W., and Kolesár, M. (2016). Robust standard errors in
small samples: Some practical advice. *Review of Economics and
Statistics*, 98(4), 701–712.

Long, J. S., and Ervin, L. H. (2000). Using heteroscedasticity
consistent standard errors in the linear regression model.
*The American Statistician*, 54(3), 217–224.

Long, J. S., and Freese, J. (2014). *Regression Models for
Categorical Dependent Variables Using Stata* (3rd ed.). Stata Press.

McFadden, D. (1974). Conditional logit analysis of qualitative
choice behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics*
(pp. 105–142). Academic Press.

Menard, S. (2004). Six approaches to calculating standardized
logistic regression coefficients. *The American Statistician*,
58(3), 218–223.

Menard, S. (2011). Standards for standardized logistic regression
coefficients. *Social Forces*, 89(4), 1409–1428.

Nagelkerke, N. J. D. (1991). A note on a general definition of the
coefficient of determination. *Biometrika*, 78(3), 691–692.

Olejnik, S., and Algina, J. (2003). Generalized eta and omega squared
statistics: Measures of effect size for some common research designs.
*Psychological Methods*, 8(4), 434–447.

Pustejovsky, J. E., and Tipton, E. (2018). Small-sample methods for
cluster-robust variance estimation and hypothesis testing in fixed
effects models. *Journal of Business & Economic Statistics*, 36(4),
672–683.

Rothman, K. J. (1990). No adjustments are needed for multiple
comparisons. *Epidemiology*, 1(1), 43–46.

Smithson, M. (2003). *Confidence Intervals*. Sage.

Steiger, J. H. (2004). Beyond the F test: Effect size confidence
intervals and tests of close fit in the analysis of variance and
contrast analysis. *Psychological Methods*, 9(2), 164–182.

Tjur, T. (2009). Coefficients of determination in logistic
regression models — A new proposal: The coefficient of
discrimination. *The American Statistician*, 63(4), 366–372.

Venables, W. N., and Ripley, B. D. (2002). *Modern Applied
Statistics with S* (4th ed.). Springer. Section 7.2 on profile
likelihood.

Wasserstein, R. L., Schirm, A. L., and Lazar, N. A. (2019). Moving
to a world beyond "p < 0.05". *The American Statistician*, 73(sup1),
1–19.
