---
title: "Publication-ready regression tables"
description: >
  Coefficient tables for fitted regression models in R with APA-aligned
  formatting: classical, heteroskedasticity-consistent and
  cluster-robust variance, four standardisation methods, partial
  effect sizes with noncentral-F CIs, average marginal effects,
  multiple-comparison adjustment, side-by-side and hierarchical
  layouts, and output to console, gt, tinytable, flextable,
  Excel, Word, or clipboard.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Publication-ready regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

`table_regression()` produces a coefficient summary table from
one or several fitted regression models. The output is
publication-ready by default and follows APA Manual 7 (American
Psychological Association 2020, Tables 7.13–7.15) formatting
conventions: paired estimate-and-CI columns, APA p-values without
leading zero, factor levels grouped under their parent variable,
fit statistics at the foot of the table, and a self-documenting
note line that names the variance estimator and any
methodological choice that affected the rendered values.

The function is **fit-first**: you pass already-fitted models,
not raw data and a formula. The same object exports cleanly to a
long broom-canonical frame (`broom::tidy()`) and to a
one-row-per-model glance summary (`broom::glance()`).

This vignette teaches the shared mechanics on `lm()` and `glm()`
fits. The *Generalised linear models* section covers the
glm-specific argument semantics; the *Mixed-effects models*
section introduces the Random effects rows that mixed-effects
fits add below the fixed effects (`lmer`, `glmer`, `glmmTMB`,
`lme`). Each further model family has a dedicated vignette that
applies these mechanics to its own estimands:
`vignette("table-regression-mixed")` for mixed-effects models,
`vignette("table-regression-counts")` for Poisson,
negative-binomial, and two-part models,
`vignette("table-regression-ordinal")` for ordinal models
(`polr`, `clm`), `vignette("table-regression-multinomial")` for
multinomial models (`multinom`, `mlogit`), and
`vignette("table-regression-survival")` for survival models
(`coxph`, `survreg`). The mechanics covered here — confidence
levels, multi-model and nested layouts, p-value adjustment,
coefficient filtering, output formats, and the broom methods —
apply across every supported class; variance-estimator
availability varies by family, and each family vignette states
its own set. One decision cuts across all of them — how to code,
test, and present categorical *predictors* — and has its own
vignette: `vignette("categorical-predictors")`.

## Basic usage

Pass a fitted `lm()` object. The default rendering returns a
single-model table with `B`, `SE`, `95% CI`, and `p` columns and
a fit-statistics footer (`n`, `R²`, `Adj.R²`):

```{r basic}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit)
```

Reading the table:

* Each predictor occupies one row; factor predictors are grouped
  under their parent variable name with one row per level. The
  reference level carries the `(ref.)` annotation and an en dash
  across the statistic columns, so the substantive comparison is
  visible in a single glance (NEJM / BMJ clinical convention).
* The `95% CI` column reports a symmetric Wald interval at the
  level set by `ci_level` (default 0.95). The CI header tracks
  the chosen level — switching to `ci_level = 0.99` re-labels the
  column accordingly.
* The footer line names the variance estimator in plain English
  so the reader can find the inferential regime without leaving
  the table. The estimator switches with `vcov` (covered in
  *Robust variance* below).
* Read substantively, one sentence per coefficient, from the
  exact numbers shown: `Male` — B = 3.86, 95% CI [2.08, 5.63],
  p < .001 — holding age and smoking constant, men score on
  average 3.9 points higher than women on the 0–100 WHO-5
  wellbeing index. `age` — B = 0.05, 95% CI [-0.01, 0.11],
  p = .130 — the interval spans effects from slightly negative to
  about a tenth of a point per year, so the data do not establish
  an adjusted age gradient.

## Standardised coefficients

Standardised coefficients (`β`) make predictors with different
natural scales comparable: a one-standard-deviation increase in
`X` predicts a `β`-standard-deviation change in `Y`. APA Manual 7
§7.13 recommends reporting both `B` and `β` so the unstandardised
effect (natural units, interpretable) stays alongside the
standardised effect (comparable across predictors).

`standardized` selects the method. Four are available; the choice
is consequential and well-documented (Cohen, Cohen, West, and
Aiken 2003 §3.4; Gelman 2008):

* `"refit"` — refit the model on z-scored outcome and predictors.
  Gold-standard convention, used by SPSS `REGRESSION` and Stata
  `regress, beta`. Both numeric and dummy-coded predictors enter
  the refit on the same scale.
* `"posthoc"` — algebraic rescaling `β = B × SD(X) / SD(Y)`,
  applied to the original fit. Numerically identical to `"refit"`
  for purely linear-additive Gaussian models; preferred when
  refitting is expensive or when `lm()` was wrapped in a pipeline
  that resists re-execution.
* `"basic"` — algebraic, but factor dummies keep their 0/1 scale
  rather than being z-scored. Useful when factor levels carry
  meaningful base rates that scale-free standardisation would
  obscure.
* `"smart"` — Gelman's (2008) input scaling: numeric predictors
  divided by `2 × SD(X)`; binary predictors and factor dummies
  kept on their 0/1 scale. The resulting `β` is the change in
  `Y`, in `SD(Y)` units, for a two-standard-deviation change in
  `X` (from mean − SD to mean + SD), which puts a continuous
  predictor's ±1 SD swing on the same footing as a binary's
  0 → 1 step. Where the paper leaves `Y` raw, spicy divides by
  `SD(Y)` so `β` stays comparable across the four methods.

When `standardized != "none"`, the `"beta"` token is auto-injected
into `show_columns` immediately after `"b"`:

```{r beta}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, standardized = "refit")
```

On the standardised scale the predictors become directly
comparable: being male predicts a 0.25-SD higher wellbeing score,
the largest standardised effect in the model, while a one-SD
increase in age predicts a change of only 0.04 SD. The raw `B`
column could not support that ranking — 3.86 versus 0.05 compares
a factor contrast with a per-year slope, two different units.

**Caveat: standardised coefficients with interactions or
transformed terms.** When the model contains a product term, an
`I()`, `poly()`, `log()`, or `splines::ns()` wrapper, the
standardised coefficient of the non-additive term has no
closed-form "one-SD change in X" reading (Aiken and West 1991;
Cohen et al. 2003 §7.7). The function emits a classed
`spicy_caveat` warning at runtime *and* prints a method-specific
caveat line in the table footer, so the limitation is exposed at
the point of use without blocking the table.

## Per-coefficient effect sizes

For each predictor, you can request a partial effect-size column.
Three measures are available — Cohen's `f²` (`partial_f2`),
Pearson's partial `η²` (`partial_eta2`), and the Olejnik–Algina
bias-corrected partial `ω²` (`partial_omega2`). Each carries a
confidence interval derived from noncentral-`F` inversion
(Smithson 2003; Steiger 2004), exposed as a separate `<token>_ci`
column. The shortcuts `"all_f2"`, `"all_eta2"`, and `"all_omega2"`
each expand to the point estimate together with its CI column.

`sochealth$education` is stored as an ordered factor, which R
codes with polynomial contrasts (`.L` = linear, `.Q` = quadratic)
— trend components across the ordered levels, not per-level
effects against a reference. For a coefficient table read
level-against-reference, convert it to a plain factor; we reuse
this copy throughout the vignette:

```{r partial}
sh <- sochealth |>
  dplyr::mutate(education = factor(education, ordered = FALSE))

fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sh)
table_regression(
  fit,
  show_columns = c("b", "p", "all_eta2", "all_omega2")
)
```

Methodology notes:

* The partial F-test is computed on a Type-II ANOVA reference
  (`car::Anova`), which respects the principle of marginality and
  is the SAS / SPSS default for unbalanced designs.
* The `ω²` point estimate is bias-corrected via the Olejnik and
  Algina (2003) formula `((F − 1) × df1) / (F × df1 + N − df1)`,
  yielding a less-biased small-sample estimator than partial `η²`.
* The CI bounds come from inverting the noncentrality parameter
  of the F-distribution at the lower and upper confidence levels
  (Steiger 2004 §4). Population partial `η²` and partial `ω²` are
  the same quantity — `ω²` is simply the less-biased small-sample
  estimator — so both columns share the single Steiger interval,
  following the MBESS convention (the `effectsize` package instead
  re-inverts an implied `F` for `ω²`, so its `ω²` bounds differ
  slightly). The interval brackets the bias-corrected point
  estimate, even when the lower bound clips at zero (common for
  near-null terms).
* For factor predictors with `k` levels, the partial F-test is
  the joint `(k − 1)` df Wald test, so the same effect-size value
  is broadcast across all non-reference dummy rows; the reference
  row leaves the effect-size cells blank.

## Average marginal effects (AME)

The `B` coefficient is the model's *structural* effect — the
change in the linear predictor `Xβ` for a one-unit change in
`x`, holding the other predictors constant. The **average
marginal effect (AME)** is the *observable* effect on the
response. For a numeric predictor it is the average partial
derivative `dE[Y|X] / dx` over the sample; for a factor level it
is the average **discrete change** — the difference in predicted
response between that level and the reference, averaged over the
sample (Long and Freese 2014 §6.2 distinguish the two). For a
purely linear-additive `lm()` both coincide with `B`. For
models with interactions or transformed terms, and for any
`glm()` with a non-identity link, the AME is the quantity the
substantive reader actually needs. The *average* in the name is
load-bearing: the effect is computed for every observation and
then averaged over the sample, not evaluated once at the
covariate means — the marginal effect at the mean (MEM), a point
at which no actual observation may sit. The two generally differ
in nonlinear models; Long and Freese (2014, §6.2.3) discuss the
choice in detail.

Add `"ame"` to `show_columns` for the point estimate. Companion
tokens display the SE, CI, and p-value: `"ame_se"`, `"ame_ci"`,
`"ame_p"`. The shortcut `"all_ame"` expands to the full set.

```{r ame-lm}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(
  fit,
  show_columns = c("b", "se", "p", "ame", "ame_ci", "ame_p")
)
```

Reading conventions:

* `"p"` always refers to the `B` (or `β`) coefficient, never to
  the AME. Use `"ame_p"` for the AME-specific p-value.
* Placing `"ame"` after `"p"` keeps the attribution of each
  p-value unambiguous — the B-block (`B / SE / p`) is closed
  before the AME-block opens.
* For non-linear models the AME is reported on the **response
  scale**, not the link scale. The *Generalised linear models*
  section below works out a logistic-regression example.

Inference is delegated to
[`marginaleffects::avg_slopes()`][marginaleffects::avg_slopes] and
respects the chosen variance estimator. Under cluster-robust
variance (covered in *Robust variance*), the AME inference shares
the coefficient's t-distribution and Satterthwaite-corrected
degrees of freedom via `clubSandwich::linear_contrast()`, so `B`
and AME are reported on the same inferential footing in the same
table. `marginaleffects` is a Suggests dependency; when
unavailable, the AME columns appear in the header but render as
NA without erroring (the rest of the table is unaffected).

## Multiple models side by side

Pass a list of `lm()` fits. The default column layout places each
model in its own panel under a centered **spanner label** showing
the model name; sub-columns (`B / SE / p`) are repeated under the
spanner. When dependent variables differ across models and the
user did not supply labels (`model_labels = ` or named list), the
bare DV name is lifted into the spanner automatically:

```{r multi}
m_wellbeing <- lm(wellbeing_score ~ age + sex + smoking,
                   data = sochealth)
m_bmi       <- lm(bmi             ~ age + sex + smoking,
                   data = sochealth)
table_regression(list(m_wellbeing, m_bmi))
```

When all models share the same DV, the DV appears in the title.
Use a named list — e.g. `list(Crude = m1, Adjusted = m2)` — to
set the spanner labels explicitly; pass `model_labels = c(...)`
to override the names from the list.

## Hierarchical / nested regression

Set `nested = TRUE` to add **in-table change-statistic rows**
(APA Table 7.13 / Stata `esttab` / SPSS Model Summary
convention). Each adjacent pair (M2 vs M1, M3 vs M2, …)
contributes one column of change stats below `R² / Adj.R²`; the
first model column gets em-dashes (no previous model to compare
to).

Hierarchical regression requires every model in the sequence to
share the same analytic sample: identical `nobs` and an identical
response variable. R's listwise deletion otherwise produces
different `n` per model as soon as a predictor introduced by M2
or M3 has missing values that M1 did not see — a silent bias on
the change statistics. We restrict to complete cases on the
union of predictors up front:

```{r sochealth-cc}
sochealth_cc <- sochealth |>
  dplyr::select(wellbeing_score, age, sex, smoking, bmi, region) |>
  na.omit()
```

```{r nested}
m1 <- lm(wellbeing_score ~ age + sex,                 data = sochealth_cc)
m2 <- lm(wellbeing_score ~ age + sex + smoking,       data = sochealth_cc)
m3 <- lm(wellbeing_score ~ age + sex + smoking + bmi, data = sochealth_cc)
table_regression(list(m1, m2, m3), nested = TRUE)
```

Reading the change rows: adding `smoking` in Model 2 does not
improve fit over the age + sex baseline (ΔR² = +0.00,
F-change = 2.28, p = .132) — with a single added term, this test
is equivalent to the test on the `smoking` coefficient itself
(p = .132 in the Model 2 column). Adding `bmi` in Model 3 does
(ΔR² = +0.02, F-change = 28.13, p < .001). The hierarchical
sequence therefore supports retaining `bmi` and provides no
incremental evidence for `smoking`.

Default change tokens auto-injected: `c("r2_change", "f_change",
"p_change")` for `lm` (APA hierarchical-regression standard),
`c("lrt_change", "p_change")` for `glm` (Hosmer & Lemeshow §3.5),
`c("aic_change", "bic_change", "lrt_change", "p_change")` for
mixed-effects fits (`lmer` / `glmer` / `glmmTMB` / `nlme::lme` —
Pinheiro & Bates 2000 §2.4.1). Customise via `show_fit_stats`;
the order of tokens controls the order of the rows. Other change
tokens are available: `"adj_r2_change"`, `"f2_change"`,
`"deviance_change"`, `"aic_change"` / `"aicc_change"` /
`"bic_change"`. Variance-explained change tokens (Δr², Δf²) are
NA for mixed-effects pairs — the F-test framework that grounds
them doesn't apply.

Validation is strict: identical `nobs` and identical response
across all models, otherwise a `spicy_invalid_input` error
explains the listwise-deletion trap and suggests refitting on the
common subset.

## Robust variance

The default `vcov = "classical"` reports the OLS standard error,
valid under homoskedastic, independent errors. Two robust
alternatives are first-class arguments; bootstrap and jackknife
variants are also available via `vcov = "bootstrap"` / `"jackknife"`.

### Heteroskedasticity-consistent (HC)

When error variance plausibly depends on the predictors (a
ubiquitous concern in cross-sectional social-science data), set
`vcov = "HC*"` for sandwich-style standard errors via
[`sandwich::vcovHC()`][sandwich::vcovHC]. The valid types are
`HC0` through `HC5`; `HC3` is the small-sample-friendly default
(Long and Ervin 2000):

```{r hc}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, vcov = "HC3")
```

The footer reads "*Std. errors: heteroskedasticity-robust (HC3)*";
the column header for the confidence interval automatically tracks
`ci_level`.

### Cluster-robust (CR)

For clustered observations (repeated measures on the same person,
students within schools, observations within regions),
`vcov = "CR*"` requests cluster-robust variance via
[`clubSandwich::vcovCR()`][clubSandwich::vcovCR], with the cluster
identifier supplied through `cluster`. Three forms are accepted:

* **Formula** — `cluster = ~region`. The variables are looked up
  in `model.frame(fit)` first, then in the model's original
  `data` argument. Recommended: independent of the dataset's
  name, composable for multi-way clustering
  (`cluster = ~region:year`).
* **String** — `cluster = "region"`. Single column name resolved
  the same way. Convenient but cannot express interactions.
* **Vector** — `cluster = df$region`. Atomic vector of length
  `nobs(fit)`. Use this when the cluster key is derived on the
  fly (`cluster = interaction(df$region, df$year)`) or pulled
  from a different dataset with matching row order.

Bare unquoted names (`cluster = region`) are **not** accepted —
they would require non-standard evaluation that breaks under
programmatic use (function wrapping, loops, dynamic column
choice).

```{r cr}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth_cc)
table_regression(
  fit,
  vcov = "CR2",
  cluster = ~region,
  show_columns = c("b", "se", "ci", "p", "ame", "ame_p")
)
```

`CR2` (the Bell-McCaffrey adjustment) is the recommended default
under few clusters (Pustejovsky and Tipton 2018; Imbens and
Kolesár 2016). Coefficient inference uses
[`clubSandwich::coef_test()`][clubSandwich::coef_test] with
Satterthwaite-corrected degrees of freedom. When AME columns are
requested, the same Satterthwaite framework is applied to the
AME contrast via `clubSandwich::linear_contrast()`, so `B` and
AME share the same t-distribution with the same df — visible in
the footer.

## Multiple-comparison adjustment

`p_adjust` applies a family-wise or false-discovery-rate
adjustment to the p-values via
[`stats::p.adjust()`][stats::p.adjust]. The family is the model's
full coefficient set (intercept and reference rows excluded);
`B` and `AME` are adjusted as independent families within the
same call. Available methods are the same as in
`stats::p.adjust()`: `"none"` (default), `"holm"`, `"hochberg"`,
`"hommel"`, `"bonferroni"`, `"BH"` / `"fdr"`, `"BY"`.

```{r p-adjust}
fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sh)
table_regression(fit, p_adjust = "bonferroni")
```

The footer documents the chosen method and the family size; the
SE column is unchanged. `p_adjust` is orthogonal to `vcov`:
combining a robust SE with a family-wise correction is fully
supported, e.g. `vcov = "HC3", p_adjust = "holm"`. The adjustment
is applied **before** `keep` / `drop` filtering, so the family
stays the model's full coefficient set and the displayed adjusted
p-values reflect the right denominator regardless of which
subset is shown.

A methodological note. Adjusting the p-values of every
coefficient of a single regression model is *not* the standard
convention in social-science or clinical reporting (Rothman 1990;
Gelman, Hill, and Yajima 2012; Greenland 2017; Harrell 2015 §5.4;
APA Manual 7 §6.46). Each coefficient tests a scientifically
distinct hypothesis on a distinct predictor, which is not the
situation that family-wise procedures were designed for. The
default `"none"` reflects this consensus.

Adjustment is nonetheless legitimate in three contexts:

1. **Mass screening** with many candidate predictors and no
   prior hypothesis — typically `"BH"` (Benjamini–Hochberg,
   false discovery rate).
2. **Pre-registered multi-endpoint confirmatory designs** —
   typically `"holm"` (Holm's step-down, strong family-wise
   error rate control).
3. **When a reviewer, an editor, or a statistical analysis plan
   explicitly requests it** — apply the requested method and
   document it in the footer.

The argument follows the same *transparency over rejection* rule
as `standardized`: the method is available, the choice among
methods is the analyst's, and the footer makes that choice
visible to the reader.

## Filtering displayed coefficients

`keep` and `drop` accept regular expressions matched against
coefficient names (as returned by [stats::coef()]). They are
mutually exclusive — use `keep` to retain only the focal
predictors, or `drop` to hide a few control variables. Multiple
patterns combine with logical OR.

```{r keep}
fit <- lm(wellbeing_score ~ age + sex + smoking + bmi + education,
          data = sh)
table_regression(fit, keep = c("^smoking", "^bmi$"))
```

```{r drop}
table_regression(fit, drop = "^education")
```

Together with `p_adjust`, this is the article-ready workflow:
adjust on the full model, display only the rows the reader cares
about — `table_regression(fit, p_adjust = "BH", keep = "^treatment")`.

## Significance stars

Stars are off by default. APA 7 §6.46 explicitly discourages
asterisks-only reporting in favour of exact p-values, and ASA's
post-2019 guidance (Wasserstein, Schirm, and Lazar 2019)
reinforces the same point. Set `stars = TRUE` for the APA preset
(`*** p < .001, ** p < .01, * p < .05`) or pass a named numeric
vector for custom thresholds:

```{r stars}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, stars = TRUE)
```

Stars suffix the `B` column (or `β` when standardisation is
requested); the threshold mapping is auto-documented in the
footer. The `p` column itself remains unstarred so the numeric
value stays readable.

## Display options

`labels` accepts a named character vector keyed by either formula
term labels (`"sex"`) or coefficient names (`"sexMale"`), so
individual contrast rows can be relabelled. `intercept_position`
switches the intercept to the bottom (Stata convention).
`reference_style = "annotation"` lifts the reference level into
the factor header (`"sex: [ref: Female]"`) and drops the explicit
reference row, for compact tables. `decimal_mark = ","` switches
to European decimal notation and automatically changes the CI
separator to `";"` to avoid ambiguity:

```{r polish}
fit <- lm(wellbeing_score ~ age + sex + education, data = sh)
table_regression(
  fit,
  labels = c(
    age            = "Age (years)",
    sex            = "Sex",
    education      = "Education"
  ),
  reference_style = "annotation",
  intercept_position = "last",
  decimal_mark = ","
)
```

## Generalised linear models (glm)

`table_regression()` accepts any `glm()` fit. Inference defaults
to the `z`-asymptotic Wald regime — the convention
`summary.glm()` applies to fixed-dispersion families (binomial,
Poisson), shared by Stata's `logit, or` and SPSS `LOGISTIC
REGRESSION`. One caveat for cross-checking: for families with an
estimated dispersion (gaussian, Gamma, inverse.gaussian, and the
quasi- families), `summary.glm()` switches to `t` statistics on
the residual degrees of freedom, so its p-values run larger than
the table's `z`-based ones in small samples — there the table
follows Stata's `glm` default rather than `summary.glm()`. The
table title becomes family-aware ("Logistic regression", "Poisson
regression", "Probit regression", …) and the default
fit-statistics block swaps in `nobs`, `pseudo_r2_mcfadden`,
`pseudo_r2_nagelkerke`, and `AIC` instead of `R²` and `Adj.R²`.

We illustrate with a logistic regression of `smoking` on `sex`,
`age`, and `education` from `sochealth`, reusing `sh` — the copy
of `sochealth` with `education` as a plain (unordered) factor,
introduced in *Per-coefficient effect sizes* — so contrasts are
dummy-coded, one row per level. The model mixes a binary factor
(`sex`), a numeric predictor (`age`), and a 3-level factor
(`education`) — enough to exercise every glm-specific feature
below.

```{r glm-basic}
fit <- glm(smoking ~ sex + age + education, data = sh,
           family = binomial)
table_regression(fit)
```

Read the fit block with care. McFadden and Nagelkerke R² are
likelihood-based analogues of R², not proportions of explained
variance — their values are not comparable to an OLS R², and
different measures disagree by construction on the same fit
(here 0.02 for McFadden versus 0.03 for Nagelkerke). Long and
Freese (2014, §3.3) find these scalar measures of limited
utility except as a rough index for comparing competing models
on the same data, and prefer the information criteria — hence
AIC in the default block. When reporting one, name the specific
measure, as the row labels here do; a generic "pseudo-R²" is
ambiguous across software. Tjur's (2009) coefficient of
discrimination, an alternative on the probability scale for
binary outcomes, is available as the opt-in token
`pseudo_r2_tjur` via `show_fit_stats`.

### Response-scale display: `exponentiate = TRUE`

Set `exponentiate = TRUE` to switch the `B` column to the ratio
scale and relabel its header according to the family and link:
`OR` for `binomial(logit)`, `IRR` for `poisson(log)`, `HR` for
`binomial(cloglog)`, `RR` for `binomial(log)`, `MR` for
`Gamma(log)`, and the generic `exp(B)` for other log-link
families (a genuine ratio of means). Links whose exponential is
*not* a ratio — probit, cauchit, inverse (the `Gamma()`
default), among others — are refused with a clear error rather
than silently mislabelled; report response-scale effects for
those models via the AME column instead. The standard error
follows the delta-method approximation
`SE_OR = OR × SE_log-odds` (the Stata `logit, or` convention).
The test statistic and the p-value stay on the link scale, where
`B = 0` and `OR = 1` are the same hypothesis and the Wald
approximation is most accurate, so they match the
unexponentiated table verbatim. (Wald tests are not invariant to
reparameterisation — a z rebuilt from `SE_OR` would give a
different statistic and p-value, which is why the link-scale
test is kept; Long & Freese 2014 §3.2.2.):

```{r glm-or}
table_regression(fit, exponentiate = TRUE)
```

Read exponentiated rows with the factor-change template (Long and
Freese 2014 §6.1.1): for `Tertiary`, OR = 0.40, 95% CI
[0.27, 0.59] — holding sex and age constant, the odds of smoking
for a tertiary-educated respondent are 0.40 times the odds for
one with lower secondary education, a 60% decrease in the odds
(percentage change = 100 × (OR − 1)). An odds ratio is a ratio of
*odds*, not of probabilities: with a smoking prevalence around
21% it sits farther from 1 than the corresponding risk ratio, so
reading it as "a 60% lower probability of smoking" overstates the
effect. The *Average marginal effects* section below reports the
probability-scale counterpart for the same contrast (−0.15, that
is, 15 percentage points).

### Term-level partial chi-square

`partial_chi2` is the glm analog of `partial_f2`: for each model
term, the partial likelihood-ratio chi-square via
`drop1(test = "LRT")` (SAS PROC GENMOD `TYPE3` likelihood-ratio
analysis; Long & Freese 2014 §3.2.2, §3.2.4). Rendered as
`value (df)` so factor terms (`k − 1` df) and numeric terms
(1 df) read at a glance:

```{r glm-partial-chi2}
fit2 <- glm(smoking ~ age + education, data = sh, family = binomial)
table_regression(fit2, show_columns = c("b", "partial_chi2", "p"))
```

The `p` column and the `χ²` column answer different questions.
Each per-level p-value compares one education level with the
reference, so it changes when the reference level changes; the
joint test — 21.70 on 2 df, p < .001 — asks whether `education`
as a whole improves the model, and is invariant to the reference
choice. When a factor's levels straddle significance, the joint
test is the one to report. Wald and likelihood-ratio tests of
the same joint hypothesis are asymptotically equivalent but
differ in finite samples; many statisticians prefer the
likelihood-ratio form when both are available (Long & Freese
2014 §3.2.2), which is what `partial_chi2` computes.

### Standardised coefficients: `refit` and `pseudo`

For `glm`, `standardized = "refit"` z-scores numeric *predictors*
only and refits the model — the response stays on its observed
scale because the link function is fixed. This is the
"x-standardization" convention (Long and Freese 2014 §4.7.2).
The other algebraic methods (`"posthoc"`, `"basic"`, `"smart"`)
apply X-only scaling using the same algebra as in the `lm` case.

`standardized = "pseudo"` (`glm` only) is the Menard (2004, 2011)
*fully* standardised coefficient, scaling by `SD(X) / SD(Y*)`
where `Y*` is the latent variable on the link scale and
`SD(Y*) = sqrt(var(linear-predictor) + var_link)` with
`var_link` = π²/3 for logit, 1 for probit, π²/6 for cloglog.
Defined for binomial families; non-binomial returns NA with a
`spicy_caveat`:

```{r glm-pseudo}
table_regression(fit, standardized = "pseudo")
```

### Average marginal effects: probability units, not log-odds

For `glm`, AME is averaged over the observed sample — the partial
derivative for numeric predictors, the discrete change for factor
levels — **on the response scale, not the link
scale**. For logistic regression this means the displayed AME is
in **probability units, not log-odds**: an AME of `-0.15` for
`education = Tertiary` reads "on average, holding sex and age
constant, a respondent with tertiary education is 15 percentage
points less likely to be a current smoker than one with lower
secondary education". This is almost always the quantity a
substantive reader wants from a logistic regression, and it is
the reason AME is increasingly recommended over odds ratios for
communicating logistic effects (Mood 2010; Long and Freese 2014
§6.2.3).

```{r glm-ame}
table_regression(fit, show_columns = c("b", "p", "ame", "ame_ci", "ame_p"))
```

Point estimate and inference are delegated to
[`marginaleffects::avg_slopes()`][marginaleffects::avg_slopes].
Under cluster-robust variance, the Satterthwaite-df handling
described in the *Robust variance* section applies to `glm` AME
as well.

### Profile-likelihood CIs: `ci_method = "profile"`

The default is `ci_method = "wald"` (symmetric
`estimate ± z × SE`, matching `summary.glm`). For `glm` you can
also request `ci_method = "profile"`, which calls
`MASS::confint.glm()` to compute the profile-likelihood CI —
asymmetric, obtained by inverting the likelihood-ratio statistic
rather than the quadratic Wald approximation, and the preferred
choice under sparse data or near-boundary estimates (Venables
& Ripley *MASS* §7.2). The estimate, SE, statistic, and p-value
all remain Wald; `"profile"` only refines the CI bounds:

```{r glm-profile}
table_regression(fit, ci_method = "profile")
```

### Hierarchical glm (LRT)

`nested = TRUE` on a list of nested glm fits appends `Δχ²` and
`p (change)` rows to the fit-statistics block — default change
tokens `c("lrt_change", "p_change")`, as in *Hierarchical /
nested regression* above — the APA-conventional layout for
hierarchical logistic regression (Hosmer & Lemeshow §3.5; Long &
Freese 2014 §3.2.4). The chi-square statistic comes from
`anova(test = "LRT")`. The variance-explained change tokens
(`r2_change`, `f_change`) are not defined for glm — requesting
them in `show_fit_stats` raises a `spicy_invalid_input` error
that points to the pseudo-R² family and to `lrt_change` +
`p_change` instead:

```{r glm-nested}
m1 <- glm(smoking ~ sex,                   data = sh, family = binomial)
m2 <- glm(smoking ~ sex + age,             data = sh, family = binomial)
m3 <- glm(smoking ~ sex + age + education, data = sh, family = binomial)
table_regression(list(m1, m2, m3), nested = TRUE)
```

### Gaussian glm caveat

A `glm()` with `family = gaussian` and `link = "identity"` is
mathematically equivalent to `lm()` but lacks the variance-
explained effect-size family (`partial_f2 / η² / ω²`) and the
Satterthwaite-corrected AME path. Following the *transparency
over rejection* rule, spicy accepts the fit and emits a
`spicy_caveat` suggesting a refit with `lm()`.

## Mixed-effects models

`table_regression()` supports four mixed-effects engines:
`lme4::lmer()` and `lme4::glmer()`, `glmmTMB::glmmTMB()`, and
`nlme::lme()`. The fixed-effects section of the table follows
the same conventions as `lm` / `glm`. Below it, a subordinate
**Random effects** block reports the variance components as rows
— one per random standard deviation (σ), correlation (ρ), and
the residual — each with its estimate, SE, and CI in the shared
coefficient columns. The number of groups — one
`N (<grouping factor>)` row per grouping factor, here
`N (Subject)` — and the Nakagawa marginal / conditional R²
(Nakagawa & Schielzeth 2013) join the fit-statistic rows, so
they align per model in a comparison table:

```{r mixed-basic, eval = requireNamespace("lme4", quietly = TRUE) && requireNamespace("merDeriv", quietly = TRUE)}
library(lme4)
fit <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
table_regression(fit)
```

No `ICC` row appears here: a random-slope fit has no single
intra-class correlation, so the row is dropped automatically
(the intercept-only case shows it).

Two deliberate conventions:

* The variance-component rows carry **no p-value**. The null
  hypothesis σ = 0 sits on the boundary of the parameter space,
  where a Wald test is invalid (Self & Liang 1987); no reporting
  guideline requests a per-component p, and the R ecosystem
  (`lme4`, `nlme`, `parameters`, `broom.mixed`), Stata, and
  MLwiN all omit it. The correct significance signal for the
  random part is the footer's **likelihood-ratio test** against
  the no-random-effects model, whose p-value applies a boundary
  correction. With a single variance component the halving
  `p = 0.5 × P(χ²₁ > LR)` is the exact Self & Liang (1987)
  50:50 mixture — the `chibar2(01)` line Stata's `mixed`
  prints. With more random parameters, as here (two variances
  and a correlation), the exact reference is a multi-component
  mixture (Stram & Lee 1994); spicy reports the pragmatic
  halved `χ²_q`, less conservative than the plain `χ²_q` that
  Stata prints for joint tests with a caveat that it is
  conservative.
* The footer also carries the estimator label — `(REML)` for
  restricted maximum likelihood (the `lmer` / `lme` default) or
  `(ML)` for full ML (`glmer`, `glmmTMB` default, and
  `lmer(..., REML = FALSE)`). The label is informational: it
  tells the reader which likelihood the variance estimates are
  anchored to, without implying that REML and ML estimates are
  interchangeable for inference.

The display and inference controls are arguments, not
conventions: `show_re = FALSE` hides the block,
`re_scale = "variance"` switches the rows from σ to σ²,
`re_columns` restricts which quantities the variance rows show,
`re_ci = "profile"` replaces the Wald CIs on variance components
with boundary-respecting profile-likelihood CIs (Wald is the
default because profiling is far slower), and `re_test = "lrt"`
or `"rlrt"` adds an opt-in per-component test ("do we need this
random slope?") with the boundary correction applied. The footer
names the fixed-effect inference method per engine —
Satterthwaite t when `lmerTest` is loaded, Wald-z for `glmer` /
`glmmTMB` and plain `lmer`, containment-df t for `nlme::lme`.
The AME tokens of `show_columns` (see *Average marginal effects*
above) work identically for mixed-effects fits — always on the
response scale, with Wald-z inference.

`nested = TRUE` works on mixed-effects fits with the change
tokens listed in *Hierarchical / nested regression*; note that
its plain-χ² reference is appropriate for steps that add fixed
effects, while random-structure steps belong to `re_test`, whose
reference respects the boundary.

For the model-building sequence, the REML/ML estimator choice,
the Nakagawa R² and ICC mechanics, per-class inference paths,
singular fits, testing random components, and the `glmmTMB` /
`nlme::lme` specifics — including a worked `glmer` AME example —
see `vignette("table-regression-mixed", package = "spicy")`.

## Output formats

The default print method draws an ASCII table to the console.
The same content is available as a plain `data.frame`
(`output = "data.frame"`), a long broom-style data.frame
(`output = "long"`), or as a rich-format table for `gt`,
`tinytable`, `flextable`, `excel`, `word`, or `clipboard`:

```{r data-frame}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "data.frame"
)
str(out)
```

```{r long}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "long"
)
out[, c("model_id", "term", "estimate_type", "estimate",
        "std.error", "p.value")]
```

```{r gt-output, eval = build_rich_tables}
pkgdown_dark_gt(
  table_regression(
    lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
    output = "gt"
  )
)
```

## broom integration

`broom::tidy()` returns a long tibble with one row per
`(model_id, term, estimate_type)` and broom-canonical column
names (`estimate`, `std.error`, `conf.low`, `conf.high`,
`statistic`, `p.value`). `broom::glance()` returns one row per
`(model_id, outcome)` with model-level statistics; `df.residual`
is preserved as a numeric so cluster-robust Satterthwaite df
flows through verbatim:

```{r broom}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
out <- table_regression(fit, standardized = "refit")

broom::tidy(out)
broom::glance(out)
```

The long format is the right entry point when the table is one
step in a larger pipeline — saving to disk for the manuscript
appendix, faceting by subgroup, or feeding a downstream
post-estimation analysis. The `tidy()` output keeps the
`estimate_type` column so the same data frame can hold rows for
B, β, AME, and per-coefficient effect-size estimates without
ambiguity.

## See also

- `vignette("categorical-predictors", package = "spicy")` for
  categorical *predictors* across all model families: dummy coding
  and reference levels, joint tests, ordinal predictors
  (scores vs dummies), and the categorization trap.

Model-family vignettes build on the mechanics shown here:

- `vignette("table-regression-mixed", package = "spicy")` for
  mixed-effects (multilevel) models: random effects as table
  rows, ICC, the boundary-corrected chi-bar-squared test, and
  opt-in per-term LRT / RLRT.
- `vignette("table-regression-counts", package = "spicy")` for
  count and two-part models: Poisson and negative-binomial rate
  ratios, offsets, and zero-inflated / hurdle components as
  labelled blocks.
- `vignette("table-regression-ordinal", package = "spicy")` for
  ordinal (proportional-odds) models: shared slopes, ordered
  thresholds as rows, and per-category marginal effects.
- `vignette("table-regression-multinomial", package = "spicy")`
  for multinomial logistic models: outcome categories as columns
  and changing the reference category.
- `vignette("table-regression-survival", package = "spicy")` for
  survival models: Cox hazard ratios with events and concordance,
  accelerated failure time models, and survival estimands.

Descriptive and reporting companions:

- `vignette("table-categorical", package = "spicy")` for the APA
  Table 1 categorical descriptors (factors, labelled variables,
  chi-squared tests, association measures).
- `vignette("table-continuous", package = "spicy")` for the APA
  Table 1 / Table 2 continuous descriptors and unadjusted
  group-comparison tests.
- `vignette("table-continuous-lm", package = "spicy")` for the
  one-predictor-by-many-outcomes linear-model counterpart
  (estimated marginal means, contrast or slope, four effect-size
  families with noncentral CIs).
- `vignette("summary-tables-reporting", package = "spicy")` for an
  end-to-end reporting workflow that combines the four spicy
  summary-table helpers along the APA Table 1 / 2 / 3 sequence.

## References

Aiken, L. S., and West, S. G. (1991).
*Multiple Regression: Testing and Interpreting Interactions*.
Sage.

American Psychological Association (2020).
*Publication Manual of the American Psychological Association*
(7th ed.). Section 6.46.

Cohen, J., Cohen, P., West, S. G., and Aiken, L. S. (2003).
*Applied Multiple Regression / Correlation Analysis for the
Behavioral Sciences* (3rd ed.). Lawrence Erlbaum.

Gelman, A. (2008). Scaling regression inputs by dividing by two
standard deviations. *Statistics in Medicine*, 27(15), 2865–2873.

Gelman, A., Hill, J., and Yajima, M. (2012). Why we (usually)
don't have to worry about multiple comparisons. *Journal of
Research on Educational Effectiveness*, 5(2), 189–211.

Greenland, S. (2017). For and against methodologies: Some
perspectives on recent causal and statistical inference debates.
*European Journal of Epidemiology*, 32(1), 3–20.

Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd
ed.). Springer. Section 5.4 on multiplicity.

Hosmer, D. W., Lemeshow, S., and Sturdivant, R. X. (2013).
*Applied Logistic Regression* (3rd ed.). Wiley.

Imbens, G. W., and Kolesár, M. (2016). Robust standard errors in
small samples: Some practical advice. *Review of Economics and
Statistics*, 98(4), 701–712.

Long, J. S., and Ervin, L. H. (2000). Using heteroscedasticity
consistent standard errors in the linear regression model.
*The American Statistician*, 54(3), 217–224.

Long, J. S., and Freese, J. (2014). *Regression Models for
Categorical Dependent Variables Using Stata* (3rd ed.). Stata
Press.

McFadden, D. (1974). Conditional logit analysis of qualitative
choice behavior. In P. Zarembka (Ed.), *Frontiers in
Econometrics* (pp. 105–142). Academic Press.

Menard, S. (2004). Six approaches to calculating standardized
logistic regression coefficients. *The American Statistician*,
58(3), 218–223.

Menard, S. (2011). Standards for standardized logistic
regression coefficients. *Social Forces*, 89(4), 1409–1428.

Mood, C. (2010). Logistic regression: Why we cannot do what we
think we can do, and what we can do about it. *European
Sociological Review*, 26(1), 67–82.

Nagelkerke, N. J. D. (1991). A note on a general definition of
the coefficient of determination. *Biometrika*, 78(3), 691–692.

Nakagawa, S., and Schielzeth, H. (2013). A general and simple
method for obtaining R² from generalized linear mixed-effects
models. *Methods in Ecology and Evolution*, 4(2), 133–142.

Olejnik, S., and Algina, J. (2003). Generalized eta and omega
squared statistics: Measures of effect size for some common
research designs. *Psychological Methods*, 8(4), 434–447.

Pinheiro, J. C., and Bates, D. M. (2000). *Mixed-Effects Models
in S and S-PLUS*. Springer.

Pustejovsky, J. E., and Tipton, E. (2018). Small-sample methods
for cluster-robust variance estimation and hypothesis testing in
fixed effects models. *Journal of Business & Economic
Statistics*, 36(4), 672–683.

Rothman, K. J. (1990). No adjustments are needed for multiple
comparisons. *Epidemiology*, 1(1), 43–46.

Self, S. G., and Liang, K.-Y. (1987). Asymptotic properties of
maximum likelihood estimators and likelihood ratio tests under
non-standard conditions. *Journal of the American Statistical
Association*, 82(398), 605–610.

Smithson, M. (2003). *Confidence Intervals*. Sage.

Steiger, J. H. (2004). Beyond the F test: Effect size confidence
intervals and tests of close fit in the analysis of variance and
contrast analysis. *Psychological Methods*, 9(2), 164–182.

Stram, D. O., and Lee, J. W. (1994). Variance components testing
in the longitudinal mixed effects model. *Biometrics*, 50(4),
1171–1177.

Tjur, T. (2009). Coefficients of determination in logistic
regression models — A new proposal: The coefficient of
discrimination. *The American Statistician*, 63(4), 366–372.

Venables, W. N., and Ripley, B. D. (2002). *Modern Applied
Statistics with S* (4th ed.). Springer. Section 7.2 on profile
likelihood.

Wasserstein, R. L., Schirm, A. L., and Lazar, N. A. (2019).
Moving to a world beyond "p < 0.05". *The American Statistician*,
73(sup1), 1–19.
