---
title: "Multinomial regression tables"
description: >
  Publication-ready tables for multinomial logistic regression in R with
  `table_regression()`: per-outcome coefficient blocks, odds ratios (what
  Stata labels relative-risk ratios), per-category average marginal
  effects, changing the reference category, nested comparison by
  likelihood-ratio test, and the `nnet::multinom` engine plus the
  `mlogit` discrete-choice engine.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multinomial regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_nnet <- requireNamespace("nnet", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_nnet
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_mlogit <- have_nnet && requireNamespace("mlogit", quietly = TRUE)
have_broom  <- have_nnet && requireNamespace("broom", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(nnet)
```

This vignette covers **multinomial logistic regression** ŌĆö models for a
*nominal* outcome, a categorical response whose categories have no
natural order: employment status, party choice, transport mode. The
companion vignette [*Publication-ready regression
tables*](table-regression.html) covers the shared mechanics (`vcov`,
`ci_level`, output formats, multi-model layouts, broom integration),
and [*Ordinal regression tables*](table-regression-ordinal.html) covers
the case where the categories *are* ordered. Here we focus on what is
specific to nominal outcomes: per-outcome coefficient blocks, the
choice of reference category, and the distance between a coefficient's
sign and its effect on a probability.

`table_regression()` supports both engines:

* **`nnet::multinom()`** ŌĆö the baseline-category logit model on
  characteristics of the *person* (case-specific covariates); the
  social-science standard.
* **`mlogit::mlogit()`** ŌĆö McFadden's conditional logit on
  characteristics of the *alternatives* (discrete-choice data).

The running example is employment status in the bundled `sochealth`
survey (`?sochealth`) ŌĆö four categories with `Employed` by far the most
frequent:

```{r data}
d <- sochealth
table(d$employment_status)
```

One recoding first. `education` is stored as an *ordered* factor, and R
gives ordered factors polynomial contrasts by default ŌĆö the model would
estimate `.L` (linear) and `.Q` (quadratic) components that fit
identically but are awkward to report. Applied tables conventionally
show treatment dummies against a reference level (Fox & Weisberg 2019),
so we drop the ordering and keep the level order:

```{r recode}
d$education <- factor(as.character(d$education),
                      levels = c("Lower secondary", "Upper secondary",
                                 "Tertiary"))
```

## The baseline-category logit model in one paragraph

For an outcome with $J$ categories, the model fits $J-1$ logit
equations simultaneously by maximum likelihood, each comparing one
category against a common reference (the *baseline*):
$\log[P(Y = j)/P(Y = J)] = \alpha_j + x^\top\beta_j$ (Agresti 2013, ch. 8).
Two consequences shape the table. Each predictor gets **one coefficient
per non-reference outcome** ŌĆö there is no single "effect of education",
only its effect on each comparison ŌĆö and the $J-1$ fitted equations
determine the logit for *every* pair of categories, so nothing is lost
by picking one baseline. Fitting separate binary logits instead of the
simultaneous model is consistent but less efficient; the loss is minor
when the baseline is the most frequent category (Begg & Gray 1984) ŌĆö
and Stata's `mlogit` defaults to the most frequent outcome.
`multinom()` uses the factor's first level ŌĆö here `Employed`, which
happens to be both first and most frequent: large and substantively
meaningful, exactly what a reference should be.

## Basic table

```{r basic}
fit <- multinom(employment_status ~ age + sex + education,
                data = d, trace = FALSE)
table_regression(fit)
```

Reading the table:

* Rows come in **per-outcome blocks** under each predictor:
  `Unemployed: Tertiary` is the tertiary-education coefficient in the
  *Unemployed-versus-Employed* equation. Every row answers "compared to
  `Employed`".
* The `(Intercept)` block is one intercept per equation ŌĆö the log-odds
  of each category against `Employed` at the reference levels and
  age 0. It anchors the equations rather than carrying substantive
  interest.
* Inference is **Wald-z** (`df = Inf`), matching `summary()`-based
  practice for ML fits and Stata `mlogit`; `nnet` itself prints no
  p-values.
* The model-fit block reports **n** and **AIC**.
* In the call, `trace = FALSE` just silences the optimizer's iteration
  log; unlike `polr()`, no `Hess = TRUE` is needed.

One scope note: `multinom` fits report **classical Wald-z inference
only** ŌĆö a robust or cluster-robust `vcov` request is refused with an
explanatory error rather than silently ignored.

## Odds ratios ŌĆö and what Stata calls them

`exponentiate = TRUE` puts the estimates and CI bounds on the ratio
scale:

```{r or}
table_regression(fit, exponentiate = TRUE, show_columns = c("b", "ci", "p"))
```

Each value is an **odds ratio against the reference outcome**:
compared with lower-secondary graduates of the same age and sex,
tertiary graduates have 0.29 times the odds of being unemployed rather
than employed ŌĆö a 71% reduction, the education gradient the table is
really about. Formally it is the odds ratio of the *conditional*
distribution restricted to the two categories in the comparison
(Agresti 2013).

One terminology note, because readers will meet it: Stata's
`mlogit, rrr` labels these same numbers **relative-risk ratios** ŌĆö
$e^\beta$ is also the multiplicative change in the probability ratio
of outcome $j$ versus the baseline ŌĆö while SAS prints "Odds Ratio
Estimates" for its generalized logits. Same quantity, two readings;
`spicy` labels it `OR`, the reading that generalizes from binary
logistic regression.

## Changing the reference category

The reference is a *presentation* choice, not a modeling one: refit
with another baseline and it is the same model, reparameterized ŌĆö
same likelihood, same AIC, same fitted probabilities; only the
comparisons displayed change.

```{r relevel}
d2 <- d
d2$employment_status <- relevel(d2$employment_status, ref = "Unemployed")
fit_unemp <- multinom(employment_status ~ age + sex + education,
                      data = d2, trace = FALSE)
c(AIC(fit), AIC(fit_unemp))
table_regression(fit_unemp, exponentiate = TRUE, show_columns = c("b", "p"))
```

Against `Employed`, education looked irrelevant for the `Student`
contrast; against `Unemployed`, every `Tertiary` contrast is strong
(`Student: Tertiary` OR 3.94, p < .001; even `Inactive: Tertiary`
reaches p = .005; only `Inactive: Upper secondary` stays
inconclusive). The p-values moved because the *questions* changed, not
the model ŌĆö a multinomial table only ever shows $J-1$ of the
$J(J-1)/2$ pairwise comparisons, and any of the others can be
recovered by releveling (Agresti 2013; Long 1997). Report the baseline
that makes your substantive comparisons direct.

## The sign trap, and marginal effects

The most common misreading of a multinomial table is to treat a
coefficient's sign as the direction of the effect on that category's
*probability*. It is not: the marginal effect of $x_k$ on $P(Y=m)$ is
$P_m(\beta_{k,m} - \sum_j \beta_{k,j} P_j)$ ŌĆö the coefficient *minus a
probability-weighted average of all categories' coefficients* ŌĆö so its
sign can differ from the coefficient's, and can even change across the
range of $x_k$ (Long & Freese 2014; Wulff 2015). The categories compete
for probability mass.

The probability-scale summary is the per-category **average marginal
effect** (AME): within each predictor block, one row per response
category:

```{r ame}
table_regression(fit, show_columns = c("b", "ame"))
```

Three things to read off, using the `Tertiary` rows:

* The **reference category is back**: `Employed` has no coefficient
  rows (it is the baseline of every equation), yet it carries the
  largest *positive* probability effect in the table (+0.14 for
  `Tertiary`, second in magnitude only to `Unemployed`'s ŌłÆ0.16). The
  log-odds side never shows this; the probability side must, because
  the effects on all $J$ categories **sum to Ōēł 0** ŌĆö the mass that
  tertiary education moves out of `Unemployed` lands mainly in
  `Employed`.
* Cells are **on the probability scale**: ŌłÆ0.16 is a drop of 16
  **percentage points** in the probability of unemployment for
  tertiary relative to lower-secondary education, averaged over the
  observed covariates (`marginaleffects::avg_slopes()`; Williams
  2012). Add `"ame_se"`, `"ame_ci"`, or `"ame_p"` to `show_columns`
  for delta-method inference on these effects.
* Here every AME happens to share its coefficient's sign; do not count
  on it. When an outcome's coefficient sits close to the
  probability-weighted average of all the coefficients, the
  probability effect can flip sign ŌĆö the reason Wulff (2015)
  recommends never interpreting a multinomial model from its
  coefficient signs alone.

## An ordinal predictor: scores or dummies?

`education` is ordered ŌĆö Lower secondary < Upper secondary < Tertiary ŌĆö
and the dummy coding above ignores that. The alternative is to enter it
as **numeric scores** (1, 2, 3): one slope per equation instead of two
dummies, a model that is both simpler and nested in the factor model.
Agresti (2007, sec. 4.4.3) gives the workflow: fit both codings and let
a likelihood-ratio test judge whether the category coefficients depart
from a linear trend in the scores. `nested = TRUE` runs exactly that
comparison:

```{r scores}
d$educ_score <- as.numeric(d$education)   # 1 / 2 / 3
fit_lin <- multinom(employment_status ~ age + sex + educ_score,
                    data = d, trace = FALSE)
table_regression(list(fit_lin, fit), nested = TRUE,
                 show_columns = c("b", "p"))
```

The change rows read: freeing education from the linear trend buys a
chi-squared of 4.52 on 3 degrees of freedom (one per equation),
p = .211 ŌĆö no evidence the dummies improve on the scores, and AIC
agrees (2514.3 vs 2515.8). On Agresti's grounds the scores model is
preferable when it fits: simpler to report, and more powerful because
it concentrates the trend in one parameter per equation instead of
two.

Three caveats keep the decision honest. A non-significant test does
not *prove* linearity ŌĆö in modest samples several codings fit
adequately (Agresti 2007). Integer scores assume the steps are equally
spaced, which is a substantive claim about education levels, not a
statistical default (Agresti 2007, sec. 2.5.4). And testing first,
then reporting the scores model as if it had been chosen in advance,
inflates the Type I error of the subsequent test (Grambsch & O'Brien
1991) ŌĆö with only three levels the dummies cost one extra parameter
per equation, and Harrell (2015) reasonably recommends just keeping
them. Present the LRT as a transparent simplification decision, not a
proof.

## Independence of irrelevant alternatives, briefly

The multinomial logit assumes the odds between any two categories do
not depend on what other alternatives exist ŌĆö the *independence of
irrelevant alternatives* (IIA), of red-bus/blue-bus fame (the critique
goes back to Debreu 1960). Formal tests exist (Hausman & McFadden
1984; Small & Hsiao 1985), but simulation evidence shows them to be
unsatisfactory for applied work ŌĆö different tests reach conflicting
verdicts on the same data (Cheng & Long 2007) ŌĆö and Long & Freese
(2014) do not recommend them. The working advice remains McFadden's
(1974): use the model when the alternatives are "distinct and weighed
independently" ŌĆö not close substitutes. For a fixed, exhaustive set
like employment status, where no alternative is ever added or removed,
IIA is rarely the practical concern it is in choice modeling; where
alternatives genuinely substitute (two bus lines), the remedy is a
different model (nested or mixed logit), not a mechanical test.

## Discrete choice: `mlogit`

`multinom()` models the chooser; McFadden's **conditional logit**
models the *choices*. When covariates describe the alternatives ŌĆö
the price of each travel mode, the catch rate of each fishing mode ŌĆö
the coefficient is one per *attribute*, shared across alternatives
(McFadden 1974; Croissant 2020). The `mlogit` engine renders both
kinds of covariate in one table. The classic `Fishing` data: 1182
anglers choose among four modes; `price` and `catch` vary by
alternative, `income_k` (income in thousands of dollars) describes
the angler:

```{r mlogit, eval = have_mlogit, message = FALSE}
library(mlogit)
data("Fishing", package = "mlogit")
Fishing$income_k <- Fishing$income / 1000
Fish <- dfidx(Fishing, varying = 2:9, choice = "mode", shape = "wide")

fit_choice <- mlogit(mode ~ price + catch | income_k, data = Fish)
table_regression(fit_choice, exponentiate = TRUE,
                 show_columns = c("b", "ci", "p"))
```

Reading it: the **alternative-specific** covariates (`price`, `catch`)
get a single row each ŌĆö one more dollar of price multiplies a mode's
odds by 0.98 *whichever mode it is* ŌĆö while the **case-specific**
covariate (`income_k`) works like `multinom` and needs one coefficient
per non-reference alternative. The footer's `n` counts **choice
situations** (1182 anglers), not the 4728 rows of the long-format
data: each angler contributes one choice, and that is the model's
sample size.

Three guardrails: average marginal effects are refused for `mlogit`
(`marginaleffects` has no slopes method for its one-row-per-choice
structure); the robust `vcov` family is `CR0`ŌĆō`CR3` only, with one
cluster value per choice situation, not per long-format row; and
`HC*` is refused outright ŌĆö `sandwich::vcovHC()` mis-scales the
sandwich for mlogit's per-choice-situation scores.

## Several models side by side

Multinomial fits compare like any other class ŌĆö here the education
gradient before and after adjustment:

```{r multi}
fit_min <- multinom(employment_status ~ education, data = d, trace = FALSE)
table_regression(list(Unadjusted = fit_min, Adjusted = fit),
                 exponentiate = TRUE, show_columns = c("b", "p"))
```

## Output formats

Everything above used the default console output. The same table
renders as a raw data frame, a long broom-style tibble, or ŌĆö with the
corresponding Suggests package ŌĆö a rich `gt`, `flextable`,
`tinytable`, Excel, or Word table:

```{r outputs-df}
head(table_regression(fit, exponentiate = TRUE, output = "data.frame"))
```

```{r outputs-gt, eval = build_rich_tables && have_nnet}
table_regression(fit, show_columns = c("b", "ame"), output = "gt")
```

[`broom::tidy()`](https://broom.tidymodels.org) returns the long
frame, one row per `(term, estimate_type)`; the per-outcome structure is
carried in the term labels, and AME rows cover all $J$ categories
including the reference:

```{r broom, eval = have_broom}
broom::tidy(table_regression(fit, show_columns = c("b", "ame")))
```

## References

* Agresti, A. (2007). *An Introduction to Categorical Data Analysis*
  (2nd ed.). Wiley.
* Agresti, A. (2013). *Categorical Data Analysis* (3rd ed.). Wiley.
* Begg, C. B., & Gray, R. (1984). Calculation of polychotomous
  logistic regression parameters using individualized regressions.
  *Biometrika*, 71(1), 11ŌĆō18.
* Cheng, S., & Long, J. S. (2007). Testing for IIA in the multinomial
  logit model. *Sociological Methods & Research*, 35(4), 583ŌĆō600.
* Croissant, Y. (2020). Estimation of random utility models in R: The
  mlogit package. *Journal of Statistical Software*, 95(11), 1ŌĆō41.
* Debreu, G. (1960). Review of R. D. Luce, *Individual Choice
  Behavior*. *American Economic Review*, 50(1), 186ŌĆō188.
* Fox, J., & Weisberg, S. (2019). *An R Companion to Applied
  Regression* (3rd ed.). Sage.
* Grambsch, P. M., & O'Brien, P. C. (1991). The effects of
  transformations and preliminary tests for non-linearity in
  regression. *Statistics in Medicine*, 10(5), 697ŌĆō709.
* Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.).
  Springer.
* Hausman, J. A., & McFadden, D. (1984). Specification tests for the
  multinomial logit model. *Econometrica*, 52(5), 1219ŌĆō1240.
* Long, J. S. (1997). *Regression Models for Categorical and Limited
  Dependent Variables*. Sage.
* Long, J. S., & Freese, J. (2014). *Regression Models for Categorical
  Dependent Variables Using Stata* (3rd ed.). Stata Press.
* McFadden, D. (1974). Conditional logit analysis of qualitative
  choice behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics*
  (pp. 105ŌĆō142). Academic Press.
* Small, K. A., & Hsiao, C. (1985). Multinomial logit specification
  tests. *International Economic Review*, 26(3), 619ŌĆō627.
* Williams, R. (2012). Using the margins command to estimate and
  interpret adjusted predictions and marginal effects. *The Stata
  Journal*, 12(2), 308ŌĆō331.
* Wulff, J. N. (2015). Interpreting results from the multinomial logit
  model: Demonstrated by foreign market entry. *Organizational
  Research Methods*, 18(2), 300ŌĆō325.
