---
title: "Count and two-part regression tables"
description: >
  Publication-ready tables for count regression in R with
  `table_regression()`: Poisson and negative-binomial rate ratios,
  zero-inflated and hurdle models with their zero component as a
  labelled block, per-block exponentiation, combined average marginal
  effects, cluster-robust variance across both components, and
  zero-inflated mixed models with `glmmTMB`.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Count and two-part regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_counts <- requireNamespace("pscl", quietly = TRUE) &&
  requireNamespace("MASS", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_counts
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_glmmTMB <- have_counts && requireNamespace("glmmTMB", quietly = TRUE)
have_sandwich <- have_counts && requireNamespace("sandwich", quietly = TRUE)
have_broom   <- have_counts && requireNamespace("broom", quietly = TRUE)
have_me      <- have_counts && requireNamespace("marginaleffects", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(pscl)     # zeroinfl(), hurdle()
```

This vignette covers **count regression** — models for a non-negative
integer outcome such as number of publications, doctor visits, or
species counts. The companion vignette [*Publication-ready regression
tables*](table-regression.html) covers the shared mechanics (`vcov`,
`ci_level`, output formats, multi-model layouts, broom integration);
here we focus on what is specific to counts: rate ratios,
overdispersion, and above all the **two-part models** whose zero
component `table_regression()` renders as its own labelled block.

`table_regression()` supports the full count-model escalation:

* **`stats::glm(family = poisson())`** — the Poisson baseline;
* **`MASS::glm.nb()`** — negative binomial, for overdispersion;
* **`pscl::zeroinfl()`** and **`pscl::hurdle()`** — the two-part
  models (Zeileis, Kleiber & Jackman 2008);
* **`glmmTMB::glmmTMB()`** — all of the above with random effects.

The running example is `pscl::bioChemists` (Long 1990): the number of
articles published by 915 biochemistry PhD students in the last three
years of their doctorate, with gender, marital status, children under
six, and the mentor's article count as predictors. Thirty percent of
the students published nothing:

```{r data}
data("bioChemists", package = "pscl")
mean(bioChemists$art == 0)
```

## Count models in one paragraph

A Poisson regression models the *rate* of events through a log link,
so `exponentiate = TRUE` turns each coefficient into an **incidence
rate ratio** (IRR) — a multiplicative effect on the expected count.
Real count data usually violate the Poisson assumption in one or both
of two ways: the variance exceeds the mean (**overdispersion** — the
negative binomial's job), and there are more zeros than the count
process predicts (**excess zeros** — the job of the zero-inflated and
hurdle models, which add a second, binary submodel for the zeros).
Each extension changes what the table must show, and the sections
below follow that escalation.

## Poisson baseline

```{r poisson}
fit_pois <- glm(art ~ fem + mar + kid5 + ment, data = bioChemists,
                family = poisson())
table_regression(fit_pois, exponentiate = TRUE)
```

Each IRR multiplies the expected article count: women publish about
20% less than men (IRR 0.80), and each additional mentor publication
raises a student's rate by 3% (IRR 1.03). Inference is Wald-z on the
link scale; the footer records the SE and CI conventions.

One remark for rate data: the observation window here is a fixed three
years for everyone, so no exposure adjustment is needed. With unequal
exposure — person-years, plot area, time at risk — add
`offset(log(exposure))` to the formula as usual; the offset is
absorbed silently (no spurious coefficient row) and the IRRs read as
rate ratios per unit of exposure.

## Overdispersion: the negative binomial

The articles' variance is far above their mean, and a Poisson model
reacts by **understating the standard errors**. `MASS::glm.nb()` adds
a dispersion parameter:

```{r negbin}
fit_nb <- MASS::glm.nb(art ~ fem + mar + kid5 + ment, data = bioChemists)
table_regression(fit_nb, exponentiate = TRUE)
```

The IRRs barely move, but the inference does: marriage drops from
p = .013 under Poisson to p = .072 here — the Poisson significance was
an artifact of ignored overdispersion. AIC falls from 3312 to 3134,
and that comparison returns below.

## Excess zeros: zero-inflation and hurdle

Two-part models tell two different stories about the zeros, and the
difference matters for interpretation:

* A **zero-inflated** model (`zeroinfl()`; Lambert 1992) says some
  zeros are *structural*: a latent group that would never publish
  regardless of the count process. Its zero component models the
  probability of being such a **structural zero**.
* A **hurdle** model (`hurdle()`; Mullahy 1986) says all zeros come
  from one hurdle: first you publish at all or you do not, then a
  truncated count process decides how much. Its zero component models
  the probability of a **nonzero count** — the *opposite direction*.

`table_regression()` renders the count coefficients first and the zero
component as a labelled block, with a footer line stating exactly what
the block models. Under `exponentiate = TRUE` the exponentiation is
**per block**: count coefficients become IRR, and the logit zero
component becomes odds ratios:

```{r zeroinfl}
fit_zip <- zeroinfl(art ~ fem + mar + kid5 + ment | ment,
                    data = bioChemists)
table_regression(fit_zip, exponentiate = TRUE)
```

```{r hurdle}
fit_hur <- hurdle(art ~ fem + mar + kid5 + ment | ment,
                  data = bioChemists)
table_regression(fit_hur, exponentiate = TRUE)
```

Compare the `ment` row in the two zero blocks — the same predictor,
the same substantive finding, and *opposite-looking* odds ratios:

* Zero-inflation block: OR 0.88 — each mentor publication **lowers the
  odds of being a structural zero**.
* Zero hurdle block: OR 1.08 — each mentor publication **raises the
  odds of publishing at all**.

Both say productive mentors get students over the zero. Reading either
number without its block label invites a sign error; the block labels
and their footer lines exist to prevent it. As for choosing between
the two: fit rarely decides — here the AICs are nearly tied (3225.5
vs 3233.8) — so let the substantive story choose. Zero-inflation fits
when some units plausibly *can never* experience the event (a latent
never-publisher class); a hurdle fits when everyone faces the same
all-or-nothing first stage. Zero-component
coefficients are substantive hypotheses: they take significance stars
and join the `p_adjust` family like any other coefficient. To display
only the count part, set `show_components = FALSE` (the model is still
estimated in full).

## Which count model? Side by side

The escalation is a model-comparison question, and the table lays it
out. Per Long (1997) and Cameron & Trivedi (2013), much of what looks
like zero-inflation *is* overdispersion — the AIC row makes the point.
(The once-standard Vuong test is not used here: it is no longer
recommended for testing zero-inflation — Wilson 2015 — so information
criteria and substantive reasoning carry the comparison.)

```{r side-by-side}
fit_zinb <- zeroinfl(art ~ fem + mar + kid5 + ment | ment,
                     data = bioChemists, dist = "negbin")
table_regression(
  list(Poisson = fit_pois, "Neg. binomial" = fit_nb, ZINB = fit_zinb),
  show_columns = c("b", "p"), exponentiate = TRUE
)
```

The negative binomial alone recovers most of the fit (AIC
3312 → 3134); adding the inflation part on top buys a further, modest
improvement (3122). The zero block simply stays empty in the columns
of models that have none. Substantively the IRRs agree across all
three — what changes is the inference and the account of the zeros.

## One number per predictor: the combined AME

Rate ratios live inside one component. For a single response-scale
summary that spans *both* parts, request average marginal effects: the
AME of a two-part model is the effect on the **overall expected count**
E(Y), combining the count and zero processes (computed by
`marginaleffects::avg_slopes()` on the full model):

```{r ame, eval = have_me}
table_regression(fit_zip, show_columns = c("b", "ame", "p"))
```

Being a woman costs about a third of an article (AME −0.36) once both
channels are combined; the zero-component rows carry no AME of their
own because the combined effect already includes them.

## Cluster-robust variance

Students cluster in labs, journals, cohorts. The `CR*` family covers
**both components with one estimator** — `sandwich::vcovCL()` on the
full score matrix, so the count and zero rows shift together. For
two-part models every `CR*` variant maps to that same estimator, and
the footer records it as `CL`:

```{r cr, eval = have_sandwich, message = FALSE}
set.seed(1)
bioChemists$lab <- factor(sample(1:60, nrow(bioChemists), replace = TRUE))
table_regression(fit_zip, vcov = "CR0", cluster = bioChemists$lab,
                 show_columns = c("b", "se", "p"))
```

With clustering acknowledged (here a synthetic 60-lab assignment for
illustration), marriage loses its significance (p = .041 → .147) — the
usual fate of borderline effects under a proper dependence structure.
`HC*` and the resampling estimators have no two-part backend and are
refused with a clear error.

## Mixed counts: `glmmTMB`

When counts are also grouped, `glmmTMB` (Brooks et al. 2017) combines
everything above with random effects: the zero-inflation block, the random-effects block,
and per-block exponentiation appear in one table — here salamander
counts in mined and unmined streams (Price et al. 2016):

```{r glmmtmb, eval = have_glmmTMB}
data("Salamanders", package = "glmmTMB")
fit_mix <- glmmTMB::glmmTMB(
  count ~ mined + (1 | site),
  ziformula = ~ mined,
  family = glmmTMB::nbinom2, data = Salamanders
)
table_regression(fit_mix, exponentiate = TRUE)
```

Unmined streams host over four times the salamander rate (IRR 4.34)
*and* have far lower odds of a structural zero (OR 0.10) — the two
channels tell one coherent ecological story. The random-effects block
and its conventions are covered in the companion vignette
[*Mixed-effects regression tables*](table-regression-mixed.html).

## Output formats

Everything above used the default console output. The same structure —
zero block included, as the last rows below — carries to a raw data
frame, a long broom-style tibble, and the rich `gt`, `flextable`,
`tinytable`, Excel, or Word targets:

```{r outputs-df}
table_regression(fit_zip, output = "data.frame")
```

```{r outputs-gt, eval = build_rich_tables && have_counts}
table_regression(fit_zip, exponentiate = TRUE, output = "gt")
```

In [`broom::tidy()`](https://broom.tidymodels.org) the zero-component
terms carry a `zero_` prefix, so the two parts separate cleanly:

```{r broom, eval = have_broom}
td <- broom::tidy(table_regression(fit_zip))
td[, c("term", "estimate_type", "estimate", "p.value")]
```

## References

* Brooks, M. E., Kristensen, K., van Benthem, K. J., Magnusson, A.,
  Berg, C. W., Nielsen, A., Skaug, H. J., Mächler, M., & Bolker,
  B. M. (2017). glmmTMB balances speed and flexibility among packages
  for zero-inflated generalized linear mixed modeling. *The R
  Journal*, 9(2), 378–400.
* Cameron, A. C., & Trivedi, P. K. (2013). *Regression Analysis of
  Count Data* (2nd ed.). Cambridge University Press.
* Lambert, D. (1992). Zero-inflated Poisson regression, with an
  application to defects in manufacturing. *Technometrics*, 34(1),
  1–14.
* Long, J. S. (1990). The origins of sex differences in science.
  *Social Forces*, 68(4), 1297–1316.
* Long, J. S. (1997). *Regression Models for Categorical and Limited
  Dependent Variables*. Sage.
* Mullahy, J. (1986). Specification and testing of some modified
  count data models. *Journal of Econometrics*, 33(3), 341–365.
* Price, S. J., Muncy, B. L., Bonner, S. J., Drayer, A. N., & Barton,
  C. D. (2016). Effects of mountaintop removal mining and valley
  filling on the occupancy and abundance of stream salamanders.
  *Journal of Applied Ecology*, 53(2), 459–468.
* Wilson, P. (2015). The misuse of the Vuong test for non-nested
  models to test for zero-inflation. *Economics Letters*, 127, 51–53.
* Zeileis, A., Kleiber, C., & Jackman, S. (2008). Regression models
  for count data in R. *Journal of Statistical Software*, 27(8).
