---
title: "GEE regression tables (population-averaged models)"
description: >
  Publication-ready tables for generalized estimating equations in R
  with `table_regression()`: population-averaged coefficients with
  the fit's own cluster-robust sandwich inference, the working
  correlation structure disclosed in the footer, QIC-based comparison
  of working correlations, jackknife variants for few clusters,
  odds ratios with event counts, and population-averaged marginal
  effects.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GEE regression tables (population-averaged models)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_geepack <- requireNamespace("geepack", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_geepack
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_me <- have_geepack && requireNamespace("marginaleffects", quietly = TRUE)
have_ft <- have_geepack && requireNamespace("flextable", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(geepack) # geeglm()
```

This vignette covers **generalized estimating equations** (GEE) --
marginal models for clustered and longitudinal data, fitted with
`geepack::geeglm()` (Liang & Zeger, 1986; Halekoh, Højsgaard & Yan,
2006). The companion vignette [*Publication-ready regression
tables*](table-regression.html) covers the shared mechanics (output
engines, labels, `tidy()`); [*Mixed-effects regression
tables*](table-regression-mixed.html) shows the population-averaged
and subject-specific answers side by side in a single table.

## When GEE: the population-averaged question

Clustered data can be modelled to answer two different questions.
A mixed model answers the **subject-specific** (conditional) one:
what happens to *this* subject's outcome when their predictor
changes, holding their random effect fixed. GEE answers the
**population-averaged** (marginal) one: how does the *average*
outcome differ between populations that differ in the predictor
(Zeger, Liang & Albert, 1988). The marginal estimand is the natural
one for policy and epidemiology -- a treatment's effect on a
population's risk, not on one patient's odds (Hubbard et al., 2010).

For a linear model the two coincide. On a logit link they do not:
with a random-intercept variance $\sigma^2$, the population-averaged
coefficient is attenuated by approximately
$1/\sqrt{1 + 0.346\,\sigma^2}$ relative to the subject-specific one
(Zeger et al., 1988), so the conditional odds ratio is always
farther from 1. Neither is "biased" -- they answer different
questions. The side-by-side table in the
[mixed-effects vignette](table-regression-mixed.html) shows the gap
on real data.

Two properties make GEE attractive when the marginal question is the
right one. The coefficients stay consistent even if the assumed
within-cluster correlation is wrong, and the standard errors are
computed with a cluster-robust sandwich estimator *by construction*
-- robustness is not an option added after the fact, it is the
method (Liang & Zeger, 1986).

## A first table

`geepack`'s `dietox` data follow the weight of 72 pigs over 12
weeks. A gaussian GEE with an exchangeable working correlation:

```{r first-table}
data(dietox, package = "geepack")

fit <- geeglm(
  Weight ~ Time + Cu,
  id     = Pig,
  data   = dietox,
  family = gaussian,
  corstr = "exchangeable"
)
table_regression(fit)
```

Three things to read off this table:

* **The title** says *Population-averaged linear regression (GEE)* --
  the estimand is named, not implied.
* **The fit statistics** describe the sample the way the GEE
  literature (and Stata's `xtgee` header) does: the number of
  observations, the number of clusters (`N (Pig)`), and the largest
  cluster. These are the clusters the sandwich variance is actually
  computed over -- see the section on unsorted data below.
* **The footer** carries two disclosures. `Std. errors: Robust
  sandwich (GEE), clusters by Pig` names the inference: the fit's own
  sandwich estimator, clustered on its `id =` variable. `GEE working
  correlation: exchangeable (alpha = ...)` names the working
  correlation structure with its estimated parameter -- two GEE fits
  with the same formula but different structures are different
  models, so the structure is model-defining and always disclosed.

## Choosing the working correlation

The working correlation encodes how observations within a cluster
are assumed to correlate: `"independence"` (none), `"exchangeable"`
(all pairs equally), `"ar1"` (decaying with distance), or
`"unstructured"`. Point estimates are consistent under any of them;
the choice affects efficiency. Because GEE has no likelihood, AIC is
undefined -- the comparison tool is Pan's (2001) quasi-likelihood
information criterion, available as opt-in fit statistics:

```{r qic}
fit_ind <- update(fit, corstr = "independence")
fit_ar1 <- update(fit, corstr = "ar1")

table_regression(
  list(
    Independence = fit_ind,
    Exchangeable = fit,
    "AR(1)"      = fit_ar1
  ),
  show_fit_stats = c("nobs", "n_groups", "qic", "qicu")
)
```

Lower QIC is better for choosing the working correlation; QICu is
the variant for comparing mean models under a fixed structure (Pan,
2001). The footer discloses each model's structure on its own line.
Note that the estimates barely move across columns -- that is the
GEE consistency property at work -- while the standard errors
differ, which is exactly what the working correlation is for.

`qic` and `qicu` are computed only when you ask for them:
`geepack::QIC()` silently refits the independence model, and the
default table should not pay that cost for a number it does not
display.

## The displayed clusters are the clusters of the inference

`geepack` defines clusters as **consecutive runs** of identical
`id =` values. If the data are not sorted by `id`, a subject whose
rows are scattered becomes many small "clusters", and the sandwich
variance is computed over those fragments -- a classic and silent
GEE pitfall:

```{r footgun}
set.seed(3)
panel <- data.frame(id = rep(1:30, each = 4), x = rnorm(120))
panel$y <- panel$x + rep(rnorm(30), each = 4) + rnorm(120)
shuffled <- panel[sample(nrow(panel)), ]

fit_shuffled <- geeglm(y ~ x, id = id, data = shuffled, family = gaussian)
table_regression(fit_shuffled, show_fit_stats = c("nobs", "n_groups", "max_cluster_size"))
```

The table reports well over a hundred clusters for what the analyst
knows to be 30 subjects -- because that is what the model actually
used. `table_regression()` deliberately reads the cluster structure
from the fit (`geese$clusz`, what `summary()` reports as *Number of
clusters*) rather than counting unique `id` values, so the table
*diagnoses* the mistake instead of papering over it. The remedy is
to sort by `id` before fitting (and to use geeglm's `waves =`
argument when series are incomplete):

```{r footgun-fixed}
sorted <- shuffled[order(shuffled$id), ]
fit_sorted <- geeglm(y ~ x, id = id, data = sorted, family = gaussian)
table_regression(fit_sorted, show_fit_stats = c("nobs", "n_groups", "max_cluster_size"))
```

## Few clusters: the jackknife variants

The sandwich estimator is asymptotic in the number of clusters; with
only a few dozen, it tends to be anti-conservative. `geeglm` offers
jackknife approximations as alternatives (`std.err = "jack"`,
`"j1s"`, `"fij"`; Halekoh et al., 2006). The estimator choice lives
on the fit, and the table reads and names whatever the fit computed:

```{r jackknife}
fit_j1s <- geeglm(
  Weight ~ Time + Cu,
  id      = Pig,
  data    = dietox,
  family  = gaussian,
  corstr  = "exchangeable",
  std.err = "j1s"
)
table_regression(fit_j1s)
```

The footer now reads *One-step jackknife (GEE), clusters by Pig* --
the displayed uncertainty is always the fit's own, never a spicy
recomputation.

## A binomial GEE end to end

The `respiratory` trial (111 patients, four visits) illustrates the
full workflow for a binary outcome. `exponentiate = TRUE` displays
population-averaged odds ratios, and the opt-in `"n_events"` column
adds events/N per group -- computed on the estimation sample:

```{r binomial}
data(respiratory, package = "geepack")
respiratory$outcome <- as.integer(respiratory$outcome)
respiratory$subject <- interaction(respiratory$center, respiratory$id)
# Placebo as the reference level, so the displayed row is the
# active-treatment odds ratio.
respiratory$treat <- relevel(respiratory$treat, ref = "P")

fit_resp <- geeglm(
  outcome ~ treat + age + baseline,
  id     = subject,
  data   = respiratory,
  family = binomial,
  corstr = "exchangeable"
)
table_regression(
  fit_resp,
  exponentiate = TRUE,
  show_columns = c("n_events", "b", "ci", "p")
)
```

This odds ratio is marginal: it compares the odds of a positive
outcome between the treated and untreated *populations*. The
subject-specific odds ratio from the matching mixed model is larger
-- the attenuation of the first section -- and the
[mixed-effects vignette](table-regression-mixed.html) shows the two
side by side.

Average marginal effects translate the odds ratio onto the
probability scale, and for a GEE fit their uncertainty inherits the
sandwich covariance automatically:

```{r ame, eval = have_me}
table_regression(
  fit_resp,
  show_columns = c("b", "p", "ame", "ame_ci")
)
```

The AME column reads as percentage-point differences in the
population-averaged probability of a positive outcome -- for the
treatment row, between the treated and placebo populations.

## What spicy refuses for GEE, and why

`table_regression()` treats requests that GEE cannot honour as hard,
explained errors rather than silent approximations:

* **`vcov = "HC1"` / `"CR2"` and `cluster =`.** GEE inference is
  robust by construction: the sandwich, clustered on the model's own
  `id =`, *is* the default display. There is nothing for spicy's
  robust-variance machinery to add; to change the estimator, refit
  with geeglm's `std.err =` option, and to change the clustering,
  refit with a different `id =`.
* **`standardized`.** There is no established convention for
  standardizing population-averaged coefficients: the response SD
  mixes within- and between-cluster variance. Standardize predictors
  before fitting if magnitudes must be compared.
* **`nested = TRUE` and likelihood-based statistics.** GEE is
  estimated by quasi-likelihood: there is no likelihood, hence no
  likelihood-ratio change test, no AIC and no pseudo-R². Compare
  working models with QIC, or use `geepack`'s `anova()` Wald tests
  outside the table.

Each refusal names its reason and its remedy in the error message --
in a table meant for publication, a wrong convention silently
applied is worse than an error.

```{r refusal, error = TRUE}
table_regression(fit, vcov = "HC3")
```

## Presentation-ready output

Everything above renders identically through the rich output
engines. A final example with `flextable`, ready for a Word
manuscript:

```{r rich, eval = build_rich_tables && have_ft}
table_regression(
  fit_resp,
  exponentiate = TRUE,
  show_columns = c("n_events", "b", "ci", "p"),
  output = "flextable"
)
```

```{r rich-fallback, eval = !build_rich_tables, echo = FALSE}
# Console fallback so the vignette stays self-contained outside
# pkgdown builds.
```

## References

* Halekoh, U., Højsgaard, S., & Yan, J. (2006). The R package
  geepack for generalized estimating equations. *Journal of
  Statistical Software*, 15(2), 1–11.
* Hubbard, A. E., Ahern, J., Fleischer, N. L., Van der Laan, M.,
  Lippman, S. A., Jewell, N., Bruckner, T., & Satariano, W. A.
  (2010). To GEE or not to GEE: Comparing population average and
  mixed models for estimating the associations between neighborhood
  risk factors and health. *Epidemiology*, 21(4), 467–474.
* Liang, K.-Y., & Zeger, S. L. (1986). Longitudinal data analysis
  using generalized linear models. *Biometrika*, 73(1), 13–22.
* Pan, W. (2001). Akaike's information criterion in generalized
  estimating equations. *Biometrics*, 57(1), 120–125.
* Zeger, S. L., Liang, K.-Y., & Albert, P. S. (1988). Models for
  longitudinal data: A generalized estimating equation approach.
  *Biometrics*, 44(4), 1049–1060.
