---
title: "Survival regression tables"
description: >
  Publication-ready tables for survival regression in R with
  `table_regression()`: Cox hazard ratios with events and concordance,
  cluster-robust (Lin-Wei) variance for multi-centre data, hierarchical
  Cox comparisons by partial-likelihood ratio test, accelerated failure
  time models with time ratios, and the `survival`, `rms`, and
  `flexsurv` engines.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Survival regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_survival <- requireNamespace("survival", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_survival
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_rms      <- have_survival && requireNamespace("rms", quietly = TRUE)
have_flexsurv <- have_survival && requireNamespace("flexsurv", quietly = TRUE)
have_broom    <- have_survival && requireNamespace("broom", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(survival)
```

This vignette covers **survival (time-to-event) regression** — models
for the **time until an event occurs**, with censoring for the
subjects whose event is not observed within follow-up. The companion vignette [*Publication-ready
regression tables*](table-regression.html) covers the shared mechanics
(`vcov`, `ci_level`, output formats, multi-model layouts, broom
integration); here we focus on what is specific to survival fits:
hazard ratios and time ratios, event counts and concordance, and the
variance estimators of multi-centre studies.

`table_regression()` supports four survival engines:

* **`survival::coxph()`** — the semiparametric Cox proportional
  hazards model, the field default;
* **`survival::survreg()`** — parametric accelerated failure time
  (AFT) models;
* **`rms::cph()`** — the Cox model in the `rms` ecosystem;
* **`flexsurv::flexsurvreg()`** — fully parametric models over a wide
  family of distributions.

The running example is `survival::lung`: survival of 228 patients with
advanced lung cancer, with age, sex, and the ECOG performance score
(0 = fully active, higher is worse) as predictors. One row lacks
`ph.ecog` and one lacks the enrolling institution, so we declare the
analytic sample explicitly — the same 226 rows serve every model and,
later, supply the cluster variable:

```{r data}
lung2 <- na.omit(lung[, c("time", "status", "age", "sex",
                          "ph.ecog", "inst")])
lung2$sex <- factor(lung2$sex, levels = 1:2,
                    labels = c("Male", "Female"))
nrow(lung2)
```

## The Cox model in one paragraph

The Cox model works on the **hazard** — the instantaneous event rate
among those still at risk. A one-unit increase in a predictor
multiplies that hazard by a constant factor — the exponentiated
coefficient — at every point in time (the **proportional-hazards
assumption**), and the baseline hazard itself is left completely
unspecified: that is the semiparametric trick that made the model
ubiquitous in applied work (Cox 1972; Therneau & Grambsch 2000).
Everything below also assumes censoring is **uninformative**: given
the covariates, subjects censored at time *t* must be representative
of those still at risk at *t* — an assumption the observed data
cannot verify. Two consequences shape
the table. There is **no
intercept row** — the baseline hazard absorbs it — and there is no
residual variance, so the model-fit block reports what a survival
reader expects instead: the **number of events** and the
**concordance**.

## Basic table: hazard ratios

```{r cox}
cx <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung2)
table_regression(cx, exponentiate = TRUE)
```

Reading the table:

* `exponentiate = TRUE` turns each log-hazard coefficient into a
  **hazard ratio**. Women's hazard is 0.57 times men's — a 43% lower
  hazard at any given moment, all else equal; each ECOG point
  multiplies the hazard by 1.60. Leave `exponentiate` off for the
  log-hazard `B` column.
* Inference is **Wald-z** on the log scale, matching `summary.coxph()`
  and Stata `stcox`; the statistic and p-value are invariant under the
  exponentiation.
* The footer reports **events** (163 deaths among 226 patients — the
  precision and power of a survival model are governed chiefly by its
  event count, not its n) and the **concordance** C = 0.64: the probability that, of
  two comparable patients, the one predicted to fail earlier actually
  does. 0.5 is a coin flip; 0.64 is modest discrimination
  (Harrell 2015).

The table reports the model, not its assumption: check
proportionality with `survival::cox.zph()` before publishing
(Therneau & Grambsch 2000):

```{r cox-zph}
cox.zph(cx)
```

Each covariate gets its own test of the proportional-hazards
assumption, plus a global test. The smallest p here is .12 (sex) and
the global test gives p = .17 — little evidence against
proportionality. When the test rejects, the two standard remedies
flow through the same table without ceremony: stratify the offending
categorical factor (`strata()` terms are absorbed into the baseline
hazard, no coefficient row) or give the covariate a time-varying
effect via counting-process data (`Surv(tstart, tstop, event)`);
both render like any other fit.

## Why there is no AME column

An `"ame"` request on a Cox fit is refused with an explanation
rather than silently honoured: an average marginal effect needs a
response scale, and a Cox model deliberately does not commit to one —
a per-coefficient "effect on survival" would be ambiguous (at what
time? on which scale?), and the delta-method standard errors the
generic AME machinery would attach to it are unreliable for Cox
fits. The hazard ratio *is* the per-coefficient summary of this
model. Absolute-scale summaries exist, but they require the analyst
to choose a **time horizon** first — which is exactly what the next
section does, with bootstrap inference in place of the delta method.

## Absolute effects: RMST and risk differences

A hazard ratio answers "at what relative rate?", never "how much
longer?" or "how many fewer?". Women's hazard is 0.57 times men's —
but how many extra months of life is that, and how many fewer deaths
by the end of the first year? Those are the questions patients and
policy makers actually ask, and the methodological literature
increasingly calls for these answers to be reported alongside the
hazard ratio (Royston & Parmar 2013; Uno et al. 2014). They also
settle a workflow question the assumption check left open: when
`cox.zph()` rejects proportionality, the hazard ratio degrades into
a follow-up-weighted average of a changing effect, while the RMST
difference over `[0, tau]` remains a well-defined estimand — in that
case the absolute summaries lead the report rather than accompany
it. Two absolute summaries do the job:

* the **restricted mean survival time (RMST) difference**: the
  difference in mean event-free time over a fixed window `[0, tau]`
  — the area between the two survival curves up to `tau`;
* the **risk difference**: the difference in the probability of
  having had the event by a landmark time.

`table_regression()` computes both from the fitted Cox model by
**g-computation** (regression standardization): every patient's
survival curve is predicted twice, once under each level of the
predictor, keeping their other covariates as observed; the curves
are averaged, and the two standardized curves are integrated to
`tau` (RMST) or evaluated at `at_time` and complemented
(risk = 1 − survival). Standard errors come from
a nonparametric bootstrap, so expect the chunk to take a few
seconds; the 200 replicates below keep the vignette light, and the
default `boot_n = 1000` is the better choice for a publication
table.

```{r cox-rmst}
set.seed(7)
table_regression(cx,
                 show_columns = c("b", "rmst", "rmst_ci",
                                  "risk_diff", "risk_diff_ci"),
                 tau = 365, at_time = 365,
                 exponentiate = TRUE, boot_n = 200)
```

The absolute columns re-express the same model on interpretable
scales, each averaged over the sample's covariate mix. That cuts
both ways: computed from the fit, they inherit its assumptions —
with proportionality in doubt, estimate the RMST nonparametrically
(per-arm Kaplan–Meier, as in `survRM2`) rather than standardizing a
misspecified model. Stratified fits work too: standardization keeps
each subject's own stratum baseline — only the exposure is set
counterfactually — and the note discloses it; the strata variable
itself gets no contrast row, exactly as it has no hazard ratio. Being female adds an average of **40.5
event-free days** over the first year (95% CI 17 to 64) and lowers
the one-year risk of death by **19 percentage points** (95% CI 8 to
30); each additional ECOG point costs 38 days and adds 15 percentage
points to one-year risk. And the age row teaches that the contrast
unit matters as much as the scale: per **year** of age the effect is
invisible (HR 1.01, under a day of RMST), but per **decade** — the
clinically natural unit — the same fit implies HR 1.12 and about
nine fewer event-free days. Rescale the predictor when the per-unit
contrast is not the meaningful one.

Two design decisions shape these columns. First, **the horizon is
mandatory**: an RMST without its `tau` is not an estimand, so
`table_regression()` refuses to guess — pass a clinically meaningful
time, or `tau = "minmax"` for the largest window every compared
group can support (the value used is disclosed in the table note).
Second, the contrast is level against reference for factor
predictors and a +1-unit shift for continuous predictors, mirroring
the AME convention for other model families — so the columns read
the same way across your tables.

## Multi-centre data: cluster-robust variance

The `lung` patients were enrolled by 18 institutions, and outcomes
within an institution may correlate. The `CR*` request routes to the
**Lin & Wei (1989)** robust variance — the grouped score (dfbeta)
sandwich that `coxph(..., cluster = )` and Stata `stcox, vce(cluster)`
use:

```{r cox-cluster}
table_regression(cx, vcov = "CR0", cluster = ~inst,
                 exponentiate = TRUE)
```

Two lessons here. First, the mechanics: the cluster variable must
provide **one value per row of the model data** — declaring the
analytic sample up front (as we did) keeps the cluster aligned when
predictors carry missing values. Second, the direction: robust
standard errors are not always larger — the sex SE *shrinks* here.
Clustering corrects the variance in whichever direction the
within-centre correlation points; treating it as a conservative
inflation ritual misreads it. The robust variance corrects the
standard errors while leaving the estimates marginal
(population-averaged); the modeling alternative is a shared
**frailty** term — `coxph(Surv(time, status) ~ age + sex + ph.ecog +
frailty(inst), data = lung2)` — whose hazard ratios are instead
conditional on the centre effect (Therneau & Grambsch 2000, ch. 9).
Which estimand the analysis needs decides between them.

## Hierarchical Cox models

Does adding the performance score improve on the age-and-sex model? `nested = TRUE`
compares adjacent models by the **partial-likelihood ratio test** (the
`anova.coxph()` convention) and appends the change rows:

```{r cox-nested}
cx0 <- coxph(Surv(time, status) ~ age + sex, data = lung2)
table_regression(list(cx0, cx), nested = TRUE, exponentiate = TRUE,
                 show_columns = c("b", "p"))
```

Adding `ph.ecog` buys a chi-squared change of 16.78 on one degree of
freedom (p < .001), an AIC drop of 14.8, and a rise in concordance from
0.60 to 0.64 — three readings of the same improvement, on the
likelihood, information, and discrimination scales respectively.

## The univariable screen

Epidemiological analyses often begin one step earlier: before any
multivariable model, each candidate predictor is examined in its own
Cox model, and the two stages are reported side by side.
`table_regression_uv()` runs that workflow in one call — `outcome`
takes the `Surv()` expression, and each column group is fitted, and
its n and events counted, on its own estimation sample:

```{r cox-uv}
table_regression_uv(lung2, outcome = Surv(time, status),
                    predictors = c(age, sex, ph.ecog),
                    method = "coxph", exponentiate = TRUE,
                    show_columns = c("n", "n_events", "b", "ci", "p"))
```

The `Events/N` column carries the information a survival reader
checks first — 53 of the 90 women died against 110 of the 136 men —
and the paired columns support the screen's real question: **does
adjustment move an estimate?** Age is the instructive row. Alone it
looks significant (HR 1.02, p = .041); adjusted for sex and
performance status it is not (p = .225) — its univariable signal was
partly carried by the worse performance scores of the older
patients. Reading the two columns together is what protects the
final model from both false leads and premature exclusions: a
predictor should not enter or leave on its univariable p-value
alone. The screen also takes the estimand columns: with
`show_columns = c("b", "rmst")` and one shared numeric `tau`, each
predictor's unadjusted RMST difference sits next to the
covariate-adjusted one from the multivariable fit — the same
univariable-versus-adjusted reading, on the absolute scale.

Two cautions complete the reading. The comparison is clean here only
because the analytic sample was declared up front: every model uses
the same 226 rows, so movement between the columns is adjustment,
not sample change (with missing predictor values, each univariable
model fits on its own larger complete cases — the `N` column and the
table note disclose exactly that). And when the estimate is a hazard
ratio, movement under adjustment is not by itself proof of
confounding: hazard ratios are **non-collapsible**, so adding a
strongly prognostic covariate shifts the estimate even when it is
unrelated to the predictor.

## Accelerated failure time: `survreg()`

The parametric alternative models **time itself** (Wei 1992):
covariates stretch or compress survival time by a constant factor.
When is it the better tool? When the audience needs the time-scale
reading, when proportionality is in doubt — the AFT assumption of a
constant time ratio is a *different* assumption, not a weaker one —
or when the analysis needs a parametric baseline, for instance to
extrapolate beyond the follow-up window. Under
`exponentiate = TRUE` the coefficients become **time ratios** (TR):

```{r survreg}
sr <- survreg(Surv(time, status) ~ age + sex + ph.ecog, data = lung2,
              dist = "weibull")
table_regression(sr, exponentiate = TRUE)
```

The intercept is back — a parametric model must anchor the time
scale, and exp(intercept) ≈ 801 days is that anchor (the extrapolated
characteristic time at the reference levels and age 0), not a ratio
despite the column header.

Set the TR next to the Cox HR and the numbers *look* contradictory:
women have TR 1.50 but HR 0.57. They are the same finding on opposite
scales — a protective effect **multiplies survival time up** (50%
longer) and **multiplies the hazard down** (43% lower). Neither
direction is more correct; the Weibull is the one family
that is simultaneously AFT and proportional-hazards, and the two
scales are linked by an exact identity within the fit:
HR = TR^(−1/scale) = 1.50^(−1/0.73) ≈ 0.574, with the scale
parameter read from the footer — agreeing with the semiparametric
Cox estimate (0.573 at full precision, 0.57 as the tables display). The footer names the
distribution because the interpretation depends on it —
`exponentiate` produces time ratios only for log-scale distributions
(Weibull, exponential, lognormal, loglogistic); for an identity-scale
distribution (`dist = "gaussian"`) the request is skipped with a
warning, since the coefficients already act on the time scale.

The RMST and risk-difference columns work here exactly as for the Cox
model — same tokens, same `tau` / `at_time`, same bootstrap — with one
advantage: the AFT curves are closed-form, so the standardized
survival is smooth rather than a step function. The estimates answer
the same absolute-scale question, computed under the parametric
assumptions the footer names.

## `rms::cph()`

Analysts working in the `rms` ecosystem (Harrell 2015) — for its
validation, calibration, and nomogram tooling — fit the Cox model
with `rms::cph()`, and `table_regression()` renders it through the
same layout. One requirement carried over from `rms` itself:
cluster-robust variance uses `rms::robcov()`, which needs the model
matrix and response stored at fit time — fit with
`x = TRUE, y = TRUE` or the robust request fails with exactly that
instruction:

```{r cph, eval = have_rms}
cph_fit <- rms::cph(Surv(time, status) ~ age + sex, data = lung2,
                    x = TRUE, y = TRUE)
table_regression(cph_fit, vcov = "CR0", cluster = ~inst,
                 exponentiate = TRUE)
```

## Fully parametric: `flexsurv`

`flexsurv::flexsurvreg()` (Jackson 2016) fits parametric survival
models over a wide family of distributions (and the Royston–Parmar
2002 spline models). The location
coefficients are exponentiated under a generic `exp(B)` header — the
substantive reading (time ratio, hazard ratio) depends on the
distribution, which the footer names together with its ancillary
parameters:

```{r flexsurv, eval = have_flexsurv}
fs <- flexsurv::flexsurvreg(Surv(time, status) ~ age + sex,
                            data = lung2, dist = "weibull")
table_regression(fs, exponentiate = TRUE)
```

Two guardrails apply. A spline model on the `"normal"` scale places
its coefficients on a probit-like scale whose exponential has no
meaning, so `exponentiate = TRUE` is refused there; and covariates on
ancillary parameters (`anc =`) act on their own scales, so
`exponentiate = TRUE` is refused for the whole fit rather than
printing meaningless ratios for the ancillary rows.

## Output formats

Everything above used the default console output. The same table
renders as a raw data frame, a long broom-style tibble, or — with the
corresponding Suggests package — a rich `gt`, `flextable`,
`tinytable`, Excel, or Word table:

```{r outputs-df}
table_regression(cx, exponentiate = TRUE, output = "data.frame")
```

```{r outputs-gt, eval = build_rich_tables && have_survival}
table_regression(cx, exponentiate = TRUE, output = "gt")
```

[`broom::tidy()`](https://broom.tidymodels.org) returns the long
frame; with `exponentiate = TRUE` the estimates and CI bounds are on
the ratio scale, as displayed:

```{r broom, eval = have_broom}
broom::tidy(table_regression(cx, exponentiate = TRUE))
```

## References

* Cox, D. R. (1972). Regression models and life-tables. *Journal of
  the Royal Statistical Society, Series B*, 34(2), 187–220.
* Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.).
  Springer.
* Jackson, C. H. (2016). `flexsurv`: A platform for parametric
  survival modeling in R. *Journal of Statistical Software*, 70(8).
* Lin, D. Y., & Wei, L. J. (1989). The robust inference for the Cox
  proportional hazards model. *Journal of the American Statistical
  Association*, 84(408), 1074–1078.
* Royston, P., & Parmar, M. K. B. (2002). Flexible parametric
  proportional-hazards and proportional-odds models for censored
  survival data. *Statistics in Medicine*, 21(15), 2175–2197.
* Royston, P., & Parmar, M. K. B. (2013). Restricted mean survival
  time: an alternative to the hazard ratio for the design and
  analysis of randomized trials with a time-to-event outcome. *BMC
  Medical Research Methodology*, 13, 152.
* Uno, H., Claggett, B., Tian, L., et al. (2014). Moving beyond the
  hazard ratio in quantifying the between-group difference in
  survival analysis. *Journal of Clinical Oncology*, 32(22),
  2380–2385.
* Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data:
  Extending the Cox Model*. Springer.
* Wei, L. J. (1992). The accelerated failure time model: a useful
  alternative to the Cox regression model in survival analysis.
  *Statistics in Medicine*, 11(14–15), 1871–1879.
