---
title: "Ordinal regression tables"
description: >
  Publication-ready tables for ordinal (proportional-odds / cumulative
  logit) regression in R with `table_regression()`: shared slope
  coefficients and ordered thresholds, odds ratios, per-outcome-category
  average marginal effects laid out as a probability matrix,
  cluster-robust variance, the `MASS::polr` and `ordinal::clm` engines,
  testing the proportional-odds assumption, and the partial
  proportional odds and scale relaxations.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Ordinal regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_ordinal <- requireNamespace("ordinal", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
library(MASS)        # polr()
```

This vignette covers **ordinal regression** ŌĆö models for an *ordered*
categorical response such as a Poor &lt; Fair &lt; Good &lt; Very good
health rating or a Likert agreement scale. The companion vignette
[*Publication-ready regression tables*](table-regression.html) covers
the shared mechanics (`vcov`, `ci_level`, output formats, multi-model
layouts, broom integration); here we focus on what is *specific* to
ordinal fits.

`table_regression()` supports both ordinal engines:

* **`MASS::polr()`** ŌĆö proportional-odds cumulative logit (also probit /
  cloglog / cauchit via `method =`);
* **`ordinal::clm()`** ŌĆö the more flexible cumulative link model, with
  optional scale and partial-proportional-odds (nominal) effects.

The model used throughout is a self-rated-health rating regressed on
age, sex, smoking, and physical activity, using the bundled `sochealth`
survey data (`?sochealth`). `self_rated_health` is an ordered factor:

```{r data}
levels(sochealth$self_rated_health)
```

Check this before fitting anything. `polr()` and `clm()` estimate on
the factor's level order *as given* ŌĆö they do not warn when the outcome
is a plain (unordered) factor, or when the levels sit in the wrong
order. Data read from text files is particularly exposed: `factor()`
defaults to alphabetical levels (here `Fair, Good, Poor, Very good`),
which would silently fit a nonsensical scale. And reversing the order
negates every coefficient and marginal effect. Every interpretation
below hinges on knowing the direction of the scale ŌĆö here worst to
best (Long &amp; Freese 2014).

## The proportional-odds model

A cumulative logit model with $K$ ordered response categories estimates
**one slope per predictor** (shared across all $K-1$ cumulative
splits ŌĆö the *proportional-odds assumption*) plus $K-1$ ordered
**thresholds** (cut-points). A positive location (proportional-odds)
slope means that higher values of the predictor push the response
toward *higher* categories. There is deliberately **no per-category
coefficient**: the proportional-odds restriction collapses the effect
into a single slope per predictor, trading per-category detail for one
interpretable number. The per-category structure reappears in the
*marginal effects* (below), where a one-unit change has a different
effect on the probability of each category ŌĆö and in the relaxations of
the shared-slope restriction that `ordinal::clm()` offers (tested and
demonstrated below).

## Basic table

Pass a fitted `polr()` object. `table_regression()` renders the shared
slope coefficients with a Wald-$z$ inference regime, followed by a
subordinate **`Thresholds`** block listing the ordered cut-points with
their own B / SE / CI / p (the field convention: `summary.polr()`, SPSS
PLUM, SAS, Stata `ologit`, Bender & Grouven 1997):

```{r basic}
fit <- polr(
  self_rated_health ~ age + sex + smoking + physical_activity,
  data = sochealth, Hess = TRUE
)
table_regression(fit)
```

Reading the table:

* Each predictor has **one** coefficient row (proportional odds);
  factor predictors are grouped under their parent variable with the
  reference level carrying `(ref.)` and an em-dash.
* Inference is **Wald-$z$** (`df = Inf`): ordinal MLE has no residual
  degrees of freedom, matching `summary.polr()`, Stata `ologit`, and
  SPSS PLUM.
* A subordinate **`Thresholds`** block lists the cut-points
  (`Poor | Fair`, `Fair | Good`, `Good | Very good`) with B / SE / CI /
  p, like the predictor rows. They locate the category boundaries on the
  latent logit scale and replace the single intercept of a binary logit.
  They are reported on the log-odds scale and are **never
  exponentiated**. The $z$-test against zero is rarely of interest ŌĆö it
  only asks whether the cumulative split sits at 50/50 at the profile
  where *every* predictor equals zero, here a respondent aged 0, not a
  meaningful case ŌĆö so do not over-read a "significant" threshold; the
  substance lies in the odds ratios and marginal effects below. Hide
  the block with `show_thresholds = FALSE` to fall back to a compact
  footer line.
* Below the fit-stats rule, the model-fit block reports **N**, two
  pseudo-$R^2$ (**McFadden**, the Stata `ologit` default, and
  **Nagelkerke**, the SPSS PLUM default), and **AIC**. Both compare the
  fitted log-likelihood to the intercept-only model's, and both round
  to 0.00 here: these four predictors explain almost none of the
  variation in self-rated health. This survey model therefore
  illustrates table mechanics and interpretation *wording*; read the
  exemplary sentences below as templates, not as substantive findings.
  Override the block with `show_fit_stats`
  (e.g. `show_fit_stats = c("nobs", "AIC", "BIC")`).

## Odds ratios: `exponentiate = TRUE`

On the logit scale a coefficient is a log cumulative-odds ratio.
`exponentiate = TRUE` reports **odds ratios** instead, exponentiating
the estimate and its CI bounds and relabelling the column header
accordingly. This is link-specific: under `method = "cloglog"` the
exponentiated coefficient is a **hazard ratio** (the grouped-time
proportional-hazards reading) and the header and footer relabel to HR;
under probit or cauchit the exponential has no ratio interpretation,
and `exponentiate = TRUE` is refused with a clear error.

```{r or}
table_regression(fit, exponentiate = TRUE)
```

The OR here is a *cumulative* odds ratio: it multiplies the odds of
scoring **above** any given cut-point rather than at or below it, and ŌĆö
because the model is proportional-odds ŌĆö the same ratio applies at
every cut-point (`Poor | Fair`, `Fair | Good`, `Good | Very good`). It
is *not* the odds of landing in one specific category; that
per-category detail is what the AME matrix below provides. For a
location (proportional) coefficient, an OR above 1 raises the
cumulative odds of being in a *higher* health category and an OR below
1 lowers them ŌĆö the non-proportional block later in this vignette
follows the opposite convention. Exact reading for the one borderline
effect here: smoking's OR of 0.76 means that, at every cut-point,
smokers have about 24% lower odds of being in a higher health category
rather than a lower one, adjusting for age, sex, and physical activity.
The CI upper bound prints as 1.00 but is 0.997 before rounding; with
p = .047 this is a borderline association, so treat the sentence as a
wording template. The `Thresholds` rows stay on the **log-odds scale**
(a cut-point is not an odds ratio, so it is never exponentiated); only
the predictor coefficients become ORs, and the footer flags this.

## Average marginal effects: a probability matrix

Odds ratios are multiplicative and describe the (cumulative) odds, not
probabilities directly. For a reader-friendly, **probability-scale**
summary, request average marginal effects (AME). Because an ordinal
model predicts a probability for *every* response category, each
predictor has **one AME per category** ŌĆö the effect of the predictor on
the probability of that category. `table_regression()` lays these out
as the field-standard **matrix**: predictors in rows, response
categories in columns (Long &amp; Freese 2014; Williams 2012).

```{r ame}
table_regression(fit, show_columns = c("b", "ame"))
```

How to read it, with the `smoking = Yes` row:

* The four `AME ...` columns are the average change in the probability
  of each health category for a smoker versus a non-smoker, holding the
  other predictors at their observed values.
* The effects **sum to Ōēł 0 across the row**: a predictor *redistributes*
  probability mass between categories (the total probability is always
  1). Here smoking moves mass out of the better categories and into the
  worse ones ŌĆö about 4.8 points out of `Very good`, 4.2 into `Fair`.
* Cells are **probabilities** (the 0ŌĆō1 scale, matching
  `marginaleffects::avg_slopes()` and the binary-`glm` AME). An AME of
  `0.04` is a change of **4 percentage points** ŌĆö *not* 4 percent. The
  footer note states this; reserve "percent" for multiplicative
  quantities such as odds ratios.

An AME is the single-number summary of a fuller object: the predicted
probability of each category as the predictor varies. When the story
deserves more than one number per category ŌĆö or to check that an
average is not masking a non-monotone pattern ŌĆö tabulate those
probabilities directly:

```{r avg-predictions}
marginaleffects::avg_predictions(fit, by = "smoking")
```

Smokers' expected distribution sits lower on the scale than
non-smokers' ŌĆö `Very good` 20.9% versus 25.8%, `Fair` 26.1% versus
21.9% ŌĆö and the smoking AME row above is exactly the difference between
these two profiles. For predicted-probability curves along a continuous
predictor, `marginaleffects::plot_predictions()` plots the same
quantities.

Request only the AME matrix (drop the coefficient column) with
`show_columns = "ame"`, or add the AME standard errors / CIs / p-values
with `"ame_se"`, `"ame_ci"`, `"ame_p"` (or the `"all_ame"` bundle):

```{r ame-only}
table_regression(fit, show_columns = "ame")
```

Marginal effects are the response-scale (probability) output of
[`marginaleffects::avg_slopes()`](https://marginaleffects.com). For a
single-outcome model (binary `glm`, `lm`, mixed) the AME stays a single
column ŌĆö the per-category matrix appears only when the response has
more than two categories.

## Cluster-robust standard errors

Ordinal fits honour the cluster-robust `vcov` family
(`"CR0"`ŌĆō`"CR3"`) via `sandwich::vcovCL()`. Pass the cluster as a
formula, a column name, or a vector (see the main vignette,
*How to specify `cluster`*):

```{r cluster}
table_regression(
  fit, vcov = "CR2", cluster = ~region,
  show_columns = c("b", "ame")
)
```

The footer switches to name the estimator and the clustering variable.
Heteroskedasticity-consistent (`HC*`) estimators and the `bootstrap` /
`jackknife` resamplers are *not* defined for ordinal fits and are
refused with a clear `spicy_unsupported_vcov` error rather than a
silent fallback.

## Standard errors and confidence intervals

The three inference regimes differ in **how the standard error and the
confidence interval relate**:

* **Wald** (default): both come from the model information matrix. The CI
  is `estimate ┬▒ z ├Ś SE` (symmetric) and the *p*-value uses the same SE ŌĆö
  SE, CI and *p* are one coherent set.
* **Robust / cluster-robust** (`vcov = "CR*"`): the whole set switches to
  the sandwich estimator ŌĆö SE, CI (`┬▒ z ├Ś SE_robust`) and *p* shift
  together, still coupled.
* **Profile likelihood** (`ci_method = "profile"`): the CI is inverted
  from the likelihood-ratio statistic and is **asymmetric** ŌĆö *not*
  `estimate ┬▒ z ├Ś SE`. Profile is a **CI-only refinement**: the estimate,
  SE, statistic and *p*-value stay Wald; only the CI changes. It covers
  the predictor coefficients (via `confint()`); the thresholds stay Wald.
  A robust `vcov` takes precedence (profile is model-based), so requesting
  both uses the robust Wald CIs.

Because a profile CI cannot be reconstructed from the SE, the footer
**discloses** it (`95% CIs: profile likelihood.`) alongside the SE method
ŌĆö following APA 7 / SAMPL / STROBE and matching
`parameters::model_parameters()`:

```{r profile}
table_regression(fit, ci_method = "profile", show_columns = c("b", "ci", "p"))
```

## The `ordinal::clm()` engine

`ordinal::clm()` is rendered through the same code path and supports the
same columns. It is the engine to reach for when you need flexible link
functions, scale effects, or a relaxation of proportional odds.

```{r clm, eval = have_ordinal}
clm_fit <- ordinal::clm(
  self_rated_health ~ age + sex + smoking + physical_activity,
  data = sochealth
)
table_regression(clm_fit, show_columns = c("b", "ame"))
```

### Testing proportional odds

The shared-slope restriction is a testable hypothesis, and the workflow
is: fit, **test**, then relax only what fails (Long &amp; Freese 2014).
`ordinal` provides a likelihood-ratio test per predictor:
`nominal_test()` refits the model freeing each predictor across the
cut-points and compares, and `scale_test()` does the same for a scale
(dispersion) effect. For `polr` there is no built-in equivalent ŌĆö the
classic logit-specific check is Brant's (1990) test (package `brant`);
`clm`'s tests are the likelihood-based generalisation.

```{r po-test, eval = have_ordinal}
ordinal::nominal_test(clm_fit)
ordinal::scale_test(clm_fit)
```

Each `nominal_test` row asks: does freeing this predictor across the
$K-1$ cut-points (here 2 extra parameters) improve the fit? No term
clearly fails. Smoking is the one borderline case (LRT = 5.15, df = 2,
p = .076); age, sex and physical activity are firmly consistent with
proportional odds (p &gt; .66), and no term shows a scale effect
(p &gt; .21). The decision rule is to relax *only* the terms that
fail ŌĆö here none clearly does, so the proportional-odds table above is
adequate as the primary analysis. Two caveats temper the rule: at
moderate sample sizes these per-term tests have limited power, and one
test per predictor invites multiplicity ŌĆö so a borderline result argues
for reporting the relaxed fit as a *sensitivity analysis*, not for
switching automatically. In large samples the opposite failure
dominates: the tests reject routinely ŌĆö Long &amp; Freese (2014) report
rejection in the majority of real applications ŌĆö while also being
sensitive to other kinds of misspecification, and the freed model's
*predictions* often barely differ from the proportional fit's. When a
test rejects, compare predicted probabilities across the two fits
before restructuring the model. And a violated assumption is never a
rationale for linear regression on the ordinal outcome, whose
assumptions are stronger still (Long &amp; Freese 2014). The next two
sections demonstrate the two relaxations, using smoking precisely
because it is the borderline term.

### Partial proportional odds: `nominal = ~`

A **nominal** component frees a predictor's effect across the
cut-points: a separate coefficient *per cut-point* instead of one
shared slope ŌĆö Peterson &amp; Harrell's (1990) *partial* proportional
odds. These render as a labelled **`Non-proportional effects`** block ŌĆö
one row per cut-point ŌĆö between the proportional coefficients and the
thresholds:

```{r clm-nominal, eval = have_ordinal}
clm_npo <- ordinal::clm(
  self_rated_health ~ age, nominal = ~ smoking,
  data = sochealth
)
table_regression(clm_npo)
```

The freed structure is read *within* the block: the smoking effect is
concentrated at the lowest split (`Poor | Fair`: B = 0.82, p = .003)
and near-null at the two higher splits (B = 0.22 and 0.24, p &gt; .16).
That heterogeneity is what the shared slope averaged away, and what
relaxing proportional odds buys.

One trap, and it matters: **nominal coefficients carry the opposite
sign to location coefficients.** `clm` folds nominal terms into the
threshold side of the model ŌĆö they shift the cut-points $\theta_j$
rather than enter the slope ŌĆö so a *positive* nominal coefficient
raises the cumulative odds of being **at or below** that cut-point,
pushing toward the *lower* categories there. Under
`exponentiate = TRUE` the smoking row at `Poor | Fair` becomes
OR = 2.28: a smoker's odds of rating health *Poor* (rather than Fair or
better) are 2.28 times a non-smoker's. That is the same worsening of
health the proportional fit expressed as OR = 0.76 (&lt; 1) on the
higher-category side ŌĆö reverse polarity, because the nominal term
enters the threshold, not the slope. Do not apply the earlier
location-coefficient sign rules to this block.

Two caveats for the non-proportional terms, both surfaced by the table:
robust / cluster SEs are **not available** (`sandwich` has no
estimating-functions method, so a robust `vcov` is refused), and
`ci_method = "profile"` covers the proportional coefficients only (the
non-proportional and threshold rows stay Wald).

### Scale effects: `scale = ~`

A **scale** component lets the *dispersion* of the latent response
depend on covariates, relaxing the constant-variance assumption of the
basic model. The motivating question is about spread, not location: is
latent health more *variable* among smokers than non-smokers? Test it
like any nested pair:

```{r clm-scale-test, eval = have_ordinal}
clm_loc <- ordinal::clm(
  self_rated_health ~ age + smoking,
  data = sochealth
)
clm_scale <- ordinal::clm(
  self_rated_health ~ age + smoking, scale = ~ smoking,
  data = sochealth
)
anova(clm_loc, clm_scale)
```

No evidence here (LR = 1.51, df = 1, p = .220). The scale coefficient
lives on the **log standard-deviation** metric: the estimate of 0.09
says smokers' latent SD is exp(0.09) Ōēł 1.10 ŌĆö about 10% larger ŌĆö but
the CI includes 0 on the log scale, so the data cannot distinguish it
from equal spread. The table renders the location coefficients, a
subordinate **`Scale effects`** block for the scale coefficients, and
the thresholds; the footer states the log-SD scale. Scale rows are
never exponentiated ŌĆö their exponential is a ratio of latent standard
deviations, not an odds ratio:

```{r clm-scale, eval = have_ordinal}
table_regression(clm_scale)
```

Cluster-robust SEs are not available for scale fits (`sandwich` has no
estimating-functions method for them), so `CR*` is refused up front
with a clear `spicy_unsupported_vcov` error:

```{r clm-scale-refuse, eval = have_ordinal, error = TRUE}
table_regression(clm_scale, vcov = "CR2", cluster = ~region)
```

### Choosing an ordinal specification

The driving question is what to do when proportional odds fails ŌĆö or
was never plausible.

* When the tests pass, the shared-slope cumulative model is the most
  parsimonious choice: one slope per predictor, one table row each.
* When a *few* predictors fail, free just those with `nominal = ~` and
  report the freed fit, at least as a sensitivity analysis.
* When the assumption fails *broadly*, or the ordering of the response
  is itself doubtful, abandon ordinality for **multinomial logit** ŌĆö
  the standard unconstrained alternative (Long &amp; Freese 2014),
  compared to the ordinal fit by AIC ŌĆö covered in the
  [multinomial vignette](table-regression-multinomial.html).
* For asymmetric responses and grouped survival times, the **cloglog**
  link (`method = "cloglog"` in `polr`, `link = "cloglog"` in `clm`)
  replaces the odds-ratio reading with a hazard-ratio one, connecting
  the link menu above to a modelling rationale.

## Several models side by side

The same fit can appear several times with different estimators, or
different fits can be compared, with `table_regression(list(...))`.
Per-model `vcov` is supplied as a list:

```{r multi}
fit_min <- polr(self_rated_health ~ smoking, data = sochealth, Hess = TRUE)
table_regression(
  list(Unadjusted = fit_min, Adjusted = fit),
  show_columns = c("b", "ame_p")
)
```

## Output formats

The default console table shown above is one of several targets. The
`output` argument also produces a raw data frame, a long broom-style
tibble, and ŌĆö with the corresponding Suggests package ŌĆö rich `gt`,
`flextable`, `tinytable`, Excel, or Word tables. The structure (the
per-category AME matrix included) carries through to every format.

```{r outputs-df}
head(table_regression(fit, show_columns = c("b", "ame"), output = "data.frame"))
```

```{r outputs-gt, eval = build_rich_tables}
table_regression(fit, show_columns = c("b", "ame"), output = "gt")
```

[`broom::tidy()`](https://broom.tidymodels.org) returns the long frame,
one row per `(term, estimate_type, outcome_level)`; each per-category
AME row is labelled by its response category in the `outcome_level`
column:

```{r broom}
broom::tidy(table_regression(fit, show_columns = c("b", "ame")))
```

## References

* Agresti, A. (2010). *Analysis of Ordinal Categorical Data* (2nd ed.).
  Wiley.
* Brant, R. (1990). Assessing proportionality in the proportional odds
  model for ordinal logistic regression. *Biometrics*, 46(4),
  1171ŌĆō1178.
* Long, J. S., &amp; Freese, J. (2014). *Regression Models for
  Categorical Dependent Variables Using Stata* (3rd ed.). Stata Press.
* Peterson, B., &amp; Harrell, F. E. (1990). Partial proportional odds
  models for ordinal response variables. *Journal of the Royal
  Statistical Society: Series C (Applied Statistics)*, 39(2), 205ŌĆō217.
* Williams, R. (2012). Using the margins command to estimate and
  interpret adjusted predictions and marginal effects. *The Stata
  Journal*, 12(2), 308ŌĆō331.
* Arel-Bundock, V., Greifer, N., &amp; Heiss, A. (2024). How to Interpret
  Statistical Models Using marginaleffects in R and Python. *Journal of
  Statistical Software*.
