---
title: "One outcome across several groupings"
description: >
  Build the transposed summary table in R with table_outcome(): one
  continuous outcome described across the levels of several
  categorical variables, one block of rows per variable, a group
  comparison per block, and a marginal Overall row — with the same
  statistic vocabulary and the same output engines as
  table_continuous().
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{One outcome across several groupings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

`table_outcome()` answers one question: **how does this outcome vary
across these groupings?** It fixes a single continuous outcome and
stacks the groupings underneath it, one block of rows per grouping.

That is the transpose of `table_continuous()`, which fixes a single
grouping and stacks the outcomes. The two tables hold the same
numbers when there is one of each; they differ in what a reader can
scan down the page, and the choice is a reporting decision rather
than a technical one.

## Which shape do I need?

The rule fits on one line.

- **Several outcomes across one grouping** — `table_continuous(select
  = , by = )`. The classic Table 1: age, BMI and well-being compared
  between two arms.
- **One outcome across one or several groupings** —
  `table_outcome(outcome = , by = )`. The classic "correlates of"
  table: who has a higher BMI — men or women, smokers or not, which
  region?

Both are legitimate with a single `by`. If you already know that more
groupings are coming, start here: adding a variable to `by` adds a
block, and nothing else in the call changes.

## A first table

```{r first}
table_outcome(sochealth, outcome = bmi, by = c(sex, smoking))
```

Four things to notice, because each is a decision:

1. **The title names the outcome.** The groupings are the rows, so a
   title listing them would repeat the stub — and with six blocks
   there would be no title left.
2. **One stub column.** The block header carries the *variable*; its
   levels are indented underneath. Nothing else is a label column.
3. **The `p` sits on the block header**, not on a level, because the
   comparison belongs to the whole block.
4. **The first row is `Overall`**, the marginal summary of the whole
   analytic sample.

## The `Overall` row and the denominators

The `Overall` row is what makes the blocks readable together: it is
the common denominator every block partitions.

Under the default `drop_na = FALSE`, that statement is exact. Each
block shows a `(Missing)` level for the rows whose grouping value is
missing, so the block's counts add up to the `Overall` count —
whatever the grouping.

```{r denominators}
tbl <- table_outcome(sochealth, bmi, by = c(sex, smoking))
raw <- as.data.frame(tbl)

# The marginal count.
raw$n[raw$.row_role == "summary"]

# And each block's, summed over its displayed levels.
described <- raw[raw$.row_role %in% c("level", "missing"), ]
tapply(described$n, described$variable, sum)
```

The outcome's own missing values are removed **once, globally**,
before any block, and the note says how many. That is the sentence
that reconciles the `Overall` count with the raw data — 1188 here,
not 1200.

`drop_na = TRUE` removes each block's missing-grouping rows instead of
showing them. The blocks then no longer share a denominator, and the
note reports the loss per variable:

```{r dropna}
table_outcome(sochealth, bmi, by = smoking, drop_na = TRUE)
```

### Why "Overall" and not "Total"

The two words are not interchangeable, and spicy keeps them apart.

*Total* is the word of a **count margin**: the column of
`table_categorical()` where frequencies add up, and where the margin
literally is the sum of the cells beside it. *Overall* is the whole
**analytic sample**: a row where a mean is recomputed over every
observation, and where nothing is added to anything. A mean is not a
total, and calling it one would be a reading error carried into every
translation of the table.

## Choosing the statistics

`show_columns` takes exactly the tokens `table_continuous()` takes,
with the same meanings — `"m"`, `"sd"`, `"med"`, `"iqr"`,
`"med_iqr"`, `"q1"`, `"q3"`, `"min"`, `"max"`, `"ci"`, `"med_ci"`,
`"n"`, `"weighted_n"`. See the *Choosing which statistics to show*
section of `vignette("table-continuous")` for the vocabulary itself;
nothing about it changes here.

```{r tokens}
table_outcome(
  sochealth,
  bmi,
  by = c(sex, education),
  show_columns = c("med_iqr", "n")
)
```

Only the character-vector form is accepted. The per-variable named
list of `table_continuous()` would name nothing here: there is one
outcome.

One consequence of that vocabulary is worth stating, because it is
automatic. **The table tests what it shows.** A display carrying a
median without a mean switches the default comparison to the
rank-based family — the Wilcoxon rank-sum test with two levels, the
Kruskal-Wallis test with three or more — so the table never tests a
mean it does not print. The note above says so. An explicit `test =`
is sovereign and warns instead.

## The group comparison, block by block

Every block is a **separate one-way comparison** of the outcome across
the levels of that variable. Nothing in this table adjusts one block
for another, and the note under the table says so in as many words.

```{r blocks}
table_outcome(
  sochealth,
  bmi,
  by = c(sex, region),
  statistic = TRUE,
  effect_size = "auto"
)
```

Read that table as a set of bivariate descriptions. It is not a model,
and the blocks are not independent of one another: sex and region are
correlated in any real sample, so a difference visible in one block
may be another block's difference seen through a third variable. When
the question is "does this hold when the others are accounted for",
the table you want is `table_regression()`, or
`table_continuous_lm()` with covariates.

Each block chooses its own test from its own level count. With
`test = "welch"` — the default — a two-level block gets the Welch
*t*-test and a three-or-more-level block gets the Welch one-way
ANOVA, which is why the note above names both. `effect_size = "auto"`
follows the test the block actually ran: Hedges' *g* beside a
*t*-test, eta-squared beside an ANOVA, their rank homologues beside a
rank test. A column can therefore mix measures from row to row, and
each cell prefixes its own glyph.

A block that cannot be compared degrades **alone**. Fewer than two
observed levels, or a level holding a single observation, and that
block keeps empty statistics while its neighbours are untouched.

```{r degraded}
thin <- data.frame(
  score = c(4, 5, 6, 7, 9, 11, 12, 14),
  arm = c("A", "A", "A", "A", "B", "B", "B", "B"),
  # One observed level: nothing to compare it with.
  site = rep("only one site", 8),
  # Two levels, but one of them holds a single observation.
  wave = c("first", "first", "first", "first", "first", "first", "first", "last")
)
table_outcome(thin, score, by = c(arm, site, wave))
```

## Choosing the `by` columns

`by` is a tidyselect expression, and the canonical forms are an
explicit enumeration or `where(is.factor)`:

```{r where-factor}
table_outcome(
  sochealth,
  wellbeing_score,
  by = where(is.factor),
  show_columns = c("m", "sd", "n")
)
```

Negation (`by = -c(x, y)`) is **not** recommended, and the reason is
worth a paragraph. It sweeps in every remaining column, and a numeric
one becomes a block with one *level* per distinct value, in order of
first appearance — a single block sixty rows long, where a reader
expects a handful of categories. The comparison still runs, so the
block is not empty; it is unreadable, which is worse, because it looks
like an answer. A variable producing more than twenty levels raises a
warning for that reason. The threshold is arbitrary and the help page
says so; the warning is a nudge, never a refusal, because the family
refuses no numeric grouping.

Labelled data deserves a word here, because importing from SPSS or
Stata is a central use of this package. A `haven_labelled` column used
as `by` shows its numeric CODES, not its value labels — the same as in
`table_continuous(by = )`. Convert it first (`haven::as_factor()`) to
get the labels in the stub.

The order of the blocks is the order you write, and the order of the
levels inside a block is the variable's own: a factor keeps its
declared level order (empty levels included — a level nobody chose is
information about the instrument), anything else takes the order of
first appearance.

## Weights

Two conventions coexist, and `rescale` is the switch between them —
the same choice `vignette("table-continuous")` sets out in full.

Without `rescale`, weights are read as **frequencies**: all weights 1
reproduces the unweighted table, and integer weights reproduce the
*statistics* of the data duplicated that many times. With
`rescale = TRUE` they are read as **sampling weights**, normalised so
that they sum to the sample size.

`sochealth$weight` is a survey design weight, so `rescale = TRUE` is
the reading you want for it:

```{r weights}
table_outcome(
  sochealth,
  bmi,
  by = sex,
  weights = weight,
  rescale = TRUE,
  p_value = FALSE,
  show_columns = c("m", "sd", "n", "weighted_n")
)
```

Note *statistics*, above, and not *the table*: the mean, SD, quantiles
and interval reproduce the expanded data exactly, while `n` stays the
raw count of rows that carried the weights. That is the point of
showing both counts, below.

Three things that call is doing on purpose.

`p_value = FALSE` is **required**. A weighted table refuses the group
comparison: the tests here have no weighted version, and a *p*-value
computed unweighted beside weighted descriptives would be silently
wrong in the one place a reader cannot check. For a weighted
comparison, use `table_continuous_lm(weights = )`.

`weighted_n` shows the sum of weights beside the raw count, because
the two answer different questions — how many people the estimate
represents, and how many rows carried it.

And `rescale` normalises over the outcome's whole surviving sample,
once, never per level. A per-level rescale would destroy the relative
weights *across* levels, which is the entire information a sampling
weight carries into this table. Every mean is unchanged by it; the
SDs move, because their denominator is the sum of weights minus one,
and that sum is exactly what rescaling changes.

## Citing a cell in running text

Every cell is addressable by identity, so a number quoted in a
sentence cannot drift from the number printed in the table. A level's
statistic needs its level; a **block's** statistic — the *p*, the
effect size, the test — has no level to name, and is addressed
without one.

```{r inline}
tbl <- table_outcome(sochealth, bmi, by = c(sex, smoking))

inline(tbl, sex, "Female", "m")
inline(tbl, sex, column = "p")
inline(tbl, bmi)
```

The last call is the most natural citation this table offers: a bare
`inline()` on the outcome cites the marginal mean, from the `Overall`
row.

Patterns compose several cells into one fragment, and follow the same
rule — every token must live on the same row:

```{r inline-pattern}
inline(tbl, sex, "Female", "{m} ({ci_label} {ci})")
inline(tbl, smoking, column = "{p}")
```

## Downstream: the raw frame and the broom views

`output = "long"` returns the compute frame: one row per displayed
row, with `.row_role` saying what each row is.

```{r long}
head(table_outcome(sochealth, bmi, by = sex, output = "long"), 4)
```

`tidy()` returns the described rows — the marginal row and one row per
(grouping x level) — and `glance()` returns one row per **block**,
carrying that block's own comparison:

```{r broom}
generics::tidy(tbl)
generics::glance(tbl)
```

Two identity columns rather than one, and deliberately: `outcome` is
constant down the frame, `variable` names the block. A single column
would have to mean two different things depending on the row.

The `glance()` schema is fixed. `smd_type` and `smd_value` are present
and `NA` from the first version, so the day a standardized mean
difference enters this table it cannot break a pipeline that indexes
the frame. Index by name, never by position.

## Comparison with `gtsummary::tbl_continuous()`

`gtsummary::tbl_continuous(data, variable = Y, include = c(A, B))`
builds the same shape. Where the two differ, each difference is a
decision:

- **the outcome is named** — in our title, on every engine; in
  gtsummary the column header reads only the sample size (`N = 1,200`
  on this data) and the outcome appears nowhere;
- **a marginal row** — we open on `Overall`, gtsummary has none;
- **missing groupings are shown** — a `(Missing)` level, so the block
  counts still add up; at its defaults gtsummary renders no such row;
- **the comparison is on by default** — gtsummary adds it with
  `add_p()`;
- **the quartiles are R's** — `med` / `q1` / `q3` equal
  `stats::median()` and `stats::quantile()` on the same vector (type
  7, R's default), while gtsummary computes type 2, the
  averaged-order-statistic definition of SAS and SPSS. The medians
  always agree; the quartiles need not;
- **the composite cell** reads `Med [Q1, Q3]` with frozen brackets,
  and its separator becomes a semicolon under a comma decimal mark,
  where a comma would otherwise serve two roles at once.

## Output engines

Everything above renders identically through six further engines. The
console is the reference; the six reproduce it cell for cell, indentation
and block rules included.

```{r engines-list, eval = FALSE}
table_outcome(sochealth, bmi, by = c(sex, smoking), output = "gt")
table_outcome(sochealth, bmi, by = c(sex, smoking), output = "tinytable")
table_outcome(sochealth, bmi, by = c(sex, smoking), output = "flextable")
table_outcome(
  sochealth,
  bmi,
  by = c(sex, smoking),
  output = "excel",
  excel_path = "bmi.xlsx"
)
table_outcome(
  sochealth,
  bmi,
  by = c(sex, smoking),
  output = "word",
  word_path = "bmi.docx"
)
table_outcome(sochealth, bmi, by = c(sex, smoking), output = "clipboard")
```

A journal style moves the defaults of the whole table, exactly as it
does elsewhere in the family:

```{r style}
table_outcome(
  sochealth,
  bmi,
  by = sex,
  style = "jama",
  show_columns = c("m", "sd", "n")
)
```

## Rendered output

```{r gt, eval = build_rich_tables}
pkgdown_dark_gt(
  table_outcome(
    sochealth,
    bmi,
    by = c(sex, education),
    effect_size = "auto",
    output = "gt"
  )
)
```

```{r tinytable, eval = build_rich_tables}
table_outcome(
  sochealth,
  wellbeing_score,
  by = c(sex, smoking),
  show_columns = c("med_iqr", "n"),
  output = "tinytable"
)
```

## See also

- `vignette("table-continuous", package = "spicy")` for the
  transposed shape — several outcomes across one grouping — and for
  the statistic vocabulary in full.
- `vignette("table-categorical", package = "spicy")` for categorical
  outcomes.
- `vignette("as-structured", package = "spicy")` for the typed view
  behind this table, and the geometry fields (`.row_role`, `.indent`)
  a custom renderer reads.
- `vignette("summary-tables-reporting", package = "spicy")` for the
  cross-function reporting workflow.
