---
title: "Continuous summary tables"
description: >
  Build continuous summary tables in R with table_continuous(),
  including grouped descriptives, group-comparison tests, effect sizes,
  and export to console, gt, tinytable, flextable, Excel, or Word —
  APA formatting by default, named journal styles (JAMA, NEJM,
  The Lancet, ...) by argument.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Continuous summary tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

`table_continuous()` summarizes continuous variables either overall or
by a categorical grouping variable. It is designed for readable summary
tables in the console and for publication-ready outputs in rendered
documents. When `by` is supplied, it can also add group-comparison
tests, test statistics, and effect sizes. Formatting follows APA
conventions by default; `style = "jama"` (or `"nejm"`, `"lancet"`,
`"annals"`, `"aer"`) switches the whole table to that journal's
published rules — the *House styles* section of
`vignette("summary-tables-reporting")` shows each style and the
guideline sentence behind every rule.

## Basic usage

Use `select` to choose the continuous variables you want to summarize:

```{r basic}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health)
)
```

If you omit `select`, `table_continuous()` scans the data frame and
keeps numeric columns. Non-numeric columns are dropped silently; set
`verbose = TRUE` to list them in a message:

```{r default-select}
table_continuous(sochealth)
```

## Grouped summaries

Add `by` to summarize the same variables across categories. When
`by` is supplied, a Welch-test *p*-value column is added automatically:

```{r grouped}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education
)
```

This is the main pattern for reporting continuous variables across
groups such as education, sex, treatment arm, or survey wave. Pass
`p_value = FALSE` to suppress the test column and keep the output
strictly descriptive.

If you want the same outcomes reported in a linear-model workflow,
with heteroskedasticity-consistent or cluster-robust standard errors,
case weights, or additive covariate adjustment, use
`table_continuous_lm()` instead:

```{r lm-companion}
table_continuous_lm(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  vcov = "HC3"
)
```

## Add test statistics and effect sizes

Grouped tables can also report the test statistic and an effect size
alongside the default *p*-value column:

```{r pvalue-effect}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  statistic = TRUE,
  effect_size = "auto",
  effect_size_ci = TRUE
)
```

Use `test = "student"` for equal-variance parametric tests or
`test = "nonparametric"` for rank-based comparisons:

```{r nonparametric}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score),
  by = education,
  test = "nonparametric",
  statistic = TRUE,
  effect_size = "auto"
)
```

`effect_size = "auto"` selects the canonical measure for the
chosen `test` and number of groups: Hedges' *g* (parametric, 2
groups), eta-squared (parametric, 3+ groups), rank-biserial *r*
(nonparametric, 2 groups), or epsilon-squared (nonparametric, 3+
groups). To pick a specific measure explicitly, pass its character
name:

```{r effect-size-explicit}
table_continuous(
  sochealth,
  select = wellbeing_score,
  by = sex,
  effect_size = "hedges_g",
  effect_size_ci = TRUE
)
```

Allowed values are `"none"` (default), `"auto"` (= legacy `TRUE`),
`"hedges_g"`, `"eta_sq"`, `"r_rb"`, and `"epsilon_sq"`. Incompatible
explicit choices (e.g. `"eta_sq"` with two groups, or `"hedges_g"`
with `test = "nonparametric"`) trigger an actionable error.

When you need the underlying columns for further processing, use
`output = "data.frame"`:

```{r raw-output}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score),
  by = education,
  statistic = TRUE,
  effect_size = "auto",
  output = "data.frame"
)
```

## Balance: the standardized mean difference

A baseline table comparing two arms is usually read for *balance*, not
for significance. `smd = TRUE` adds the standardized mean difference
of the Table 1 literature, and the idiomatic balance table drops the
*p* column at the same time:

```{r smd}
table_continuous(
  sochealth,
  select = c(age, bmi, wellbeing_score),
  by = sex,
  smd = TRUE,
  p_value = FALSE
)
```

The two arguments are independent — turning the SMD on turns nothing
else off, so you have to write both. The rule of thumb the table note
quotes is |SMD| > 0.1; spicy prints the number and never highlights
it.

The column is **signed**, group 1 minus group 2 in the order the table
displays the groups, so the direction of an imbalance is readable
without going back to the two means. It requires exactly two groups: a
`by` with three or more is refused rather than averaged over pairs,
because an average over pairs has no published reading and can sit
under 0.1 while one pair sits well over it.

### It is not the effect size

The SMD and Hedges' *g* look alike and are never the same number. On
`sochealth` the two agree to two decimals, which hides the point, so
here they are on a small unbalanced sample where they cannot hide:

```{r smd-vs-g}
small <- sochealth[c(1:6, 601:604), c("sex", "bmi")]
table_continuous(
  small,
  select = bmi,
  by = sex,
  smd = TRUE,
  effect_size = "hedges_g",
  effect_size_digits = 4
)
```

Both standardize a mean difference, but by different denominators.
Austin's SMD divides by the root **mean of the two group variances**;
*g* divides by the degrees-of-freedom pooled SD and then multiplies by
the small-sample correction *J*. Two consequences, both worth knowing:

- At **equal** group sizes the two denominators coincide, so the SMD
  *is* Cohen's *d*. At unequal sizes it is not.
- *g* is never the SMD, at any sample size. With equal groups the
  ratio *g* / SMD is exactly *J*: 0.80 at *n* = 3 per group, 0.96 at
  *n* = 10, approaching 1 only as the sample grows. Reading one for
  the other costs 20% on a small trial.

Read each for what it is: the SMD is a balance diagnostic, *g* is an
effect size. It also follows that the SMD carries **no confidence
interval and no *p*-value** — attaching one would put back the test
reasoning the balance literature asks the reader to drop.

## Selecting variables

`select` supports tidyselect helpers:

```{r tidyselect}
table_continuous(
  sochealth,
  select = starts_with("life_sat"),
  by = sex
)
```

For more programmatic selection, set `regex = TRUE`:

```{r regex}
table_continuous(
  sochealth,
  select = "^life_sat",
  regex = TRUE,
  by = education,
  output = "data.frame"
)
```

Use `exclude` when you want a broad selection with one or two explicit
removals:

```{r exclude}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health, life_sat_work),
  exclude = "life_sat_work",
  by = sex
)
```

## Handling missing values

Missing values can never enter a mean, so they are always excluded
from each variable's statistics. The table discloses the exclusion in
a note rather than staying silent:

```{r missing}
table_continuous(
  sochealth,
  select = c(bmi, life_sat_health)
)
```

Missing values in `by` are removed by default (`drop_na = TRUE`),
and the removal is again disclosed rather than silent -- here in the
note ("Rows with missing income_group removed") and in a warning:

```{r missing-by-default}
table_continuous(
  sochealth,
  select = bmi,
  by = income_group
)
```

Set `drop_na = FALSE` to display those rows as a dedicated
"(Missing)" group instead. The group-comparison test and effect size
still cover the observed groups only, matching `table_categorical()`:

```{r missing-by}
table_continuous(
  sochealth,
  select = bmi,
  by = income_group,
  drop_na = FALSE
)
```

## Weights

Survey data usually comes with case weights, and `sochealth` carries
one. Passing it weights every displayed statistic — mean, SD,
quantiles, extremes, and the CI of the mean — and the table says so
in its note:

```{r weights}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score),
  weights = weight,
  rescale = TRUE
)
```

Two conventions coexist for weighted statistics, and spicy makes the
choice explicit instead of silent. Without `rescale`, weights are
taken as *frequencies*: integer weights reproduce, exactly, the
statistics of the data with each row repeated `w` times — the reading
of SPSS's `WEIGHT BY` and Stata's `fweight`. With `rescale = TRUE`,
weights are first normalised to sum to the number of observations —
the *sampling-weights* reading, invariant to the scale of the
weights, whose SD matches Stata's `aweight` and `survey::svyvar()`.
For a survey weight like this one, `rescale = TRUE` is the reading
you want. The mean is the same under both (it is a ratio); the SD
and the `weighted_n` column differ. The exact formulas and the
cross-software correspondence are in the Weights section of
`?table_continuous`.

Two things are deliberately refused under weights rather than
silently wrong: the median confidence interval (an order-statistic
interval with no weighted version), and the group tests and effect
sizes — a *t*-test printed next to weighted descriptives would itself
be unweighted. For weighted group comparisons, `table_continuous_lm()`
takes the same `weights` argument and fits them properly:

```{r weights-by, eval = FALSE}
# Weighted descriptives by group: turn the test columns off
table_continuous(
  sochealth,
  select = bmi,
  by = sex,
  weights = weight,
  rescale = TRUE,
  p_value = FALSE
)
# Weighted group comparison: the model-based tool
table_continuous_lm(sochealth, select = bmi, by = sex, weights = weight)
```

The standardized mean difference is the exception, and passes under
weights:

```{r weights-smd}
table_continuous(
  sochealth,
  select = c(age, bmi),
  by = sex,
  weights = weight,
  rescale = TRUE,
  smd = TRUE,
  p_value = FALSE
)
```

It is a descriptive balance diagnostic with no *p* and no interval —
which is precisely why the balance literature substitutes it for the
test — and it is computed from the same weighted means and variances
the M and SD columns display, so it cannot be unweighted inference
sitting next to weighted descriptives. One consequence of that shared
producer: without `rescale`, the weighted SMD inherits the frequency
reading and is *not* invariant to the scale of the weights, exactly as
the SD column is not. With `rescale = TRUE` it is. For a survey weight,
use `rescale = TRUE`.

## Custom labels

By default, `table_continuous()` labels each variable with its label
attribute when one is present (e.g. data imported with `haven`), and
with the column name otherwise -- that is why the tables above read
"Body mass index" rather than `bmi`. Use the `labels` argument, a
named character vector keyed by column name, to override either. Only
the listed columns are relabelled; the others keep their attribute
label or column name:

```{r labels}
table_continuous(
  sochealth,
  select = c(wellbeing_score, life_sat_health),
  by = sex,
  labels = c(wellbeing_score = "Well-being score (0-100)")
)
```

## Display options

The printed ASCII table and every rendered output share the same
formatting vocabulary as `table_continuous_lm()`; `align`, `p_digits`,
and `decimal_mark` are also shared with `table_categorical()`:

- `align = "decimal"` (default) aligns numeric columns on the
  decimal mark, matching SPSS / SAS / LaTeX `siunitx` conventions.
  `"center"` and `"right"` are the alternatives.
- `p_digits = 3` (default, the APA standard) drives both the
  displayed precision of the `p` column and the small-*p* threshold:
  with `p_digits = 4`, the chunk below shows `.0176` for BMI and
  `<.0001` for the well-being score.
- `show_n = FALSE` and `ci = FALSE` drop the corresponding columns
  (and the CI spanner / borders) structurally from every output;
  the underlying `n` and `ci_lower` / `ci_upper` are always present
  in `output = "data.frame"` / `"long"` for downstream code.

```{r display-opts}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score),
  by = sex,
  ci = FALSE,
  show_n = FALSE,
  p_digits = 4
)
```

## Tidying for downstream pipelines

`table_continuous()` returns an object that can be coerced to a plain
`data.frame` / `tbl_df` (stripping the spicy formatting attributes)
or piped into `broom::tidy()` / `broom::glance()` for any downstream
tidyverse stats workflow:

```{r tidy-glance}
out <- table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score),
  by = sex
)

# Long descriptive rows: one per (variable x group) with broom-style
# columns (outcome, label, group, estimate = mean, std.error,
# conf.low / conf.high, n, min, max, sd).
broom::tidy(out)

# One row per outcome with the omnibus test + effect-size summary
# (test_type, statistic, df, df.residual, p.value, es_type, es_value,
# es_ci_lower / es_ci_upper, n_total).
broom::glance(out)

# Or just unbox to a plain data.frame (long-format underlying data)
head(as.data.frame(out))
```

## Output formats

`table_continuous()` supports the same reporting-oriented outputs as
`table_categorical()`:

| `output` value | Returned object |
|:--|:--|
| `"default"` | Styled ASCII console table |
| `"data.frame"` / `"long"` | Plain `data.frame` with the underlying long-format rows (synonyms; pick whichever reads better in your code) |
| `"tinytable"` | Formatted tinytable |
| `"gt"` | Formatted gt table |
| `"flextable"` | Formatted flextable |
| `"excel"` | Written `.xlsx` file |
| `"clipboard"` | Copied text table |
| `"word"` | Written `.docx` file |

`output = "gt"` produces a formatted gt table with APA-style borders
and column spanners:

```{r gt, eval = FALSE}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  output = "gt"
)
```

```{r gt-render, echo = FALSE, eval = build_rich_tables}
pkgdown_dark_gt(
  table_continuous(
    sochealth,
    select = c(bmi, wellbeing_score, life_sat_health),
    by = education,
    output = "gt"
  )
)
```

`output = "tinytable"` works well in Quarto and R Markdown documents:

```{r tinytable, eval = build_rich_tables}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  output = "tinytable"
)
```

## Export to Excel or Word

Use the same function to export the table directly:

```{r export, eval = FALSE}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  output = "excel",
  excel_path = "table_continuous.xlsx"
)

table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  output = "word",
  word_path = "table_continuous.docx"
)
```

## See also

- See `vignette("table-categorical", package = "spicy")` for grouped
  tables of categorical variables.
- See `vignette("table-continuous-lm", package = "spicy")` for
  model-based continuous summary tables with robust standard errors or
  case weights.
- See `vignette("table-regression", package = "spicy")` for full
  coefficient tables from one or several fitted `lm` / `glm` models
  (APA Table 3).
- See `vignette("summary-tables-reporting", package = "spicy")` for a
  cross-function reporting workflow that ties the four summary-table
  helpers together along the APA Table 1 / 2 / 3 sequence.
