---
title: "Regression coefficient tables in R"
description: >
  Build publication-ready regression coefficient tables in R with
  table_regression(): classical / HC / cluster-robust / bootstrap /
  jackknife variance, AME with Satterthwaite-corrected df under CR*
  (a first in R), four standardisation methods, per-coefficient
  effect sizes (partial f² / η² / ω² with noncentral CIs),
  multiple-comparison adjustment via stats::p.adjust() with full-
  family scope, regex-based keep / drop filters for focal-predictor
  display, side-by-side and hierarchical regression layouts, APA-
  aligned formatting, and output to console, gt, tinytable,
  flextable, Excel, Word, or clipboard.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression coefficient tables in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)

# Complete-case subset for the cluster-robust examples — the cluster
# vector must align with the rows actually used by `lm()`, so we
# drop the 25 rows with at least one NA in the predictor set up front.
sochealth_cc <- na.omit(
  sochealth[, c("wellbeing_score", "bmi", "age", "sex",
                "smoking", "education", "region")]
)
```

`table_regression()` produces a publication-ready coefficient summary
for one or several `lm()` fits, side-by-side, with the inferential
machinery and the formatting conventions normally found in Stata's
`regress` / `eststo` workflow or SPSS's `REGRESSION` output. The
function is fit-first (you pass already-fitted models, not raw data
+ formula), the long-format internal representation is built around
the `(model_id, term, estimate_type)` triplet for clean
[`broom::tidy()`][broom::tidy] / [`broom::glance()`][broom::glance]
exports, and the rendered table accepts heteroskedasticity-consistent,
cluster-robust, bootstrap, or jackknife variance, four standardisation
methods, per-coefficient effect sizes with noncentral CIs, and a
hierarchical-comparison footer.

Two design choices position `table_regression()` against the existing
R landscape (`modelsummary`, `gtsummary`, `parameters`,
`marginaleffects`):

* **AME with Satterthwaite-corrected degrees of freedom under CR\* variance.**
  When the user requests both Average Marginal Effects and a
  cluster-robust variance estimator, `table_regression()` builds the
  closed-form linear contrast representing each AME and passes it to
  [`clubSandwich::linear_contrast()`][clubSandwich::linear_contrast]
  with `test = "Satterthwaite"`. Existing R tools default to a
  z-asymptotic AME under CR\*, which is anti-conservative for few
  clusters (Pustejovsky and Tipton 2018). The B coefficient and the
  AME therefore share the same inferential regime in the same table.
* **Transparent caveat on standardised coefficients with non-additive terms.**
  When `standardized != "none"` and the model contains an interaction
  or a transformed term (`I()`, `poly()`, `log()`, `splines::ns()`),
  the function emits a classed `spicy_caveat` warning at runtime AND
  auto-documents the caveat in the table footer, with method-specific
  wording. Existing tools either compute silently or document only in
  `?` — `table_regression()` is "more pro via transparency, not
  rejection" (Cohen, Cohen, West, and Aiken 2003 §7.7).

Supports `lm` and `glm` (binomial / poisson / Gamma /
inverse.gaussian / quasi families with any link). Mixed-effects
models (`merMod`, `lmerModLmerTest`) are on the roadmap for
spicy 0.16+. See the *Generalised linear models* section below
for the glm-specific argument semantics.

## Basic usage

Pass a fitted `lm()` object. The default rendering returns a
single-model table with `B`, `SE`, `95% CI`, and `p`, plus a fit-stats
footer (`n`, `R²`, `Adj.R²`):

```{r basic}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit)
```

Reference levels of factor predictors carry the `(ref.)` annotation
and an em-dash in the statistic columns. The vcov footer line names
the variance estimator in plain English.

## Heteroskedasticity-consistent variance

Set `vcov = "HC*"` for sandwich-style standard errors via
[`sandwich::vcovHC()`][sandwich::vcovHC]. The valid types are `HC0`
through `HC5`; `HC3` is the small-sample-friendly default
(Long and Ervin 2000):

```{r hc}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, vcov = "HC3")
```

The footer reads "*Std. errors: heteroskedasticity-robust (HC3)*";
the column header for the confidence interval automatically tracks
`ci_level`.

## Cluster-robust variance

For clustered observations, `vcov = "CR*"` requests
cluster-robust variance via
[`clubSandwich::vcovCR()`][clubSandwich::vcovCR], with the cluster
identifier supplied through `cluster`. The orchestrator detects the
column name and prints it in the footer:

```{r cr}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth_cc)
table_regression(
  fit,
  vcov = "CR2",
  cluster = sochealth_cc$region
)
```

`CR2` (the Bell-McCaffrey adjustment) is the recommended default
under few clusters (Pustejovsky and Tipton 2018; Imbens and Kolesár
2016). Coefficient inference uses
[`clubSandwich::coef_test()`][clubSandwich::coef_test] with
Satterthwaite-corrected degrees of freedom — visible in the footer
when AME is also requested (next section).

## Average Marginal Effects with Satterthwaite df

Add `"AME"` to `show_columns` to display the average marginal effect of
each predictor with its compact `value [CI]` cell. When the variance
estimator is cluster-robust, `table_regression()` constructs the linear
contrast representing each AME and inverts it through
`clubSandwich::linear_contrast()` with the Satterthwaite framework, so
B and AME share the same t-distribution with the same df:

```{r ame-cr}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth_cc)
table_regression(
  fit,
  vcov = "CR2",
  cluster = sochealth_cc$region,
  show_columns = c("B", "SE", "CI", "AME", "p")
)
```

The footer now reads "*AME inference: t-distribution with
Satterthwaite-corrected df (Pustejovsky and Tipton 2018) via
`clubSandwich::linear_contrast()`*". For non-CR variance estimators
the AME column delegates to
[`marginaleffects::avg_slopes()`][marginaleffects::avg_slopes] with a
df argument coherent with B (residual df for classical and HC,
asymptotic z for bootstrap and jackknife).

## Standardised coefficients

`standardized` controls the standardisation method. Four are available;
the choice is consequential and the differences are documented in the
literature (Cohen et al. 2003 §3.4; Gelman 2008):

* `"refit"` — refit on z-scored outcome and predictors. The gold
  standard; produces β identical to `lm.beta` and to SPSS / Stata
  `Beta`.
* `"posthoc"` — post-hoc rescaling `β = B × SD(X) / SD(Y)`. Matches
  `effectsize::standardize_parameters(method = "basic")` and
  `parameters::model_parameters(standardize = "basic")`.
* `"basic"` — like posthoc but factor dummies stay on the 0/1 scale
  rather than being z-scored.
* `"smart"` — Gelman's (2008) recommendation: numeric predictors
  divided by `2 × SD(X)`; binary predictors centred only.

When non-`"none"`, the `"beta"` token is auto-injected into
`show_columns` immediately after `"B"`:

```{r beta}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, standardized = "refit")
```

For models with interactions or transformed predictors, the function
emits a `spicy_caveat` warning AND prints a method-specific caveat in
the footer (Aiken and West 1991; Cohen et al. 2003 §7.7). This is the
"transparency over rejection" choice — the table is computed and
displayed, with the limitation made explicit at the point of use.

## Per-coefficient effect sizes

Three partial effect-size tokens are available — Cohen's f²
(`partial_f2`), Pearson's partial η² (`partial_eta2`), and the
Olejnik-Algina bias-corrected partial ω² (`partial_omega2`). All three
include a confidence interval derived from noncentral-F inversion
(Steiger 2004; Smithson 2003) and render as a single
`value [CI_low, CI_high]` cell:

```{r partial}
fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sochealth)
table_regression(
  fit,
  show_columns = c("B", "p", "partial_eta2", "partial_omega2")
)
```

The η² point estimate matches `effectsize::eta_squared(partial = TRUE)`
on a Type-II reference (`car::Anova`) to machine epsilon. The ω² point
estimate matches `effectsize::omega_squared(partial = TRUE)` to
machine epsilon, using the Olejnik and Algina (2003) formula
`((F-1) × df1) / (F × df1 + N - df1)`. CIs use the Steiger (2004)
inversion bounds, which always bracket the corresponding bias-corrected
point estimate.

For factor predictors with k levels, the partial F-test is the joint
(k-1) df Wald test, so the same effect-size value is broadcast across
all non-reference dummies (and reference rows show an em-dash).

## Multiple-comparison adjustment

`p_adjust` applies a family-wise or false-discovery-rate adjustment
to the p-values via [`stats::p.adjust()`][stats::p.adjust]. The
family is the model's full coefficient set (intercept and reference
rows excluded); `B` and `AME` are adjusted as independent families
within the same call. Available methods are the same as in
`stats::p.adjust()`: `"none"` (default), `"holm"`, `"hochberg"`,
`"hommel"`, `"bonferroni"`, `"BH"` / `"fdr"`, `"BY"`.

```{r p-adjust}
fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sochealth)
table_regression(fit, p_adjust = "bonferroni")
```

The footer documents the chosen method and the family size; the SE
column is unchanged. `p_adjust` is orthogonal to `vcov`: combining
a robust SE with a family-wise correction is fully supported, e.g.
`vcov = "HC3", p_adjust = "holm"`. The adjustment is applied
**before** `keep` / `drop` filtering, so the family stays the
model's full coefficient set and the displayed adjusted p-values
reflect the right denominator regardless of which subset is shown
(matching the `modelsummary` and `parameters` convention).

A methodological note. Adjusting the p-values of every coefficient
of a single regression model is *not* the standard convention in
social-science or clinical reporting (Rothman 1990; Greenland 2017;
Harrell *Regression Modeling Strategies* §5.4; Gelman, Hill &
Yajima 2012; APA Manual 7 §6.46). Each coefficient tests a
scientifically distinct hypothesis on a distinct predictor, which
is not the situation that family-wise procedures were designed for.
The default `"none"` reflects this consensus. Adjustment is
nonetheless legitimate in three contexts: (i) mass screening with
many candidate predictors and no prior hypothesis (typically
`"BH"` / FDR), (ii) pre-registered multi-endpoint confirmatory
designs (typically `"holm"`), or (iii) when a reviewer or SAP
explicitly requests it. spicy exposes the argument under the same
"transparency over rejection" rule used for `standardized`: the
tool is here, the methodological choice is yours, and the footer
makes the choice visible to the reader.

## Filtering displayed coefficients

`keep` and `drop` accept regular expressions matched against
coefficient names (as returned by [stats::coef()]). They are
mutually exclusive — pick `keep` to whitelist focal predictors, or
`drop` to hide a few control variables. Multiple patterns combine
with logical OR.

```{r keep}
fit <- lm(wellbeing_score ~ age + sex + smoking + bmi + education,
          data = sochealth)
table_regression(fit, keep = c("^smoking", "^bmi$"))
```

```{r drop}
table_regression(fit, drop = "^education")
```

Together with `p_adjust`, this is the article-ready workflow:
adjust on the full model, display only the rows the reader cares
about — `table_regression(fit, p_adjust = "BH", keep = "^treatment")`.

## Multiple models side by side

Pass a list of `lm()` fits. The default column layout puts each model
in its own panel with its own `B / SE / CI / p`. When the dependent
variables differ across models, an "Outcome" row is auto-added at the
top of the body so each model is unambiguously labelled:

```{r multi}
m_wellbeing <- lm(wellbeing_score ~ age + sex + smoking,
                   data = sochealth)
m_bmi       <- lm(bmi             ~ age + sex + smoking,
                   data = sochealth)
table_regression(list(m_wellbeing, m_bmi))
```

When all models share the dependent variable, the outcome row is
suppressed (the DV appears in the title). Use a named list — e.g.
`list(Crude = m1, Adjusted = m2)` — to control the column-header
labels; pass `model_labels = c(...)` to override the names; pass
`outcome_labels = FALSE` to suppress the outcome row regardless.

## Hierarchical / nested regression

Set `nested = TRUE` to add a "Model comparison" footer block with
the change statistics for each adjacent pair (Cohen et al. 2003
§5.4). The default tokens for `lm` are `r2_change`, `F`, and `p` —
the APA-conventional layout for hierarchical regression:

```{r nested}
m1 <- lm(wellbeing_score ~ age + sex,                 data = sochealth_cc)
m2 <- lm(wellbeing_score ~ age + sex + smoking,       data = sochealth_cc)
m3 <- lm(wellbeing_score ~ age + sex + smoking + bmi, data = sochealth_cc)
table_regression(list(m1, m2, m3), nested = TRUE)
```

Validation is strict: identical `nobs` AND identical response variable
across all models, otherwise a `spicy_invalid_input` error explains
that R's listwise deletion may produce different `n` per model and
suggests refitting on the common subset (the reason for `sochealth_cc`
being prepared at the top of the vignette). Other comparison tokens
are available — `LRT` (likelihood-ratio χ²), `AIC` / `AICc` / `BIC`
(deltas), `f2_change`, `deviance_change` — selectable via
`nested_stats`.

## Generalised linear models (`glm`)

`table_regression()` accepts any `glm()` fit. Inference defaults to
the z-asymptotic Wald regime that matches `summary.glm`,
`parameters::model_parameters`, Stata's `logit, or` and SPSS
`LOGISTIC REGRESSION`. The title becomes family-aware ("Logistic
regression", "Poisson regression", "Probit regression", ...) and the
default footer block for `nobs / pseudo_r2_mcfadden /
pseudo_r2_nagelkerke / AIC` swaps in instead of `R² / Adj.R²`:

```{r glm-basic}
fit <- glm(am ~ mpg + wt, data = mtcars, family = binomial)
table_regression(fit)
```

### Response-scale display: `exponentiate = TRUE`

Set `exponentiate = TRUE` to switch the `B` column to the response
scale and rebrand its header per family / link: `OR` for
`binomial(logit)`, `IRR` for `poisson(log)`, `HR` for
`binomial(cloglog)`, `RR` for `binomial(log)`, `MR` for
`Gamma(log)`, and the generic `exp(B)` otherwise. The standard error
follows the delta-method approximation `SE_OR = OR × SE_log-odds`
(Stata `logit, or`; `parameters::model_parameters()`); the test
statistic and the p-value are invariant under monotone
transformation:

```{r glm-or}
table_regression(fit, exponentiate = TRUE)
```

### Term-level partial chi-square

`partial_chi2` is the glm analog of `partial_f2`: for each model
term, the partial likelihood-ratio chi-square via
`drop1(test = "LRT")` (SAS PROC LOGISTIC `TYPE3`; Long & Freese
2014 §3.5; Allison "TYPE3"). Rendered as `value (df)` so factor
terms (k−1 df) and numeric terms (1 df) read at a glance:

```{r glm-partial-chi2}
mt <- mtcars; mt$cyl <- factor(mt$cyl)
fit2 <- glm(am ~ mpg + cyl, data = mt, family = binomial)
table_regression(fit2, show_columns = c("B", "partial_chi2", "p"))
```

### Standardised coefficients: `refit` and the new `pseudo`

For `glm`, `standardized = "refit"` z-scores numeric *predictors*
only and refits the model (the response stays on its observed scale
since the link is fixed). This is Long & Freese 2014 §4.3.4
"x-standardization". The other algebraic methods (`"posthoc"`,
`"basic"`, `"smart"`) apply X-only scaling per the
`parameters` / `effectsize` convention.

`standardized = "pseudo"` (`glm` only) is the Menard (2004, 2011)
*fully* standardised coefficient, scaling by `SD(X) / SD(Y*)` where
`Y*` is the latent variable on the link scale and
`SD(Y*) = sqrt(var(linear-predictor) + var_link)` with `var_link` =
π²/3 for logit, 1 for probit, π²/6 for cloglog. Defined for binomial
families; non-binomial returns NA with a `spicy_caveat`:

```{r glm-pseudo, eval = FALSE}
table_regression(fit, standardized = "pseudo")
```

### Average Marginal Effects (AME)

The `AME` token returns response-scale marginal effects via
`marginaleffects::avg_slopes()`. For `glm`, AME is computed as
the average of `dE[Y|X]/dx` over the observed sample (so for
logistic regression the displayed AME is in probability units, not
log-odds). Under cluster-robust variance (`vcov = "CR2"` etc.), the
inference uses Satterthwaite df from
`clubSandwich::coef_test()` on the dominant underlying coefficient
— Pustejovsky & Tipton (2018) §4 approximation for nonlinear
contrasts:

```{r glm-ame, eval = FALSE}
table_regression(fit, show_columns = c("B", "AME", "p"))
```

### Profile-likelihood CIs: `ci_method = "profile"`

The default is `ci_method = "wald"` (symmetric `estimate ± z × SE`,
matching `summary.glm`). For `glm` you can also request
`ci_method = "profile"`, which calls `MASS::confint.glm()` to
compute the profile-likelihood CI — asymmetric, exact for
likelihood-based inference, and the gold standard under sparse data
or near-boundary estimates (Venables & Ripley *MASS* §7.2). The
estimate, SE, statistic, and p-value all remain Wald;
`"profile"` only refines the CI bounds:

```{r glm-profile, eval = FALSE}
table_regression(fit, ci_method = "profile")
```

### Hierarchical glm (LRT)

`nested = TRUE` on a list of nested glm fits adds a "Model
comparison" block whose default tokens are `c("LRT", "p")` — the
APA-conventional layout for hierarchical logistic regression
(Hosmer & Lemeshow §3.5; Long & Freese 2014 §3.6). The chi-square
statistic comes from `anova(test = "LRT")`. `r2_change` and `F` are
not defined for glm and return em-dashes if explicitly requested:

```{r glm-nested, eval = FALSE}
m1 <- glm(am ~ mpg,                 data = mtcars, family = binomial)
m2 <- glm(am ~ mpg + wt,            data = mtcars, family = binomial)
m3 <- glm(am ~ mpg + wt + factor(cyl), data = mtcars, family = binomial)
table_regression(list(m1, m2, m3), nested = TRUE)
```

### Gaussian glm caveat

A `glm()` with `family = gaussian` and `link = "identity"` is
mathematically equivalent to `lm()` but lacks the variance-explained
effect-size family (`partial_f2 / η² / ω²`) and the Satterthwaite-
corrected AME path. Following the *transparency over rejection*
rule, spicy accepts the fit and emits a `spicy_caveat` suggesting a
refit with `lm()`.

## Significance stars

Stars are off by default. APA 7 §6.46 explicitly discourages them, and
the modern R consensus (`modelsummary`, `gtsummary`, `parameters`,
`fixest`) defaults likewise OFF. Set `stars = TRUE` for the APA
preset (`*** p < .001, ** p < .01, * p < .05`) or pass a named numeric
vector for custom thresholds:

```{r stars}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, stars = TRUE)
```

Stars suffix the B column (or β when standardisation is requested);
the threshold mapping is auto-documented in the footer. The `p`
column itself remains unstarred so the numeric value stays readable.

## Display knobs

`labels` accepts a named character vector keyed by either formula term
labels (`"sex"`) or coefficient names (`"sexMale"`), so individual
contrast rows can be relabelled. `intercept_position` switches the
intercept to the bottom (Stata convention). `reference_style =
"annotation"` lifts the reference level into the factor header
(`"sex: [ref: Female]"`) and drops the explicit reference row, for
compact tables. `decimal_mark = ","` switches to European decimal
notation and automatically changes the CI separator to `";"` to avoid
ambiguity:

```{r polish}
fit <- lm(wellbeing_score ~ age + sex + education, data = sochealth)
table_regression(
  fit,
  labels = c(
    age            = "Age (years)",
    sex            = "Sex",
    education      = "Education"
  ),
  reference_style = "annotation",
  intercept_position = "last",
  decimal_mark = ","
)
```

## Output formats

The default print method draws an ASCII table to the console. The
same content is available as a plain `data.frame` (`output =
"data.frame"`), a long broom-style data.frame (`output = "long"`), or
as a rich-format table for `gt`, `tinytable`, `flextable`, `excel`,
`word`, or `clipboard`:

```{r data-frame}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "data.frame"
)
str(out)
```

```{r long}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "long"
)
out[, c("model_id", "term", "estimate_type", "estimate",
        "std.error", "p.value")]
```

```{r gt-output, eval = build_rich_tables}
pkgdown_dark_gt(
  table_regression(
    lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
    output = "gt"
  )
)
```

## broom integration

`broom::tidy()` returns a long tibble with one row per
`(model_id, term, estimate_type)` and broom-canonical column names
(`estimate`, `std.error`, `conf.low`, `conf.high`, `statistic`,
`p.value`). `broom::glance()` returns one row per
`(model_id, outcome)` with model-level statistics, with `df.residual`
kept numeric so cluster-robust Satterthwaite df is preserved verbatim:

```{r broom}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
out <- table_regression(fit, standardized = "refit")

broom::tidy(out)
broom::glance(out)
```

This makes `table_regression()` outputs ready for downstream pipelines
— `gtsummary`, `modelsummary`, `parameters` — without bespoke glue.

## See also

- `vignette("table-continuous-lm", package = "spicy")` for the
  one-predictor-by-many-outcomes counterpart (estimated marginal
  means, contrast or slope, four effect-size families).
- `vignette("summary-tables-reporting", package = "spicy")` for an
  end-to-end reporting workflow combining the spicy summary-table
  helpers.

## References

Aiken, L. S., and West, S. G. (1991).
*Multiple Regression: Testing and Interpreting Interactions*.
Sage.

American Psychological Association (2020).
*Publication Manual of the American Psychological Association*
(7th ed.). Section 6.46.

Cohen, J., Cohen, P., West, S. G., and Aiken, L. S. (2003).
*Applied Multiple Regression / Correlation Analysis for the
Behavioral Sciences* (3rd ed.). Lawrence Erlbaum.

Gelman, A. (2008). Scaling regression inputs by dividing by two
standard deviations. *Statistics in Medicine*, 27(15), 2865–2873.

Gelman, A., Hill, J., and Yajima, M. (2012). Why we (usually) don't
have to worry about multiple comparisons. *Journal of Research on
Educational Effectiveness*, 5(2), 189–211.

Greenland, S. (2017). For and against methodologies: Some
perspectives on recent causal and statistical inference debates.
*European Journal of Epidemiology*, 32(1), 3–20.

Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.).
Springer. Section 5.4 on multiplicity.

Hosmer, D. W., Lemeshow, S., and Sturdivant, R. X. (2013).
*Applied Logistic Regression* (3rd ed.). Wiley.

Imbens, G. W., and Kolesár, M. (2016). Robust standard errors in
small samples: Some practical advice. *Review of Economics and
Statistics*, 98(4), 701–712.

Long, J. S., and Ervin, L. H. (2000). Using heteroscedasticity
consistent standard errors in the linear regression model.
*The American Statistician*, 54(3), 217–224.

Long, J. S., and Freese, J. (2014). *Regression Models for
Categorical Dependent Variables Using Stata* (3rd ed.). Stata Press.

McFadden, D. (1974). Conditional logit analysis of qualitative
choice behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics*
(pp. 105–142). Academic Press.

Menard, S. (2004). Six approaches to calculating standardized
logistic regression coefficients. *The American Statistician*,
58(3), 218–223.

Menard, S. (2011). Standards for standardized logistic regression
coefficients. *Social Forces*, 89(4), 1409–1428.

Nagelkerke, N. J. D. (1991). A note on a general definition of the
coefficient of determination. *Biometrika*, 78(3), 691–692.

Olejnik, S., and Algina, J. (2003). Generalized eta and omega squared
statistics: Measures of effect size for some common research designs.
*Psychological Methods*, 8(4), 434–447.

Pustejovsky, J. E., and Tipton, E. (2018). Small-sample methods for
cluster-robust variance estimation and hypothesis testing in fixed
effects models. *Journal of Business & Economic Statistics*, 36(4),
672–683.

Rothman, K. J. (1990). No adjustments are needed for multiple
comparisons. *Epidemiology*, 1(1), 43–46.

Smithson, M. (2003). *Confidence Intervals*. Sage.

Steiger, J. H. (2004). Beyond the F test: Effect size confidence
intervals and tests of close fit in the analysis of variance and
contrast analysis. *Psychological Methods*, 9(2), 164–182.

Tjur, T. (2009). Coefficients of determination in logistic
regression models — A new proposal: The coefficient of
discrimination. *The American Statistician*, 63(4), 366–372.

Venables, W. N., and Ripley, B. D. (2002). *Modern Applied
Statistics with S* (4th ed.). Springer. Section 7.2 on profile
likelihood.

Wasserstein, R. L., Schirm, A. L., and Lazar, N. A. (2019). Moving
to a world beyond "p < 0.05". *The American Statistician*, 73(sup1),
1–19.
