---
title: "Categorical summary tables with table_categorical()"
description: >
  Build categorical summary tables in R with table_categorical(),
  including grouped cross-tabulations, effect sizes, confidence
  intervals, and export to gt, tinytable, flextable, Excel, or Word.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical summary tables with table_categorical()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spicy)
```

`table_categorical()` builds publication-ready categorical tables suitable for
APA-style reporting in social science and data science research. With
`by`, it produces grouped cross-tabulation tables with chi-squared
\(p\)-values, effect sizes, confidence intervals, and multi-level
headers. Without `by`, it produces one-way frequency-style tables for
the selected variables. Export to gt, tinytable, flextable, Excel, or
Word. This vignette walks through the main features.

## Basic usage

For grouped tables, provide a data frame, one or more selected
variables, and a grouping variable:

```{r basic}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education
)
```

The default output is `"wide"` with `style = "raw"`, which returns a
data frame with numeric values suitable for further processing.

## One-way tables

Omit `by` to build a frequency-style table for the selected variables:

```{r oneway}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  output = "default"
)
```

## Output formats

`table_categorical()` supports several output formats. The table below
summarizes the options:

| Format | Description |
|---|---|
| `"wide"` | Data frame, one row per modality (default) |
| `"default"` | Styled ASCII table in the console |
| `"long"` | Data frame, one row per modality x group |
| `"gt"` | Formatted gt table |
| `"tinytable"` | Formatted tinytable |
| `"flextable"` | Formatted flextable |
| `"excel"` | Excel file (requires `excel_path`) |
| `"clipboard"` | Copy to clipboard |
| `"word"` | Word document (requires `word_path`) |

### gt output

The `"gt"` format produces a table with APA-style borders, column
spanners, and proper alignment:

```{r gt}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  output = "gt"
)
```

### tinytable output

```{r tinytable}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = sex,
  output = "tinytable"
)
```

### Data frame output

Use `style = "report"` with `"wide"` or `"long"` to get formatted
character columns (ready for display), or `style = "raw"` for numeric
values (ready for analysis):

```{r wide-report}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  output = "wide",
  style = "report"
)
```

Use `output = "default"` for a styled ASCII table in the console. Set
`styled = FALSE` with `output = "default"` to recover the underlying
wide raw `data.frame`.

## Custom labels

By default, `table_categorical()` uses variable names as row headers. Use the
`labels` argument to provide human-readable labels:

```{r labels}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  labels = c("Smoking status", "Regular physical activity"),
  output = "gt"
)
```

## Association measures and confidence intervals

By default, `table_categorical()` reports Cramer's V for nominal variables and
automatically switches to Kendall's Tau-b when both variables are
ordered factors. Override with `assoc_measure`:

```{r assoc-measure}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  assoc_measure = "phi",
  output = "tinytable"
)
```

Add confidence intervals with `assoc_ci = TRUE`. In rendered formats
(gt, tinytable, flextable), the CI is shown inline:

```{r ci-rendered}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  assoc_ci = TRUE,
  output = "gt"
)
```

In data formats (wide, long, excel, clipboard), separate `CI lower` and
`CI upper` columns are added:

```{r ci-data}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  assoc_ci = TRUE,
  output = "wide",
  style = "report"
)
```

## Weighted tables

Pass survey weights with the `weights` argument. Use `rescale = TRUE` so
the total weighted N matches the unweighted N:

```{r weighted}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  weights = "weight",
  rescale = TRUE,
  output = "gt"
)
```

## Handling missing values

By default, rows with missing values are dropped (`drop_na = TRUE`).
Set `drop_na = FALSE` to display them as a "(Missing)" category:

```{r missing}
table_categorical(
  sochealth,
  select = income_group,
  by = education,
  drop_na = FALSE,
  output = "gt"
)
```

## Filtering and reordering levels

Use `levels_keep` to display only specific modalities. The order you
specify controls the display order, which is useful for placing
"(Missing)" last:

```{r levels-keep}
table_categorical(
  sochealth,
  select = income_group,
  by = education,
  drop_na = FALSE,
  levels_keep = c("Low", "High", "(Missing)"),
  output = "gt"
)
```

## Formatting options

Control the number of digits for percentages, p-values, and the
association measure:

```{r formatting}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  percent_digits = 2,
  p_digits = 4,
  v_digits = 3,
  output = "gt"
)
```

## Exporting to Excel, Word, or clipboard

For Excel export, provide a file path:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  output = "excel",
  excel_path = "my_table.xlsx"
)
```

For Word, use `output = "word"`:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  output = "word",
  word_path = "my_table.docx"
)
```

You can also copy directly to the clipboard for pasting into a
spreadsheet or a text editor:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  output = "clipboard"
)
```
