---
title: "Bayesian regression tables"
description: >
  Publication-ready tables for Bayesian regression in R with
  `table_regression()`: posterior medians and credible intervals, the
  deliberate absence of p-values, convergence checks before reading,
  prior disclosure, random effects from the posterior draws, and the
  `rstanarm` and `brms` engines.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Bayesian regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_rstanarm <- requireNamespace("rstanarm", quietly = TRUE) &&
  requireNamespace("posterior", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_rstanarm
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
# The brms chunk compiles C++ (minutes, and a working toolchain) --
# too fragile for CI site builds. Opt in locally with
# Sys.setenv(SPICY_BUILD_BRMS = "true") before knitting.
run_brms <- identical(Sys.getenv("SPICY_BUILD_BRMS"), "true") &&
  requireNamespace("brms", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(rstanarm)
```

This vignette covers **Bayesian regression tables** — the same
`table_regression()` call on a posterior instead of a maximum-likelihood
fit. The companion vignette [*Publication-ready regression
tables*](table-regression.html) covers the shared mechanics; here we
focus on what genuinely changes when the estimates are draws from a
posterior distribution: what the columns mean, which familiar columns
are *deliberately absent*, and what to check before reading anything.

`table_regression()` supports two engines:

* **`rstanarm`** (`stan_glm()`, `stan_lmer()`, `stan_glmer()`) — ships
  precompiled models, so fits run in seconds and every chunk in this
  vignette executes live;
* **`brms`** (`brm()`) — a wider model space, at the cost of compiling
  each model; shown at the end (the code is identical in spirit).

The running example is the logistic regression of `smoking` from the
main vignette's glm section — same data (`sochealth`), same predictors
— so the Bayesian table can be read against its frequentist twin.

```{r fit}
sh <- sochealth
sh$education <- factor(sh$education, ordered = FALSE)
fit <- stan_glm(smoking ~ sex + age + education, data = sh,
                family = binomial(),
                iter = 1000, chains = 2, seed = 123, refresh = 0)
```

The sampler settings are trimmed so this vignette builds in seconds —
for real work, keep rstanarm's defaults (four chains of 2,000
iterations): R-hat compares chains, so it needs several to compare,
and credible bounds are tail quantiles, estimated exactly where draws
are scarcest (Vehtari et al. 2021).

## What changes in a Bayesian table

```{r basic}
table_regression(fit)
```

Reading the columns against their frequentist counterparts:

* **B is the posterior median** — the 50% quantile of each
  coefficient's draws, robust to skewed posteriors (the `rstanarm`
  convention). It is not a maximum-likelihood point estimate, although
  under the default weakly informative priors and this sample size the
  two are nearly identical.
* **SE is the posterior MAD SD** — the median absolute deviation of
  the draws, rescaled so it equals the standard deviation for a
  normal posterior — and the footer says so (`Std. errors: posterior
  MAD SD (scaled median absolute deviation).`) rather than borrowing
  the frequentist term. Median + MAD SD is the robust pairing of
  *Regression and Other Stories* (§5.3) and of `rstanarm`'s own
  print: both members resist skewed posteriors, where a mean + SD
  pair can be dominated by one heavy tail. As Gelman, Hill &amp;
  Vehtari put it, "for convenience we sometimes call the posterior
  mad sd the 'standard error'".
* **The interval is a credible interval**, and the header says so:
  `95% CrI`, the equal-tailed 2.5% and 97.5% posterior quantiles. Its reading is the direct
  probability statement that confidence intervals are so often
  misread as making: *given model, priors, and data, the coefficient
  lies in [-1.31, -0.56] with 95% probability* (the `Tertiary` row).
  No appeal to repeated sampling is required. The equal-tailed
  interval is the default because it is transformation-invariant
  (which is what makes the exact odds-ratio bounds below possible)
  and matches `rstanarm`; the Kruschke tradition's **highest-density
  interval** is one argument away — `ci_method = "hdi"` computes the
  shortest interval containing 95% of the draws (Kruschke 2015) and
  relabels the header `95% HDI`. For these near-symmetric posteriors
  the two differ only in the second decimal; they part ways on skewed
  posteriors, where the HDI shifts toward the mode.
* **There is no p column, by design.** A posterior has no p-value,
  and the Bayesian Analysis Reporting Guidelines (Kruschke 2021) ask
  that any decision rule be stated explicitly and kept separate from
  the posterior summary itself — so the default table reports the
  posterior and leaves decisions to the reader (requesting `"p"`
  explicitly is refused; in a mixed frequentist–Bayesian table the
  shared p column simply dashes the Bayesian rows). The closest posterior summary —
  the **probability of direction** (`pd`, the share of draws on the
  dominant side of zero; Makowski et al. 2019) — is opt-in:

```{r pd}
table_regression(fit, show_columns = c("b", "ci", "pd"))
```

Read against the table: `Male`, at 0.62, is barely better than a coin
flip on sign; `age`, at 0.89, leans positive without settling the
direction; the education contrasts display 1.00 — essentially every
retained draw is negative (all 1,000 for `Tertiary`; 999 of 1,000 for
`Upper secondary`, which rounds up). A `pd` of 0.50 would say the
draws split evenly, with zero at the posterior median. It reads as
evidence of *direction*, never of practical size — and one disclosure
keeps this section honest: under weak priors `pd` maps almost
one-to-one onto the frequentist one-sided p-value (two-sided
p ≈ 2(1 − pd), so `age`'s 0.89 corresponds to p ≈ 0.23; Makowski et
al. 2019). `pd` is not a route around the p-value's pathologies; it
is the same directional information stated as a posterior
probability.

For a claim of practical size, the Kruschke tradition asks how much
of the posterior falls inside a **region of practical equivalence**
(ROPE) around zero. spicy deliberately renders no ROPE column: the
reporting guidelines require the equivalence limits to be *justified*
(BARG step 4.C), and limits are substantive, per-coefficient claims —
±0.05 log-odds may be negligible for one predictor and consequential
for another — that a package cannot default on the analyst's behalf.
When you can defend limits, compute the shares outside the table with
`bayestestR::rope(fit, range = c(-0.18, 0.18))` (or
`bayestestR::describe_posterior()`), state the limits and their
justification in the text, and keep the table's posterior summaries
as they are.

## Check convergence before reading anything

A frequentist table is wrong only if the model is wrong; a Bayesian
table can additionally be wrong because the *sampler* has not
converged — the draws then misrepresent the very posterior the columns
summarize. spicy therefore checks every Bayesian fit before rendering:
when any sampled parameter misses its target — R-hat at or above 1.01,
effective sample size below 100 per chain (never less than 400, so
fewer chains do not weaken the bar), any divergent transition,
E-BFMI below 0.2 (Vehtari et al. 2021) — the table gains a
`Sampler diagnostics:` footer line naming the failure and a warning
fires (classed `spicy_bayes_diagnostics`, so scripts can catch or
mute this guard selectively with `withCallingHandlers()`). A clean
fit prints nothing: the publication table stays lean, and silence
means the checks passed. The guard reads *all* sampled parameters —
including the group-level ones a multilevel table summarizes — not
just the displayed coefficients.

For parameter-level triage the same three diagnostics are available
as opt-in columns in all-Bayesian tables, together with a fourth,
`mcse` — the Monte Carlo standard error of the displayed posterior
median:

```{r diag-cols}
table_regression(fit, show_columns = c("b", "ci", "rhat", "mcse"))
```

The `mcse` column answers a question the other diagnostics do not:
*how many of the printed digits are real?* MCMC output is stochastic
— rerun the sampler with another seed and the third significant
digit of most estimates will move. The working rule (Gelman, Vehtari,
McElreath et al. 2026, §11.6): a displayed digit is Monte-Carlo
stable when **twice the MCSE stays below one unit of that digit**.
Here twice the MCSE of the coefficients is on the order of their
second decimal: that digit sits near the stability edge — honest to
report, and a reminder that a two-decimal table is close to the
resolution limit of this deliberately short chain. Most of
the time two *significant* digits are all a posterior summary can
support, halving the MCSE costs four times the iterations, and tail
quantities (the CrI bounds) are noisier than the median. When the
displayed digits matter — a boundary against a clinical threshold,
say — check `mcse` before trusting the last one.

Under the hood these are three numbers per parameter (Vehtari et al.
2021): R-hat, which compares chains (at convergence, 1.00; worry
above roughly 1.01), and two effective sample sizes that discount
autocorrelation — bulk ESS, which governs point estimates such as the
posterior median, and tail ESS, which governs the credible-interval
bounds:

```{r convergence}
posterior::summarise_draws(
  posterior::subset_draws(posterior::as_draws_array(fit),
                          variable = names(fit$coefficients))
)[, c("variable", "rhat", "ess_bulk", "ess_tail")]
```

Read the output rather than glossing it. R-hat sits at 1.00 for every
slope but prints 1.01 for the intercept — exactly the alert level,
and exactly what a deliberately short two-chain run invites; a
manuscript fit would simply use the defaults. Bulk effective sizes
exceed the 1,000 retained draws — legitimate, since Hamiltonian Monte
Carlo can produce anticorrelated draws that beat independence — while
the tail effective sizes, near 700–800, are what guard the 2.5% and
97.5% bounds the CI column reports: comfortable for two-decimal
reporting. When these diagnostics genuinely fail, the remedy is more
iterations or a reparameterized model — never a more optimistic
reading of the table.

## Priors are part of the model — disclose them

The posterior blends data with priors, so a reported Bayesian model is
incompletely specified until the priors are stated. `rstanarm`'s
defaults are weakly informative (centred at zero, scaled to the data),
which is why the table above sits so close to its frequentist twin —
but that is a property to *verify and report*, not assume:

```{r priors}
prior_summary(fit)
```

The *Adjusted prior* line is the disclosure that matters: `rstanarm`
rescales the default `normal(0, 2.5)` by each predictor's spread
(scale 2.5/sd(x)), so the priors actually applied here are
normal(0, 5.00) on the sex contrast and normal(0, 0.17) on age —
equivalent to placing normal(0, 2.5) on the coefficients of
*standardized* predictors. For a manuscript, one sentence suffices
("rstanarm defaults: independent normal(0, 2.5) priors, autoscaled by
each predictor's SD — the *Adjusted prior* scales above"), and a
sensitivity re-fit with wider or substantive priors belongs in the
appendix when the priors could plausibly matter.

## Odds ratios: `exponentiate = TRUE`

The ratio scale works exactly as for `glm`, with one pleasant
difference: every displayed quantity is computed from the
exponentiated draws themselves, so all of them are exact posterior
statements about the odds ratio — the credible bounds are the
exponentiated quantiles, and the SE is the posterior MAD SD of
`exp(draws)` rather than the delta-method approximation
(SE_OR = OR × SE_log-odds), which understates the ratio-scale spread
severalfold when the posterior is wide. Ratio-scale posteriors are
right-tail-heavy, so the interval remains the primary uncertainty
summary — never reconstruct one as B ± 2 SE:

```{r or}
table_regression(fit, exponentiate = TRUE)
```

Reading the `Tertiary` row with the factor-change template: holding
sex and age constant, the odds of smoking for a tertiary-educated
respondent are 0.40 times the odds for one with lower secondary
education — and given model, priors, and data, that odds ratio lies
between 0.27 and 0.57 with 95% probability.

## The frequentist twin, side by side

Because the layout machinery is shared, the comparison table is one
`list()` away — and with weak priors and n = 1,175 it mostly documents
agreement, which is itself worth showing to a skeptical reader:

```{r side-by-side}
gf <- glm(smoking ~ sex + age + education, data = sh,
          family = binomial)
table_regression(list("ML (glm)" = gf, "Bayes (stan_glm)" = fit),
                 show_columns = c("b", "ci"))
```

The intervals differ in *reading*, not much in width: the frequentist
column's CI is a coverage statement about the procedure, the Bayesian
column's CrI a probability statement about the parameter. The shared
header stays `95% CI` here — the honest common label — and the footer
discloses per model that the Bayesian column is an equal-tailed
posterior credible interval (the all-Bayesian table above relabels
the header itself, `95% CrI`). In the fit-statistics block, `n` and
`R² (Bayes)` fill per model, but the likelihood-based rows — the two
pseudo-R² and AIC — fill only the frequentist one, because a
posterior has none of them. Which brings us to what the table
refuses.

## What is refused, and why

Several familiar requests are meaningless for a posterior, and spicy
refuses them with an explanatory error rather than rendering something
silently wrong:

* `p_adjust` — there are no p-values to adjust;
* likelihood-based fit statistics (`"aic"`, `"bic"`, pseudo-R²) —
  the Bayesian table has its own instead: the default block reports
  `R² (Bayes)`, the posterior median of the draws-based R² (Gelman
  et al. 2019), and `show_fit_stats = c("nobs", "r2_bayes",
  "elpd_loo", "looic")` adds the PSIS-LOO expected log predictive
  density and its deviance-scale twin, with the elpd standard error
  disclosed in the footer (Vehtari, Gelman & Gabry 2017; a few
  seconds per model). Prefer the log scale: the −2 of LOOIC persists
  for historical reasons only (Gelman, Vehtari, McElreath et al.
  2026, App. B). A `"waic"` token exists too — its SE is
  disclosed the same way — though PSIS-LOO is generally preferred.
  These estimates come with their own reliability diagnostics, and
  spicy never mutes them: when PSIS-LOO flags observations with
  Pareto k above the sample-size-specific threshold — min(1 −
  1/log10(S), 0.7), the same bound `loo::loo()` itself prints
  (Vehtari et al. 2024) — or WAIC flags p_waic above 0.4, the footer
  says the estimate is unreliable and a `spicy_bayes_diagnostics`
  warning fires. The remediation ladder (Gelman, Vehtari, McElreath
  et al. 2026, ch. 24): `loo::loo_moment_match()` first, then
  refitting the flagged folds (`k_threshold = 0.7`, brms
  `reloo = TRUE`), then K-fold cross-validation. Model *comparison* belongs to
  `loo::loo_compare()` outside the table: the comparison uncertainty
  is the standard error of the elpd *difference* (not the per-model
  SEs shown here), and differences smaller than about 4 carry no
  practical meaning (Gelman, Vehtari, McElreath et al. 2026, ch. 9).
  The canonical fit *check* is the graphical posterior predictive
  check (`pp_check(fit)`; Gelman et al. 2013, ch. 6) — Bayes factors
  stay out by design (Gelman et al. 2013, §7.4);
* robust / cluster-robust `vcov` — a sandwich estimator corrects a
  misspecified likelihood's standard errors; nothing standard plays
  that role for a posterior (misspecification-robust
  Bayesian procedures exist but remain research-grade), so model the
  clustering instead (next section);
* `ci_method = "profile"` / `"boot_percentile"` and the AME family —
  profile likelihood and bootstrap replicates are frequentist
  constructions with no posterior analogue (the credible interval is
  computed from the draws; `"hdi"` is the one alternative flavor),
  and a draws-based AME needs its own design (planned);
* variational and optimizing fits (`stan_glm(..., algorithm =
  "meanfield")`, `"optimizing"`) — they carry an *approximate*
  posterior on which the MCMC diagnostics above would be vacuous;
  refit with `algorithm = "sampling"`;
* a ROPE column — deliberately absent, as discussed in the `pd`
  section: equivalence limits are substantive claims the analyst
  must justify per coefficient (`bayestestR::rope()` renders them
  once justified).

```{r refused, error = TRUE}
table_regression(fit, p_adjust = "holm")
```

## Multilevel models: random effects from the draws

Clustered data is where the Bayesian machinery earns its keep — the
priors regularize the group-level variance exactly where a
likelihood-based fit struggles (few groups, boundary estimates). The
random effects render as the same block as for `lmer`, with every
number taken directly from the posterior draws: the estimate is the
posterior median of each SD, the interval its posterior quantiles —
and there is no boundary-corrected test because nothing is being
tested — the chi-bar-squared machinery exists only because the
frequentist null σ = 0 sits on the edge of the parameter space, a
situation a posterior quantile never encounters:

```{r multilevel}
mfit <- stan_lmer(Reaction ~ Days + (Days | Subject),
                  data = lme4::sleepstudy,
                  iter = 1000, chains = 2, seed = 7, refresh = 0)
table_regression(mfit)
```

Note the guard from the convergence section doing its job: this
deliberately trimmed run (two short chains, a correlated random
slope) misses the R-hat and ESS targets, so the table says so in its
footer and a `spicy_bayes_diagnostics` warning fires — a manuscript
fit with rstanarm's default sampler settings renders this same table
clean.

Compare this block with the `lmer` table of the
[mixed-effects vignette](table-regression-mixed.html): same layout,
but the credible interval on the `Days` slope SD — the
`σ Subject Days` row — is a genuine
posterior statement — no chi-bar-squared correction, no profile
refit — and the footer reads `Random effects (MCMC)`. The per-group
deviations themselves (one per subject) are deliberately *not*
listed as coefficient rows; they are shrinkage estimates, and their
summary is the block's SD.

## `brms`

`brms` reaches the same table through the same code path; the only
practical difference is that each model compiles before sampling
(minutes, and a C++ toolchain), which is why this chunk is shown
without being evaluated. Run it locally to see the table — it matches
the `rstanarm` one above row for row, as the package's shared frame
tests verify:

```{r brms, eval = run_brms}
bfit <- brms::brm(Reaction ~ Days + (Days | Subject),
                  data = lme4::sleepstudy,
                  chains = 2, iter = 1000, seed = 7, refresh = 0,
                  silent = 2)
table_regression(bfit)
```

Coefficients are posterior medians with equal-tailed credible
intervals, the random-effect block is built from the `sd_*` / `cor_*`
draws, and the refusals above apply identically. What `brms` buys is
model space: distributional regression, non-linear formulas, and
families neither `glm` nor `rstanarm` covers — all rendered through
the same table.

## Output formats

The default console table is one of several targets. The `output`
argument also produces a raw data frame, a long broom-style tibble,
and — with the corresponding Suggests package — rich `gt`,
`flextable`, `tinytable`, Excel, or Word tables. The Bayesian
structure carries through: the p column stays a dash, and the
posterior summaries keep full precision in the tidy frame. In
[`broom::tidy()`](https://broom.tidymodels.org) the `estimate` column
is the posterior median, `std.error` the posterior MAD SD, and
`conf.low` / `conf.high` the credible
bounds; `p.value`, `statistic`, and `df` are `NA` — pipelines that
filter on `p.value` must switch to the interval (or `pd`):

```{r broom}
td <- broom::tidy(table_regression(fit))
td[, c("term", "estimate", "std.error", "conf.low", "conf.high",
       "p.value")]
```

```{r outputs-gt, eval = build_rich_tables}
table_regression(fit, exponentiate = TRUE, output = "gt")
```

## See also

- `vignette("table-regression", package = "spicy")` for the shared
  mechanics (columns, layouts, filtering).
- `vignette("table-regression-mixed", package = "spicy")` for the
  frequentist multilevel counterpart of the random-effects block.

## References

* Gelman, A., Hill, J., &amp; Vehtari, A. (2020). *Regression and
  Other Stories*. Cambridge University Press.
* Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari,
  A., &amp; Rubin, D. B. (2013). *Bayesian Data Analysis* (3rd ed.).
  CRC Press.
* Gelman, A., Goodrich, B., Gabry, J., &amp; Vehtari, A. (2019).
  R-squared for Bayesian regression models. *The American
  Statistician*, 73(3), 307–309.
* Gelman, A., Vehtari, A., McElreath, R., et al. (2026). *Bayesian
  Workflow*. Chapman &amp; Hall / CRC Press.
* Kruschke, J. K. (2015). *Doing Bayesian Data Analysis* (2nd ed.).
  Academic Press.
* Kruschke, J. K. (2021). Bayesian analysis reporting guidelines.
  *Nature Human Behaviour*, 5(10), 1282–1291.
* Makowski, D., Ben-Shachar, M. S., Chen, S. H. A., &amp; Lüdecke, D.
  (2019). Indices of effect existence and significance in the Bayesian
  framework. *Frontiers in Psychology*, 10, 2767.
* Vehtari, A., Gelman, A., &amp; Gabry, J. (2017). Practical Bayesian
  model evaluation using leave-one-out cross-validation and WAIC.
  *Statistics and Computing*, 27(5), 1413–1432.
* Vehtari, A., Simpson, D., Gelman, A., Yao, Y., &amp; Gabry, J.
  (2024). Pareto smoothed importance sampling. *Journal of Machine
  Learning Research*, 25(72), 1–58.
* Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., &amp;
  Bürkner, P.-C. (2021). Rank-normalization, folding, and
  localization: An improved R-hat for assessing convergence of MCMC.
  *Bayesian Analysis*, 16(2), 667–718.
