---
title: "Survival regression tables"
description: >
  Publication-ready tables for survival regression in R with
  `table_regression()`: Cox hazard ratios with events and concordance,
  cluster-robust (Lin-Wei) variance for multi-centre data, hierarchical
  Cox comparisons by partial-likelihood ratio test, accelerated failure
  time models with time ratios, and the `survival`, `rms`, and
  `flexsurv` engines.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Survival regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_survival <- requireNamespace("survival", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_survival
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_rms      <- have_survival && requireNamespace("rms", quietly = TRUE)
have_flexsurv <- have_survival && requireNamespace("flexsurv", quietly = TRUE)
have_broom    <- have_survival && requireNamespace("broom", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(survival)
```

This vignette covers **survival (time-to-event) regression** — models
for how long until an event occurs, with censoring for the subjects
who never reach it. The companion vignette [*Publication-ready
regression tables*](table-regression.html) covers the shared mechanics
(`vcov`, `ci_level`, output formats, multi-model layouts, broom
integration); here we focus on what is specific to survival fits:
hazard ratios and time ratios, event counts and concordance, and the
variance estimators of multi-centre studies.

`table_regression()` supports four survival engines:

* **`survival::coxph()`** — the semiparametric Cox proportional
  hazards model, the field default;
* **`survival::survreg()`** — parametric accelerated failure time
  (AFT) models;
* **`rms::cph()`** — the Cox model in the `rms` ecosystem;
* **`flexsurv::flexsurvreg()`** — fully parametric models over a wide
  family of distributions.

The running example is `survival::lung`: survival of 228 patients with
advanced lung cancer, with age, sex, and the ECOG performance score
(0 = fully active, higher is worse) as predictors. One row lacks
`ph.ecog` and one lacks the enrolling institution, so we declare the
analytic sample explicitly — the same 226 rows serve every model and,
later, the cluster variable:

```{r data}
lung2 <- na.omit(lung[, c("time", "status", "age", "sex",
                          "ph.ecog", "inst")])
lung2$sex <- factor(lung2$sex, levels = 1:2,
                    labels = c("Male", "Female"))
nrow(lung2)
```

## The Cox model in one paragraph

The Cox model works on the **hazard** — the instantaneous event rate
among those still at risk. Each coefficient multiplies that hazard by
a constant factor at every point in time (the **proportional-hazards
assumption**), and the baseline hazard itself is left completely
unspecified: that is the semiparametric trick that made the model
universal (Cox 1972; Therneau & Grambsch 2000). Two consequences shape
the table. There is **no
intercept row** — the baseline hazard absorbs it — and there is no
residual variance, so the model-fit block reports what a survival
reader expects instead: the **number of events** and the
**concordance**.

## Basic table: hazard ratios

```{r cox}
cx <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung2)
table_regression(cx, exponentiate = TRUE)
```

Reading the table:

* `exponentiate = TRUE` turns each log-hazard coefficient into a
  **hazard ratio**. Women's hazard is 0.57 times men's — a 43% lower
  hazard at any given moment, all else equal; each ECOG point
  multiplies the hazard by 1.60. Leave `exponentiate` off for the
  log-hazard `B` column.
* Inference is **Wald-z** on the log scale, matching `summary.coxph()`
  and Stata `stcox`; the statistic and p-value are invariant under the
  exponentiation.
* The footer reports **events** (163 deaths among 226 patients — the
  precision and power of a survival model are governed chiefly by its
  event count, not its n) and the **concordance** C = 0.64: the probability that, of
  two comparable patients, the one predicted to fail earlier actually
  does. 0.5 is a coin flip; 0.64 is modest discrimination
  (Harrell 2015).

The table reports the model, not its assumption: test proportionality
with `survival::cox.zph()` before reporting (here the global test
gives p = .17 — no evidence against it; Therneau & Grambsch 2000).

## Why there is no AME column

Requesting an `"ame"` column on a Cox fit is refused with an
explanation rather than answered: an average marginal effect needs a
response scale, and a Cox model deliberately does not commit to one —
a per-coefficient "effect on survival" would be ambiguous (at what
time? on which scale?) with unreliable standard errors. The hazard
ratio *is* the per-coefficient summary of this model. Absolute-scale
summaries (risk differences at a horizon, restricted mean survival
time) are a different estimand family, not a per-coefficient column.

## Multi-centre data: cluster-robust variance

The `lung` patients were enrolled by 18 institutions, and outcomes
within an institution may correlate. The `CR*` request routes to the
**Lin & Wei (1989)** robust variance — the grouped score (dfbeta)
sandwich that `coxph(..., cluster = )` and Stata `stcox, vce(cluster)`
use:

```{r cox-cluster}
table_regression(cx, vcov = "CR0", cluster = ~inst,
                 exponentiate = TRUE)
```

Two lessons here. First, the mechanics: the cluster variable must
provide **one value per row of the model data** — declaring the
analytic sample up front (as we did) keeps the cluster aligned when
predictors carry missing values. Second, the direction: robust
standard errors are not always larger — the sex SE *shrinks* here.
Clustering corrects the variance in whichever direction the
within-centre correlation points; treating it as a conservative
inflation ritual misreads it.

## Hierarchical Cox models

Does the performance score improve on age and sex? `nested = TRUE`
compares adjacent models by the **partial-likelihood ratio test** (the
`anova.coxph()` convention) and appends the change rows:

```{r cox-nested}
cx0 <- coxph(Surv(time, status) ~ age + sex, data = lung2)
table_regression(list(cx0, cx), nested = TRUE, exponentiate = TRUE,
                 show_columns = c("b", "p"))
```

Adding `ph.ecog` buys a chi-squared change of 16.78 on one degree of
freedom (p < .001), an AIC drop of 14.8, and lifts the concordance from
0.60 to 0.64 — three readings of the same improvement, on the
likelihood, information, and discrimination scales respectively.

## Accelerated failure time: `survreg()`

The parametric alternative models **time itself** (Wei 1992):
covariates stretch or compress survival time by a constant factor.
Under `exponentiate = TRUE` the coefficients become **time ratios**
(TR):

```{r survreg}
sr <- survreg(Surv(time, status) ~ age + sex + ph.ecog, data = lung2,
              dist = "weibull")
table_regression(sr, exponentiate = TRUE)
```

The intercept is back — a parametric model must anchor the time
scale, and exp(intercept) ≈ 801 days is that anchor (the extrapolated
characteristic time at the reference levels and age 0), not a ratio
despite the column header.

Set the TR next to the Cox HR and the numbers *look* contradictory:
women have TR 1.50 but HR 0.57. They are the same finding on opposite
scales — a protective effect **multiplies survival time up** (50%
longer) and **multiplies the hazard down** (43% lower). Neither
direction is more correct; the Weibull family is the one distribution
that is simultaneously AFT and proportional-hazards, and the two
scales are linked by an exact identity within the fit:
HR = TR^(−1/scale) = 1.50^(−1/0.73) ≈ 0.57, with the scale parameter
read from the footer — matching the semiparametric Cox estimate to
displayed precision (0.575 vs 0.573). The footer names the
distribution because the interpretation depends on it —
`exponentiate` produces time ratios only for log-scale distributions
(Weibull, exponential, lognormal, loglogistic); for an identity-scale
distribution (`dist = "gaussian"`) the request is skipped with a
warning, since the coefficients already act on the time scale.

## `rms::cph()`

The `rms` Cox engine renders through the same layout. One requirement
carried over from `rms` itself: cluster-robust variance uses
`rms::robcov()`, which needs the design matrices stored at fit time —
fit with `x = TRUE, y = TRUE` or the robust request errors with that
exact instruction:

```{r cph, eval = have_rms}
cph_fit <- rms::cph(Surv(time, status) ~ age + sex, data = lung2,
                    x = TRUE, y = TRUE)
table_regression(cph_fit, vcov = "CR0", cluster = ~inst,
                 exponentiate = TRUE)
```

## Fully parametric: `flexsurv`

`flexsurv::flexsurvreg()` (Jackson 2016) fits parametric survival
models over a wide family of distributions (and the Royston–Parmar
2002 spline models). The location
coefficients exponentiate to the generic `exp(B)` — the substantive
reading (time ratio, hazard ratio) depends on the distribution, which
the footer names together with its ancillary parameters:

```{r flexsurv, eval = have_flexsurv}
fs <- flexsurv::flexsurvreg(Surv(time, status) ~ age + sex,
                            data = lung2, dist = "weibull")
table_regression(fs, exponentiate = TRUE)
```

Two guardrails apply. A spline model on the `"normal"` scale places
its coefficients on a probit-like scale whose exponential has no
meaning, so `exponentiate = TRUE` is refused there; and covariates on
ancillary parameters (`anc =`) act on their own scales, so they refuse
exponentiation as a whole rather than print meaningless ratios.

Two Cox extensions flow through the same table without ceremony:
stratified models (`strata()` terms are absorbed into the baseline, no
coefficient row) and counting-process data
(`Surv(tstart, tstop, event)` for time-varying covariates) render like
any other fit.

## Output formats

Everything above used the default console output. The same table
renders as a raw data frame, a long broom-style tibble, or — with the
corresponding Suggests package — a rich `gt`, `flextable`,
`tinytable`, Excel, or Word table:

```{r outputs-df}
table_regression(cx, exponentiate = TRUE, output = "data.frame")
```

```{r outputs-gt, eval = build_rich_tables && have_survival}
table_regression(cx, exponentiate = TRUE, output = "gt")
```

[`broom::tidy()`](https://broom.tidymodels.org) returns the long
frame; with `exponentiate = TRUE` the estimates and CI bounds are on
the ratio scale, as displayed:

```{r broom, eval = have_broom}
broom::tidy(table_regression(cx, exponentiate = TRUE))
```

## References

* Cox, D. R. (1972). Regression models and life-tables. *Journal of
  the Royal Statistical Society, Series B*, 34(2), 187–220.
* Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.).
  Springer.
* Jackson, C. H. (2016). `flexsurv`: A platform for parametric
  survival modeling in R. *Journal of Statistical Software*, 70(8).
* Lin, D. Y., & Wei, L. J. (1989). The robust inference for the Cox
  proportional hazards model. *Journal of the American Statistical
  Association*, 84(408), 1074–1078.
* Royston, P., & Parmar, M. K. B. (2002). Flexible parametric
  proportional-hazards and proportional-odds models for censored
  survival data. *Statistics in Medicine*, 21(15), 2175–2197.
* Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data:
  Extending the Cox Model*. Springer.
* Wei, L. J. (1992). The accelerated failure time model: a useful
  alternative to the Cox regression model in survival analysis.
  *Statistics in Medicine*, 11(14–15), 1871–1879.
