---
title: "Mixed-effects regression tables"
description: >
  Publication-ready tables for mixed-effects (multilevel) regression in R
  with `table_regression()`: the textbook model-building sequence on the
  High School & Beyond data (ICC, contextual effects, random slopes,
  cross-level interactions), random effects as table rows, the
  boundary-corrected chi-bar-squared test, opt-in per-term LRT / RLRT,
  Nakagawa R-squared, Satterthwaite inference, three-level and crossed
  designs, and the `lme4`, `glmmTMB`, and `nlme` engines.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Mixed-effects regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
have_mixed <- requireNamespace("lme4", quietly = TRUE) &&
  requireNamespace("lmerTest", quietly = TRUE) &&
  requireNamespace("merDeriv", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = have_mixed
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")
have_glmmTMB <- have_mixed && requireNamespace("glmmTMB", quietly = TRUE)
have_nlme    <- have_mixed && requireNamespace("nlme", quietly = TRUE)
have_rlrsim  <- have_mixed && requireNamespace("RLRsim", quietly = TRUE)
have_clubsw  <- have_mixed && requireNamespace("clubSandwich", quietly = TRUE)
have_broom   <- have_mixed && requireNamespace("broom", quietly = TRUE)
have_margins <- have_mixed && requireNamespace("marginaleffects", quietly = TRUE)
source("_pkgdown-helpers.R")
```

```{r setup, message = FALSE}
library(spicy)
library(lmerTest)   # masks lme4::lmer() with the Satterthwaite-aware version
```

This vignette covers **mixed-effects (multilevel) regression** — models
for data with a grouping structure, such as repeated measurements
nested in subjects or respondents nested in regions. The companion
vignette [*Publication-ready regression
tables*](table-regression.html) covers the shared mechanics (`vcov`,
`ci_level`, output formats, multi-model layouts, broom integration);
here we focus on what is *specific* to mixed fits — above all, how the
**random effects** are reported.

`table_regression()` supports four mixed engines:

* **`lme4::lmer()`** — linear mixed models. Load **`lmerTest`** before
  fitting to get Satterthwaite degrees of freedom; plain `lme4` fits
  fall back to a large-sample Wald-z (the footer says which you got).
* **`lme4::glmer()`** — generalized linear mixed models (Wald-z).
* **`glmmTMB::glmmTMB()`** — a wider family space, plus zero-inflation
  and dispersion components.
* **`nlme::lme()`** — the classical engine (containment-df t-tests).

Two datasets carry the main narrative. The table anatomy is shown on
`lme4::sleepstudy` — reaction times of 18 subjects measured on 10
consecutive days of a sleep-restriction study, the canonical case for
a random intercept and slope. The model-*building* sequence — from a
naive OLS to cross-level interactions — then runs on the High School &
Beyond data of the multilevel textbooks (7,185 pupils in 160 schools;
the data ship with `nlme`, but the models are `lmer()` fits).

## The mixed model in one paragraph

A mixed model splits the coefficients in two. **Fixed effects** are the
population-level slopes you would report from any regression. **Random
effects** let selected coefficients vary across groups: `(Days |
Subject)` gives every subject their own intercept and their own `Days`
slope, drawn from a common distribution. What the model *estimates* for
the random part is not 18 pairs of coefficients but the **variance
components** of that distribution — how much subjects differ at
baseline, how much their slopes differ, and how the two go together. A
regression table therefore needs two kinds of rows, and that is exactly
what `table_regression()` prints.

## Basic table

```{r basic}
fit <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
table_regression(fit)
```

Reading the table, top to bottom:

* **Fixed effects** first, as in any regression table. With `lmerTest`
  loaded, inference is a **t-test with Satterthwaite degrees of
  freedom** — the `lmerTest` default (Kuznetsova et al. 2017), offered
  by SAS PROC MIXED as `DDFM=SATTERTHWAITE`, and the method the
  small-sample comparison of Luke (2017) recommends when groups are
  few — and the footer says so.
* The **`Random effects:`** block reports the variance components as
  rows, each with its own SE and CI: the between-subject SD of the
  intercept and of the `Days` slope (`σ`), their correlation (`ρ`), and
  the residual SD. The SD scale is the default because each SD is on
  the same scale as the coefficient it modifies (Gelman & Hill 2007) —
  ms for the intercept, ms per day for the `Days` slope: sleep
  deprivation adds about 10.5 ms per day on average (fixed `Days`),
  while individual slopes spread around that mean with an SD of about
  5.9 ms per day.
* The p column of the random-effect rows shows a **dash, by design**
  — see the next section.
* Below the rule: `n` (observations), **`N (Subject)`** (groups),
  and the **Nakagawa marginal / conditional R²** — the variance
  explained by the fixed effects alone versus fixed and random
  together (Nakagawa & Schielzeth 2013). The gap between the two
  (0.28 vs 0.80 here) is itself a summary of how much the grouping
  structure matters.
* The footer closes with the **likelihood-ratio test of the whole
  random part** against the same model without random effects,
  referred to a **chi-bar-squared** mixture distribution.

## Why the variance components carry no p-value

A variance cannot be negative, so the null hypothesis "this variance
component is zero" sits **on the boundary** of the parameter space.
There, the sampling distribution of a variance estimate is skewed and
truncated at zero, so the Wald z has no valid normal reference and its
p-value is unreliable. The likelihood-ratio test misbehaves too:
referred to a plain chi-squared, it is conservative at the boundary
(Self & Liang 1987) — which is what motivates the corrected reference
below. Printing a Wald p next to each `σ` would look rigorous and be
meaningless. None of the dedicated mixed-model
engines prints one by default — `lme4`, `nlme`, Stata `mixed`, and
MLwiN all decline; SAS PROC MIXED offers a Wald Z only behind its
`COVTEST` option, with documented cautions.

What *is* valid is a **likelihood-ratio test with a boundary-corrected
reference**: a mixture of chi-squared distributions (chi-bar-squared;
Self & Liang 1987; Stram & Lee 1994) — for a single tested component, a
50:50 mixture of adjacent degrees of freedom. The footer reports such a
test for the random part as a whole. Across several parameters jointly
the mixture weights depend on the model's geometry, so the footer falls
back on a pragmatic halved chi-squared; Stata's `mixed` approaches the
same case from the opposite side, printing the chi-squared without
halving, noting that the test "is conservative and provided only for
reference". The per-term tests below each constrain a single component
and refer the statistic to the 50:50 mixture itself. Here
chi-bar-squared(3) = 150.04, p < .001: the random structure earns its
place.

## From OLS to a multilevel model, step by step

The sections above dissect one fitted table. This one walks the
model-building path most multilevel analyses follow — the sequence
codified by Raudenbush & Bryk (2002) and taught across the multilevel
literature (Hox et al. 2018; Snijders & Bosker 2012; Bressoux 2010) —
on the data those books made canonical: the High School & Beyond
sample of 7,185 pupils in 160 U.S. schools, modelling mathematics
achievement from pupil socio-economic status (SES) and school sector.
SES is a standardized composite of parental education, occupation, and
income (sample SD 0.78, so a one-point difference is a large, roughly
1.3-SD contrast). Each step adds one ingredient, and one table shows
what it changes.

Two derived predictors do a lot of work below: the school mean of SES,
and the pupil's SES *centered within the school*. Keeping the two
apart is what lets a multilevel model estimate separate within- and
between-school effects (step 3):

```{r hsb-data, eval = have_nlme}
data("MathAchieve", package = "nlme")
data("MathAchSchool", package = "nlme")
hsb <- merge(MathAchieve[, c("School", "SES", "MathAch")],
             MathAchSchool[, c("School", "Sector")], by = "School")
hsb$School  <- factor(hsb$School)
hsb$meanses <- ave(hsb$SES, hsb$School)   # school mean of SES
hsb$cses    <- hsb$SES - hsb$meanses      # SES centered within school
```

### Step 0: the regression you would have fitted anyway

```{r hsb-ols, eval = have_nlme}
ols <- lm(MathAch ~ SES, data = hsb)
```

A one-point difference in SES is associated with 3.18 more achievement
points (SE 0.10; its column appears in the step-2 table). The price is
one silent assumption: that 7,185 pupils are 7,185 independent
observations. They are not; pupils share
schools, teachers, and neighbourhoods. The next steps make the
grouping structure part of the model.

### Step 1: the empty model — how much do schools matter?

The multilevel analysis starts with a model containing *no predictors
at all*, only a school random intercept. Its job is a single number:

```{r hsb-empty, eval = have_nlme, warning = FALSE}
# (the accompanying console warning is suppressed in this rendering)
m_empty <- lmer(MathAch ~ 1 + (1 | School), data = hsb)
table_regression(m_empty)
```

The **ICC of 0.18** is nothing but the two σ rows combined:
2.93² / (2.93² + 6.26²) ≈ 0.18. Eighteen percent of the achievement
variance lies *between* schools — the variance partition that
justifies everything that follows (and matches the number Raudenbush &
Bryk report for these data). With 160 schools, the design sits
comfortably above the roughly 50 groups Maas & Hox (2005) find
necessary for reliable standard errors and interval coverage of the
variance components — the point estimates themselves hold up with
fewer.

One display note, visible here for the first time: at this sample size
`table_regression()` skips the Wald SE and CI of the variance
components because their computation grows superlinearly with n. The
table note states the omission, and the build-time warning names the
override (`options(spicy.re_se_max_n = )`). Estimates, ICC, R², and the
LR test are unaffected; the random-part inference below runs through
`re_test`, and `re_ci = "profile"` restores boundary-respecting
profile-likelihood CIs at any sample size (a couple of seconds per
variance parameter — the route `lme4` itself recommends over SEs).
The same note (and suppressed warning) accompanies every HSB table in
this sequence.

### Step 2: a random intercept, next to the OLS

```{r hsb-ri, eval = have_nlme, warning = FALSE}
m_ri <- lmer(MathAch ~ SES + (1 | School), data = hsb)
table_regression(list(OLS = ols, Multilevel = m_ri),
                 show_columns = c("b", "se", "p"))
```

Reading the two columns against each other, five things change:

* The **intercept SE more than doubles** (0.08 → 0.19). OLS treats
  7,185 pupils as independent evidence about school-shared quantities;
  the multilevel model counts the 160 schools it actually has. The
  OLS precision was borrowed, not earned.
* The **SES coefficient itself moves** (3.18 → 2.39): a raw SES slope
  mixes two sources of information — pupils compared within schools,
  and schools compared with schools — and the multilevel fit reweights
  them toward the within-school comparison. Step 3 separates the two
  outright.
* The **ICC row reads 0.11**, not step 1's 0.18: it partitions the
  variance that SES leaves unexplained — not a contradiction, a
  conditional version of the same quantity.
* The footer's chi-bar-squared test (458.9 on 1 df) says the school
  effect is not optional.
* **One pair of rows not to read across:** the OLS AIC rests on the
  full ML likelihood, the mixed model's on the REML likelihood —
  different objectives, not a common scale. The valid whole-model
  comparison is the chi-bar-squared test above; ML-refit information
  criteria are `nested = TRUE`'s job.
* And the **marginal R² (0.08) sits below the OLS R² (0.13)** not
  because the model fits worse, but because the fixed part now carries
  a smaller, within-school-weighted slope while the between-school
  variance moves to the random part, where the Nakagawa denominator
  still counts it.

### Step 3: one variable, two effects — the contextual split

A pupil's SES and a school's SES composition are different variables
with different effects, and entering raw `SES` estimates an
uninterpretable blend of the two. The standard decomposition enters
the within-school part and the school mean separately (Enders &
Tofighi 2007; Snijders & Bosker 2012, §4.6):

```{r hsb-wb, eval = have_nlme, warning = FALSE}
m_wb <- lmer(MathAch ~ cses + meanses + (1 | School), data = hsb)
table_regression(m_wb, show_columns = c("b", "se", "p"))
```

Within a school, a one-point difference in SES is associated with 2.19
more points of achievement (`cses`); *between* schools, a one-point
difference in composition is associated with 5.87 (`meanses`) — more
than twice the individual effect. The **contextual effect** proper —
what better-off schoolmates add for the *same* pupil, over and above
their own SES — is the difference, 5.87 − 2.19 = 3.68 (Raudenbush &
Bryk 2002). To give it its own inference, refit with raw `SES` in place
of `cses` — an equivalent reparameterization of the same model — and
the `meanses` coefficient *is* the contextual effect: 3.68 (SE 0.38,
p < .001). The OLS slope of 3.18 in step 0 was a weighted blend of the
within and between effects, answering neither question.

### Step 4: does the SES slope vary across schools?

So far every school shares one SES slope. Freeing it — `(cses |
School)` — asks whether SES matters *more in some schools than
others*, and `re_test = "lrt"` attaches a p-value to the answer with a
boundary-corrected likelihood-ratio test: the model is refit without
the varying slope, and the REML deviance difference is referred to the
chi-bar-squared mixture (the mechanics — and which components are
testable at all — are dissected in *Testing individual components*
below):

```{r hsb-slope, eval = have_nlme, warning = FALSE}
m_rs <- lmer(MathAch ~ cses + meanses + (cses | School), data = hsb)
table_regression(m_rs, re_test = "lrt", show_columns = c("b", "p"))
```

The slope SD is 0.83 achievement points around the average slope of
2.19, and its test reads p = .003 — solid evidence that the slope
genuinely varies across schools — so the random slope stays. Taking the
school slopes as roughly normal, 95% of them lie within
2.19 ± 1.96 × 0.83 — about 0.6 to 3.8, the plausible value range of
Raudenbush & Bryk (2002): SES pays off in virtually every school, but
six to seven times more in the steepest than in the flattest. (The full
anatomy of this decision — what the side-by-side comparison does and
does not test, how the R² pair moves, what happens to the fixed SE —
is dissected on `sleepstudy` in *Random intercept or random slope?*
below.)

### Step 5: why do slopes differ? A cross-level interaction

A varying slope is a question, not an answer: what is it *about a
school* that flattens or steepens its SES gradient? Interacting the pupil-level
slope with a school-level predictor — a **cross-level interaction** —
is the multilevel move that answers it, here with school sector:

```{r hsb-cross, eval = have_nlme, warning = FALSE}
m_cl <- lmer(MathAch ~ cses * Sector + meanses + (cses | School),
             data = hsb)
table_regression(m_cl, show_columns = c("b", "se", "p"))
```

In public schools the within-school SES slope is 2.79; in Catholic
schools it is 2.79 − 1.35 = 1.44 — half as steep, while their
intercept sits 1.25 points higher (both conditional on school
composition, since `meanses` stays in the model). This is the classic
finding of Raudenbush & Bryk's own analysis of these data: Catholic
schooling weakens the link between a pupil's background and
achievement. The interaction row's p tests the *difference* between the
two slopes, not the Catholic slope itself; 1.44 as printed carries no
standard error — refit with `Sector` releveled
(`relevel(hsb$Sector, "Catholic")`) and it becomes a first-row
coefficient with its own inference (1.44, SE 0.18). Note where the
answer lives: the interaction is a *fixed* effect, carried by its own
row and Satterthwaite p; the slope heterogeneity it accounts for shows
in the σ `cses` row shrinking from 0.83 to 0.52.

**The likelihood rule this sequence obeyed.** `lmer()` estimates by
REML by default, and a REML likelihood is only comparable between
models with the *same fixed part* — which is why the random-structure
decisions above could read REML deviances and AICs directly (steps 3
and 4: 46,578.6 → 46,571.7). Comparing fixed parts requires ML:
`nested = TRUE` performs that refit for you, silently, as
`lme4::anova()` does. The footer's `(REML)`/`(ML)` tag on the
random-effects test line always states which likelihood the fitted
model — and hence that test — used.

**Before reporting any of these tables.** The sequence above runs
estimation, testing, and interpretation; every textbook it follows puts
three checks before the reporting step. Confirm that the fit converged
without warnings and was not singular (see *When the fit is singular*
below); inspect the level-1 residuals for pattern (`plot(fit)`); and
check that the estimated random effects look roughly normal, with no
lone school driving the variance (`qqnorm()` on `ranef(fit)`). The
table reports what the model claims, not whether the model is adequate
(Snijders & Bosker 2012, ch. 10); `performance::check_model()` runs the
battery in one call.

## Three levels, and crossed designs

Nothing above is limited to two levels. A third level enters through
the grouping formula — `(1 | batch/cask)` reads "casks nested in
batches" — and the table gains one variance row and one group count
per level:

```{r pastes, eval = have_mixed}
p3 <- lmer(strength ~ 1 + (1 | batch/cask), data = Pastes)
table_regression(p3, show_columns = c("b", "se"))
```

Both levels of nesting get their σ row, `N (groups)` counts each
level, and the footer's chi-bar-squared test covers the random part
jointly (2 df here); at n = 60, far below the size cap, the Wald SE of
the variance components is back — and the CI with it, trimmed from this
display by `show_columns`. Random slopes can sit at any
level — `(x | school) + (1 | class)` renders each block under its
grouping factor — and **crossed** (non-nested) structures such as
`(1 | plate) + (1 | sample)` flow through the same machinery: one
block per factor, no extra syntax.

## When the fit is singular

The boundary is not only a testing problem — estimates land *on* it.
When a variance component is estimated at exactly zero (`lme4` calls
this a **singular fit**), its Wald SE and CI are meaningless, so
`table_regression()` omits them and says why in the footer:

```{r singular, message = FALSE, warning = FALSE}
# (the accompanying console warning is suppressed in this rendering)
set.seed(2026)
d <- data.frame(x = rnorm(120), g = factor(rep(1:12, each = 10)))
d$y <- 2 + 0.5 * d$x + rnorm(120)   # no group effect at all
sfit <- lmer(y ~ x + (1 | g), data = d)
table_regression(sfit)
```

The chi-bar-squared test reads p = 1.000 — the boundary case. The
likelihood-ratio statistic is exactly zero, and half of the
chi-bar-squared null distribution is a point mass *at* zero, so the
observed value carries no evidence whatever against the no-group-effect
null (Stata prints `Prob >= chibar2 = 1.0000` in the same situation).
The note states the fact for the table's reader; the actionable advice arrives as a console warning
when the table is built: simplify the random structure, or test the
component with `re_test = "lrt"`, rather than report a zero variance
with invented uncertainty.

## Testing individual components: `re_test`

When a reviewer asks about one *specific* component — "do you really
need the random slope?" — request an opt-in per-term test.
`re_test = "lrt"` refits the model without each testable component and
fills the p column of its row with the boundary-corrected
likelihood-ratio test. The statistic is the one `lmerTest::ranova()`
computes (a REML deviance difference); `ranova` refers it to a plain
χ², which is conservative at the boundary, so spicy applies the
chi-bar-squared mixture instead:

```{r re-test-lrt}
table_regression(fit, re_test = "lrt")
```

Only the `Days` slope row gains a p-value: following the marginality
rule of `lmerTest::ranova()`, an intercept is tested only when it
stands alone on the left of the `|` — as in `(1 | g)` — and a
correlation is not removable on its own.
The footer names the test.

For a model with a **single** variance component — here the
random-intercept-only baseline, which returns in the comparison below —
`re_test = "rlrt"` instead uses the exact restricted likelihood-ratio
test of Crainiceanu & Ruppert (2004), with the null distribution
simulated by `RLRsim::exactRLRT()`. This exact test is preferable in
small samples, where the chi-bar-squared approximation is at its
weakest:

```{r fit-ri}
fit_ri <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
```

```{r re-test-rlrt, eval = have_rlrsim}
table_regression(fit_ri, re_test = "rlrt")
```

## The random-effects block on your terms

Three arguments control the block without touching the estimates.
`re_scale = "variance"` switches the rows from SD (`σ`) to variance
(`σ²`); `re_columns` restricts which quantities the variance-component
rows display (the SE and CI on a variance component come from
`merDeriv` and are themselves boundary-fragile — dropping them is a
defensible editorial choice); `show_re = FALSE` removes the block and
all random-effects footer lines, leaving a fixed-effects-only table
(the `N (groups)` and R² fit-stat rows remain):

```{r re-scale}
table_regression(fit, re_scale = "variance", re_columns = "est")
```

Where those SEs come from depends on the engine: `lmer` / `glmer`
rows use the `merDeriv` Hessian (Wang & Merkle 2018; the cells fall
back to NA when the package is absent or the fit is singular),
`glmmTMB` its native Wald `confint()`, and `nlme::lme` its
`intervals()`. Correlation (`ρ`) rows carry a multivariate
delta-method SE for `lmer`, native intervals elsewhere. Switching
`re_scale` transports SE and CI between the SD and variance scales by
the delta method (`SE(σ) = SE(σ²) / 2σ`), so the displayed values
agree with `VarCorr()` on either scale up to rounding.

Whatever the display, `broom::tidy()` always carries the full set
(estimate, SE, CI) for the variance-component rows — display and data
are decoupled.

## Random intercept or random slope? Models side by side

The HSB sequence above made this decision on real school data with one
`re_test` call; this section dissects its full anatomy on
`sleepstudy`, an example small enough that every quantity can be
checked by hand.
Model comparison is where the row layout pays off: the variance
components align across columns like any coefficient, and each model
gets its own footer line.

```{r side-by-side}
table_regression(
  list("Intercept only" = fit_ri, "+ random slope" = fit),
  show_columns = c("b", "se", "p")
)
```

**What is — and is not — being tested here.** Each footer's
chi-bar-squared line answers one question only: *does this model need
its random part at all?* — the fit against a plain linear regression.
Neither footer compares the two columns to each other. That comparison
— *does the slope earn its place on top of the intercept?* — is the
per-term test of the previous section: `re_test = "lrt"` on the slope
model refits exactly the intercept-only model of column 1 and refers
the REML deviance difference (42.8) to the 50:50 mixture of χ²(1) and
χ²(2), p ≈ 3 × 10⁻¹⁰ — the textbook procedure for retaining a random
slope (Snijders & Bosker 2012, §6.2; Stram & Lee 1994). AIC and BIC
agree (1794.5 → 1755.6; 1807.2 → 1774.8). For hierarchies that grow
the *fixed* part instead, `nested = TRUE` adds chi-squared, AIC, and BIC
change rows — an ML-refit test with a plain χ² reference,
appropriate there but conservative for random-structure changes, which
is why the slope decision belongs to `re_test`.

**Reading the two R² rows — without over-reading them.** The Nakagawa
pair is descriptive, and each member answers a different question.
*Marginal* R² is the share of variance explained by the fixed effects
alone: it reads 0.28 in both columns — 0.280 versus 0.279 before
rounding — because the two models share a fixed part and the balanced
design leaves the numerator identical. The near-equality is a property
of this fit, not a theorem: the random-effect variances enter the
denominator of the marginal R², so it can shift when the random
structure changes. *Conditional* R² adds the
random effects: it rises (0.70 → 0.80) because individual slopes let
each subject's own trajectory absorb variance the first model left in
the residual. The trap: a conditional R² can hardly go *down* when the
random structure grows, so its increase is **not** evidence that the
slope is needed — it measures how much the grouping structure
explains, never whether a component earns its place. Selection belongs
to the boundary-corrected test and the information criteria; the R²
pair then describes the model you selected. (Where the numbers come
from: spicy computes the pair natively from the closed-form
decomposition for Gaussian, single-trial binomial — logit, probit,
cloglog — and Poisson log fits, cross-validated against
`performance::r2_nakagawa()` to 10⁻¹⁰, and delegates the remaining
families — multi-trial binomial, negative binomial, beta,
zero-inflation — to `performance`, a Suggests dependency: absent, the
R² rows render as NA and the rest of the table is unaffected.)

**The ICC row is an interpretation, not a test.** For the
intercept-only model it comes straight from the two σ rows above it:
ICC = σ²(Subject) / (σ²(Subject) + σ²(Residual)) =
37.12² / (37.12² + 30.99²) ≈ 0.59 — squares, because the block
displays standard deviations (`re_scale = "variance"` shows the variances directly).
Read it two equivalent ways: 59% of the variance remaining after the
fixed effects lies *between* subjects rather than within them — the
variance partition coefficient of the multilevel textbooks (Snijders &
Bosker 2012, §3.3) — and, equivalently, two observations from the same
subject are expected to correlate at 0.59. It appears only for the
intercept-only model: with a random slope, the correlation between two
observations from the same subject depends on *when* they were taken,
so a single ICC no longer exists — none is printed.

One further detail repays attention: the fixed `Days` SE **nearly
doubles** (0.80 → 1.55). A model without the random slope overstates the
precision of the average slope by ignoring between-subject slope
variation (Schielzeth & Forstmeier 2009) — the inferential stake of
the decision, visible in the table itself.

Whether to *test* the slope at all is its own debate. The design-driven
position includes every slope the design permits (Barr et al. 2013);
parsimony advocates prune components the data cannot support
(Matuschek et al. 2017); Gelman & Hill (2007) would rather estimate
every component and let partial pooling do the pruning: when a variance
is near zero, the group-level coefficients shrink almost entirely
toward the common one, and the model collapses to the simpler fit on
its own. The table serves
all three positions: the components are always displayed with their
uncertainty, and the test stays opt-in.

## Generalized linear mixed models

`glmer()` fits render identically, with two family-driven changes:
inference is **Wald-z** (no Satterthwaite for GLMMs), and
`exponentiate = TRUE` is link-gated exactly as for `glm` — odds ratios
under logit, rate ratios under log, and a hard error on links whose
exponential has no meaning (probit). The example is `lme4::cbpp`,
herd-level incidence of a cattle disease across four periods:

```{r glmer}
gfit <- glmer(
  cbind(incidence, size - incidence) ~ period + (1 | herd),
  data = cbpp, family = binomial()
)
table_regression(gfit, exponentiate = TRUE)
```

The random-effect row is now on the **log-odds scale** (the linear
predictor's scale), not the outcome scale: σ = 0.64 means a herd one SD
above the average carries about e^0.64 ≈ 1.9 times the baseline *odds*
of infection, so between-herd variation in susceptibility is far from
negligible. Incidence falls across the trial — relative to period 1,
the odds of new infection are 0.37 (period 2), 0.32 (period 3), and
0.21 (period 4). The footer's chi-bar-squared test (14.01 on 1 df,
p < .001) confirms the herd effect, and the Nakagawa R² pair carries
over unchanged.

No ICC row appears. The latent-scale formula σ² / (σ² + π²/3) describes
a *single* Bernoulli trial, and it would still return a number here
(0.11). But an aggregated `cbind()` count is not a single trial: with
denominators of varying size, the correlation between two rows no
longer matches that single-trial quantity, so one printed ICC would be
ambiguous and none is shown. A Bernoulli 0/1 `glmer`, where every
observation *is* one trial, shows it — computed on the latent logit
scale, whose residual variance is fixed at π²/3. The same adjusted-ICC
rule covers the other single-trial links — σ²_d = 1 under probit,
π²/6 under cloglog — and Poisson log fits, where
σ²_d = log(1 + 1/λ) with λ taken from the null model (Nakagawa,
Johnson & Schielzeth 2017); the printed value matches
`performance::icc()`'s adjusted ICC to 10⁻⁶.

## Average marginal effects

An odds ratio answers "by what factor do the odds change"; the question
a substantive reader more often asks is "by how many percentage points
does the probability change". The `"ame"` / `"ame_se"` / `"ame_ci"` /
`"ame_p"` tokens of `show_columns` work for mixed-effects fits exactly
as for `glm`, delegated to `marginaleffects::avg_slopes()`, always on
the **response scale**:

* `glmer` binomial: **probability scale** (percentage points), not
  log-odds;
* `glmer` / `glmmTMB` Poisson: **count scale** (units of the outcome),
  not log-rate;
* Gaussian fits (identity link): the AME equals the `B` coefficient —
  the column is filled but redundant.

```{r mixed-ame, eval = have_margins}
set.seed(1)
n <- 500
g <- factor(rep(1:25, length.out = n))
x <- rnorm(n)
cat <- factor(sample(c("A", "B", "C"), n, replace = TRUE))
y <- rbinom(n, 1, plogis(0.5 + 0.8 * x + (cat == "B") * 0.3 +
                           (cat == "C") * -0.5 + rnorm(25)[g]))
afit <- lme4::glmer(y ~ x + cat + (1 | g), family = binomial)
table_regression(afit, show_columns = c("b", "se", "p", "ame", "ame_p"))
```

Factor predictors are handled level-by-level: AME rows align under the
same factor header (`cat: A (ref.) / B / C`), each level sharing a row
with its `B` coefficient — for a factor level the AME is the average
*discrete* change against the reference, not a derivative.

Inference on the AME rows is Wald-z asymptotic (`df = Inf`) with the
model-based vcov. That matches the B-row inference for `glmer` and
`glmmTMB`, which are Wald-z themselves; it does *not* match the
t-based B rows of `lmerTest` fits (Satterthwaite df) or `nlme::lme`
(containment df) — there the AME row keeps its large-sample z while
the B row keeps its t, so under an identity link the same estimate can
carry two different p-values in the same row. Treat the AME p as a
large-sample approximation for those classes.

## `glmmTMB`: beyond `lme4`

`glmmTMB` fits reach the same table through the same code path, and
bring two extra components. A zero-inflation (`ziformula =`) or
dispersion (`dispformula =`) component renders as its own labelled
block of rows, with a footer line stating what the component models and
on which scale — here with the `Salamanders` count data. Sixty percent
of these stream surveys count no salamanders, and the zeros cluster in
mined sites (84% zero, against 38% elsewhere): some are ordinary
sampling zeros, others sites that mining has rendered uninhabitable.
The `ziformula` models the probability of that structural absence
separately from abundance where salamanders can occur:

```{r glmmtmb, eval = have_glmmTMB}
data("Salamanders", package = "glmmTMB")
zfit <- glmmTMB::glmmTMB(
  count ~ mined + (1 | site),
  ziformula = ~ mined,
  family = poisson(), data = Salamanders
)
table_regression(zfit, exponentiate = TRUE)
```

Note the per-block exponentiation: the count coefficients become
**IRR**, while the logit zero-inflation coefficients become **odds
ratios of a structural zero**. Read one number from each: relative to
mined streams, unmined ones carry about three times the abundance
(IRR 3.13) where salamanders occur at all, and about a fifth of the
odds of structural absence (OR 0.18). Component blocks are covered in
depth in the counts documentation (`?table_regression_counts`); opt out
with `show_components = FALSE`.

## `nlme::lme`

Beyond legacy code, `lme()` remains the engine of choice when the
residuals need structure `lme4` cannot fit — heteroscedasticity across
groups (`weights = varIdent()`) or serial correlation within them
(`correlation = corAR1()`). Such fits flow through the same layout, and
the footer's boundary-corrected test refits the null carrying the same
residual structure. Its fixed-effect inference is the **containment-df
t-test** native to `nlme`, and the footer says so:

```{r nlme, eval = have_nlme}
lfit <- nlme::lme(Reaction ~ Days, random = ~ 1 | Subject,
                  data = sleepstudy)
table_regression(lfit)
```

## Cluster-robust and other variance estimators

Mixed fits honour the cluster-robust family (`"CR0"`–`"CR3"`) through
`clubSandwich`, with Satterthwaite small-sample degrees of freedom
computed from the robust covariance — the footer attributes them
accordingly. Reach for `CR*` when you suspect the random-effects
structure does not fully capture the within-group dependence:

```{r cr2, eval = have_clubsw}
table_regression(fit_ri, vcov = "CR2", cluster = ~Subject)
```

Requests that have no valid backend are refused with a clear error
rather than silently ignored: `HC*` (a single-level estimator whose sandwich
assumes independent observations — the very dependence the random
effects exist to model; use `CR*` instead), `CR*` on `glmer` (no `clubSandwich` method), and the
resampling estimators. For `glmmTMB`, `CR*` covers the conditional
component only, and the footer discloses it.

## Standardized coefficients

Mixed fits support `standardized = "refit"` — the model is refit on
z-scored data (Cohen et al. 2003), for mixed models the only well-posed
approach, since the marginal SD(y) has no unique decomposition across
levels. The β rows inherit the same Satterthwaite reference
distribution as their unstandardized counterparts, so B and β carry one
consistent p per term:

```{r standardized}
table_regression(fit, standardized = "refit", show_columns = c("b", "beta", "p"))
```

The algebraic shortcuts (`"posthoc"`, `"basic"`, `"smart"`) are
therefore not defined for mixed models and fall back to `"refit"` with
a warning. Formulas with inline transforms (`log(x)`, `poly()`) decline
the refit and omit the β rows, with a warning; pre-compute the
transformed column in `data` for an exact refit.

## Output formats

Everything above used the default console output. The same table
renders as a raw data frame, a long broom-style tibble, or — with the
corresponding Suggests package — a rich `gt`, `flextable`,
`tinytable`, Excel, or Word table; the random-effects block carries
through to every format.

```{r outputs-df}
head(table_regression(fit, output = "data.frame"), 8)
```

```{r outputs-gt, eval = build_rich_tables && have_mixed}
table_regression(fit, output = "gt")
```

In [`broom::tidy()`](https://broom.tidymodels.org) the
variance-component rows are marked `estimate_type = "vc"` and their
terms are prefixed `re::` (correlations carry a `::cor` suffix), so
they filter cleanly:

```{r broom, eval = have_broom}
td <- broom::tidy(table_regression(fit))
td[td$estimate_type == "vc", c("term", "estimate", "std.error", "conf.low", "conf.high")]
```

## References

* Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random
  effects structure for confirmatory hypothesis testing: Keep it
  maximal. *Journal of Memory and Language*, 68(3), 255–278.
* Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting
  linear mixed-effects models using lme4. *Journal of Statistical
  Software*, 67(1).
* Bressoux, P. (2010). *Modélisation statistique appliquée aux sciences
  sociales* (2nd ed.). De Boeck.
* Cohen, J., Cohen, P., West, S. G., & Aiken, L. S. (2003). *Applied
  Multiple Regression/Correlation Analysis for the Behavioral
  Sciences* (3rd ed.). Lawrence Erlbaum.
* Crainiceanu, C. M., & Ruppert, D. (2004). Likelihood ratio tests in
  linear mixed models with one variance component. *Journal of the Royal
  Statistical Society, Series B*, 66(1), 165–185.
* Enders, C. K., & Tofighi, D. (2007). Centering predictor variables in
  cross-sectional multilevel models: A new look at an old issue.
  *Psychological Methods*, 12(2), 121–138.
* Gelman, A., & Hill, J. (2007). *Data Analysis Using Regression and
  Multilevel/Hierarchical Models*. Cambridge University Press.
* Hox, J. J., Moerbeek, M., & van de Schoot, R. (2018). *Multilevel
  Analysis: Techniques and Applications* (3rd ed.). Routledge.
* Luke, S. G. (2017). Evaluating significance in linear mixed-effects
  models in R. *Behavior Research Methods*, 49(4), 1494-1502.
* Kuznetsova, A., Brockhoff, P. B., & Christensen, R. H. B. (2017).
  lmerTest package: Tests in linear mixed effects models. *Journal of
  Statistical Software*, 82(13).
* Maas, C. J. M., & Hox, J. J. (2005). Sufficient sample sizes for
  multilevel modeling. *Methodology*, 1(3), 86–92.
* Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D.
  (2017). Balancing Type I error and power in linear mixed models.
  *Journal of Memory and Language*, 94, 305–315.
* Nakagawa, S., & Schielzeth, H. (2013). A general and simple method
  for obtaining R² from generalized linear mixed-effects models.
  *Methods in Ecology and Evolution*, 4(2), 133–142.
* Raudenbush, S. W., & Bryk, A. S. (2002). *Hierarchical Linear
  Models: Applications and Data Analysis Methods* (2nd ed.). Sage.
* Schielzeth, H., & Forstmeier, W. (2009). Conclusions beyond support:
  overconfident estimates in mixed models. *Behavioral Ecology*,
  20(2), 416–420.
* Self, S. G., & Liang, K.-Y. (1987). Asymptotic properties of maximum
  likelihood estimators and likelihood ratio tests under nonstandard
  conditions. *JASA*, 82(398), 605–610.
* Snijders, T. A. B., & Bosker, R. J. (2012). *Multilevel Analysis: An
  Introduction to Basic and Advanced Multilevel Modeling* (2nd ed.).
  Sage.
* Stram, D. O., & Lee, J. W. (1994). Variance components testing in
  the longitudinal mixed effects model. *Biometrics*, 50(4),
  1171–1177.
* Wang, T., & Merkle, E. C. (2018). merDeriv: Derivative computations
  for linear mixed effects models with application to robust standard
  errors. *Journal of Statistical Software*, 87(Code Snippet 1), 1-16.
