---
title: "Publication-ready regression tables"
description: >
  Coefficient tables for `lm` and `glm` fits with APA-aligned
  formatting: classical, heteroskedasticity-consistent and
  cluster-robust variance, four standardisation methods, partial
  effect sizes with noncentral-F CIs, average marginal effects,
  multiple-comparison adjustment, side-by-side and hierarchical
  layouts, and output to console, gt, tinytable, flextable,
  Excel, Word, or clipboard.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Publication-ready regression tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

`table_regression()` produces a coefficient summary table from
one or several fitted `lm()` or `glm()` models. The output is
publication-ready by default and follows APA Manual 7 (American
Psychological Association 2020, Tables 7.13–7.15) formatting
conventions: paired estimate-and-CI columns, APA p-values without
leading zero, factor levels grouped under their parent variable,
fit statistics at the foot of the table, and a self-documenting
note line that names the variance estimator and any
methodological choice that affected the rendered values.

The function is **fit-first**: you pass already-fitted models,
not raw data and a formula. The same object exports cleanly to a
long broom-canonical frame (`broom::tidy()`) and to a
one-row-per-model glance summary (`broom::glance()`).

`lm` and `glm` are both supported. The *Generalised linear
models* section covers the glm-specific argument semantics; the
*Mixed-effects models* section covers the random-effects panel
that mixed-effects fits add below the fit-statistics footer
(`lmer`, `glmer`, `glmmTMB`, `lme`). Everything else applies to
all four families.

## Basic usage

Pass a fitted `lm()` object. The default rendering returns a
single-model table with `B`, `SE`, `95% CI`, and `p` columns and
a fit-statistics footer (`n`, `R²`, `Adj.R²`):

```{r basic}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit)
```

Reading the table:

* Each predictor occupies one row; factor predictors are grouped
  under their parent variable name with one row per level. The
  reference level carries the `(ref.)` annotation and an em-dash
  across the statistic columns, so the substantive comparison is
  visible in a single glance (NEJM / BMJ clinical convention).
* The `95% CI` column reports a symmetric Wald interval at the
  level set by `ci_level` (default 0.95). The CI header tracks
  the chosen level — switching to `ci_level = 0.99` re-labels the
  column accordingly.
* The footer line names the variance estimator in plain English
  so the reader can find the inferential regime without leaving
  the table. The estimator switches with `vcov` (covered in
  *Robust variance* below).

## Standardised coefficients

Standardised coefficients (`β`) make predictors with different
natural scales comparable: a one-standard-deviation increase in
`X` predicts a `β`-standard-deviation change in `Y`. APA Manual 7
§7.13 recommends reporting both `B` and `β` so the unstandardised
effect (natural units, interpretable) stays alongside the
standardised effect (comparable across predictors).

`standardized` selects the method. Four are available; the choice
is consequential and well-documented (Cohen, Cohen, West, and
Aiken 2003 §3.4; Gelman 2008):

* `"refit"` — refit the model on z-scored outcome and predictors.
  Gold-standard convention, used by SPSS `REGRESSION` and Stata
  `regress, beta`. Both numeric and dummy-coded predictors enter
  the refit on the same scale.
* `"posthoc"` — algebraic rescaling `β = B × SD(X) / SD(Y)`,
  applied to the original fit. Numerically identical to `"refit"`
  for purely linear-additive Gaussian models; preferred when
  refitting is expensive or when `lm()` was wrapped in a pipeline
  that resists re-execution.
* `"basic"` — algebraic, but factor dummies keep their 0/1 scale
  rather than being z-scored. Useful when factor levels carry
  meaningful base rates that scale-free standardisation would
  obscure.
* `"smart"` — Gelman's (2008) recommendation: numeric predictors
  divided by `2 × SD(X)`; binary predictors centred only. The
  resulting `β` is the predicted change in `Y` for one within-
  sample standard deviation of `X`, on the original `Y` scale,
  which keeps factor and numeric coefficients on roughly
  comparable footing.

When `standardized != "none"`, the `"beta"` token is auto-injected
into `show_columns` immediately after `"b"`:

```{r beta}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, standardized = "refit")
```

**Caveat: standardised coefficients with interactions or
transformed terms.** When the model contains a product term, an
`I()`, `poly()`, `log()`, or `splines::ns()` wrapper, the
standardised coefficient of the non-additive term has no
closed-form "one-SD change in X" reading (Aiken and West 1991;
Cohen et al. 2003 §7.7). The function emits a classed
`spicy_caveat` warning at runtime AND prints a method-specific
caveat line in the table footer, so the limitation is exposed at
the point of use without blocking the table.

## Per-coefficient effect sizes

For each predictor, you can request a partial effect-size column.
Three measures are available — Cohen's `f²` (`partial_f2`),
Pearson's partial `η²` (`partial_eta2`), and the Olejnik–Algina
bias-corrected partial `ω²` (`partial_omega2`). Each carries a
confidence interval derived from noncentral-`F` inversion
(Smithson 2003; Steiger 2004), exposed as a separate `<token>_ci`
column. The shortcuts `"all_f2"`, `"all_eta2"`, `"all_omega2"`
expand to the point estimate and its CI pair in one go:

```{r partial}
fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sochealth)
table_regression(
  fit,
  show_columns = c("b", "p", "all_eta2", "all_omega2")
)
```

Methodology notes:

* The partial F-test is computed on a Type-II ANOVA reference
  (`car::Anova`), which respects the principle of marginality and
  is the SAS / SPSS default for unbalanced designs.
* The `ω²` point estimate is bias-corrected via the Olejnik and
  Algina (2003) formula `((F − 1) × df1) / (F × df1 + N − df1)`,
  yielding a less-biased small-sample estimator than partial `η²`.
* The CI bounds come from inverting the noncentrality parameter
  of the F-distribution at the lower and upper confidence levels
  (Steiger 2004 §4). The Steiger inversion always brackets the
  bias-corrected point estimate, even when the lower bound clips
  at zero (common for near-null terms).
* For factor predictors with `k` levels, the partial F-test is
  the joint `(k − 1)` df Wald test, so the same effect-size value
  is broadcast across all non-reference dummy rows; the reference
  row shows an em-dash.

## Average marginal effects (AME)

The `B` coefficient is the model's *structural* effect — the
change in the linear predictor `Xβ` for a one-unit change in
`x`, holding the other predictors constant. The **average
marginal effect (AME)** is the *observable* effect on the
response: the average partial derivative `dE[Y|X] / dx` over the
sample. For a purely linear-additive `lm()` the two coincide. For
models with interactions or transformed terms, and for any
`glm()` with a non-identity link, the AME is the quantity the
substantive reader actually needs.

Add `"ame"` to `show_columns` for the point estimate. Companion
tokens display the SE, CI, and p-value: `"ame_se"`, `"ame_ci"`,
`"ame_p"`. The shortcut `"all_ame"` expands to the full set.

```{r ame-lm}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(
  fit,
  show_columns = c("b", "ame", "ame_ci", "ame_p")
)
```

Reading conventions:

* `"p"` always refers to the `B` (or `β`) coefficient, never to
  the AME. Use `"ame_p"` for the AME-specific p-value.
* Placing `"ame"` after `"p"` keeps the "which p belongs to what"
  reading unambiguous — the B-block (`B / SE / p`) is closed
  before the AME-block opens.
* For non-linear models the AME is reported on the **response
  scale**, not the link scale. The *Generalised linear models*
  section below works out a logistic-regression example.

Inference is delegated to
[`marginaleffects::avg_slopes()`][marginaleffects::avg_slopes] and
respects the chosen variance estimator. Under cluster-robust
variance (covered in *Robust variance*), the AME inference shares
the coefficient's t-distribution and Satterthwaite-corrected
degrees of freedom via `clubSandwich::linear_contrast()`, so `B`
and AME are reported on the same inferential footing in the same
table.

## Multiple models side by side

Pass a list of `lm()` fits. The default column layout places each
model in its own panel under a centered **spanner label** showing
the model name; sub-columns (`B / SE / p`) are repeated under the
spanner. When dependent variables differ across models and the
user did not supply labels (`model_labels = ` or named list), the
bare DV name is lifted into the spanner automatically:

```{r multi}
m_wellbeing <- lm(wellbeing_score ~ age + sex + smoking,
                   data = sochealth)
m_bmi       <- lm(bmi             ~ age + sex + smoking,
                   data = sochealth)
table_regression(list(m_wellbeing, m_bmi))
```

When all models share the same DV, the DV appears in the title.
Use a named list — e.g. `list(Crude = m1, Adjusted = m2)` — to
set the spanner labels explicitly; pass `model_labels = c(...)`
to override the names from the list.

## Hierarchical / nested regression

Set `nested = TRUE` to add **in-table change-statistic rows**
(APA Table 7.13 / Stata `esttab` / SPSS Model Summary
convention). Each adjacent pair (M2 vs M1, M3 vs M2, …)
contributes one column of change stats below `R² / Adj.R²`; the
FIRST model column gets em-dashes (no previous model to compare
to).

Hierarchical regression requires every model in the sequence to
share the same analytic sample: identical `nobs` AND identical
response variable. R's listwise deletion otherwise produces
different `n` per model as soon as a predictor introduced by M2
or M3 has missing values that M1 did not see — a silent bias on
the change statistics. We restrict to complete cases on the
union of predictors up front:

```{r sochealth-cc}
sochealth_cc <- sochealth |>
  dplyr::select(wellbeing_score, age, sex, smoking, bmi, region) |>
  na.omit()
```

```{r nested}
m1 <- lm(wellbeing_score ~ age + sex,                 data = sochealth_cc)
m2 <- lm(wellbeing_score ~ age + sex + smoking,       data = sochealth_cc)
m3 <- lm(wellbeing_score ~ age + sex + smoking + bmi, data = sochealth_cc)
table_regression(list(m1, m2, m3), nested = TRUE)
```

Default change tokens auto-injected: `c("r2_change", "f_change",
"p_change")` for `lm` (APA hierarchical-regression standard),
`c("lrt_change", "p_change")` for `glm` (Hosmer & Lemeshow §3.5),
`c("aic_change", "bic_change", "lrt_change", "p_change")` for
mixed-effects fits (`lmer` / `glmer` / `glmmTMB` / `nlme::lme` —
Pinheiro & Bates 2000 §2.4.1). Customise via `show_fit_stats`;
the order of tokens controls the order of the rows. Other change
tokens are available: `"adj_r2_change"`, `"f2_change"`,
`"deviance_change"`, `"aic_change"` / `"aicc_change"` /
`"bic_change"`. Variance-explained change tokens (Δr², Δf²) are
NA for mixed-effects pairs — the F-test framework that grounds
them doesn't apply.

Validation is strict: identical `nobs` AND identical response
across all models, otherwise a `spicy_invalid_input` error
explains the listwise-deletion trap and suggests refitting on the
common subset.

## Robust variance

The default `vcov = "classical"` reports the OLS standard error,
valid under homoskedastic, independent errors. Two robust
alternatives are first-class arguments; bootstrap and jackknife
variants are also available via `vcov = "bootstrap"` / `"jackknife"`.

### Heteroskedasticity-consistent (HC)

When error variance plausibly depends on the predictors (a
ubiquitous concern in cross-sectional social-science data), set
`vcov = "HC*"` for sandwich-style standard errors via
[`sandwich::vcovHC()`][sandwich::vcovHC]. The valid types are
`HC0` through `HC5`; `HC3` is the small-sample-friendly default
(Long and Ervin 2000):

```{r hc}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, vcov = "HC3")
```

The footer reads "*Std. errors: heteroskedasticity-robust (HC3)*";
the column header for the confidence interval automatically tracks
`ci_level`.

### Cluster-robust (CR)

For clustered observations (repeated measures on the same person,
students within schools, observations within regions),
`vcov = "CR*"` requests cluster-robust variance via
[`clubSandwich::vcovCR()`][clubSandwich::vcovCR], with the cluster
identifier supplied through `cluster`. Three forms are accepted:

* **Formula** — `cluster = ~region`. The variables are looked up
  in `model.frame(fit)` first, then in the model's original
  `data` argument. Recommended: independent of the dataset's
  name, composable for multi-way clustering
  (`cluster = ~region:year`).
* **String** — `cluster = "region"`. Single column name resolved
  the same way. Convenient but cannot express interactions.
* **Vector** — `cluster = df$region`. Atomic vector of length
  `nobs(fit)`. Use this when the cluster key is derived on the
  fly (`cluster = interaction(df$region, df$year)`) or pulled
  from a different dataset with matching row order.

Bare unquoted names (`cluster = region`) are **not** accepted —
they would require non-standard evaluation that breaks under
programmatic use (function wrapping, loops, dynamic column
choice).

```{r cr}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth_cc)
table_regression(
  fit,
  vcov = "CR2",
  cluster = ~region,
  show_columns = c("b", "se", "ci", "p", "ame", "ame_p")
)
```

`CR2` (the Bell-McCaffrey adjustment) is the recommended default
under few clusters (Pustejovsky and Tipton 2018; Imbens and
Kolesár 2016). Coefficient inference uses
[`clubSandwich::coef_test()`][clubSandwich::coef_test] with
Satterthwaite-corrected degrees of freedom. When AME columns are
requested, the same Satterthwaite framework is applied to the
AME contrast via `clubSandwich::linear_contrast()`, so `B` and
AME share the same t-distribution with the same df — visible in
the footer.

## Multiple-comparison adjustment

`p_adjust` applies a family-wise or false-discovery-rate
adjustment to the p-values via
[`stats::p.adjust()`][stats::p.adjust]. The family is the model's
full coefficient set (intercept and reference rows excluded);
`B` and `AME` are adjusted as independent families within the
same call. Available methods are the same as in
`stats::p.adjust()`: `"none"` (default), `"holm"`, `"hochberg"`,
`"hommel"`, `"bonferroni"`, `"BH"` / `"fdr"`, `"BY"`.

```{r p-adjust}
fit <- lm(wellbeing_score ~ age + sex + smoking + education,
          data = sochealth)
table_regression(fit, p_adjust = "bonferroni")
```

The footer documents the chosen method and the family size; the
SE column is unchanged. `p_adjust` is orthogonal to `vcov`:
combining a robust SE with a family-wise correction is fully
supported, e.g. `vcov = "HC3", p_adjust = "holm"`. The adjustment
is applied **before** `keep` / `drop` filtering, so the family
stays the model's full coefficient set and the displayed adjusted
p-values reflect the right denominator regardless of which
subset is shown.

A methodological note. Adjusting the p-values of every
coefficient of a single regression model is *not* the standard
convention in social-science or clinical reporting (Rothman 1990;
Gelman, Hill, and Yajima 2012; Greenland 2017; Harrell 2015 §5.4;
APA Manual 7 §6.46). Each coefficient tests a scientifically
distinct hypothesis on a distinct predictor, which is not the
situation that family-wise procedures were designed for. The
default `"none"` reflects this consensus.

Adjustment is nonetheless legitimate in three contexts:

1. **Mass screening** with many candidate predictors and no
   prior hypothesis — typically `"BH"` (Benjamini–Hochberg,
   false discovery rate).
2. **Pre-registered multi-endpoint confirmatory designs** —
   typically `"holm"` (Holm's step-down, strong family-wise
   error rate control).
3. **When a reviewer, an editor, or a statistical analysis plan
   explicitly requests it** — apply the requested method and
   document it in the footer.

The argument is exposed under the same *transparency* rule used
for `standardized`: the tool is here, the methodological choice
is yours, and the footer makes the choice visible to the reader.

## Filtering displayed coefficients

`keep` and `drop` accept regular expressions matched against
coefficient names (as returned by [stats::coef()]). They are
mutually exclusive — pick `keep` to whitelist focal predictors,
or `drop` to hide a few control variables. Multiple patterns
combine with logical OR.

```{r keep}
fit <- lm(wellbeing_score ~ age + sex + smoking + bmi + education,
          data = sochealth)
table_regression(fit, keep = c("^smoking", "^bmi$"))
```

```{r drop}
table_regression(fit, drop = "^education")
```

Together with `p_adjust`, this is the article-ready workflow:
adjust on the full model, display only the rows the reader cares
about — `table_regression(fit, p_adjust = "BH", keep = "^treatment")`.

## Generalised linear models (glm)

`table_regression()` accepts any `glm()` fit. Inference defaults
to the `z`-asymptotic Wald regime — the convention used by
`summary.glm()`, Stata's `logit, or`, and SPSS `LOGISTIC
REGRESSION`. The table title becomes family-aware ("Logistic
regression", "Poisson regression", "Probit regression", …) and
the default fit-statistics block swaps in `nobs`,
`pseudo_r2_mcfadden`, `pseudo_r2_nagelkerke`, and `AIC` instead
of `R²` and `Adj.R²`.

We illustrate with a logistic regression of `smoking` on `sex`,
`age`, and `education` from `sochealth`. The `education` variable
is an ordered factor in the source data; we convert it to a plain
factor so contrasts are dummy-coded (one row per level) rather
than the polynomial trends (`.L`, `.Q`, …) R applies by default to
ordered factors. The model mixes a binary factor (`sex`), a
numeric predictor (`age`), and a 3-level factor (`education`) —
enough to exercise every glm-specific feature below.

```{r glm-basic}
sh <- sochealth |>
  dplyr::mutate(education = factor(education, ordered = FALSE))

fit <- glm(smoking ~ sex + age + education, data = sh,
           family = binomial)
table_regression(fit)
```

### Response-scale display: `exponentiate = TRUE`

Set `exponentiate = TRUE` to switch the `B` column to the
response scale and rebrand its header per family and link: `OR`
for `binomial(logit)`, `IRR` for `poisson(log)`, `HR` for
`binomial(cloglog)`, `RR` for `binomial(log)`, `MR` for
`Gamma(log)`, and the generic `exp(B)` otherwise. The standard
error follows the delta-method approximation
`SE_OR = OR × SE_log-odds` (the Stata `logit, or` convention).
The test statistic and the p-value are invariant under any
monotone transformation, so they remain on the link scale and
match the unexponentiated table verbatim:

```{r glm-or}
table_regression(fit, exponentiate = TRUE)
```

### Term-level partial chi-square

`partial_chi2` is the glm analog of `partial_f2`: for each model
term, the partial likelihood-ratio chi-square via
`drop1(test = "LRT")` (SAS PROC LOGISTIC `TYPE3`; Long & Freese
2014 §3.5). Rendered as `value (df)` so factor terms (`k − 1`
df) and numeric terms (1 df) read at a glance:

```{r glm-partial-chi2}
fit2 <- glm(smoking ~ age + education, data = sh, family = binomial)
table_regression(fit2, show_columns = c("b", "partial_chi2", "p"))
```

### Standardised coefficients: `refit` and the new `pseudo`

For `glm`, `standardized = "refit"` z-scores numeric *predictors*
only and refits the model — the response stays on its observed
scale because the link function is fixed. This is the
"x-standardization" convention (Long and Freese 2014 §4.3.4).
The other algebraic methods (`"posthoc"`, `"basic"`, `"smart"`)
apply X-only scaling using the same algebra as in the `lm` case.

`standardized = "pseudo"` (`glm` only) is the Menard (2004, 2011)
*fully* standardised coefficient, scaling by `SD(X) / SD(Y*)`
where `Y*` is the latent variable on the link scale and
`SD(Y*) = sqrt(var(linear-predictor) + var_link)` with
`var_link` = π²/3 for logit, 1 for probit, π²/6 for cloglog.
Defined for binomial families; non-binomial returns NA with a
`spicy_caveat`:

```{r glm-pseudo}
table_regression(fit, standardized = "pseudo")
```

### Average marginal effects: probability units, not log-odds

For `glm`, AME is computed as the average of `dE[Y|X] / dx` over
the observed sample — **on the response scale, not the link
scale**. For logistic regression this means the displayed AME is
in **probability units, not log-odds**: an AME of `-0.15` for
`education = Tertiary` reads "on average, holding sex and age
constant, a respondent with tertiary education is 15 percentage
points less likely to be a current smoker than one with lower
secondary education". This is almost always the quantity a
substantive reader wants from a logistic regression, and it is
the reason AME is increasingly recommended over odds ratios for
communicating logistic effects (Mood 2010; Long and Freese 2014
§5.3).

```{r glm-ame}
table_regression(fit, show_columns = c("b", "p", "ame", "ame_ci", "ame_p"))
```

Point estimate and inference are delegated to
[`marginaleffects::avg_slopes()`][marginaleffects::avg_slopes].
Under cluster-robust variance, the Satterthwaite-df handling
described in the *Robust variance* section applies to `glm` AME
as well.

### Profile-likelihood CIs: `ci_method = "profile"`

The default is `ci_method = "wald"` (symmetric
`estimate ± z × SE`, matching `summary.glm`). For `glm` you can
also request `ci_method = "profile"`, which calls
`MASS::confint.glm()` to compute the profile-likelihood CI —
asymmetric, exact for likelihood-based inference, and the gold
standard under sparse data or near-boundary estimates (Venables
& Ripley *MASS* §7.2). The estimate, SE, statistic, and p-value
all remain Wald; `"profile"` only refines the CI bounds:

```{r glm-profile}
table_regression(fit, ci_method = "profile")
```

### Hierarchical glm (LRT)

`nested = TRUE` on a list of nested glm fits adds a "Model
comparison" block whose default tokens are `c("LRT", "p")` — the
APA-conventional layout for hierarchical logistic regression
(Hosmer & Lemeshow §3.5; Long & Freese 2014 §3.6). The
chi-square statistic comes from `anova(test = "LRT")`.
`r2_change` and `F` are not defined for glm and return em-dashes
if explicitly requested:

```{r glm-nested}
m1 <- glm(smoking ~ sex,                   data = sh, family = binomial)
m2 <- glm(smoking ~ sex + age,             data = sh, family = binomial)
m3 <- glm(smoking ~ sex + age + education, data = sh, family = binomial)
table_regression(list(m1, m2, m3), nested = TRUE)
```

### Gaussian glm caveat

A `glm()` with `family = gaussian` and `link = "identity"` is
mathematically equivalent to `lm()` but lacks the variance-
explained effect-size family (`partial_f2 / η² / ω²`) and the
Satterthwaite-corrected AME path. Following the *transparency
over rejection* rule, spicy accepts the fit and emits a
`spicy_caveat` suggesting a refit with `lm()`.

## Mixed-effects models

`table_regression()` supports four mixed-effects classes:

* `lme4::lmer()` and `lme4::glmer()`
* `glmmTMB::glmmTMB()`
* `nlme::lme()`

The fixed-effects section of the table follows the same
conventions as `lm` / `glm`. Below the fit-statistics footer
(`n`, `AIC`, `BIC`, ...), a **Random effects panel** reports
the variance components, the model's intra-class correlation
(ICC) when defined, and the sample size at each grouping level:

```{r mixed-basic, eval = requireNamespace("lme4", quietly = TRUE) && requireNamespace("merDeriv", quietly = TRUE)}
library(lme4)
fit <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
table_regression(fit)
```

The panel header carries the estimator label — `(REML)` for
restricted maximum likelihood (the `lmer` / `lme` default) or
`(ML)` for full ML (`glmer`, `glmmTMB` default, and
`lmer(..., REML = FALSE)`). The label is informational: it tells
the reader which likelihood the variance estimates are anchored
to, without implying that REML and ML estimates are
interchangeable for inference.

### Fit statistics: Nakagawa marginal and conditional R²

Classical R² is not defined for mixed-effects models: the
within-group correlation breaks the variance-decomposition
identity that lm's R² relies on. The publication-standard
substitute is the **Nakagawa & Schielzeth** family of R²
(Nakagawa & Schielzeth 2013; Nakagawa, Johnson & Schielzeth
2017), which decomposes the response variance into

* `R² (marginal)` — variance explained by the **fixed effects
  alone**;
* `R² (conditional)` — variance explained by **fixed plus
  random effects**.

For Gaussian fits the formula is closed-form:

\[ R^2_{\text{marg}} = \frac{\sigma^2_f}{\sigma^2_f + \sum_g \sigma^2_g + \sigma^2_\epsilon}, \quad R^2_{\text{cond}} = \frac{\sigma^2_f + \sum_g \sigma^2_g}{\sigma^2_f + \sum_g \sigma^2_g + \sigma^2_\epsilon} \]

For non-Gaussian families (binomial logit / probit / cloglog,
Poisson log, etc.) the residual term σ²_ε is replaced by the
**link-scale distribution-specific variance** σ²_d
(π²/3 for logit, 1 for probit, π²/6 for cloglog, a lognormal
approximation for Poisson log; Nakagawa et al. 2017 §3).

spicy ships its own native implementation for the three
combinations covered by the closed-form formula in Nakagawa &
Schielzeth (2013) + Nakagawa et al. (2017) sec. 3:

* Gaussian (identity link) — all four engines;
* Bernoulli binomial with logit / probit / cloglog link
  (single trial per observation);
* Poisson with log link (lognormal approximation, Nakagawa et
  al. 2017 sec. 3.6).

These values are cross-validated against
`performance::r2_nakagawa()` to 1 × 10⁻¹⁰ on every supported
combination.

For families with more involved link-scale variance formulas —
binomial with multiple trials (`cbind(succ, fail)` /
`weights`), negative binomial, beta, Tweedie, dispersion
models, zero-inflated GLMM — spicy delegates to
`performance::r2_nakagawa()`. `performance` is a Suggests
dependency: when absent the fit-stat rows render as NA without
erroring (the table still prints; only the R² rows go blank).
The default `show_fit_stats` for mixed fits is
`c("nobs", "r2_marginal", "r2_conditional", "AIC", "BIC")` —
set `show_fit_stats` explicitly to override.

### Display scale: σ vs σ²

By default the panel reports the random-effect **standard
deviation** σ rather than the variance σ². Following Gelman
(2005, p. 5), the SD scale is "*directly interpretable as the
size of the variation across groups*" and lives on the same
scale as the response variable, which is what readers usually
want to compare to fixed-effect coefficients. Set
`re_scale = "variance"` to switch back to σ²:

```{r mixed-scale, eval = FALSE}
table_regression(fit, re_scale = "variance")
```

The conversion is internally consistent: SE and CI are
transported between scales by the Delta method
(`SE(σ) = SE(σ²) / (2σ)`, `CI(σ) = √CI(σ²)`), so the panel
values agree with `arm::sigma.hat()` (SD) and with
`VarCorr(fit)` (variance) up to rounding.

### Standard errors and confidence intervals

Each variance row of the panel carries a Wald SE and 95 % CI
(`est ± z · SE`, clamped at 0 for variances). The source
depends on the class:

* `lmer` / `glmer`: Hessian-based SE via `merDeriv` (Wang and
  Merkle 2018). Requires the optional `merDeriv` package
  (declared in `Suggests`); the column reads NA gracefully
  when the package is missing or when the fit is singular
  (`lme4::isSingular(fit) == TRUE`).
* `glmmTMB`: native Wald CI from `confint(fit, method = "Wald")`,
  with SE backed out by Delta method on σ.
* `nlme::lme`: native Wald CI from `nlme::intervals(fit)`.

For **correlations** between random terms (ρ rows in the
panel), only the point estimate is shown — SE / CI for
correlations need a full Delta method on the Cholesky
factorisation, deferred to a future release.

A methodological note. Wald CIs on variance components are
known to be optimistic near the boundary σ = 0 (Self and
Liang 1987 chi-bar-squared mixture; Bolker FAQ 2024). When
robustness is critical, the recommended fallback is the
profile-likelihood CI directly on the fit:

```{r mixed-profile, eval = FALSE}
confint(fit, method = "profile", parm = "theta_")  # lmer
confint(fit, method = "profile")                   # glmmTMB
nlme::intervals(fit)                                # lme (Wald-equivalent)
```

Profile CIs are not exposed as a `table_regression()` option
because they are 10–100× slower than Wald on random-slope
models (≈ 45 s on `sleepstudy` random slope vs ≈ 100 ms for
Wald), which is incompatible with the function's "fast enough
for a knit" goal. The panel reports the SE so the reader can
form their own boundary-safe judgment; the `re_columns`
argument (below) lets you drop SE and CI from the rendered
output entirely if your reporting standard prefers
**estimation without inference** on variance components
(Gelman 2005; Bates et al. 2015).

p-values are not reported for variance components. The null
σ² = 0 sits on the boundary of the parameter space, so the
standard Wald χ²₁ statistic has a chi-bar-squared limiting
distribution (Self and Liang 1987; Stram and Lee 1994). LRT-
based tests on a chi-bar-squared scale (`lmerTest::ranova()`)
remain the recommended path when the question is genuinely
"do we need this random term?". Note that `lmerTest`'s
Satterthwaite p-values apply to the **fixed effects** of an
`lmer` fit (and `table_regression()` will pick them up
automatically when the fit object inherits from `lmerModLmerTest`);
they do not apply to variance components.

### Fixed-effect p-values and CIs across classes

The footer carries a one-line annotation naming the inference
method used for the fixed-effect p-value and CI columns. The
mapping is:

* `lmerModLmerTest` (= `lmerTest::lmer()`, or `lme4::lmer()`
  after `library(lmerTest)`): Satterthwaite t-test. p and CI use
  `summary(fit)$coefficients[, c("df", "Pr(>|t|)")]` from
  lmerTest. This is the recommended default whenever the
  number of groups is below ~30 (Luke 2017).
* `lmerMod` without lmerTest loaded: Wald-z, large-sample
  approximation. df is reported as ∞ in the broom-canonical
  frame; p-values come from `pnorm`. This matches
  `parameters::model_parameters(ci_method = "wald")`, SAS
  PROC MIXED with `ddfm = z`, and Stata `xtmixed`'s default.
  The footer suggests loading lmerTest for small samples.
* `glmerMod`: Wald-z asymptotic from `summary(fit)$coefficients[,
  "Pr(>|z|)"]`. No t-distribution applies to GLMMs in general
  (the link-scale residual variance is fixed for binomial /
  Poisson families).
* `glmmTMB`: Wald-z asymptotic from `summary(fit)$coefficients[,
  "Pr(>|z|)"]`. Same logic as `glmerMod`.
* `lme` (nlme): t-test with **containment degrees of freedom**.
  p and df come from `summary(fit)$tTable`. Pinheiro & Bates
  (2000) §2.4.2 — between-group covariates get df = n_groups −
  rank, within-group covariates get the residual df.

The footer line lets the reader see at a glance what's being
reported. The behavioural choice is stable across classes:
spicy uses each engine's own inference path, falling back to
Wald-z only when no better df strategy is available.

### glmer (logistic / Poisson) and the adjusted ICC

`glmer` fits render the same panel as `lmer` — σ rows with Wald
SE and 95 % CI from `merDeriv`, an ICC row, and the N (group)
row. Two things differ from the lmer case:

* No "σ (Residual)" row. `lme4::sigma()` returns the fixed
  dispersion parameter (1 for `binomial` and `poisson`) by
  convention, but this is not an estimated parameter with its
  own SE — reporting it as a residual variance is misleading.
  spicy suppresses the row for non-Gaussian fits.
* The ICC uses the **link-scale distribution variance** in the
  denominator (Nakagawa, Johnson & Schielzeth 2017 — adjusted
  ICC):

  | Family · link    | distribution variance σ²_d |
  |------------------|----------------------------|
  | binomial logit   | π²/3 ≈ 3.29                |
  | binomial probit  | 1                          |
  | binomial cloglog | π²/6 ≈ 1.64                |
  | poisson  log     | log(1 + 1/λ), λ = exp(β₀_null + 0.5 σ²_g_null) |

  with `ICC = σ²_g / (σ²_g + σ²_d)`. The value matches
  `performance::icc(fit)$ICC_adjusted` to 1 × 10⁻⁶.

For binomial fits with the `cbind(succ, fail)` (multi-trial)
form, the closed-form σ²_d formula above does not apply directly
— spicy returns `NA` for the ICC row rather than emit a
misleading value. The Nakagawa marginal / conditional R² rows
still render via the `performance` fallback.

### Hierarchical / nested mixed-effects

`table_regression(list(m1, m2, m3), nested = TRUE)` also works
on mixed-effects fits. The change rows are populated from
`anova(m1, m2)` (likelihood-ratio test):

* `ΔAIC` — `AIC(m2) − AIC(m1)`, negative when m2 fits better
* `ΔBIC` — `BIC(m2) − BIC(m1)`
* `Δχ²` — likelihood-ratio statistic
* `p (change)` — LRT p-value

```{r mixed-nested, eval = requireNamespace("lme4", quietly = TRUE)}
m1 <- lme4::lmer(Reaction ~ 1     + (1 | Subject), data = lme4::sleepstudy, REML = FALSE)
m2 <- lme4::lmer(Reaction ~ Days  + (1 | Subject), data = lme4::sleepstudy, REML = FALSE)
m3 <- lme4::lmer(Reaction ~ Days  + (Days | Subject), data = lme4::sleepstudy, REML = FALSE)
table_regression(list(m1, m2, m3), nested = TRUE)
```

Variance-explained change tokens (`Δr²`, `Δf²`) are NA — the
F-test framework that grounds them doesn't apply to mixed
models. `ΔR² (marginal)` and `ΔR² (conditional)` are not
reported either: the row-by-row difference would be conceptually
mixing two quantities (a marginal and a conditional gain), so
the reader can compute it from the absolute R² rows directly if
needed.

Methodological caveat. `lme4::anova()` auto-refits REML fits
with ML before the LRT (spicy suppresses the message). The LRT
is a **fixed-effect-only** comparison: testing additional random
terms with naive χ² is conservative — the formally correct test
is the chi-bar-squared mixture (Self & Liang 1987), best run
directly via `lmerTest::ranova()`. spicy surfaces fixed-effect
nesting comparisons; random-structure model selection lives
outside the table footer.

### Average marginal effects (AME)

The `"ame"` / `"ame_se"` / `"ame_ci"` / `"ame_p"` tokens of
`show_columns` are wired for mixed-effects fits, delegated to
`marginaleffects::avg_slopes(fit, df = Inf)`. The AME column is
the publication-substantive quantity when the link is non-linear:

* `glmer` binomial logit: AME is on the **probability scale**
  (percentage points), not the log-odds scale. For the typical
  "what does a one-unit change in `x` do to the predicted
  probability?" reading, the AME column is what you want.
* `glmer` Poisson log: AME is on the **count scale** (units of
  the outcome), not log-rate.
* `lmer` / `lme` / `glmmTMB` Gaussian: AME = B coefficient
  (identity link); the column is filled but redundant.

```{r mixed-ame, eval = requireNamespace("lme4", quietly = TRUE) && requireNamespace("marginaleffects", quietly = TRUE)}
set.seed(1)
n <- 500
g <- factor(rep(1:25, length.out = n))
x <- rnorm(n)
cat <- factor(sample(c("A", "B", "C"), n, replace = TRUE))
y <- rbinom(n, 1, plogis(0.5 + 0.8 * x + (cat == "B") * 0.3 +
                           (cat == "C") * -0.5 + rnorm(25)[g]))
fit <- lme4::glmer(y ~ x + cat + (1 | g), family = binomial)
table_regression(fit, show_columns = c("b", "se", "p", "ame", "ame_p"))
```

Factor predictors are handled level-by-level: AME rows align
under the same factor header (`cat: A (ref.) / B / C`), each
level sharing a row with its B coefficient. Inline `factor()`
calls and ordered factors are reconstructed from the
model-frame column when the coef-name lookup misses (same
fallback logic used for `lm` / `glm`).

Inference is Wald-z asymptotic (`df = Inf`) with the model-based
vcov -- matching the inference path of the B-row p-values in the
same table. `marginaleffects` is a Suggests dependency; when
unavailable, the AME columns appear in the header but render as
NA without erroring (the rest of the table is unaffected).

### Hiding parts of the panel

Three arguments control panel rendering:

* `show_re = FALSE` suppresses the panel entirely. Useful when
  the random structure is documented in the manuscript text
  and the table only needs the fixed-effects block.
* `re_scale = "variance"` switches to σ² (see *Display scale*
  above).
* `re_columns` is a character subset of `c("est", "se", "ci")`.
  `"est"` is mandatory. Drop SE and CI when reporting a
  minimal table (`re_columns = "est"`) or drop only CI when
  following a journal style that uses parenthesised SE
  alone (`re_columns = c("est", "se")`).

```{r mixed-cols, eval = FALSE}
table_regression(fit, re_columns = "est")               # σ only
table_regression(fit, re_columns = c("est", "se"))      # σ (SE)
table_regression(fit, show_re = FALSE)                  # fixed effects only
```

## Significance stars

Stars are off by default. APA 7 §6.46 explicitly discourages
asterisks-only reporting in favour of exact p-values, and ASA's
post-2019 guidance (Wasserstein, Schirm, and Lazar 2019)
reinforces the same point. Set `stars = TRUE` for the APA preset
(`*** p < .001, ** p < .01, * p < .05`) or pass a named numeric
vector for custom thresholds:

```{r stars}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
table_regression(fit, stars = TRUE)
```

Stars suffix the `B` column (or `β` when standardisation is
requested); the threshold mapping is auto-documented in the
footer. The `p` column itself remains unstarred so the numeric
value stays readable.

## Display knobs

`labels` accepts a named character vector keyed by either formula
term labels (`"sex"`) or coefficient names (`"sexMale"`), so
individual contrast rows can be relabelled. `intercept_position`
switches the intercept to the bottom (Stata convention).
`reference_style = "annotation"` lifts the reference level into
the factor header (`"sex: [ref: Female]"`) and drops the explicit
reference row, for compact tables. `decimal_mark = ","` switches
to European decimal notation and automatically changes the CI
separator to `";"` to avoid ambiguity:

```{r polish}
fit <- lm(wellbeing_score ~ age + sex + education, data = sochealth)
table_regression(
  fit,
  labels = c(
    age            = "Age (years)",
    sex            = "Sex",
    education      = "Education"
  ),
  reference_style = "annotation",
  intercept_position = "last",
  decimal_mark = ","
)
```

## Output formats

The default print method draws an ASCII table to the console.
The same content is available as a plain `data.frame`
(`output = "data.frame"`), a long broom-style data.frame
(`output = "long"`), or as a rich-format table for `gt`,
`tinytable`, `flextable`, `excel`, `word`, or `clipboard`:

```{r data-frame}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "data.frame"
)
str(out)
```

```{r long}
out <- table_regression(
  lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
  output = "long"
)
out[, c("model_id", "term", "estimate_type", "estimate",
        "std.error", "p.value")]
```

```{r gt-output, eval = build_rich_tables}
pkgdown_dark_gt(
  table_regression(
    lm(wellbeing_score ~ age + sex + smoking, data = sochealth),
    output = "gt"
  )
)
```

## broom integration

`broom::tidy()` returns a long tibble with one row per
`(model_id, term, estimate_type)` and broom-canonical column
names (`estimate`, `std.error`, `conf.low`, `conf.high`,
`statistic`, `p.value`). `broom::glance()` returns one row per
`(model_id, outcome)` with model-level statistics; `df.residual`
is preserved as a numeric so cluster-robust Satterthwaite df
flows through verbatim:

```{r broom}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
out <- table_regression(fit, standardized = "refit")

broom::tidy(out)
broom::glance(out)
```

The long format is the right entry point when the table is one
step in a larger pipeline — saving to disk for the manuscript
appendix, faceting by subgroup, or feeding a downstream
post-estimation analysis. The `tidy()` output keeps the
`estimate_type` column so the same data frame can hold rows for
B, β, AME, and per-coefficient effect-size estimates without
ambiguity.

## See also

- `vignette("table-categorical", package = "spicy")` for the APA
  Table 1 categorical descriptors (factors, labelled variables,
  chi-squared tests, association measures).
- `vignette("table-continuous", package = "spicy")` for the APA
  Table 1 / Table 2 continuous descriptors and unadjusted
  group-comparison tests.
- `vignette("table-continuous-lm", package = "spicy")` for the
  one-predictor-by-many-outcomes linear-model counterpart
  (estimated marginal means, contrast or slope, four effect-size
  families with noncentral CIs).
- `vignette("summary-tables-reporting", package = "spicy")` for an
  end-to-end reporting workflow that combines the four spicy
  summary-table helpers along the APA Table 1 / 2 / 3 sequence.

## References

Aiken, L. S., and West, S. G. (1991).
*Multiple Regression: Testing and Interpreting Interactions*.
Sage.

American Psychological Association (2020).
*Publication Manual of the American Psychological Association*
(7th ed.). Section 6.46.

Bates, D., Mächler, M., Bolker, B., and Walker, S. (2015).
Fitting linear mixed-effects models using lme4.
*Journal of Statistical Software*, 67(1), 1–48.

Bolker, B. (2024). GLMM FAQ. <https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html>.
Random-effects p-values and CI section.

Cohen, J., Cohen, P., West, S. G., and Aiken, L. S. (2003).
*Applied Multiple Regression / Correlation Analysis for the
Behavioral Sciences* (3rd ed.). Lawrence Erlbaum.

Gelman, A. (2005). Analysis of variance — why it is more
important than ever. *The Annals of Statistics*, 33(1), 1–53.

Gelman, A. (2008). Scaling regression inputs by dividing by two
standard deviations. *Statistics in Medicine*, 27(15), 2865–2873.

Gelman, A., Hill, J., and Yajima, M. (2012). Why we (usually)
don't have to worry about multiple comparisons. *Journal of
Research on Educational Effectiveness*, 5(2), 189–211.

Greenland, S. (2017). For and against methodologies: Some
perspectives on recent causal and statistical inference debates.
*European Journal of Epidemiology*, 32(1), 3–20.

Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd
ed.). Springer. Section 5.4 on multiplicity.

Hosmer, D. W., Lemeshow, S., and Sturdivant, R. X. (2013).
*Applied Logistic Regression* (3rd ed.). Wiley.

Imbens, G. W., and Kolesár, M. (2016). Robust standard errors in
small samples: Some practical advice. *Review of Economics and
Statistics*, 98(4), 701–712.

Long, J. S., and Ervin, L. H. (2000). Using heteroscedasticity
consistent standard errors in the linear regression model.
*The American Statistician*, 54(3), 217–224.

Long, J. S., and Freese, J. (2014). *Regression Models for
Categorical Dependent Variables Using Stata* (3rd ed.). Stata
Press.

Luke, S. G. (2017). Evaluating significance in linear mixed-
effects models in R. *Behavior Research Methods*, 49(4), 1494–
1502. (Empirical comparison of Wald-z, Wald-t, Kenward-Roger,
and Satterthwaite; recommends Satterthwaite or KR for samples
with few groups.)

McFadden, D. (1974). Conditional logit analysis of qualitative
choice behavior. In P. Zarembka (Ed.), *Frontiers in
Econometrics* (pp. 105–142). Academic Press.

Menard, S. (2004). Six approaches to calculating standardized
logistic regression coefficients. *The American Statistician*,
58(3), 218–223.

Menard, S. (2011). Standards for standardized logistic
regression coefficients. *Social Forces*, 89(4), 1409–1428.

Nakagawa, S., Johnson, P. C. D., and Schielzeth, H. (2017).
The coefficient of determination R² and intra-class correlation
coefficient from generalized linear mixed-effects models
revisited and expanded. *Journal of the Royal Society Interface*,
14(134), 20170213.

Nakagawa, S., and Schielzeth, H. (2013). A general and simple
method for obtaining R² from generalized linear mixed-effects
models. *Methods in Ecology and Evolution*, 4(2), 133–142.

Mood, C. (2010). Logistic regression: Why we cannot do what we
think we can do, and what we can do about it. *European
Sociological Review*, 26(1), 67–82.

Nagelkerke, N. J. D. (1991). A note on a general definition of
the coefficient of determination. *Biometrika*, 78(3), 691–692.

Olejnik, S., and Algina, J. (2003). Generalized eta and omega
squared statistics: Measures of effect size for some common
research designs. *Psychological Methods*, 8(4), 434–447.

Pinheiro, J. C., and Bates, D. M. (2000). *Mixed-Effects Models
in S and S-PLUS*. Springer. §2.4.2 on containment degrees of
freedom for the nlme implementation.

Pustejovsky, J. E., and Tipton, E. (2018). Small-sample methods
for cluster-robust variance estimation and hypothesis testing in
fixed effects models. *Journal of Business & Economic
Statistics*, 36(4), 672–683.

Rothman, K. J. (1990). No adjustments are needed for multiple
comparisons. *Epidemiology*, 1(1), 43–46.

Self, S. G., and Liang, K.-Y. (1987). Asymptotic properties of
maximum likelihood estimators and likelihood ratio tests under
non-standard conditions. *Journal of the American Statistical
Association*, 82(398), 605–610.

Smithson, M. (2003). *Confidence Intervals*. Sage.

Steiger, J. H. (2004). Beyond the F test: Effect size confidence
intervals and tests of close fit in the analysis of variance and
contrast analysis. *Psychological Methods*, 9(2), 164–182.

Stram, D. O., and Lee, J. W. (1994). Variance components testing
in the longitudinal mixed effects model. *Biometrics*, 50(4),
1171–1177.

Tjur, T. (2009). Coefficients of determination in logistic
regression models — A new proposal: The coefficient of
discrimination. *The American Statistician*, 63(4), 366–372.

Venables, W. N., and Ripley, B. D. (2002). *Modern Applied
Statistics with S* (4th ed.). Springer. Section 7.2 on profile
likelihood.

Wang, T., and Merkle, E. C. (2018). merDeriv: Derivative
computations for linear mixed effects models with application
to robust standard errors. *Journal of Statistical Software*,
87(Code Snippet 1), 1–16.

Wasserstein, R. L., Schirm, A. L., and Lazar, N. A. (2019).
Moving to a world beyond "p < 0.05". *The American Statistician*,
73(sup1), 1–19.
