---
title: "Explore variables and build codebooks in R"
description: >
  Explore variables, inspect labels, and build interactive codebooks in R
  with spicy. Learn how to use varlist(), vl(), code_book(), and
  label_from_names() for survey and labelled datasets.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore variables and build codebooks in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spicy)
```

Before you build frequency tables or cross-tabulations, it is often worth
checking how your variables are named, labelled, and coded.

spicy provides a simple workflow for variable exploration and
documentation in R. You can derive labels from imported column names,
inspect variables with `varlist()` or `vl()`, and build an interactive
codebook with `code_book()`.

This vignette focuses on three common tasks:

- clean imported column names and recover variable labels with
  `label_from_names()`
- inspect variables, labels, values, classes, and missing data with
  `varlist()` and `vl()`
- generate an interactive codebook for review or export with
  `code_book()`

These tools are especially useful for survey datasets, labelled data,
and imported files where variable names and labels need to be checked
before analysis.

## Why inspect variables before analysis?

Variable inspection helps catch common problems early: unclear names,
missing labels, unexpected coding, and variables with many missing
values. A quick review of your dataset also makes it easier to choose
which variables to tabulate, summarize, or report later.

## Recover labels from imported column names

Some imported files store both a variable name and a variable label in
the column header. `label_from_names()` splits names of the form
`name<sep>label`, renames the columns, and stores the label as a proper
variable label.

```{r label-from-names}
df <- tibble::tibble(
  "age. Age of respondent" = c(25, 30, 41),
  "edu. Highest education level" = c("Lower", "Upper", "Tertiary"),
  "smoke. Current smoker" = c("No", "Yes", "No")
)

out <- label_from_names(df)
labelled::var_label(out)
```

This is especially useful for LimeSurvey-style exports and similar files
where labels arrive in the column names instead of the variable
metadata.

## Inspect variables with varlist()

`varlist()` gives a compact summary of each variable, including its
name, label, representative values, class, number of distinct values,
number of valid observations, and missing values.

```{r varlist-all}
varlist(sochealth, tbl = TRUE)
```

For a focused inspection, select only the variables you want to review:

```{r varlist-selected}
varlist(sochealth, smoking, education, income_group, tbl = TRUE)
```

This is often enough to confirm that labels, factor levels, and missing
values look correct before moving on to tabulations.

## Select subsets of variables

`varlist()` supports tidyselect, which makes it easy to inspect a subset
of variables by name pattern or type.

```{r varlist-tidyselect}
varlist(sochealth, starts_with("life_sat"), tbl = TRUE)
```

```{r varlist-numeric}
varlist(sochealth, where(is.numeric), tbl = TRUE)
```

If you prefer a shorter call in interactive work, `vl()` is a shorthand
for `varlist()`:

```{r vl-example}
vl(sochealth, starts_with("bmi"), tbl = TRUE)
```

## Build an interactive codebook

When you want a searchable and exportable overview of the whole dataset,
`code_book()` builds an interactive codebook in the Viewer.

```{r code-book-basic}
if (requireNamespace("DT", quietly = TRUE)) {
  code_book(sochealth)
}
```

You can also request a fuller display of values or include missing
values explicitly in the summary:

```{r code-book-values}
if (requireNamespace("DT", quietly = TRUE)) {
  code_book(sochealth, values = TRUE, include_na = TRUE)
}
```

This is useful when reviewing a dataset with collaborators or preparing
documentation before analysis.

## When to use varlist() and code_book()

Use `varlist()` when you want a quick summary in a script or a tibble
you can inspect directly.

Use `vl()` when you want the same summary with a shorter call in
interactive work.

Use `code_book()` when you want a searchable, interactive codebook for
review or export.
