---
title: "The structured view: programmatic access to spicy tables"
description: >
  `as_structured()` exposes the typed representation behind every
  `table_regression()`, `table_categorical()`, `table_continuous()`,
  `table_continuous_lm()`, and `table_outcome()` output: a numeric body with per-row
  identity columns, per-column metadata, per-cell status, and the
  format specification. Filter rows programmatically, aggregate raw
  values, or build a custom renderer -- against one schema for all
  five families.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

`table_regression()` returns a *display* representation: a character
table with `(ref.)` annotations, en-dashes, bracketed `[LL, UL]`
intervals, and APA-formatted p-values. That is the right thing to
print and the right thing to publish -- and the wrong thing to
compute on. The companion accessor `as_structured()` returns the
*typed* view underneath: a fully numeric body, confidence intervals
pre-split into their own columns, `NA` where a cell is structurally
empty, plus the metadata a renderer needs to reconstruct the
display. For regression tables it is the very contract spicy's own
output engines (Excel, gt, tinytable, flextable, clipboard) consume
internally -- nothing about it is second-class -- and the
descriptive families (`table_categorical()`, `table_continuous()`,
`table_continuous_lm()`, `table_outcome()`) expose the same schema, covered at the end
of this vignette. The tour below uses a regression table; everything
it shows reads identically on the other four.

Two sibling tools serve different needs, and choosing well saves
work. `broom::tidy()` returns the *statistical* long form -- one row
per term and estimate type per model, with `estimate`, `std.error`,
`conf.low`, `p.value`. A default table carries a single estimate
type, so that is one row per coefficient; requesting standardized
betas or AMEs adds a row per term for each. It is the right input
for meta-analysis, further computation on estimates, or ggplot. `as_structured()` returns the *display
geometry* -- rows and columns exactly as the rendered table lays
them out, factor headers and fit statistics included -- and is the
right input for filtering what a reader will see, post-processing
numbers in table shape, or rendering the table with an engine spicy
does not ship. This vignette covers the second tool. The shared
mechanics of `table_regression()` are in
[*Publication-ready regression
tables*](table-regression.html); the class-by-class map is
[*Supported models*](table-regression-supported-models.html).

## The schema on one model

Start from the display view, so the typed view has something to be
compared against:

```{r schema}
fit <- lm(wellbeing_score ~ age + sex + smoking, data = sochealth)
tbl <- table_regression(fit)
tbl
```

Every display trait named above is visible here: `(ref.)` markers
and en-dashes on the reference rows, blank factor header rows, one
bracketed `[LL, UL]` column, APA p-values (`<.001`, no leading
zero). The typed view underneath:

```{r components}
s <- as_structured(tbl)
names(s)
```

The centrepiece is `body`: a data frame with the `Variable` column,
one numeric column per displayed statistic, and four dot-prefixed
identity columns at the end (the next section). The confidence
interval that prints as one bracketed column arrives split into
`LL` / `UL`:

```{r body}
s$body
```

Read it against the printed table and the structure becomes
obvious: factor header rows (`sex:`, `smoking:`) and reference
rows carry `NA` across the statistics -- structurally empty, not
missing data -- and the fit-statistics rows (`n`, `R²`, ...) put
their value in the first numeric column. What each row *is* travels
with the row itself, in four dot-prefixed identity columns at the
end of `body` -- never in positions you would have to track:

```{r indices}
s$body[, c("Variable", ".variable", ".level", ".row_role", ".indent")]
```

`.variable` is the source variable (or, on a fit-statistics row,
the statistic's token), `.level` the factor level, `.row_role` one
of `"coef"`, `"factor_header"`, `"level"`, `"reference"`,
`"fit_stat"`, `"outcome"`, `"vc"`, and `.indent` the display indent
depth. A row is addressable by `(.variable, .level)` rather than by
its position -- and because identity is data, not indices, it
survives whatever you do to the body: subset it, reorder it, stack
two tables with `rbind()`, and every row still says what it is.

```{r identity}
s$body[which(s$body$.variable == "smoking"), c("Variable", ".level", "B")]
which(s$body$.row_role == "reference")
```

`col_meta` describes each column: which token produced it, which
model it belongs to, its display precision, and -- for p-value
columns -- the APA style and below-threshold rule, plus per-row
`fit_stat_overrides` where a fit-statistic row departs from the
column's precision (put to work in the renderer below).
`format_spec` carries the global defaults (decimal mark, digits,
CI level):

```{r meta}
s$col_meta$B
s$format_spec$ci_level
```

That is five of the nine components. `spanners` appears in the
multi-model section below; `stars`, `cell_status` and
`col_meta$display_cells` in the section after it. The last two are
`ci_pairs`, which records which `LL`/`UL` columns pair up under one
interval label (here, columns 4 and 5 under `95% CI`), and
`outcome_labels_by_col`, which labels the outcome header row of
multi-outcome tables (empty here). `version` names the contract the
object carries -- see *The contract* below. The complete schema,
component by component, is in `?as_structured`.

## Filtering and aggregating

Because the body is numeric, ordinary data-frame idioms work. Keep
the coefficient rows a reader would call significant:

```{r filter}
s$body[which(s$body$p < 0.05), c("Variable", "B", "p")]
```

Note the `which()`: the structurally empty rows are `NA`, and
`which()` drops them where a bare logical index would keep them as
`NA` rows. To restrict attention to coefficient rows before any
filter, select on the role:

```{r coefrows}
coef_rows <- which(s$body$.row_role %in% c("coef", "level"))
s$body[coef_rows, c("Variable", "B", "95% CI: LL", "95% CI: UL")]
```

Aggregation is just as direct -- the mean absolute coefficient, the
range of CI widths:

```{r aggregate}
mean(abs(s$body$B[coef_rows]))
range(s$body[coef_rows, "95% CI: UL"] - s$body[coef_rows, "95% CI: LL"])
```

## Multi-model tables: spanners and per-model columns

With several models, columns are prefixed by the model label (and
the interval columns compact away by default, which is why each
model shows `B` / `SE` / `p`). Two more components become useful.
`spanners` maps each model label to its column indices in `body`;
`cell_status` says, cell by cell, which ones carry the reference
en-dash -- only the models that actually contain the factor do. To
see it discriminate, the `Minimal` model below omits `smoking`:

```{r multi}
fit2 <- lm(wellbeing_score ~ age + sex + smoking + bmi, data = sochealth)
fit0 <- lm(wellbeing_score ~ age + sex, data = sochealth)
s2 <- as_structured(
  table_regression(list(Minimal = fit0, Base = fit, Extended = fit2))
)
s2$spanners
names(s2$body)
ref <- which(s2$body$.row_role == "reference")
s2$body[ref, c("Variable", ".variable", ".level")]
vapply(s2$cell_status, function(x) x[ref[2]], character(1))
```

Rows 4 and 7 are the two reference rows (`Female (ref.)`,
`No (ref.)`). All three models contain `sex`, so every column of
row 4 is marked `"reference"`; only `Base` and `Extended` contain
`smoking`, so on row 7 the `Minimal` columns are unmarked -- a
renderer draws the reference en-dash where the status says
`"reference"` and leaves the rest blank.

Model-wise extraction follows from the spanner map:

```{r modelwise}
s2$body[, c(1, s2$spanners$Extended)]
```

## Cells a number cannot express

Two things a reader sees are not values of a statistic, and the
typed body alone cannot express them: a cell built from two counts,
and a marker attached to a cell. Both travel in the contract, so a
renderer never has to reconstruct them from the printed table.

The event counts requested with `show_columns = "n_events"` print
as `events/N`. The body keeps the numerator as a number, and the
column's `display_cells` carries the string of each cell -- `NA`
where the number formats normally, as on the fit-statistics rows in
that same column:

```{r composite}
counts <- table_regression(
  glm(dentist_12m ~ age + sex, data = sochealth, family = binomial()),
  show_columns = c("n_events", "b", "p")
)
sc <- as_structured(counts)
sc$body[["Events/N"]]
sc$col_meta[["Events/N"]]$display_cells
```

Read the two together on the reference row: the level has 437
events out of 620, and it keeps them -- the en-dash of a reference
row means "no estimate by design", which says nothing about a
count. A renderer that ignores `display_cells` prints a numerator
under a header promising a ratio.

Significance stars work the same way. `stars = TRUE` leaves the
body untouched and fills the `stars` component: the thresholds the
footer legend documents, and the marker of each cell that takes
one.

```{r stars}
starred <- table_regression(
  lm(wellbeing_score ~ age + sex, data = sochealth),
  stars = TRUE
)
ss <- as_structured(starred)
ss$stars$thresholds
ss$stars$markers$B
```

The rule for which column takes the markers is the console's: the
raw coefficient, the standardized coefficient only when `B` is not
displayed beside it, and the average marginal effect on its own
p-value. `stars` is `NULL` when the table has none.

A third case is a cell that has no number to show at all, and
`cell_status` distinguishes the two reasons. `"reference"` is the
reference level of a factor, in that estimate block and that model.
`"undefined"` is a statistic that applies to the row but is not
computable -- the standard error of a variance component, or one a
`re_columns` selection left out. Both display as an en-dash. A cell
with no status whose value is `NA` is simply absent, and displays
blank.

```{r status}
sc$cell_status$B
```

One more display convention lives in the row identity rather than
in a cell. A model with absorbed fixed effects discloses them as a
block: the `Fixed effects:` row takes `.row_role == "factor_header"`
and each absorbed factor a `"level"` row with `.indent` 1, named as
the model names it. Treat that block exactly like a factor group in
the coefficients and it renders the way it prints.

## The descriptive tables

Everything above extends beyond regression. The same accessor reads
`table_categorical()`, `table_continuous()`,
`table_continuous_lm()` and `table_outcome()`, and returns the same schema -- the same
identity columns, the same `cell_status`, the same `display_cells`
mechanism for composite cells. A consumer written for one family
reads the other four. `table_categorical()` lays out one factor
header row per variable and one row per category:

```{r desc-categorical}
ct <- table_categorical(sochealth, c(sex, smoking), by = education)
sc <- as_structured(ct)
sc$body[, c("Variable", ".variable", ".level", ".row_role")]
sc$spanners
```

Row identity is data here too: the `"(Missing)"` row carries
`.row_role == "missing"` rather than a label a consumer would have to
match, and the margin column is flagged in `col_meta` rather than
found by the string `"Total"`.

```{r desc-categorical-keys}
sc$body[sc$body$.row_role == "missing", c("Variable", ".variable", ".level")]
sc$col_meta[["Total n"]]$total
```

`table_continuous()` puts one row per variable, or one row per level
of `by` — the level lives in `.level`, with role `"group"`. The
`col_meta` tokens cover the `show_columns` vocabulary and add one per
group-comparison column: `"statistic"`, `"p"`, `"es"`, `"smd"`. Those
four are not `show_columns` values and cannot be, because every token
of that vocabulary is a statistic OF a group, while these compare two:

```{r desc-continuous}
cn <- table_continuous(sochealth, c(age, bmi), by = sex, effect_size = "auto")
scn <- as_structured(cn)
scn$body[, c("Variable", ".variable", ".level", ".row_role", "M", "SD")]
vapply(scn$col_meta, `[[`, character(1), "token")
```

Cells the console builds from more than one number keep their numeric
anchor in `body` and their printed string in `display_cells` — the
effect size and its interval, a compact `Med [Q1, Q3]`, a test gloss:

```{r desc-composite}
scn$col_meta$ES$display_cells
```

`table_continuous_lm()` reports one row per outcome, with the levels
of `by` as columns; each marginal-mean column says which level it
carries, so nothing has to parse `M (Male)`:

```{r desc-lm}
lm_tbl <- table_continuous_lm(sochealth, c(age, bmi), by = sex)
slm <- as_structured(lm_tbl)
slm$body[, c("Variable", ".variable", ".row_role")]
slm$col_meta[["M (Male)"]]
```

`table_outcome()` is the one shape where three roles share
`.indent == 0`: a `"summary"` row for the whole analytic sample, a
`"factor_header"` row per grouping, and the indented `"level"` rows
under it. A consumer that read "a level is anything that is not a
header" would swallow the marginal row; the `.indent` field is the
one to read, and it is why the block geometry of all three
descriptive families is computed from it:

```{r desc-outcome}
ot <- table_outcome(sochealth, bmi, select = c(sex, smoking))
so <- as_structured(ot)
so$body[, c("Variable", ".variable", ".level", ".row_role", ".indent")]
```

The statistics of the BLOCK sit on its header row and the statistics
of the OUTCOME on the level rows, so a cell that is blank on one of
them is an absence, not an undefined value -- `cell_status` stays
empty there, and is reserved for what the table itself marks:

```{r desc-outcome-blanks}
so$body[, c("Variable", "M", "p")]
```

## Building your own renderer

The structured view carries everything a renderer needs. A compact
`knitr::kable()` rendering takes a few lines: each column formats at
its `col_meta` precision, and the `fit_stat_overrides` seen earlier
supply the per-row exceptions -- that is what renders the `n` row as
`1175` rather than `1175.00`:

```{r kable}
render_kable <- function(tbl) {
  s <- as_structured(tbl)
  # Value columns only: the dot-prefixed identity columns describe the
  # rows, they are not cells of the table.
  value_cols <- setdiff(names(s$body), c("Variable", grep("^\\.", names(s$body), value = TRUE)))
  out <- s$body[, c("Variable", value_cols)]
  for (nm in value_cols) {
    meta <- s$col_meta[[nm]]
    prec <- meta$precision
    if (is.null(prec)) prec <- s$format_spec$digits
    row_prec <- rep(prec, nrow(out))
    for (ov in meta$fit_stat_overrides) {
      if (!is.null(ov$precision)) row_prec[ov$row] <- ov$precision
    }
    out[[nm]] <- vapply(
      seq_len(nrow(out)),
      function(i) {
        if (is.na(out[[nm]][i])) {
          ""
        } else {
          formatC(out[[nm]][i], format = "f", digits = row_prec[i])
        }
      },
      character(1)
    )
  }
  knitr::kable(out, align = c("l", rep("r", ncol(out) - 1)))
}
render_kable(tbl)
```

Anything more ambitious -- a `DT::datatable()` with significant
rows highlighted, a LaTeX booktabs layout, an HTML widget --
follows the same pattern: `body` for the cells, `col_meta` for
per-column formatting, `.row_role` / `.indent` for styling regions,
`cell_status` for the cells no number expresses, and
`format_spec` for the global conventions. The note and title stay
available on the original object (`attr(tbl, "note")`,
`attr(tbl, "title")`), so a custom renderer can reproduce spicy's
self-documenting footer verbatim.

## The contract

Three properties make the structured view safe to build on:

* **Engine parity.** For regression tables, the built-in engines
  consume this exact structure: what you read in `body` is, cell for
  cell, what the Excel workbook contains and what the gt table
  displays at its precision. For the descriptive families the typed
  view is built from the same computation frames the display is --
  value fidelity is pinned by tests, and the engines migrate onto
  the view as the contract settles.
* **Stability tier.** `as_structured()` sits in the *stabilising*
  tier of the API (see `?spicy`, section *API stability*). The rule
  the contract actually keeps in the `0.y` series is this: nothing
  ever changes *silently*. Components are added by default; when one
  must be removed or reshaped -- as version 3 did with the row-index
  vectors -- the removal is announced under *Breaking changes* in
  `NEWS.md` with its replacement named, and the version guard below
  turns any version mismatch into an error instead of a wrong table.
* **Version guard.** `version` names the contract an object
  carries; it is `3` here. A table built by a spicy that speaks a
  different contract -- older or newer -- is refused rather than
  mis-read, with the correspondence between the two named in the
  error. Objects from before the structured view existed are
  refused outright.


For the statistical long form -- estimates with standard errors and
unformatted p-values, one row per term and estimate type across
models -- reach for `broom::tidy()` on the same object; for one-line model summaries,
`broom::glance()`. The three views are built from the same frames,
so their numbers always agree.
