---
title: "Frequency tables and cross-tabulations"
description: >
  Create frequency tables and cross-tabulations with chi-squared tests,
  effect sizes, and weighted counts in R using the spicy package.
  Covers sorting, percentages, grouping, and labelled data.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spicy)
```

`freq()` and `cross_tab()` are the core tabulation functions in spicy.
They handle factors, labelled variables (from haven or labelled),
weights, and missing values out of the box. This vignette covers the
main options using the bundled `sochealth` dataset.

## Frequency tables with freq()

### Basic usage

Pass a data frame and a variable name to get counts and percentages:

```{r freq-basic}
freq(sochealth, education)
```

### Sorting

Sort by frequency with `sort = "-"` (decreasing) or `sort = "+"`
(increasing). Sort alphabetically with `sort = "name+"` or
`sort = "name-"`:

```{r freq-sort}
freq(sochealth, education, sort = "-")
```

Sort alphabetically:

```{r freq-sort-name}
freq(sochealth, education, sort = "name+")
```

### Cumulative percentages

Add cumulative columns with `cum = TRUE`:

```{r freq-cum}
freq(sochealth, smoking, cum = TRUE)
```

### Weighted frequencies

Supply a weight variable with `weights`. By default
(`rescale = FALSE`), the table shows the raw weighted counts:

```{r freq-weighted-raw}
freq(sochealth, education, weights = weight)
```

Set `rescale = TRUE` to adjust the weighted total to match the
unweighted sample size:

```{r freq-weighted}
freq(sochealth, education, weights = weight, rescale = TRUE)
```

### Labelled variables

When a variable has value labels (e.g., imported from SPSS or Stata
with haven), `freq()` shows them by default with the `[code] label`
format. Control this with `labelled_levels`:

```{r freq-labelled}
# Create a labelled version of the smoking variable
sh <- sochealth
sh$smoking_lbl <- labelled::labelled(
  ifelse(sh$smoking == "Yes", 1L, 0L),
  labels = c("Non-smoker" = 0L, "Current smoker" = 1L)
)

# Default: [code] label
freq(sh, smoking_lbl)

# Labels only (no codes)
freq(sh, smoking_lbl, labelled_levels = "labels")

# Codes only (no labels)
freq(sh, smoking_lbl, labelled_levels = "values")
```

### Custom missing values

Raw data files often store nonresponse as an ordinary level -- a
"Refused" or "Don't know" answer sitting next to the substantive
categories. `na_val` reclassifies such a value as missing: it moves
to the Missing block and valid percentages are computed without it.

```{r freq-naval}
# Simulate a raw file where nonresponse was typed as a level
sh$income_raw <- as.character(sochealth$income_group)
sh$income_raw[is.na(sh$income_raw)] <- "Refused"

# "Refused" counts as a valid answer...
freq(sh, income_raw)

# ...until na_val reclassifies it as missing
freq(sh, income_raw, na_val = "Refused")
```

### Declared missing values

Survey files imported with haven often declare codes as missing at
the source -- `8 = Refused`, `9 = Don't know` -- through
`na_values`/`na_range` (SPSS-style) or tagged NAs (Stata-style).
The two behave differently in R. SPSS-style codes are *not* `NA`,
so most packages silently treat them as valid answers. Stata-style
tagged NAs are genuine `NA`s, but most packages collapse them into
a single undifferentiated `NA` count. spicy honors both
declarations by default: the codes are excluded from valid
percentages and statistics, and `freq()` shows them as labelled
rows of its Missing block instead of erasing them.

```{r freq-user-na}
sh2 <- sochealth
sh2$trust <- labelled::labelled_spss(
  c(1, 2, 8, 1, 9, 2, 1, 8, 2, 1, rep(c(1, 2), 595)),
  labels = c("Low" = 1, "High" = 2, "Refused" = 8, "Don't know" = 9),
  na_values = c(8, 9)
)
freq(sh2, trust)
```

Set `user_na = FALSE` to treat SPSS-style declared codes as
ordinary categories instead; tagged NAs are genuine `NA`s either
way, so for them it only collapses the per-tag breakdown into the
regular `NA` count. The same argument, with the same default,
exists in `cross_tab()`, the descriptive helpers
`table_categorical()`, `table_continuous()` and
`table_continuous_lm()`, and the row-wise functions `mean_n()`,
`sum_n()` and `count_n()`. (`table_regression()` takes a fitted
model, not raw data, so the argument does not apply there.)
`cross_tab()` and the descriptive helpers disclose the exclusion in
the table note; the row-wise functions return bare numeric vectors,
so they count the declared codes as missing without any note. See
the "Declared missing values" section of `?freq` for the full
contract.

## Cross-tabulations with cross_tab()

### Basic two-way table

Cross two variables to get a contingency table with a chi-squared test
and effect size:

```{r cross-basic}
cross_tab(sochealth, smoking, education)
```

### Row and column percentages

Use `percent = "row"` or `percent = "column"` to display percentages
instead of raw counts:

```{r cross-col}
cross_tab(sochealth, smoking, education, percent = "column")
```

```{r cross-row}
cross_tab(sochealth, smoking, education, percent = "row")
```

### Grouping with by

Stratify the table by a third variable:

```{r cross-by}
cross_tab(sochealth, smoking, education, by = sex)
```

For more than one grouping variable, use `interaction()`:

```{r cross-by-interaction}
cross_tab(sochealth, smoking, education,
          by = interaction(sex, age_group))
```

### Ordinal variables

When both variables are ordered factors, `cross_tab()` automatically
switches from Cramer's V to Kendall's Tau-b:

```{r cross-ordinal}
cross_tab(sochealth, self_rated_health, education)
```

You can override the automatic selection with `assoc_measure`:

```{r cross-assoc}
cross_tab(sochealth, self_rated_health, education, assoc_measure = "gamma")
```

### Confidence intervals for effect sizes

Add a 95% confidence interval for the association measure with
`assoc_ci = TRUE`:

```{r cross-ci}
cross_tab(sochealth, smoking, education, assoc_ci = TRUE)
```

### Weighted cross-tabulations

Weights are supplied via `weights`, just like in `freq()`, and both
functions share the same default: `rescale = FALSE` (raw weighted
counts).

```{r cross-weighted-raw}
cross_tab(sochealth, smoking, education, weights = weight)
```

Pass `rescale = TRUE` to make the weighted total match the
unweighted sample size:

```{r cross-weighted}
cross_tab(sochealth, smoking, education, weights = weight, rescale = TRUE)
```

### Monte Carlo simulation

The chi-squared test relies on an asymptotic approximation that
degrades when expected cell counts are small. `cross_tab()` checks
Cochran's rule and warns when a table is too sparse -- as in this
student subsample:

```{r cross-sparse}
students <- subset(sochealth, employment_status == "Student")
cross_tab(students, self_rated_health, education)
```

The warning itself names the remedy: `simulate_p = TRUE` replaces
the asymptotic p-value with one estimated by Monte Carlo simulation
(`simulate_B` random tables). A simulated test has no degrees of
freedom, which is why the statistic prints as `Chi-2(NA)`:

```{r cross-simulate}
cross_tab(students, self_rated_health, education,
          simulate_p = TRUE, simulate_B = 5000)
```

### Data frame output

Set `output = "data.frame"` to get a plain data frame for further
processing. The same argument works in `freq()`, and its two values
("default", "data.frame") are shared with the `output` argument of
the `table_*()` family -- but the rendered engines that family also
accepts ("tinytable", "gt", "flextable", ...) are not available
here:

```{r cross-df}
cross_tab(sochealth, smoking, education,
          percent = "column", output = "data.frame")
```

## Setting global defaults

You can set tabulation defaults with `options()` so you don't have
to repeat arguments. `spicy.rescale` is read by `cross_tab()`,
`freq()`, and `table_categorical()`; `spicy.percent` and
`spicy.simulate_p` are read by `cross_tab()`:

```{r options, eval = FALSE}
options(
  spicy.percent   = "column",
  spicy.simulate_p = TRUE,
  spicy.rescale   = TRUE
)
```

## Learn more

- [*Cramer's V, Phi, and association measures*](association-measures.html) - choosing the right effect size
  for your contingency table.
- [*Categorical summary tables*](table-categorical.html) - building publication-ready
  categorical tables.
- [*Summary tables for reporting*](summary-tables-reporting.html) - end-to-end reporting
  workflow that ties the four `table_*` helpers together along the
  APA Table 1 / 2 / 3 sequence (including `table_regression()` for
  the coefficient table).
- `?freq` and `?cross_tab` for the full argument reference.
