---
title: "Summary tables for reporting"
description: >
  Learn when to use each function of the spicy table family --
  from table_categorical() to table_regression(), including
  table_outcome() and the survey-design twins -- how their shared
  arguments fit together, and which output format to choose for
  console, Quarto, Word, or Excel workflows.
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

build_rich_tables <- identical(Sys.getenv("IN_PKGDOWN"), "true")

source("_pkgdown-helpers.R")
```

```{r setup}
library(spicy)
```

spicy's reporting helpers cover the full APA Manual 7 table
sequence used in empirical articles:

- `table_categorical()` and `table_continuous()` build **Table 1**
  (sample characteristics) and **Table 2** (group comparisons);
- `table_continuous_lm()` extends Table 2 to the linear-model regime
  when group means need robust or cluster-robust SE, covariate
  adjustment, or group tests under case weights — the descriptive
  families take `weights =` natively for the estimates themselves,
  and `table_continuous()` points here the moment weighted *means*
  are to be tested;
- `table_continuous_svy()` and `table_categorical_svy()` build the
  same Tables 1 and 2 from a `survey::svydesign` object, when the
  sample is stratified, clustered, or carried by replicate weights —
  design-based SEs and CIs, design degrees of freedom, design tests
  (see [*Summary tables from a survey design*](survey-tables.html));
- `table_regression()` builds **Table 3** (the coefficient table) from
  one or several fitted models — `lm()` / `glm()` through the 36
  supported classes listed by `table_regression_models()`.

The four functions share the same output grammar — the same `output`
formats (`default` console ASCII, `gt`, `tinytable`, `flextable`,
`word`, `excel`, `clipboard`),
the same `decimal_mark`, `p_digits`, `labels`, and `align`
arguments, and the same `digits` control for numeric cells (the
categorical table's cells are percentages, so it spells the argument
`percent_digits`) — so a single reporting workflow can move smoothly from
descriptive to inferential without juggling different APIs. This
article focuses on that shared logic; the function-specific articles
cover the methodological options in depth.

## Choose the right function

Use the function that matches the unit you want to report:

| Function | Reports | Selection grammar | Typical additions |
|:--|:--|:--|:--|
| `table_categorical()` | Categorical variables (factors, labelled) | `select`, `by` | Chi-squared test, association measure (`phi`, `cramer_v`, `tau_b`, ...), confidence interval |
| `table_continuous()` | Numeric / continuous variables | `select`, `by` | Group-comparison test (Student / Welch *t*, Wilcoxon, ANOVA, Kruskal–Wallis), effect size (Hedges' *g*, η², rank-biserial *r*, ε²) |
| `table_continuous_lm()` | Numeric outcomes through one linear model per outcome | `select`, `by` (single predictor) | Robust / cluster-robust / bootstrap / jackknife SE, case weights, additive covariate adjustment, four effect-size measures with noncentral CIs |
| `table_outcome()` | *One* numeric outcome across the levels of *several* categorical variables | `outcome` (one), `select` (many) | One group comparison per block, an `Overall` marginal row, the same statistic tokens as `table_continuous()` |
| `table_continuous_svy()`, `table_categorical_svy()` | The same descriptive tables from a `survey::svydesign` (stratified, clustered, replicate-weight samples) | design-first: `design`, then `select`, `by` | Design-based SE and CI, design df, Rao–Scott and design *t*/*F* tests, design effects, observed and weighted counts |
| `table_regression()` | One or several fitted models — 36 classes, from `lm()` / `glm()` to mixed, ordinal, survival and Bayesian engines (see [*Supported models*](table-regression-supported-models.html)) | Fit-first: pass the model object(s) directly, no `select` / `by` | APA-aligned coefficient table with `B`, `β`, `95% CI`, `p`, AME, robust variance, side-by-side and hierarchical layouts |

In practice, follow the APA sequence:

- start with `table_categorical()` for smoking, education, or
  activity — APA Table 1 categorical descriptors;
- use `table_continuous()` for BMI, well-being, or income — Table 1
  continuous descriptors and Table 2 unadjusted group comparisons;
- transpose to `table_outcome()` when the question is about *one*
  outcome and many groupings rather than many outcomes and one
  grouping -- a well-being score described by sex, by education and
  by region, block after block;
- move to `table_continuous_svy()` / `table_categorical_svy()` when
  the data are a survey design rather than a plain sample -- the
  numbers then come from the survey package, with the design's SEs,
  degrees of freedom and tests;
- switch to `table_continuous_lm()` when the same comparison must
  account for case weights, robust SE, or covariate adjustment;
- finish with `table_regression()` once the substantive model is
  fitted — APA Table 3 with all predictors, factor groupings,
  reference rows, and (optionally) standardised coefficients,
  marginal effects, or nested model comparisons.

The descriptive functions share one selection grammar — on a data
frame, or on a `survey::svydesign` for the `_svy` twins;
`table_regression()` is **fit-first** — you build the model
the usual R way (`lm()`, `glm()`, or any other supported engine) and
hand the object in. All of them share the post-construction grammar
(`output`, `labels`, `decimal_mark`, `align`, and the digits
controls), so swapping functions never breaks your rendering
pipeline.

## A shared interface

The examples below use `sochealth`, the dataset bundled with spicy: a
simulated social-health survey of 1200 respondents and 24 variables,
every one of them carrying a variable label (see `?sochealth`).

The three descriptive functions share the same core arguments:

```{r grammar-categorical}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  labels = c(
    smoking           = "Smoking status",
    physical_activity = "Regular physical activity"
  )
)
```

```{r grammar-continuous}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  labels = c(
    bmi = "Body mass index",
    wellbeing_score = "Well-being score",
    life_sat_health = "Satisfaction with health"
  )
)
```

```{r grammar-continuous-lm}
table_continuous_lm(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  weights = weight,
  vcov = "HC3"
)
```

Two words on the weighted example. `weights` supplies **case
weights**, passed to `lm(weights = )` — appropriate for weighted
article tables, but not a substitute for a full complex-survey design
(strata, clusters, calibration), which is the `survey` package's
domain. And because `sochealth$weight` holds calibrated sampling
weights, the example pairs them with a heteroskedasticity-robust
variance (`vcov = "HC3"`): the default `"classical"` WLS variance
would treat the weights as precision weights, which sampling weights
are not.

The same argument pattern is used in all three cases:

- `select` chooses the reported variables;
- `by` defines the grouping structure;
- `labels` cleans up the row labels;
- `output` decides how the result is rendered or exported.

For model-based continuous tables, the same pattern applies, but `by`
must be a single predictor because one linear model is fit per outcome.

`table_regression()` joins the same `labels` / `output` /
`decimal_mark` / `digits` grammar but is **fit-first**: rather than
expressing model structure inline through `select` and `by`, you pass
one or several already-fitted `lm()` or `glm()` objects:

```{r grammar-regression}
fit <- lm(
  wellbeing_score ~ age + sex + smoking + physical_activity,
  data = sochealth
)
table_regression(
  fit,
  labels = c(
    age               = "Age (years)",
    sex               = "Sex",
    smoking           = "Smoking status",
    physical_activity = "Regular physical activity"
  )
)
```

This split is intentional. The descriptive trio (categorical,
continuous, continuous_lm) reports the *data* — `select` and `by`
describe what you want to see. `table_regression()` reports the
*model* — the model formula has already declared which predictors,
interactions, polynomials, transformations, splines, and contrasts to
report, so passing those again through `select` / `by` would
duplicate the model object's information and risk diverging from it.

## A practical reporting sequence

A common report contains both table types, often with the same grouping
variable. For example, you might first summarize categorical health
behaviors, then summarize continuous well-being indicators.

### Categorical variables

```{r report-categorical}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  labels = c(
    smoking           = "Smoking status",
    physical_activity = "Regular physical activity",
    dentist_12m       = "Visited a dentist in the last 12 months"
  )
)
```

### Continuous variables

```{r report-continuous}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  labels = c(
    bmi = "Body mass index",
    wellbeing_score = "Well-being score",
    life_sat_health = "Satisfaction with health"
  ),
  p_value = TRUE,
  effect_size = TRUE
)
```

This keeps the reporting structure consistent while still using the
function that fits each variable type.

### Model-based continuous variables

```{r report-continuous-lm}
table_continuous_lm(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = sex,
  vcov = "HC3",
  statistic = TRUE
)
```

This is the better summary-table path when the article is already
organized around simple linear models, weighted analyses, or robust
standard errors.

### A balance table instead of a significance table

When the two groups are trial arms or a treated / control contrast,
the baseline table is read for *balance*, and the convention of that
literature is the standardized mean difference rather than the
*p*-value. Both descriptive families take `smd = TRUE`, with the same
meaning and the same refusals:

```{r report-balance-continuous}
table_continuous(
  sochealth,
  select = c(age, bmi, wellbeing_score),
  by = sex,
  smd = TRUE,
  p_value = FALSE
)
```

```{r report-balance-categorical}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = sex,
  smd = TRUE
)
```

Exactly two groups, no confidence interval and no *p*-value on the
column itself, and the usual rule of thumb (|SMD| > 0.1) quoted in the
table note without any cell being highlighted. Note the asymmetry
between the two calls: `table_continuous()` can drop its *p* column,
`table_categorical()` currently cannot.

### The coefficient table

Once the substantive model is fitted, `table_regression()` produces
the APA Table 3 coefficient summary. The same `output` argument
controls rendering, so the regression table sits in the same
reporting pipeline as the descriptive ones above:

```{r report-regression}
fit <- lm(
  wellbeing_score ~ age + sex + smoking + physical_activity,
  data = sochealth
)
table_regression(
  fit,
  standardized = "refit",
  show_columns = c("b", "beta", "ci", "p"),
  vcov = "HC3"
)
```

The default footer documents the variance estimator, flags
standardised coefficients (`β = standardised coefficient`), and
reports any multiplicity correction, so the inferential regime is
visible without leaving the table. One thing it does not carry: the
*name* of the standardisation method. Here `standardized = "refit"`
produced the β column, but the footer would read the same under any
of the five methods, so record the method in the table note or the
text of the article.

Side-by-side reporting of competing specifications (e.g., unadjusted
vs. covariate-adjusted, or `lm` vs. `glm`) is supported by passing a
list of fits:

```{r report-regression-multi}
fit_unadj <- lm(wellbeing_score ~ smoking, data = sochealth)
fit_adj   <- lm(
  wellbeing_score ~ smoking + age + sex + physical_activity,
  data = sochealth
)
table_regression(
  list("Unadjusted" = fit_unadj, "Adjusted" = fit_adj),
  show_columns = c("b", "ci", "p")
)
```

For binary or count outcomes, swap `lm()` for `glm()` and request
response-scale reporting (odds ratios, incidence rate ratios, etc.):

```{r report-regression-glm}
fit_glm <- glm(
  smoking ~ age + sex + physical_activity,
  data = sochealth,
  family = binomial()
)
table_regression(
  fit_glm,
  exponentiate = TRUE,
  show_columns = c("b", "ci", "p", "ame", "ame_ci", "ame_p")
)
```

Average marginal effects (`ame`) are useful next to the odds ratio
because they report a probability-scale change for each predictor —
the quantity most reviewers want to interpret directly. Note the two
`p` columns: the first tests the coefficient (the log odds ratio),
`ame_p` tests the average marginal effect itself. The two can differ
under non-linear links or interactions, which is why
`table_regression()` warns if you request `ame` alongside `p` without
also requesting `ame_p`.

For the epidemiological variant of Table 2 — a univariable screen of
every candidate predictor set against the multivariable model —
`table_regression_uv()` builds the whole two-part layout in one
call; see the *Univariable screening* section of
[*Publication-ready regression tables*](table-regression.html).

## Choose the output format

All four functions support the same reporting formats:

| Output | Best use |
|:--|:--|
| `"default"` | Quick console review in plain ASCII |
| `"tinytable"` | Quarto or R Markdown documents |
| `"gt"` | HTML output with styled reporting tables |
| `"flextable"` | Office-first workflows; also renders in HTML |
| `"excel"` | Spreadsheet handoff or downstream editing |
| `"word"` | Direct `.docx` export |
| `"clipboard"` | Fast pasting into another application |

Pick the output based on where the table is going, not on the analysis
itself. The underlying selection and grouping pattern stays the same.

If you want an object that fits naturally into Word and PowerPoint
workflows but can also be rendered in HTML documents, `flextable` is a
good choice:

```{r output-flextable, eval = FALSE}
if (requireNamespace("flextable", quietly = TRUE)) {
  table_continuous(
    sochealth,
    select = c(bmi, wellbeing_score, life_sat_health),
    by = education,
    output = "flextable"
  )
}
```

## House styles

Everything above uses spicy's defaults: two decimals, APA-style
*p* values (leading zero dropped, `< .001` floor), confidence
intervals in brackets. Those defaults are one house style among
several. `spicy_style()` names the others — each rule sourced from
the journal's published author guidelines:

```{r house-styles}
table_categorical(sochealth, select = smoking, by = sex, style = "jama")
```

The visible change here is the *p* column: JAMA rounds *p* values to
two decimals (`.71`), where the default reports three (`.713`).

Two properties make styles safe to adopt. A style only moves
*display* defaults — decimals, *p* notation, interval punctuation —
never the statistics underneath; and an argument you set explicitly
always wins over it, so `style = "jama"` with `p_digits = 3` keeps
your three decimals. The style travels with the table: the same call
rendered to Word, Excel, or HTML follows it, and
`options(spicy.style = "lancet")` sets one for a whole document.
`?spicy_style` lists every style and every rule with the sentence it
comes from.

## Table language

A style is what a journal asks for. A language is what your reader
reads. `options(spicy.language = "fr")` prints the table in French —
headers, row labels, titles and table notes, and the numbers with
them:

```{r language-fr}
options(spicy.language = "fr")
table_continuous(sochealth, select = bmi, by = sex)
```

One gesture, one coherent table. A language brings its typography with
it: the decimal comma, and the leading zero French typography keeps on
a *p* value (`0,003` where the English default writes `.003`). The
language of a report is a property of the report, not of a call, so it
is set once in the setup chunk. `"en"` is the default and is unchanged
by any of this. The language reaches every table, the exploration pair
included: `freq()` and `cross_tab()` take the comma from it too, and an
argument you type wins. One boundary is worth knowing: figures are
frozen when a table is built, words when it is printed, which is one
more reason the language belongs in the setup chunk.

A journal style composes with the language rather than replacing it. A
theme encodes only what its own author guidelines state, so whatever it
leaves open the language still fills — JAMA fixes no decimal mark, and
a French JAMA table keeps the comma:

```{r language-fr-style}
table_continuous(sochealth, select = bmi, by = sex, style = "jama")
```

Where the two do meet, the theme wins — you asked for it by name.
JAMA's own rule drops the leading zero of a *p* value, so the table
above writes `,02` rather than `0,02`; `style = "lancet"` keeps the
journal's midline decimal point instead of the comma. To keep a
theme's other rules and restore the zero, compose the variant:
`style = spicy_style("jama", p_style = "standard")`.

An argument you type wins over both, which is the escape hatch for a
bilingual table — French words, decimal point. It moves only the mark:
the *p* value keeps its French leading zero (`0.018`), since the
leading-zero rule is a style lever, not an argument:

```{r language-fr-mark}
table_continuous(sochealth, select = bmi, by = sex, decimal_mark = ".")
```

```{r language-reset, include = FALSE}
options(spicy.language = NULL)
```

When one word has to change and a language does not — a questionnaire
where a missing category means a refusal, not an absent value —
`options(spicy.labels = )` overrides labels one at a time, on top of
whatever language is in force:

```{r language-labels}
options(spicy.labels = list(row_missing_level = "(No answer)"))
table_categorical(sochealth, select = sex, by = smoking)
options(spicy.labels = NULL)
```

`spicy_labels()` is how you find the key for a label you want to
change: it returns every key with the text it currently resolves to.

Two things a language deliberately does not move. The column names of
the exported frames are a contract your code indexes into, so
`out[["Yes %"]]` resolves whatever the language — spicy translates its
own vocabulary, never your data. And errors and warnings stay in
English, because they are read by developers and quoted in bug reports.

The one column name that *does* follow the language comes from the same
rule read the other way: a column named after a level of `by` takes that
level's spelling, and spicy's own missing category is a level. So
`table_categorical(by = )` on a variable with missing values gives
`(Missing) n` in English and `(Manquant) n` in French. If your code
selects that column, build its name from `spicy_labels()[["row_missing_level"]]`
rather than typing it.

## Citing table values in the text

The number a sentence quotes should be the number the table prints —
retyping it is how a manuscript ends up saying 3.9 where the table
says 3.90, or keeping a *p* value a revision changed. `inline()`
returns one cell of a spicy table as text, formatted by the same
machinery that formatted the table:

```{r inline-single}
fit <- lm(wellbeing_score ~ age + sex, data = sochealth)
tbl <- table_regression(fit)
inline(tbl, sex, "Male", "b")
```

so in Quarto you write
`` `r knitr::inline_expr('inline(tbl, sex, "Male", "b")')` ``
inside the sentence. A `{token}` pattern quotes a full fragment in
one call:

```{r inline-pattern}
inline(tbl, sex, "Male", "{b} ({ci_label} {ci}; p {p})")
```

Two properties carry the guarantee. The text follows the table:
under `style = "jama"`, `decimal_mark = ","` or a table language, the
cited string changes with the printed one. And the addressing survives relabeling:
rows are found by the source variable and level — not by the
displayed label — so `labels = c(sex = "Administrative sex")` changes
the table, not your calls. Misaddressing never fails silently: an
unknown variable, level, or column token errors with the list of
available choices, and a reference or non-estimable cell refuses with
its reason instead of pasting a dash into a sentence.

## Post-process the returned table object

All four summary-table helpers return regular `gt`, `tinytable`, or
`flextable` objects, so you can keep styling them with the native
package API. This includes `table_regression()`: nothing about the
fit-first interface changes what the rendering engine produces.

Use `gt::` functions when you want to keep the `gt` workflow:

```{r postprocess-gt, eval = build_rich_tables}
tab <- pkgdown_dark_gt(table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  labels = c(
    smoking           = "Smoking status",
    physical_activity = "Regular physical activity"
  ),
  output = "gt"
))

tab |>
  gt::tab_header(
    title = "Health behaviors by education",
    subtitle = "Categorical summary table"
  ) |>
  gt::tab_source_note(
    gt::md("*Percentages are computed within each education group.*")
  )
```

Use `tinytable::` functions when you want lightweight table-specific
styling:

```{r postprocess-tinytable, eval = build_rich_tables}
tab <- table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  labels = c(
    smoking           = "Smoking status",
    physical_activity = "Regular physical activity"
  ),
  output = "tinytable"
)

tab |>
  tinytable::style_tt(
    i = 2:3,
    j = 2:5,
    background = "red",
    color = "white",
    bold = TRUE
  )
```

Use `flextable::` functions when you want to keep working toward Office
or HTML document output. The example is shown as code here because the
dark pkgdown theme is not a reliable preview of the final `flextable`
HTML rendering:

```{r postprocess-flextable, eval = FALSE}
if (requireNamespace("flextable", quietly = TRUE)) {
  tab <- table_continuous(
    sochealth,
    select = c(bmi, wellbeing_score),
    by = education,
    output = "flextable"
  )

  tab |>
    flextable::theme_booktabs() |>
    flextable::autofit() |>
    flextable::fontsize(size = 10, part = "all")
}
```

## Keep the detailed options in the function-specific articles

The dedicated articles go deeper into each function:

- `table_categorical()` covers missing values, level filtering,
  association measures, and one-way frequency-style tables.
- `table_continuous()` covers grouped descriptive statistics,
  parametric and nonparametric tests, and effect sizes.
- `table_continuous_lm()` covers estimated marginal means or slopes
  from linear models, robust / cluster-robust / bootstrap / jackknife
  variance, case weights, additive covariate adjustment
  (G-computation or equal-weight), and four effect-size measures
  with noncentral CIs.
- `table_outcome()` covers the outcome-first layout: block
  structure, the `Overall` row, per-block comparisons and effect
  sizes, and the comparison with `gtsummary::tbl_continuous()`.
- `table_continuous_svy()` and `table_categorical_svy()` cover the
  survey-design regime: delegation to the survey package, design
  degrees of freedom, quantile rules, and the weights-vs-design
  estimand boundary.
- `table_regression()` covers single- and multi-model regression
  tables across 30+ model classes (the map is
  [*Supported models*](table-regression-supported-models.html)), five
  standardisation methods (four for linear models, plus the glm-only
  pseudo-standardisation), partial effect sizes with noncentral-F
  CIs, average marginal effects, hierarchical (`nested = TRUE`)
  comparisons, multiplicity correction, and response-scale reporting
  for GLMs.

Use this article as the final reporting overview, then consult the
function-specific articles when you need the detailed controls.
