---
title: "Supported models"
description: >
  The model classes `table_regression()` supports, in one table:
  linear and generalized linear models, robust and IV estimators,
  mixed effects, ordinal and multinomial outcomes, two-part count
  models, survival models, survey-weighted and Bayesian fits -- with
  each family's marginal-effects estimand, exponentiate semantics
  and row blocks, plus the robust-variance backends and
  standardized-coefficient support each family takes.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 110)
```

```{r setup, message = FALSE}
library(spicy)
```

`table_regression()` takes fitted model objects -- never raw data plus
a formula -- and renders them as publication-ready coefficient tables.
"Supported" is a commitment, not a list of classes that happen to run:
every number a supported class produces is validated against a field
reference (the model's own `summary()`, `sandwich`, `clubSandwich`,
`marginaleffects`, `effectsize`, `performance`, Stata or SPSS
conventions), and every request a class cannot honour is **refused
with a classed error** that names what is available -- never rendered
as a silently empty or approximate column.

This article is the map. Each family below links to a dedicated
article that walks through its behaviour in depth.

## Choosing a model

Most readers arrive with the opposite question: not "what does spicy
support?" but "I have this outcome -- which model do I fit?". The
table below answers it for the situations applied work meets most
often. It *recommends and explains*; it never chooses for you --
model choice depends on your design and your estimand, and spicy will
render whichever defensible model you fit.

| Your outcome | The situation | Reach for | Notes |
|---|---|---|---|
| Continuous | roughly symmetric errors, independent observations | `lm()` | the default screen; `beta` for standardized effects |
| Continuous | positive and right-skewed | `glm(family = Gamma("log"))` | `exponentiate = TRUE` gives mean ratios (MR) |
| Continuous | censored at a bound (floor/ceiling) | `AER::tobit()` | |
| Continuous | outlying responses distort the fit | `MASS::rlm()` | M-estimation: resistant estimates, not just fixed SEs |
| Continuous | the median or another quantile is the estimand | `quantreg::rq()` | one model per quantile; effects are quantile-specific |
| Binary | independent observations | `glm(family = binomial())` | OR via `exponentiate`; add `"ame"` for probability effects |
| Binary | overdispersed grouped binomial data | `glm(family = quasibinomial())` | |
| Ordered categories | a Likert scale, severity grades | `MASS::polr()` or `ordinal::clm()` | thresholds render as a labelled block; AME per category; `clm()` also fits partial proportional odds |
| Unordered categories | 3+ nominal outcomes | `nnet::multinom()`; `mlogit::mlogit()` for alternative-specific predictors | outcome categories render as column groups |
| Count | mean roughly equal to variance | `glm(family = poisson())` | IRR via `exponentiate`; rates via an `offset()` |
| Count | variance well above the mean | `MASS::glm.nb()` | models the overdispersion; `quasipoisson` merely widens the SEs |
| Count | more zeros than the count part explains | `pscl::zeroinfl()` / `pscl::hurdle()` | both components render as labelled blocks |
| Proportion in (0, 1) | rates, indices, shares | `betareg::betareg()` | |
| Time-to-event | independent observations | `survival::coxph()` | HR -- and adjusted RMST / risk-difference columns beyond it |
| Time-to-event | a parametric survival-time model | `survival::survreg()` | time ratios (TR) via `exponentiate` |
| Any of the above | clustered or repeated measures | `lme4::lmer()` / `glmer()` **or** `geepack::geeglm()` | *the distinction below matters* |
| Any of the above | many absorbed fixed effects (panel) | `fixest::feols()` / `feglm()` | the absorbed factors render as a Yes/No block |
| Any of the above | a complex survey design | `survey::svyglm()` | design-based SEs are the inference |
| Any of the above | smooth non-linear terms | `mgcv::gam()` / `bam()` | |
| Any of the above | a Bayesian analysis | `rstanarm::stan_glm()` / `brms::brm()` | posterior medians, MAD SD, credible intervals -- no p-values |
| Any of the above | an endogenous predictor, an instrument | `AER::ivreg()` / `estimatr::iv_robust()` | |

Three distinctions the table cannot compress:

**Clustered data: marginal or conditional?** For repeated or
clustered measurements, `glmer()` and `geeglm()` answer *different
questions*. A mixed model is *subject-specific*: its odds ratio
compares outcomes within the same cluster. GEE is
*population-averaged*: its odds ratio compares whole subpopulations
-- usually the public-health question. Under a logit link the two
genuinely differ -- the population-averaged OR is attenuated toward
the null relative to the subject-specific one -- so the choice is
about your estimand, not about taste. spicy renders both with their
own correct inference (boundary-corrected random-effect tests for
the former, the native sandwich SEs and working correlation for the
latter).

**Binary outcomes: OR, RR, or a probability effect?** The odds ratio
is the default because the logit is; it is also routinely
misread as a risk ratio. When the RR is the estimand, fit the log
link (`binomial("log")`) and spicy labels the exponentiated
coefficient RR -- knowing that log-binomial models can fail to
converge on common data, which is a property of the model, not of
the table. When what you need is "how many percentage points does
this predictor change the probability", add
`show_columns = c("b", "ame")` -- the average marginal effect is the
quantity most readers actually want, it sidesteps both ratio
debates, and for ordinal and multinomial models spicy gives it per
outcome category.

**Survival: beyond the hazard ratio.** The HR is non-collapsible --
adjusting for a prognostic covariate changes it even without any
confounding -- and when proportional hazards fail, the single
reported HR is an average whose weights depend on censoring and
follow-up, not a stable quantity. spicy's `rmst` and `risk_diff`
columns give covariate-adjusted differences in restricted mean
survival time and in cumulative incidence by g-computation --
absolute, time-anchored quantities a reader can act on -- for Cox
and parametric AFT fits alike, in single tables and in the
univariable screen.

And when you fit something the guide does not cover: if the class is
supported, the registry below renders it; if it is not, spicy
refuses with a classed error that names the nearest supported route
(see *When a class is not supported*) -- it never renders an
approximation of a model it does not understand.

## The registry

The table is generated from the same internal registry that the
package itself uses, so it cannot drift from the code. Call
`table_regression_models()` to get it as a data frame.

```{r registry, echo = FALSE}
reg <- table_regression_models()
reg$family[duplicated(reg$family)] <- ""
reg$class <- paste0("`", reg$class, "`")
reg$engine <- vapply(
  strsplit(reg$engine, ", ", fixed = TRUE),
  function(fns) paste0("`", fns, "`", collapse = ", "),
  character(1)
)
knitr::kable(
  reg,
  col.names = c("Family", "Class", "Engine", "AME", "Exponentiate",
                "Blocks")
)
```

How to read the columns:

* **AME** -- what `show_columns = c("b", "ame")` adds. `yes` is an
  average marginal effect on the response scale (a probability or
  rate effect for GLM families, the slope itself under identity).
  `per category` (ordinal) is the effect on each P(Y = k);
  `per outcome` (multinomial) is one effect per non-reference
  outcome. `survival::coxph` refuses AME -- the hazard scale has no
  marginal-probability effect -- and provides covariate-adjusted
  `rmst` and `risk_diff` columns instead (also available for
  `survreg`; `rms::cph` and `flexsurvreg` support neither AME nor
  the estimand columns). `yes (draws)` means the
  effect is computed per posterior draw and summarized as a
  posterior median with MAD SD and credible interval.
* **Exponentiate** -- the labelled ratio `exponentiate = TRUE`
  produces. The label follows the link: OR under logit, IRR for
  count log-links, RR for the binomial log link, MR (mean ratio)
  for Gamma log links, HR for proportional hazards, TR (time ratio)
  for accelerated-failure-time models. Identity-link fits warn and
  stay untouched; links whose exponential is not a ratio (probit,
  cauchit, inverse) are refused.
* **Blocks** -- labelled subordinate row blocks rendered inside the
  same table (random effects, thresholds, zero components,
  per-outcome segments), each explained by a footer line.

## Cross-cutting arguments

The same arguments work across families, each through the family's
field-standard backend -- or a clear refusal.

**Robust and cluster-robust standard errors** (`vcov`, `cluster`).
Family by family:

* **`lm`, `glm`** -- `HC0`--`HC5` (`sandwich`) and `CR0`--`CR3`
  (`clubSandwich`, bias-reduced with Satterthwaite df), plus
  `"bootstrap"` / `"jackknife"` resampling estimators.
* **Mixed effects** -- `lmer` and `nlme::lme` take `CR*` via
  `clubSandwich`; `glmer`, `glmmTMB` and `gls` are model-based only
  (`clubSandwich` has no working backend for the first two, and its
  `gls` backend is not yet wired and validated in spicy).
* **Ordinal (`polr`, `clm`)** -- `CR0`--`CR3`, no `HC*`; the
  cut-point thresholds are reweighted from the same clustered vcov.
  A `clm` with a scale or nominal partial-proportional-odds
  component is model-based only.
* **Categorical** -- `multinom` takes `CR*` (one cluster value per
  observation), `mlogit` takes `CR*` (one per choice situation).
  Both refuse `HC*`: `multinom` has no working-residual form for a
  multi-equation model, and for `mlogit`, `sandwich::vcovHC()`
  computes a result but silently mis-scales the meat for its
  per-chooser score structure.
* **Quantile (`rq`)** -- its own estimator family: `"classical"`
  resolves to the robust `nid` sandwich, `iid` / `ker` / `rank` are
  opt-ins, and clustering goes through the native wild gradient
  bootstrap (`vcov = "bootstrap"` + `cluster`). `HC*` / `CR*` are
  refused.
* **Survival** -- Cox models use the Lin-Wei grouped-dfbeta
  sandwich; the `rms` fits take `CR*` via `rms::robcov()` (refit
  with `x = TRUE, y = TRUE`); `survreg` takes `CR*` via
  `sandwich::vcovCL()`.
* **`gam` / `bam`, `betareg`, `pscl` two-part** -- `CR*` via
  `sandwich::vcovCL()`; zero-inflated and hurdle fits cluster both
  components.
* **Own-estimator classes** -- `estimatr` fits keep the robust SEs
  they were computed with; `fixest` fits keep their estimator (the
  footer carries fixest's own label -- clustered, Newey-West,
  Conley, ...; fixest's "IID" is normalised to "Classical"). spicy's
  `HC*` / `CR*` tokens are refused for both.
* **Robust by construction** -- `svyglm` is design-based (Taylor /
  replicate): the design variance *is* the robust variance, and
  clustering belongs in the design itself
  (`survey::svydesign(ids = )`). `geeglm` displays the sandwich SEs
  the fit computed over its own `id =` clusters; change the
  estimator by refitting with `std.err =`. Both refuse spicy's
  `HC*` / `CR*` tokens and `cluster`.
* **Bayesian** -- `vcov` is refused: nothing standard plays the
  sandwich role for a posterior.

Whatever the backend, the footer names the estimator actually
applied, and a robust vcov also flows into the AME uncertainty.

**Standardized coefficients** (`standardized`). Available for `lm`,
`glm` (including `MASS::glm.nb`), the mixed engines (`lmer` /
`glmer` / `glmmTMB` / `nlme::lme`), and fixed-effects Bayesian fits
-- `stan_glm`-style models and standard-formula `brm()` models --
where `"posthoc"`, `"basic"` and `"smart"` are exact affine
rescales of the posterior draws. Other classes refuse: the
Bayesian refusals (multilevel fits, brms formulas with
distributional or special terms) hint to standardize predictors
before fitting, and the frequentist ones point to the AME columns
instead.

**Confidence intervals** (`ci_method`).

* Wald everywhere by default;
* `"profile"` (profile likelihood) for `glm`, `polr` and `clm`;
* `"boot_percentile"` (with `vcov = "bootstrap"`) replaces the
  bounds with equal-tailed percentile intervals of the bootstrap
  replicates;
* `"hdi"` (highest-density interval) for Bayesian fits, which
  otherwise report equal-tailed credible intervals.

**Model comparison and multiplicity.** `nested = TRUE` compares
nested fits by the family's change-test convention: Delta R-squared
with the partial F test for `lm`, the likelihood-ratio test for
`glm`, mixed, `multinom` and Cox models, and `anova.rq`'s Wald-type
F for quantile regressions (all fits at one tau). `p_adjust` applies a
multiplicity correction across the displayed p-values -- and is
refused for Bayesian tables, which carry no p-values at all.

## The families in brief

**Linear and generalized linear.** The core engines: `lm`, `glm`
(with profile CIs on request), `MASS::glm.nb` (with opt-in `theta` /
`alpha` dispersion statistics), `MASS::rlm`, `stats::nls`. Start
with `vignette("table-regression")` and
`vignette("categorical-predictors")`.

**Robust, IV, quantile, panel.** `estimatr::lm_robust()` /
`iv_robust()`, `AER::ivreg()` and `AER::tobit()`, `quantreg::rq()`
(defaulting to the heteroskedasticity-robust `nid` sandwich --
quantreg's own large-sample default -- with `iid`, `ker`, `rank`
CIs and a native clustered bootstrap as `vcov` options), and the
`fixest` estimators, whose absorbed fixed effects render as a
default-on `Fixed effects:` block -- one Yes / No row per factor,
blank for non-fixest models in a mixed table -- with the within
R-squared among the default fit statistics and per-factor
`N (<factor>)` counts through the opt-in `"n_groups"` token.

**Mixed effects.** `lmer` (Satterthwaite t via `lmerTest`), `glmer`,
`glmmTMB` (with zero-inflation and dispersion blocks), `nlme::lme`
and `nlme::gls`. Random effects render as rows -- SD, correlations,
residual -- deliberately without per-row p-values; the footer
carries the boundary-correct chi-bar-squared test, and
`re_test = "lrt"` / `"rlrt"` adds per-term tests. See
`vignette("table-regression-mixed")`.

**Population-averaged (GEE).** `geepack::geeglm()`: marginal
(population-averaged) coefficients with the fit's own sandwich SEs
as the default inference, the working correlation structure disclosed
in the footer (with its estimated alpha), cluster-structure fit
statistics, and opt-in QIC / QICu. See
`vignette("table-regression-gee")` for the full workflow;
`vignette("table-regression-mixed")` contrasts GEE with
subject-specific mixed models in a single table.

**Ordinal.** `MASS::polr` and `ordinal::clm`: proportional odds
ratios, a `Thresholds` block for the cut-points (log-odds scale,
never exponentiated), partial-proportional-odds terms as a
`Non-proportional effects` block, and per-category AME. See
`vignette("table-regression-ordinal")`.

**Categorical.** `nnet::multinom` renders outcome categories as
columns with per-outcome AME; `mlogit::mlogit` renders
per-alternative rows for discrete-choice designs. See
`vignette("table-regression-multinomial")`.

**Counts and two-part.** Poisson and negative binomial through
`glm` / `glm.nb` / `glmmTMB`, plus `pscl::zeroinfl()` and
`pscl::hurdle()` with their zero components as labelled blocks and a
combined-response AME. See `vignette("table-regression-counts")`.

**Survival.** `survival::coxph` and `rms::cph` (hazard ratios,
`strata()` supported), `survival::survreg` (time ratios) and
`flexsurv::flexsurvreg`. Absolute effects come as covariate-adjusted
RMST and risk differences by g-computation for `coxph` and
`survreg` fits. The baseline hazard those estimands standardize
follows the tie-handling convention of the fit, exactly as
`survfit()` and `basehaz()` do -- a `ties = "breslow"` fit gives a
Breslow baseline, the default Efron fit an Efron one -- so the
hazard-ratio column and the dRMST column always come from the same
likelihood. See `vignette("table-regression-survival")`.

**Survey-weighted.** `survey::svyglm()`: design-based inference;
the unweighted n is a default fit statistic and the sum of design
weights is available as `show_fit_stats = "weighted_nobs"`.

**Additive, proportions, selection.** `mgcv::gam()` / `bam()`,
`betareg::betareg()` (odds ratios on the logit mean link under
`exponentiate`; the precision is the opt-in `phi` fit statistic),
and `sampleSelection::selection()` with its selection component as
a block.

**rms.** `ols`, `lrm`, `Glm` and `cph` are first-class citizens, so
Harrell-style workflows drop in directly.

**Bayesian.** `rstanarm` and `brms` fits are summarized from their
posterior draws: posterior median, MAD SD, credible intervals,
draws-native exponentiation and AME, sampler diagnostics checked on
every fit (with opt-in `pd`, `rhat`, `ess_bulk` / `ess_tail` and
`mcse` columns). No p-values, by design. See
`vignette("table-regression-bayesian")`.

## When a class is not supported

An unsupported class fails fast with a classed error
(`spicy_unsupported`):

```{r refusal, error = TRUE}
fit <- loess(dist ~ speed, data = cars)
table_regression(fit)
```

The same contract applies inside a family: a request a class cannot
honour -- `HC*` for `multinom`, AME for `mlogit`, `exponentiate` on
a probit link, `p_adjust` on a Bayesian table -- is refused with the
reason and the supported alternative, never silently degraded.

## Programmatic access

`table_regression_models()` returns the registry as a plain data
frame -- convenient to filter, join, or cite:

```{r accessor}
subset(table_regression_models(), family == "Survival")
```

The per-family reference sections live on its help page:
`?table_regression_models` (also reachable as
`?table_regression_mixed`, `?table_regression_ordinal`,
`?table_regression_survival`, and the other family aliases).
