---
title: "Explore variables and build codebooks"
description: >
  Inspect variables and labels, and build a codebook in the console,
  Excel, or PDF, with varlist(), vl(), code_book(), and
  label_from_names(), for survey and labelled data in R.
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 97)
```

```{r setup}
library(spicy)
```

Before you build frequency tables or cross-tabulations, check how your
variables are named, labelled, and coded: unclear names, missing labels,
unexpected codes, and variables with many missing values are easier to
fix now than in a finished table. This article covers three tasks:

- recover variable labels from imported column names with
  `label_from_names()`
- inspect variables, labels, values, classes, and missing data with
  `varlist()` and `vl()`
- build a codebook, in the console, an Excel file, or a PDF, with
  `code_book()`

## Recover labels from imported column names

Some imported files store both a variable name and a variable label in
the column header. `label_from_names()` splits names of the form
`name<sep>label`, renames the columns, and stores the label as a proper
variable label.

```{r label-from-names}
df <- tibble::tibble(
  "age. Age of respondent" = c(25, 30, 41),
  "edu. Highest education level" = c("Lower", "Upper", "Tertiary"),
  "smoke. Current smoker" = c("No", "Yes", "No")
)

out <- label_from_names(df)
labelled::var_label(out)
```

The default separator `". "` was chosen to match **LimeSurvey CSV
exports** taken with *Export results -> Export format: CSV ->
Headings: Question code & question text*, which produce column names
of the form `"code. question text"`. Pass `sep =` to use any other
literal separator.

## Inspect variables with varlist()

`varlist()` gives a compact summary of each variable, including its
name, label, representative values, class, number of distinct values,
number of valid observations, and missing values.

In RStudio or Positron, the main way to use `varlist()` is
interactively. With its default behavior, it opens the overview in the
Viewer, where you can search, sort, and filter the variables.

```{r varlist-interactive, eval = FALSE}
varlist(sochealth)
```

If you prefer a shorter call in interactive work, `vl()` is a shortcut
for `varlist()`:

```{r vl-interactive, eval = FALSE}
vl(sochealth)
```

If you want the same summary returned as a tibble, use `tbl = TRUE`:

```{r varlist-all}
varlist(sochealth, tbl = TRUE)
```

If you want the `Values` column to include explicit missing values, use
`include_na = TRUE`. Selecting a few variables and keeping only the
relevant columns lets the `Values` strings print in full, so the added
`<NA>` entry is actually visible:

```{r varlist-include-na}
varlist(sochealth, smoking, income_group, self_rated_health,
  include_na = TRUE, tbl = TRUE
)[, c("Variable", "Values", "NAs")]
```

By default, the `Values` column elides longer value lists with `...`.
If you want to display all unique non-missing values instead, use
`values = TRUE`. This is especially useful for variables with a small
number of distinct values, where the full list fits on one line.
Compare the default display with the `values = TRUE` display for the
same two variables:

```{r varlist-values}
varlist(sochealth, life_sat_health, social_class,
  tbl = TRUE
)[, c("Variable", "Values", "N_distinct")]

varlist(sochealth, life_sat_health, social_class,
  values = TRUE, tbl = TRUE
)[, c("Variable", "Values", "N_distinct")]
```

For a focused inspection, select only the variables you want to review:

```{r varlist-selected}
varlist(sochealth, smoking, education, income_group, tbl = TRUE)
```

Declared missing values (haven's `na_values`/`na_range`, tagged NAs)
count as missing here too: `N_valid`, `NAs`, and `N_distinct` all
honor the declaration, consistently with the tabulation functions
(see the "Declared missing values" section of `?freq`).

This is often enough to confirm that labels, factor levels, and missing
values look correct before moving on to tabulations.

## Select subsets of variables

`varlist()` supports tidyselect, which makes it easy to inspect a subset
of variables by name pattern or type.

```{r varlist-tidyselect}
varlist(sochealth, starts_with("life_sat"), tbl = TRUE)
```

```{r varlist-numeric}
varlist(sochealth, where(is.numeric), tbl = TRUE)
```

`vl()` also works with tidyselect in the same way:

```{r vl-example}
vl(sochealth, starts_with("bmi"), tbl = TRUE)
```

## Build a codebook

A codebook is the document that travels with a data file: what each
variable measures, how it is coded, and how many observations carry
each value. `code_book()` builds it from the data frame itself. It
returns an object that prints as the list of variables and that you can
keep, inspect, or write to Excel or PDF. Counts and percentages are
unweighted: they describe the data file and are not estimates for a
population.

`code_book(sochealth)` documents every variable. The same tidyselect
selectors as in `varlist()` narrow it down:

```{r code-book-basic}
code_book(sochealth, sex, age, income_group, starts_with("bmi"))
```

The type is read off the R class, never guessed: a factor is
*categorical (nominal)*, an ordered factor *categorical (ordinal)*, a
labelled vector *categorical (labelled codes)*, an integer or double
vector *numeric*, a logical, character, or `Date` vector *logical*,
*text*, or *date*, a `POSIXct` or `POSIXlt` vector *date-time*, and an
`hms` vector *time*. The level of measurement comes from the declaration
alone: a factor whose order was not declared with `ordered()` is
nominal. A labelled vector without value labels is *numeric* (or
*text*), like one whose labels all mark missing codes (see below). The R
class itself stays in the object.

### The codebook object

Keep the result to work with its parts:

```{r code-book-object}
cb <- code_book(sochealth, sex, age, income_group, starts_with("bmi"))
```

`variables` has one row per variable: its position in the data frame,
name, label, type, R class, valid and missing counts, the count and
codes of its declared missing values, number of distinct values, and,
for numeric variables, the minimum, maximum, mean, standard deviation,
and median (dates get their earliest and latest value instead):

```{r code-book-variables}
cb$variables[, c("name", "type", "n_valid", "n_missing", "n_distinct",
                 "min", "max", "mean")]
```

`values` has one row per category of the categorical and logical
variables, with its count and its percentages of all observations and
of the valid ones, then a row for the system missing values (`NA`) of
each variable that has any:

```{r code-book-values}
cb$values
```

Numeric, text, and date variables have no category rows: `variables`
summarizes them, and they appear in `values` only through their
declared missing values (see below). An integer Likert item is numeric
too: give it value labels, or make it a factor, to list its categories.
A variable with more categories than `values` (100 by default), such as
a list of country or occupation codes, lists its first 100 categories in
their order, with their counts and percentages of the whole variable,
then its declared and system missing values; `n_categories` in
`variables` gives the total, and the PDF adds a row for the categories
not listed, so that its table still adds up. `values = Inf` lists them
all. Under the default `factor_levels = "all"`, unused levels count
toward that limit, and they are listed with a count of 0.

### Title, subtitle, authors, and notes

The header carries what a reader needs to cite and trust the document:
a title and a subtitle, the authors with their affiliations, the date,
the numbers of observations and variables, and your notes on the data
(source, exclusions, coding rules), one note per element. In the PDF, a
note that starts with `"- "` is a list item, `*italics*`, `**bold**`,
and `` `code` `` are set as such, and a web address or a `doi:` becomes
a link; the console and the Excel file show the words alone:

```{r code-book-header}
code_book(
  sochealth,
  starts_with("bmi"),
  title = "Social health survey",
  subtitle = "Body mass index",
  authors = c("Jane Doe" = "University of Somewhere"),
  notes = c(
    "Simulated data shipped with spicy (see ?sochealth).",
    "BMI in kg/m², to one decimal."
  )
)
```

`authors` also accepts a list of lists with `name`, `affiliation`, and
`orcid`. When the variables were renamed after import, `source` records
the code each one had in the source file, such as a LimeSurvey question
code: with `sex` and `age` selected, `source = c(sex = "Q1", age = "Q2")`
fills the `source` column of `variables`.

The words the codebook adds (column headers, types, header fields)
follow `options(spicy.language)`: set it to `"fr"` for a French
codebook. The labels of the data are never translated.

### Declared missing values

Data imported from SPSS often declare missing codes, such as 8 =
*Don't know* and 9 = *Refused*. `haven::read_sav()` keeps them only
with `user_na = TRUE`: its default turns them into `NA`. Stata files
carry extended missing values (`.a`, `.b`, ...) instead, which haven
reads as tagged `NA`s. `code_book()` honors both declarations as the
tabulation functions do (see the "Declared missing values" section of
`?freq`): the declared codes count as missing, are listed in `values`
with `declared_missing = TRUE`, and are summarized in the header with
the variables that carry them.

```{r code-book-declared}
trust <- haven::labelled_spss(
  c(1, 2, 2, 3, 4, 8, 9, 1, NA, 2),
  labels = c(
    "Not at all" = 1, "A little" = 2, "Somewhat" = 3, "A lot" = 4,
    "Don't know" = 8, "Refused" = 9
  ),
  na_values = c(8, 9),
  label = "Trust in the health system"
)
income <- haven::labelled_spss(
  c(3200, 4100, 99998, 5600, 99999, 2900, NA, 7400, 3800, 4500),
  labels = c("Don't know" = 99998, "Refused" = 99999),
  na_values = c(99998, 99999),
  label = "Monthly household income (CHF)"
)

cbi <- code_book(tibble::tibble(trust, income))
cbi
cbi$values
cbi$variables[, c("name", "type", "declared_codes", "min", "max", "mean")]
```

`trust` has labels on its valid codes: it is *categorical (labelled
codes)*, and `values` lists its four answers, then its two declared
missing codes and its system missing value. Every value label of
`income` sits on a declared missing code and its valid values are
measurements, so `code_book()` reads it as *numeric*: its statistics
are computed on the valid values, and its only rows in `values` are its
missing values. With `user_na = FALSE`, the declaration is ignored: the
declared codes count as valid values, and `income` becomes a
categorical variable whose every observed amount is a category.

### Write the codebook to Excel

`output = "<path>.xlsx"` writes the codebook to an Excel workbook (this
needs the `openxlsx2` package): a first worksheet with the header, then
the `variables` and `values` tables. Each table starts on row 1 with a
frozen header and filters, so it sorts cleanly and reads back without
skipping rows. Numbers stay numeric cells, and dates are text.

```{r code-book-xlsx, eval = FALSE}
code_book(
  sochealth,
  title = "Social health survey",
  authors = c("Jane Doe" = "University of Somewhere"),
  output = "sochealth_codebook.xlsx"
)
```

### Write the codebook to PDF

`output = "<path>.pdf"` writes the codebook as a document to share: a
cover (title, subtitle, authors, date), a page about the data (numbers
of observations and variables, notes, declared missing values), the
list of variables with their pages, one sheet per variable, and an
index of the variables sorted by name. It needs the quarto package and
Quarto 1.7 or later, whose bundled Typst compiles the PDF. Without
Quarto, `output = "<path>.typ"` writes the Typst source, which
`typst compile` turns into the same PDF on any machine with Typst 0.12
or later. `paper = "letter"` changes the paper size, `font`,
`font_code`, and `colors` the look, and `index_columns` the index of
variables, which takes two columns past 40 variables unless told
otherwise. spicy does not write DDI-XML: for a
DDI description of the data file, see the
[DDIwR](https://CRAN.R-project.org/package=DDIwR) package.

```{r code-book-pdf, eval = FALSE}
code_book(
  sochealth,
  title = "Social health survey",
  authors = c("Jane Doe" = "University of Somewhere"),
  output = "sochealth_codebook.pdf"
)
```

## When to use varlist() and code_book()

Use `varlist()` when you want a quick summary in a script or a tibble
you can inspect directly.

Use `vl()` when you want the same summary with a shorter call in
interactive work.

Use `code_book()` when you want a codebook to keep or share: every
variable with the counts of its categories, in an object, an Excel
file, or a PDF.

The two tools differ in one default: `varlist()` shows only the
**observed** factor levels in its `Values` column (the convention of
Stata `tab` and of the SPSS `FREQUENCIES` default, which both list
only values present in the data), whereas `code_book()` lists **all
declared** levels in `values`, unused ones with a count of 0 -- the
full coding scheme, as a data dictionary or SPSS `CTABLES` would
report it. Pass `factor_levels =` explicitly to either function to
override the default.

Their `values` arguments differ as well: in `varlist()`,
`values = TRUE` prints every value in the `Values` column; in
`code_book()`, `values` is the maximum number of categories listed per
variable (100 by default).
